---
title: "Assemble plant species list for Barro Colorado Island, Panama"
author: "Paula Uzcategui and Helene Muller-Landau"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

This Rmarkdown merges information from multiple high-quality plant species lists for Barro Colorado Island (BCI), Panama, updates to the best current woody plant species names based on the current Smithsonian ForestGEO woody species lists, utilizes prior checks of species names against the World Checklist of Vascular Plants (WCVP) version 13, and outputs the combined high-quality list. The reason for using WCVP version 13 is because that is the version currently in use by PlantNet.  The contributing lists are Garwood et al. 2009, Schnitzer et al. 2024, Zotz 2024, the Smithsonian ForestGEO Panama woody plant species and ViewTaxonomy lists (Perez et al. 2025), and a file of tree species found in forest inventory plots on BCI.  These lists are considered high-quality because they were each assembled by one or more botanists familiar with the local flora, and build on the work of Croat 1978. Combining the lists involves checking overlap not only in the original names on each list, but also on two lists of "synonyms" for BCI plant species, as well as in the current accepted name.  The synonym lists include one from Garwood 2009 and two from Smithsonian ForestGEO.  In all cases, these include not only synonyms in the botanical sense, but also prior names used for the same BCI morphospecies (including in cases where species determination was revised).  

This Rmarkdown additionally outputs a list of plant species that appear in panamabiota.org's list of plants of Barro Colorado, that are *not* in the combined high-quality list. This list is generated for the purpose of later review.  

The component and output lists are all available in the public GitHub site https://github.com/PanamaForestGEO/Panama_plant_species_lists 


## References 

Croat, Thomas B. 1978. *Flora of Barro Colorado Island*. Stanford University Press.

Garwood, Nancy C., Nancy Hensold, and Robin B. Foster. 2009. Appendix 1.  Revised Species List for Barro Colorado Island, with Synonyms. In *Seedlings of Barro Colorado Island and the Neotropics*. Cornell University Press.

Pérez, Rolando, Salomón Aguilar and David Mitre.  2025. Lista de árboles, arbustos, palmas, lianas y enredaderas de la flora de Panamá.  Version 2025-12-04.  

Schnitzer, Stefan, David DeFilippis, Rigoberto Rivera-Camaña, et al. 2024. Liana Species of the Barro Colorado Island 50-Ha Plot 2007 and 2017 Censuses. Version 7. *Dryad* https://doi.org/10.5061/DRYAD.Q2BVQ83V4.

Zotz, Gerhard.  2024.  Table S1. Updated list of epiphyte species on BCI. Online Supplementary Material for Zotz, G. 2024. Epiphyte Research on BCI. In *The First 100 Years of Research on Barro Colorado: Plant and Ecosystem Science*, ed. H. C. Muller-Landau and S. J. Wright. Smithsonian Institution Scholarly Press. https://doi.org/10.5479/si.26880865

### To do 

When reading in files, it would be good to rename columns right away if necessary to get rid of blanks, /’s, and other problematic characters, so subsequent references can be simplified.

for kable(zotznocroatbutmatched %>% filter(firstmatchtype=="orig-syn"),
      caption="Zotz species that Zotz says do not appear in Croat 1978 that nonetheless
      matched to a Garwood synonym. ")
      
      change to giving the name or names that show up in Garwood for this taxon.  
      
    
## Setup

### Load libraries
```{r setup2,}

rm(list=ls())

library(readxl)
library(writexl)
library(dplyr)
library(tidyr)
library(stringi)
library(stringr)
library(knitr)
library(DT)
```

### Read input files and initial data prep

```{r readprepdata, echo=TRUE, results="hide"}

garwood_list_orig  <- read_excel("splists_out/Garwood_2009_Appendix1HM2_WCVP_2026-03-15.xlsx", col_types = "text")

zotz_list <- read_excel("splists_out/83_Zotz_Epiphytes_TableS1_WCVP_2026-02-26.xlsx", col_types = "text") %>%
  rename_with(~ gsub("[()]", "", gsub(" ", "_", .x)))

lianas_list <- read_excel("splists_out/Lianas_Species_complete_list_WCVP_2026-03-16.xlsx", col_types = "text") %>%
  mutate(sp6 = tolower(sp6)) %>%
  rename(wcvp_accepted_lifeform=wcvp_lifeform_description)

pbiota_list <- read_excel("splists_out/PanamaBiotaBCI_WCVP_2026-02-26.xlsx", col_types = "text")

panwoody_list <- read_excel("splists_out/CurrentPanamaWoody_2026-03-24.xlsx", col_types = "text")
panwoody_synonyms <- read_excel("splists_out/alternates_and_synonyms/CurrentPanamaWoody_SYNONYMS_2026-03-24.xlsx", col_types="text")

viewtaxa_list <- read_excel("splists_raw/Forestgeo/2026-02-25FromSuzanne/ViewTaxonomy_dec4_25_HM2.xlsx")
viewtaxa_synonyms <- read_excel("splists_out/alternates_and_synonyms/ViewTaxonomy_dec4_25_SYNONYMS.xlsx")

alternate_names <- read_excel("splists_out/alternates_and_synonyms/AltNames_2026-03-24.xlsx")
alternate_names_garwood <- read_excel("splists_out/alternates_and_synonyms/alternate_names_garwood_2026-03-24.xlsx")

garwood_synonyms <- read_excel("splists_raw/Garwood/Garwood_2009_Appendix1_Synonyms.xlsx") 

garwood_synonyms <- garwood_synonyms %>% mutate(
  sp_name = trimws(paste0(Genus, " ", Species)),
  synonym_sp = trimws(paste0(SynonymGenusSingle, " ", SynonymSpeciesSingle)),
  synonym_binomial = paste0(SynonymGenusSingle, " ", SynonymSpeciesSingle)) 
# note that synonym_sp and synonym_binomial are identical; retaining for clarity of completeness

garwood_list <- garwood_list_orig %>%
  mutate(wcvp_accepted_binomial = 
           paste0(wcvp_accepted_name_genus, " ", wcvp_accepted_name_species),
         wcvp_matched_binomial = word(wcvp_matched_name,)) %>%
  rename(wcvp_accepted_lifeform = wcvp_lifeform_description)
```

We have a list of alternate_names_garwood is a list of alternate names that match to garwood. It is generated in the alternate_species_names.Rmd to be used in this script.

```{r buildaltmatchlookup}
# ─────────────────────────────────────────────────────────────────────────────
# Build a flat lookup table used downstream to resolve BCI plot species and
# lianas list entries that did not match Garwood directly.
# Both the preferred name and the alternate name are indexed as possible keys
# so that either form can be used to retrieve the Garwood match.
# ─────────────────────────────────────────────────────────────────────────────
alt_match_lookup <- bind_rows(
  alternate_names_garwood %>%
    select(match_key = prefname, garwood_orig_match, garwood_synonym_match,
           firstmatchtype, alt_name = altname, pref_name = prefname),
  alternate_names_garwood %>%
    select(match_key = altname,  garwood_orig_match, garwood_synonym_match,
           firstmatchtype, alt_name = altname, pref_name = prefname)
) %>%
  filter(!is.na(match_key)) %>%
  distinct(match_key, .keep_all = TRUE)

```

### Output file names 

MAIN OUTPUTS: 

- Combined, high-quality list of BCI plant species.  

- List of BCI species that might be woody according to WCVP lifeform. 

ADDITIONAL OUTPUTS OF FILES TO CHECK: 

- List of Zotz species not in Garwood

- List of BCI plot species not in Garwood

- List of BCI liana list species not in Garwood

- List of PanamaBiota BCI species that are not in the combined high-quality list

```{r outfilenames}
save_plantlists_dir = "splists_out/Plantnet/"

FILECOMBINEDLIST <- paste0(save_plantlists_dir, 
                               "BCI_SPECIES_BOTANISTS_LIST_",Sys.Date(),".xlsx")
FILEBCIWOODY <- paste0("splists_out/needs_revision/BCI_SPECIES_POSSIBLY_WOODY_",Sys.Date(),".xlsx")

FILECHECKZOTZ <- "splists_out/needs_revision/zotz_species_not_in_garwood.xlsx"

FILECHECKPLOTS <-  "splists_out/needs_revision/plotsbci_not_garwood.xlsx"

FILECHECKLIANAS <- "splists_out/needs_revision/lianasbci_not_garwood.xlsx"

FILECHECKPANAMABIOTA <- "splists_out/needs_revision/panamabiota_not_garwood.xlsx"

```



## Assembling a combined BCI high-quality botanist-determined plant species list

### Garwood 2009 list



### Remove non-BCI species from Garwood list

Remove species that are listed as being in the Barro Colorado Nature Monument but not on BCI itself (chg=="BCNM").

```{r garwoodnobcnm}


# filter out species that are in the mainland peninsulas but not BCI, except for those that turn out to be on BCI after all...

yesonbcibinomial <- c("Matayba oppositifolia", "Myrcia zetekiana")

garwood_list <- garwood_list %>% filter(chg != "BCNM" | is.na(chg) | orig_name %in% yesonbcibinomial)
garwoodtoremove <- subset(garwood_list_orig, chg=="BCNM" & 
                            !orig_binomial %in% yesonbcibinomial)

kable(garwoodtoremove %>% select(orig_name,SYNONYMS),
      caption = "Species dropped from Garwood because they are in the mainland peninsulas but not on BCI itself (at least according to Garwood).")

kable(data.frame(binomial=yesonbcibinomial),caption="Species that Garwood lists as only in the mainland peninsulas that are nonetheless kept because they turn out to be in other lists for BCI")
```



```{r garwoodnoid}

# view unidentified species
garwoodnobinomial <- subset(garwood_list_orig,is.na(word(orig_binomial,2)))

kable(garwoodnobinomial %>% select(orig_name,Syn_notes) %>% rename(garwood_name=orig_name),
      caption="Garwood species lacking a complete binomial")

```

This leaves `r nrow(garwood_list)` plant species on the Garwood list.  Note that in `r nrow(garwoodnobinomial)` cases these species lack proper binomials, indicating their species identity remains uncertain (including possible new species).  These are left in anyway because they are known morphospecies.  

### Zotz 2024 epiphyte species list

There are `r nrow(zotz_list)` species on the Zotz 2024 epiphyte list.  

Match Zotz species to the Garwood list as far as possible, including by synonynms in the broad sense, and identify Zotz species not on the Garwood list.  

Note that we make one manual correction to a Zotz misspelling of a Croat name to enable matching with the Garwood list for that name.  Other misspellings to the Zotz names themselves are addressed by the matching with WCVP.  
```{r zotznogarwood}
# removing morphospecies identified only at the genus level (there is only one)
zotz_list <- zotz_list %>% filter(is.na(orig_binomial) | !is.na(word(orig_binomial,2) ))

# when the name listed is the same as in Croat 1978, the column Sp_Croat_1978 is blank (NA)
# so create a new name_Croat field that is the name of the species in Croat 1978
zotz_list <- zotz_list %>%
  mutate(wcvp_accepted_binomial = 
           paste0(wcvp_accepted_name_genus," ",wcvp_accepted_name_species),
         wcvp_matched_binomial = word(wcvp_matched_name,1,2),
         croat_name = ifelse(!is.na(Sp_Croat_1978),Sp_Croat_1978,orig_name)) %>%
  mutate(croat_binomial = word(croat_name,1,2)) %>%
  rename(wcvp_accepted_lifeform=wcvp_lifeform_description)

# make one manual correction so matches work right
zotz_list <- zotz_list %>%
  mutate(croat_binomial =ifelse(croat_binomial=="Bolbiltis nicotianifolia",
         "Bolbitis nicotianifolia",croat_binomial))

# matching is done on the binomials (not full names including subspecies)

zotz_list <- zotz_list %>%
  mutate(matchorig = 
           orig_binomial %in% garwood_list$orig_binomial,
         matchcroat = 
           croat_binomial %in% garwood_list$orig_binomial,
         matchorigsynonym = 
           orig_binomial %in% garwood_synonyms$synonym_binomial,
         matchcroatsynonym =
           croat_binomial %in% garwood_synonyms$synonym_binomial,
         matchwcvpmatch = 
           !is.na(wcvp_matched_name) & 
              (wcvp_matched_binomial %in% c(garwood_list$wcvp_matched_binomial,
                                        garwood_list$orig_binomial,
                                        garwood_synonyms$synonym_binomial)),
         matchwcvpaccept = 
           !is.na(wcvp_accepted_name) & 
              wcvp_accepted_binomial %in% c(garwood_list$wcvp_matched_binomial,
                                        garwood_list$orig_binomial,
                                        garwood_synonyms$synonym_binomial)
  ) %>%
  mutate(matchany = matchorig | matchcroat | matchorigsynonym |matchcroatsynonym | 
           matchwcvpmatch | matchwcvpaccept) %>% 
  mutate(firstmatchtype=ifelse(matchorig,"orig-orig",
                               ifelse(matchcroat,"croat-orig",
                                      ifelse(matchorigsynonym,"orig-syn",
                                             ifelse(matchcroatsynonym,"croat-syn",
                                                    ifelse(matchwcvpmatch,"wcvpmatch",
                                                           ifelse(matchwcvpaccept,"wcvpaccpt",
                                                                  NA)))))))
           
zotz_nomatchgarwood <- zotz_list %>%  filter(!matchany)

write_xlsx(zotz_nomatchgarwood, FILECHECKZOTZ)

# check the species that match only on the WCVP names
checkmatch1 <- subset(zotz_list, matchany & 
                        !matchcroat & !matchorigsynonym & !matchorig & !matchcroatsynonym) %>%
  select(orig_name,match_similarity,wcvp_matched_name,wcvp_accepted_name,Sp_Croat_1978)
kable(checkmatch1,
      caption="Species in Zotz that match Garwood only via WCVP matched or accepted names.  These are all cases in which the name is misspelled in Zotz.")

# see species that matched by a synonym and not by original or Croat names
checkmatch2a <- zotz_list %>% filter(matchorigsynonym & !matchorig & !matchcroat) %>%  
  left_join(
    garwood_synonyms %>% select(garwood_orig = sp_name, synonym_binomial, syn_text = SynonymTextALL) %>% distinct(),
    by = c("orig_binomial" = "synonym_binomial")
  )%>% 
  select(orig_binomial,croat_binomial,garwood_orig,firstmatchtype,syn_text)

checkmatch2b <- zotz_list %>% filter(matchcroatsynonym & !matchorig & !matchcroat) %>%  
  left_join(
    garwood_synonyms %>% select(garwood_orig = sp_name, synonym_binomial, syn_text = SynonymTextALL) %>% distinct(),
    by = c("croat_binomial" = "synonym_binomial")
  ) %>% mutate(matchtype="croat-to-synonym") %>%
  select(orig_binomial,croat_binomial,garwood_orig,firstmatchtype,syn_text)

checkmatch2 <- rbind(checkmatch2a,
                     subset(checkmatch2b,!orig_binomial %in% checkmatch2a$orig_binomial))

kable(checkmatch2, 
          caption="Zotz species that matched Garwood only by synonym. Note that in some cases there was more than one match by synonym.  Here orig_binomial means the current name in Zotz.",
          options = list(scrollX = TRUE, pageLength = 10))

zotznocroatbutmatched <- zotz_list %>%
  filter(Sp_Croat_1978=="**" & matchany==T) %>%
  arrange(firstmatchtype,orig_name) %>%
  select(orig_name,wcvp_matched_name,wcvp_accepted_name,firstmatchtype,Sp_Croat_1978)

kable(zotznocroatbutmatched %>% filter(firstmatchtype=="orig-orig"),
      caption="Zotz species that Zotz says do not appear in Croat 1978 that nonetheless
      matched in Garwood, and appear with the same name in both")
  

kable(zotznocroatbutmatched %>% filter(firstmatchtype=="orig-syn"),
      caption="Zotz species that Zotz says do not appear in Croat 1978 that nonetheless
      matched to a Garwood synonym. ")

kable(zotz_nomatchgarwood  %>% select(orig_name, wcvp_matched_name, wcvp_accepted_name, Sp_Croat_1978), 
          caption="Zotz species that don't match anything in Garwood.  Note that in all cases Sp_Croat_1978 has value **, indicating that Zotz 2024 states that the species does not appear in Croat 1978.")

```

In total there are `r nrow(zotz_nomatchgarwood)` species on the Zotz list not already on the Garwood list.  

Add these Zotz species to the combined BCI list, which starts with Garwood.
```{r}
# using garwood as a base
garwood_list$source_list = "Garwood2009"
zotz_nomatchgarwood$source_list = "Zotz"

listscol <- c("orig_name",
              "orig_binomial",
              "wcvp_accepted_binomial",
              "wcvp_matched_name",
              "wcvp_matched_plant_name_id",
              "wcvp_matched_ipni_id",
              "wcvp_matched_status",
              "wcvp_accepted_name",
              "wcvp_accepted_family", 
              "wcvp_accepted_authority",
              "wcvp_accepted_plant_name_id", 
              "wcvp_accepted_powo_id",
              "wcvp_accepted_ipni_id",
              "wcvp_accepted_lifeform",
              "source_list"
              )

bcilist <- bind_rows(
  garwood_list %>% 
    select(all_of(c(listscol, "SYNONYMS", "LF_RF", "BCI cult"))),
  zotz_nomatchgarwood %>% select(all_of(listscol))
)

```

The pre-existing "altnerate_names" file combines species names in the synonyms or "old names" columns of currentwoody, viewtaxa and garwood. 

These are not necessarily synonyms in the botanical sense; they are any names formerly used for a given BCI morphospecies (e.g., due to misidentification) 

If a species in the BCI plots or in the lianas list does not match the garwood directly, an alternate name related to the species could match. 


### List of tree species recorded in censuses of forest plots on BCI 

Check that all the tree species recorded from forest plots on BCI are also in the combined species list.  

```{r plotslist}
# Preparing plotlist

# 1. Read data
bci_plots <- read_excel("splists_raw/Plots/abundance_plots_HM.xlsx")

bci_plots <- bci_plots %>% select(sp, binomial, bci, `bci 10 ha plot`, P11, P14)

bci_other_plots <- read.delim(
  "splists_raw/Plots/doi_10_5061_dryad_1g1jwsvc3__v20260226/mainfile.txt"
)

# 2. Summarise other plots safely
bci_other_plots <- bci_other_plots %>%
  group_by(sp6) %>%
  summarise(
    bigplot = n_distinct(tag),
    .groups = "drop"
  )

# 3. Clean and rename BEFORE joining
bci_plots <- bci_plots %>%
  rename(
    sp6 = sp,
    sp_name_plot = binomial,
    bci50ha = bci,
    bci10ha = `bci 10 ha plot`
  )

# 4. Join safely (keeps all bci_plots species)
bci_plots <- bci_plots %>%
  left_join(bci_other_plots, by = "sp6") %>%
  mutate(
    bigplot = coalesce(bigplot, 0)   # replace NA with 0
  )

# 5. Filter safely (handle NA just in case)
bci_plots <- bci_plots %>%
  mutate(across(c(bci50ha, bci10ha, P14, P11), ~coalesce(., 0))) %>%
  filter(
    bci50ha > 0 |
    bci10ha > 0 |
    P14 > 0 |
    P11 > 0 |
    bigplot > 0
  )

# 6. Create plot_name column safely
bci_plots <- bci_plots %>%
  rowwise() %>%
  mutate(
    plot_name = paste(
      c(
        if (bci50ha > 0) "bci50ha",
        if (bigplot > 0) "bcimultiplots",
        if (bci10ha > 0) "bci10ha",
        if (P14 > 0) "P14",
        if (P11 > 0) "P11"
      ),
      collapse = ", "
    )
  ) %>%
  ungroup()

# the plots files don't have wcvp accepted name, merge them by 6-letter species code in the Panama woody species list
match21 <- match(bci_plots$sp6, panwoody_list$sp6)            
nomatchview <- bci_plots[is.na(match21),] %>% select(sp6, sp_name_plot) # Those that don't match are all unidentified

if (nrow(nomatchview)>0){
knitr::kable(nomatchview, caption="Warning: Species recorded in plots on BCI whose 6-letter codes do not have a match in the Panama ForestGEO woody plant species database.")} else{
  print("All species in the plot have a match to a sp6 code in the current woody list.")
}

bci_plots <- bci_plots %>% left_join(
  panwoody_list, by="sp6"
)

# if no orig name add the binomial
bci_plots <- bci_plots %>% mutate(orig_name = ifelse(is.na(orig_name), sp_name_plot, orig_name))

# remove the row for "unidentified", if present
bci_plots <- bci_plots %>% filter(sp6!="uniden")

bcidups <- bci_plots %>% filter(duplicated(sp6)) %>% select(sp6, sp_name_plot)

if (nrow(bcidups)>0)
knitr::kable(bcidups, caption="Warning: Some 6-letter species codes are repeated in the bci_plots data.frame.")

# create binomial first if not already
bci_plots$wcvp_accepted_binomial <- paste0(
  bci_plots$wcvp_accepted_genus, " ",
  bci_plots$wcvp_accepted_species
)

```

*In total there are `r nrow(bci_plots)` tree morphospecies recorded for forest plot censuses on BCI.*

Matching to original garwood names (direct match) or to synonyms or alternate names that match to garwood.
 
#### 1. Direct match to garwood names

```{r bciplotdirectmatch}
# 1. direct name/binomial match
bci_plots_garwood_name_match <- bci_plots %>%
  filter(
    orig_name %in% bcilist$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist$wcvp_accepted_name) |

    # match binomials
    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist$orig_binomial 
    )) |

    (!is.na(wcvp_accepted_binomial) & (
        wcvp_accepted_binomial %in% bcilist$wcvp_accepted_binomial 
    ))
  )
```

*Matched `r nrow(bci_plots_garwood_name_match)` species in the plots directly to garwood.*

#### 2. Alternate name match 
```{r plotalternatematchtry}


# 2. everything that did not match directly to garwood, match to alternate names
bci_plots_garwood_synonym_match <- bci_plots %>% 
  filter(
      !orig_name %in% bci_plots_garwood_name_match$orig_name &
    (
      orig_name %in% alt_match_lookup$match_key |
      orig_binomial %in% alt_match_lookup$match_key |
      
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% alt_match_lookup$match_key) |
        
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% alt_match_lookup$match_key)
    )
  )


# 3. print the altname it matched to and why that name is in garwood

bci_plots_garwood_synonym_print <- bci_plots_garwood_synonym_match %>%
  # Join using a chain of coalesce to check the most reliable fields first
  left_join(alt_match_lookup, by = c("orig_name" = "match_key")) %>%
  left_join(alt_match_lookup, by = c("orig_binomial" = "match_key"), suffix = c("", "_bi")) %>%
  left_join(alt_match_lookup, by = c("wcvp_matched_name" = "match_key"), suffix = c("", "_match")) %>%
  left_join(alt_match_lookup, by = c("wcvp_accepted_name" = "match_key"), suffix = c("", "_acc")) %>%
  
  # Now, condense everything into final fields
  mutate(
    orig_garwood    = coalesce(garwood_orig_match, garwood_orig_match_bi, garwood_orig_match_match, garwood_orig_match_acc),
    orig_garwood_synonym = coalesce(garwood_synonym_match, garwood_synonym_match_bi, garwood_synonym_match_match, garwood_synonym_match_acc),
    firstmatchtype  = coalesce(firstmatchtype, firstmatchtype_bi, firstmatchtype_match, firstmatchtype_acc),
    alternate_name  = coalesce(alt_name, alt_name_bi, alt_name_match, alt_name_acc),
    prefered_name   = coalesce(pref_name, pref_name_bi, pref_name_match, pref_name_acc)
  ) %>%
  select(-ends_with("_bi"), -ends_with("_match"), -ends_with("_acc")) # Clean up extra columns

```

*Matched `r nrow(bci_plots_garwood_synonym_match)` species in the plots by synonyms or alternate names to garwood.*

The table below shows BCI plot species that could not be matched directly to Garwood (2009) but were linked through an alternate name. The `firstmatchtype` column indicates which name pairing produced the link:

Definitions:

- F=ForestGEO

- G=Garwood

- Fcur = ForestGEO current name from most recent list

- Falt = ForestGEO alternate name from most recent list for Panama woody plants in general

- Gorig = Garwood 2009 original name

- Galt = Garwood 2009 alternate name

- **direct synonym** — the name appears directly in the Garwood list as a synonym

 Definitions:

F=ForestGEO

G=Garwood

Fcur = ForestGEO current name from most recent list

Falt = ForestGEO alternate name from most recent list for Panama woody plants in general

Gorig = Garwood 2009 original name

Galt = Garwood 2009 alternate name

 
- **Fcur-Gorig** — the current ForestGEO name matches a name listed by Garwood

- **Fcur-Gorigbi** — the current ForestGEO name matches the binomial of a Garwood original name.

- **Falt-Gorig** — a ForestGEO alternate name matches a name originally listed by Garwood

- **Falt-Gorigbi** — a ForestGEO alternate name matches the binomial of a Garwood original name.

- **Fcur-Galt** — the current ForestGEO name matches a synonym documented by Garwood that is conflicting with a forestgeo alternate-prefered pair

- **Fcur-Galtbi** — the current ForestGEO name matches the binomial of a Garwood synonym

- **Falt-Galt** — a historical ForestGEO alternate name matches a synonym documented by Garwood

- **Falt-Galtbi** — a historical ForestGEO alternate name matches the binomial of a Garwood synonym

 

Note that "Garwood original name" refers to the name as it appears in the Garwood (2009) revision, not necessarily the currently accepted name.

```{r printaltmatches}
datatable(
  bci_plots_garwood_synonym_print %>% 
  select(sp6, orig_name, wcvp_matched_name, wcvp_accepted_name, firstmatchtype, orig_garwood,
         alternate_name, prefered_name, orig_garwood_synonym), 
          caption="BCI plot species where an alternate name of the species matched to garwood",
          options = list(scrollX = TRUE, pageLength = 10))


```

Printing species in the plots that are not in Garwood. 
```{r allnomatchbciplot}

# combine direct matches and matches by alternate names or synonyms
bci_plots_garwood_match <- bind_rows(
  bci_plots_garwood_name_match,
  bci_plots_garwood_synonym_match
) %>%
  distinct(sp6, .keep_all = TRUE)

# filter unmatched rows
bci_plots_no_garwood_at_all <- bci_plots %>%
  filter(!sp6 %in% bci_plots_garwood_match$sp6)

kable(bci_plots_no_garwood_at_all  %>% select(sp6, orig_name, wcvp_matched_name, 
                                              wcvp_accepted_name, plot_name), 
          caption="Species recorded in BCI plots census that don't match anything in Garwood")


plotsingarwoodorig <- garwood_list_orig %>% 
  filter(orig_name %in% bci_plots_no_garwood_at_all$orig_name)

if (nrow(bci_plots_no_garwood_at_all)>0 & nrow(plotsingarwoodorig)>0) {
  datatable(plotsingarwoodorig %>%
    select(sp6, orig_name, chg), 
            caption="In some cases, these species were on the original Garwood list but recorded there as being only in the BCNM, not on BCI.")
    
}

write_xlsx(bci_plots_no_garwood_at_all, FILECHECKPLOTS)

```

Added `r nrow(bci_plots_no_garwood_at_all)` completely new rows to the BCI list


```{r addplottobci}

bcilist <- bind_rows(
  bcilist,
  bci_plots_no_garwood_at_all %>% 
    mutate(source_list="plotcensuses",
           forestgeo_current_name = current_name) %>%
    select(all_of(c(listscol, "plot_name", "bci50ha", "bci10ha", "P11", "P14", "sp6", "sp4", "forestgeo_current_name"))
))


# add a field for the names that matched by alternate names
bcilist <- bcilist %>% mutate(
  pref_forestgeo = NA_character_,
  orig_garwood = NA_character_
)

# Add the prefered name in forestgeo to the species that matched with the alternate names file. 
# This is to facilitate finding the current_forestgeo name later. Since some of this matches are entirely trought alternate names
bcilist <- bcilist %>% 
  rows_update(
    bci_plots_garwood_synonym_print %>% 
      # Rename 'orig_garwood' to 'orig_name' to match bcilist exactly
    select(orig_name = orig_garwood, pref_forestgeo = prefered_name),
    by = "orig_name"
  )

```


### Lianas BCI list

Check whether they are in Garwood directly or by alternate names.  
```{r lianasnobcilist}

lianas_list <- lianas_list %>% filter(isbci=="TRUE")
lianas_list <- lianas_list %>% distinct(orig_name, .keep_all = TRUE)
lianas_list <- lianas_list %>% 
  mutate(wcvp_accepted_binomial = paste(wcvp_accepted_name_genus, wcvp_accepted_name_species),
         former_binomial = ifelse(is.na(GenusFormer),NA,paste(GenusFormer,SpeciesOnlyFormer))) %>%
  mutate(former_name = paste(ifelse(is.na(SubspeciesFormer),former_binomial,
                                    paste(former_binomial,SubspeciesFormer))))
```

Working with `r nrow(lianas_list)` liana species. 

```{r lianasdirect}

# 1. direct name/binomial match
lianas_garwood_name_match <- lianas_list %>%
  filter(

    orig_name %in% bcilist$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist$wcvp_accepted_name) |

    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist$orig_binomial 
    )) |

    (!is.na(wcvp_accepted_binomial) & (
        wcvp_accepted_binomial %in% bcilist$wcvp_accepted_binomial 
    ))
  )


#2 match of former liana name to current Garwood name 
lianas_formername_match <- lianas_list %>%
  filter(
    !orig_name %in% c(lianas_garwood_name_match$orig_name)
                      &
    (former_name %in% bcilist$orig_name | 
       former_binomial %in% bcilist$orig_binomial)  )

kable(lianas_formername_match %>% select(former_name,orig_name),
      caption="Liana species in which former names matched to Garwood original name.")
```

Matched `r nrow(lianas_garwood_name_match) + nrow(lianas_formername_match)` species in the lianas list that are directly in Garwood. 

```{r lianaaltmatch}

#4 match of former liana name to Garwood synonym
lianas_formertosynonym_match <- lianas_list %>%
  filter(
    !orig_name %in% c(lianas_garwood_name_match$orig_name)&
    (former_name %in% garwood_synonyms$synonym_sp |
       former_binomial %in% garwood_synonyms$synonym_binomial))

# see species that matched by a synonym 
lianas_formertosynonym_print <- lianas_formertosynonym_match %>%  left_join(
    garwood_synonyms %>% select(garwood_name = sp_name, synonym_sp, syn_text = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) 

kable(lianas_formertosynonym_print %>% 
            select(orig_name, former_name, garwood_name, wcvp_accepted_name,syn_text), 
          caption="Liana species whose former names matched a Garwood synonym.",
          options = list(scrollX = TRUE, pageLength = 10))

# 5. everything that did not match directly to garwood, match to alternate names
lianas_garwood_alternate_match <- lianas_list %>% 
  filter(
       !orig_name %in% c(lianas_garwood_name_match$orig_name, lianas_formertosynonym_match$orig_name) &
    (
      orig_name %in% alt_match_lookup$match_key |
      orig_binomial %in% alt_match_lookup$match_key |
      
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% alt_match_lookup$match_key) |
        
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% alt_match_lookup$match_key)
    )
  )


# print the altname it matched to and why that name is in garwood

lianas_garwood_alternate_print <- lianas_garwood_alternate_match %>%
  # Join using a chain of coalesce to check the most reliable fields first
  left_join(alt_match_lookup, by = c("orig_name" = "match_key")) %>%
  left_join(alt_match_lookup, by = c("orig_binomial" = "match_key"), suffix = c("", "_bi")) %>%
  left_join(alt_match_lookup, by = c("wcvp_matched_name" = "match_key"), suffix = c("", "_match")) %>%
  left_join(alt_match_lookup, by = c("wcvp_accepted_name" = "match_key"), suffix = c("", "_acc")) %>%
  
  # Now, condense everything into your final fields
  mutate(
    orig_garwood    = coalesce(garwood_orig_match, garwood_orig_match_bi, garwood_orig_match_match, garwood_orig_match_acc),
    orig_garwood_synonym = coalesce(garwood_synonym_match, garwood_synonym_match_bi, garwood_synonym_match_match, garwood_synonym_match_acc),
    firstmatchtype  = coalesce(firstmatchtype, firstmatchtype_bi, firstmatchtype_match, firstmatchtype_acc),
    alternate_name  = coalesce(alt_name, alt_name_bi, alt_name_match, alt_name_acc),
    prefered_name   = coalesce(pref_name, pref_name_bi, pref_name_match, pref_name_acc)
  ) %>%
  select(-ends_with("_bi"), -ends_with("_match"), -ends_with("_acc")) # Clean up extra columns

```

The table below shows liana species that could not be matched directly to Garwood (2009) but were linked through an alternate name. The `firstmatchtype` column indicates which name pairing produced the link:


Note that "Garwood original name" refers to the name as it appears in the Garwood (2009) revision, not necessarily the currently accepted name.

```{r showlianaalternate}
datatable(
  lianas_garwood_alternate_print %>% 
  select(orig_name, wcvp_matched_name, wcvp_accepted_name, firstmatchtype, orig_garwood,
         alternate_name, prefered_name, orig_garwood_synonym), 
          caption="Liana species where an alternate name of the species matched to garwood",
          options = list(scrollX = TRUE, pageLength = 10))
```

Printing liana species that are not in the list so far and need to be added: 
```{r addlianas}
# additional species still unmatched

lianas_garwood_match <- bind_rows(
  lianas_garwood_name_match,
  lianas_formername_match,
  lianas_formertosynonym_match,
  lianas_garwood_alternate_match
) %>%
  distinct(orig_name, .keep_all = TRUE)

lianas_no_garwood_at_all <- lianas_list %>%
  filter(!orig_name %in% lianas_garwood_match$orig_name)


# 6. Output and Display
kable(lianas_no_garwood_at_all %>%
  select(former_name, orig_name, wcvp_matched_name, wcvp_accepted_name), 
          caption = "Lianas species recorded in the 50-ha plot that don't have a match in Garwood",
          options = list(scrollX = TRUE, pageLength = 10))

write_xlsx(lianas_no_garwood_at_all, FILECHECKLIANAS)

```

Add to bci list
```{r lianasbciadd}

lianas_no_garwood_at_all$source_list = "lianas"

bcilist <- bind_rows(
  bcilist,
  lianas_no_garwood_at_all %>% select(all_of(listscol))
)

# Add the forestegeo names of the species that matched by alternate names
bcilist <- bcilist %>% 
  rows_update(
    lianas_garwood_alternate_print %>% 
      # Rename 'orig_garwood' to 'orig_name' to match bcilist exactly
      select(orig_name = orig_garwood, pref_forestgeo = prefered_name),
    by = "orig_name"
  )

```

### Check consistency 
```{r savebotanists}

# Check consistency
# removing rows that are not identified at species level
kable(bcilist %>% filter(is.na(word(orig_binomial, 2))) %>% select(orig_name, wcvp_accepted_name), caption="Species to be removed from final list because they are not fully identified" )

bcilist <- bcilist %>% filter(!is.na(word(orig_binomial, 2))) 

bcidupnames <- bcilist %>% group_by(orig_name) %>% filter(n()> 1) 
if (nrow(bcidupnames)>0) {
  datatable(bcidupnames%>%
          select(orig_name, wcvp_accepted_name) %>% arrange(orig_name),
          caption="Species with duplicated original name",
  options = list(scrollX = TRUE, pageLength = 10)
          )
  
}

bcidupwcvp <- bcilist %>% group_by(wcvp_accepted_name) %>% 
            filter(n()> 1)
if (nrow(bcidupwcvp)>0) {
datatable(bcidupwcvp %>% select(orig_name, wcvp_accepted_name)%>% 
            arrange(wcvp_accepted_name), 
          caption="Species duplicated wcvp_accepted name",
  options = list(scrollX = TRUE, pageLength = 10)
          )
}

bcilist %>%
  count(source_list) %>%
  arrange(desc(n)) %>%
  kable(col.names = c("source_list", "count"),caption="Sources of morphospecies in the BCI plant species list.")

datatable(
  bcilist %>%
    select(wcvp_accepted_name, orig_name, source_list) %>% 
    arrange(source_list),
    caption="Species list for BCI, with the source of each entry",
  options = list(scrollX = TRUE, pageLength = 10))

```


### Use the Panama woody plant morphospecies list to link in current ForestGEO names and 6-letter and 4-letter species codes for all species.

```{r addforestgeocurrentname}

# For all the species in the bcilist
# check if they are directly on the forestgeo /woody list
# check if they have an alternate name on that list

# Add a column with the forestgeo name
# Flag when the forestgeo_current_name != orig_garwood name

# 1. Create lookup table for woody species
forestgeo_names_lookup <- bind_rows(
    panwoody_list %>% mutate(matchby = "currentname") %>% select(match_key = current_name, forestgeo_current_name = current_name, sp6, sp4, wcvp_accepted_duplicated, matchby),
  panwoody_list %>% mutate(matchby = "origname") %>% select(match_key = orig_name, forestgeo_current_name = current_name, sp6, sp4, wcvp_accepted_duplicated, matchby),
  panwoody_list %>% mutate(matchby = "wcvpaccepted") %>% select(match_key = wcvp_accepted_name, forestgeo_current_name = current_name, sp6, sp4, wcvp_accepted_duplicated, matchby),
  panwoody_list %>% mutate(matchby = "wcvpmatch") %>% select(match_key = wcvp_matched_name, forestgeo_current_name = current_name, sp6, sp4, wcvp_accepted_duplicated, matchby)
) %>% 
  filter(!is.na(match_key)) %>%
  # Group and collapse to keep provenance if a key matched in multiple ways
  group_by(match_key, forestgeo_current_name, sp6, sp4, wcvp_accepted_duplicated) %>%
  summarise(
    matchby = paste(matchby, collapse = ", "),
    .groups = "drop"
  )

forestgeo_names_lookup <- forestgeo_names_lookup %>% mutate(row_id = row_number())

### CHECKS ###
# check if one match_key points to different forestgeo_current_name
# This happens because the wcvp accepted name is different from the forestgeo_current_name. 
# The botanist are confident these species are different. But to wcvp v13 they are the same. 
duplicate_matchkey1 <- forestgeo_names_lookup %>%
  group_by(match_key) %>%
  filter(n_distinct(forestgeo_current_name) > 1 ) %>%
  arrange(match_key)

datatable(duplicate_matchkey1,
          options = list(scrollX = TRUE, pageLength = 10),
  caption="The same key maps to a different current name. For this cases the current forestgeo name is different from the wcvp accepted name. And the current forestgeo name is believed to be the accepted name.")

# In this case we remove the species were the match_key is different from the forestgeo_current_name.
alt_names_remove <- duplicate_matchkey1 %>% filter(forestgeo_current_name != match_key)
knitr::kable(alt_names_remove, caption="Alternate names to remove from the list, the match key is duplicated.")

forestgeo_names_lookup <- forestgeo_names_lookup %>% filter(!row_id %in% alt_names_remove$row_id)
```


Now we add the forestegeo current name and codes with direct matches. 
```{r foresgeoadddirect}

# Prepare for the matches: 

bcilist <- bcilist %>% mutate(row_id = row_number())

# The species added directly from the plot list already have a sp6 code
datatable(bcilist %>%
            filter(!is.na(sp6)) %>%
            select(sp6,sp4, orig_name, forestgeo_current_name, source_list),
          caption="Species that are only in the plot list already have a foresgeo code"
          )

# 2. Match forestgeo names and codes directly from the woody list

bcilist_fgeomatch <- bcilist %>% filter(is.na(sp6)) 


bcilist_fgeomatch <- bcilist_fgeomatch %>% 
  left_join(forestgeo_names_lookup, by = c("pref_forestgeo"= "match_key"), suffix = c("", "_preffg")) %>%
  left_join(forestgeo_names_lookup, by = c("orig_name"= "match_key"), suffix = c("", "_orig")) %>%
  left_join(forestgeo_names_lookup, by = c("wcvp_accepted_name"= "match_key"), suffix = c("", "_wcvpacc")) %>%
  left_join(forestgeo_names_lookup, by = c("wcvp_matched_name"= "match_key"), suffix = c("", "_wcvpmatch"))

  
bcilist_fgeomatch <- bcilist_fgeomatch %>%  mutate(
    forestgeo_current_name =     coalesce(
      forestgeo_current_name_preffg,
      forestgeo_current_name_orig,
      forestgeo_current_name_wcvpacc,
      forestgeo_current_name_wcvpmatch
    ),
    sp6 = coalesce(
      sp6_preffg,
      sp6_orig,
      sp6_wcvpacc,
      sp6_wcvpmatch
    ),
    sp4 = coalesce(
      sp4_preffg,
      sp4_orig,
      sp4_wcvpacc,
      sp4_wcvpmatch
    )) %>%   select(-ends_with(c("_orig",  "_wcvpacc", "_wcvpmatch", "_preffg"))) 

#No many-to-many. It is a good sign, no duplicate matches.  

bcilist_fgeomatch_direct_match <- bcilist_fgeomatch %>% filter(!is.na(sp6))

# Print species matching directly. 
datatable(
  bcilist_fgeomatch_direct_match %>% select(sp6, sp4, forestgeo_current_name, pref_forestgeo, orig_name, wcvp_accepted_name, wcvp_matched_name),
  options = list(scrollX = TRUE, pageLength = 10),
  caption="BCI species matching directly to forestgeo names"
)

# Set the forestgeo fields in the main list to the ones matched directly. 
forestgeo_columns <- c("row_id","sp6","sp4","forestgeo_current_name")
bcilist <- bcilist %>% rows_update(
  bcilist_fgeomatch_direct_match %>% select(all_of(forestgeo_columns)),
  by="row_id"
  )

```
Found forestgeonames for `r nrow(bcilist_fgeomatch_direct_match)` matching directly the garwood original names to the forestgeo names. 


```{r forestgeosynonymmatch}

# 3. Match to forestgeo by synonyms.  
bcilist_nodirectmatch <- bcilist_fgeomatch %>% filter(is.na(sp6))
knitr::kable(bcilist_nodirectmatch%>% group_by(source_list) %>% summarise(n()), caption="Source of species that still dont have a match to forestgeo")

# Species without a match come mainly from Garwood and Zotz.
bcilist_nodirectmatch <- bcilist_nodirectmatch %>% 
  left_join(panwoody_synonyms %>% select(current_name, sp6, sp4, 
                                         SynonymTextSingle_join = SynonymTextSingle,
                                         SynonymTextSingle_preffg = SynonymTextSingle), 
            by=c("pref_forestgeo" = "SynonymTextSingle_join"), suffix=c("","_preffg")) %>%
  left_join(panwoody_synonyms %>% select(current_name, sp6, sp4, 
                                         SynonymTextSingle_join = SynonymTextSingle,
                                         SynonymTextSingle_orig = SynonymTextSingle), 
            by=c("orig_name" = "SynonymTextSingle_join"), suffix=c("","_orig")) %>%
  left_join(panwoody_synonyms %>% select(current_name, sp6, sp4,                            
                                         SynonymTextSingle_join = SynonymTextSingle,
                                         SynonymTextSingle_wcvpacc = SynonymTextSingle), 
            by=c("wcvp_accepted_name" = "SynonymTextSingle_join"), suffix=c("","_wcvpacc")) %>%
  left_join(panwoody_synonyms %>% select(current_name, sp6, sp4,                                  
                                         SynonymTextSingle_join = SynonymTextSingle,
                                         SynonymTextSingle_wcvpmatch = SynonymTextSingle), 
            by=c("wcvp_matched_name" = "SynonymTextSingle_join"), suffix=c("","_wcvpmatch"))

# No many-to-many warning. Good sign

bcilist_nodirectmatch <- bcilist_nodirectmatch %>% mutate(
    forestgeo_current_name =     coalesce(
      current_name,
      current_name_orig,
      current_name_wcvpacc,
      current_name_wcvpmatch
    ),
    sp6 = coalesce(
      sp6_preffg,
      sp6_orig,
      sp6_wcvpacc,
      sp6_wcvpmatch
    ),
    sp4 = coalesce(
      sp4_preffg,
      sp4_orig,
      sp4_wcvpacc,
      sp4_wcvpmatch
    ),
    synonym_matched = coalesce(
      SynonymTextSingle_preffg,
      SynonymTextSingle_orig,
      SynonymTextSingle_wcvpacc,
      SynonymTextSingle_wcvpmatch
    )
    ) %>%   select(-ends_with(c("_orig",  "_wcvpacc", "_wcvpmatch", "_preffg"))) 
  

# Print species matching to synonyms in forestgeo
bcilist_synonymmatch <- bcilist_nodirectmatch %>% filter(!is.na(sp6))
datatable(bcilist_synonymmatch %>% 
            select(sp6, sp4, forestgeo_current_name,pref_forestgeo, orig_name, wcvp_accepted_name, wcvp_matched_name, synonym_matched),
          caption="Species matching to forestgeo name by a synonym",
          options = list(scrollX = TRUE, pageLength = 10))

# Set the forestgeo fields in the main list to the ones matched by synonym 
bcilist <- bcilist %>% rows_update(
  bcilist_synonymmatch %>%  select(all_of(forestgeo_columns)),
  by="row_id"
  )

```

Found forestgeonames for `r nrow(bcilist_synonymmatch)` matching by synonyms. 


Next we check for matches by binomials. In this case we have to be careful because there can be multiple forestgeo matches for the same species. 
```{r foresgtematchbinomial}

# 4. See species that would match by binomial

bcilist_nosynonymmatch <- bcilist_nodirectmatch %>% filter(is.na(sp6)) %>% select(
  -sp6, -sp4, -forestgeo_current_name, -matchby, -synonym_matched, -current_name
)

nrow(bcilist_nosynonymmatch)

# Create binomial lookup
forestgeo_binomial_lookup <- bind_rows(
  forestgeo_names_lookup,
  panwoody_list %>%
    mutate(matchby = "origbinomial") %>%
    select(match_key = orig_binomial, forestgeo_current_name = current_name, sp6, sp4, wcvp_accepted_duplicated, matchby),
  
  panwoody_list %>% 
    mutate(matchby = "wcvpacceptedbinomial") %>% 
    select(match_key = wcvp_accepted_binomial, forestgeo_current_name = current_name, sp6, sp4, wcvp_accepted_duplicated, matchby)
) %>% 
  filter(!is.na(match_key)) %>%
  # Group and collapse to keep provenance if a key matched in multiple ways
  group_by(match_key, forestgeo_current_name, sp6, sp4, wcvp_accepted_duplicated) %>%
  summarise(
    matchby = paste(matchby, collapse = ", "),
    .groups = "drop"
  )

bci_binomialmatch_orig <- bcilist_nosynonymmatch %>%
  left_join(
  forestgeo_binomial_lookup, by=c("orig_binomial" = "match_key"), relationship = "many-to-many") %>% 
    filter(!is.na(sp6))

# Now we can have 3 cases:

nrow(bci_binomialmatch_orig)

# Case 1. Species 1 on 1 matches.  
case1_binomial <- bci_binomialmatch_orig %>%
  group_by(orig_binomial) %>%
    filter(n() == 1) %>%
  group_by(forestgeo_current_name) %>%
    filter(n() == 1) %>%
    ungroup() 

# print case 1
datatable(
  case1_binomial %>%
    select(orig_name, orig_binomial, sp6, sp4,  forestgeo_current_name) %>% arrange(forestgeo_current_name),
  caption="Species binomial matched to one forestgeo name. This means Garwood makes a distinction by a subspecies while forestgeo current name indicates there is no need for such distinction. "
  )

# In this case we keep the forestgeo current name for those species.

# Case 2: we can have two species pointing to the same forestgeo name, it means they appear as two different subspecies in Garwood, but for forestgeo they are currently the same. 

case2_binomial <- bci_binomialmatch_orig %>%
    group_by(forestgeo_current_name) %>%
    filter(n() > 1) %>%
    ungroup() 

# print case 2
datatable(
  case2_binomial %>%
    select(orig_name, orig_binomial, sp6, sp4,  forestgeo_current_name) 
  %>% arrange(forestgeo_current_name),
  caption="Species that appear as different subspecies but forestgeo makes no distinction. "
  )

# In this case we can also leave the forestgeo current name from this match


# Case 3. this is a result of the many-to-many. Multiple rows with the same orig name pointing to two different forestgeo_current name.
case3_binomial <-   bci_binomialmatch_orig %>%
    group_by(orig_name) %>%
    filter(n() > 1) %>%
  ungroup() 
  
# print case 3
datatable(
case3_binomial %>%
    select(orig_name, orig_binomial, sp6, sp4,  forestgeo_current_name) 
  %>% arrange(orig_binomial),
  caption="Species binomial matches to different species in forestgeo."
  ) 

# lets see what is happening with that: 

knitr::kable(panwoody_list %>% filter(current_name %in% c("Protium mcleodii", "Protium tenuifolium")) %>% select(orig_name, wcvp_accepted_name, orig_synonyms, current_name),
             caption="Species current name points to the same binomial name.")

# keeping only Protium tenuifolium as the forestgeo current name
#case3keep <- case3_binomial %>% filter(forestgeo_current_name == "Protium tenuifolium")

# Turns out wcvp indicated that the subspecies is now considered a different species. That is why they match by binomial to two different. What to do here???
# For now, for this case the forestgeo current name is NA. Since we dont have a match for the subspecies listed here. 
bci_binomialmatch_orig <- bind_rows(case1_binomial, case2_binomial)

# For only case 1 and 2 set the forestgeo fields in the main list to the ones matched by binomials
bcilist <- bcilist %>% rows_update(
 bci_binomialmatch_orig %>%  select(all_of(forestgeo_columns)),
  by="row_id"
  )


```
Found forestgeonames for `r nrow(bci_binomialmatch_orig)` by the binomial name.

The case of Protium tenuifolium subsp. sessiliflorum. The binomial name matches to two forestgeo names, Protium mcleodii and Protium tenuifolium. This happens because in forestgeo we have Protium tenuifolium and Protium tenuifolium subsp. mcleodii. The subspecies mcleodii then turns out to be a whole different species, Protium mcleodii according to wcvp v13. So when matching by the binomial name, we end up with a conflict.

In this case, since the name in garwood is also a subspecies Protium tenuifolium subsp. sessiliflorum, and forestgeo does make the distinction of different subspecies, I will leave the forestgeo name as NA.

I think this should be the case for all binomial matches where the subspecies is different from the species name. Since forestgeo makes the distintion of subspecies names in the code... For now I will keep matching by binomials but this needs revision. 

Nex we do the same check but for the wcvp accepted binomial.
```{r wcvpacceptedbinomialcheck}

# 4. See species that would match by binomial
bci_nobinomialmatch <- bcilist %>% filter(is.na(sp6)) %>% select(
  -sp6, -sp4, -forestgeo_current_name
)

bci_binomialmatch_wcvp <- bci_nobinomialmatch %>% left_join(
  forestgeo_binomial_lookup, by=c("wcvp_accepted_binomial" = "match_key"), relationship = "many-to-many") %>% 
    filter(!is.na(sp6))

# Now we can have 3 cases:

nrow(bci_binomialmatch_wcvp)

# Case 1. Species 1 on 1 matches.  
case1_binomial <- bci_binomialmatch_wcvp %>%
  group_by(wcvp_accepted_binomial) %>%
    filter(n() == 1) %>%
  group_by(forestgeo_current_name) %>%
    filter(n() == 1) %>%
    ungroup() 

# print case 1
datatable(
  case1_binomial %>%
    select(wcvp_accepted_name, wcvp_accepted_binomial, sp6, sp4,  forestgeo_current_name) %>% arrange(forestgeo_current_name),
  caption="Species binomial matched to one forestgeo name. This means Garwood makes a distinction by a subspecies while forestgeo current name indicates there is no need for such distinction. "
  )

# In this case we keep the forestgeo current name for those species.

# Case 2: we can have two species pointing to the same forestgeo name, it means they appear as two different subspecies in Garwood, but for forestgeo they are currently the same. 

case2_binomial <- bci_binomialmatch_wcvp %>%
    group_by(forestgeo_current_name) %>%
    filter(n() > 1) %>%
    ungroup() 

# print case 2
datatable(
  case2_binomial %>%
    select(wcvp_accepted_name, wcvp_accepted_binomial,  sp6, sp4,  forestgeo_current_name) 
  %>% arrange(forestgeo_current_name),
  caption="Species that appear as different subspecies but forestgeo makes no distinction. "
  )

# In this case we can also leave the forestgeo current name from this match


# Case 3. this is a result of the many-to-many. Multiple rows with the same orig name pointing to two different forestgeo_current name.
case3_binomial <-   bci_binomialmatch_wcvp %>%
    group_by(orig_name) %>%
    filter(n() > 1) %>%
  ungroup() 
  
# print case 3
datatable(
case3_binomial %>%
    select(wcvp_accepted_name, wcvp_accepted_binomial, sp6, sp4,  forestgeo_current_name) 
  %>% arrange(wcvp_accepted_binomial),
  caption="Species binomial matches to different species in forestgeo."
  ) 

# the same two species as before...
bci_binomialmatch_orig <- bind_rows(case1_binomial, case2_binomial)

# For only case 1 and 2 set the forestgeo fields in the main list to the ones matched by wcvp binomials
bcilist <- bcilist %>% rows_update(
 bci_binomialmatch_orig %>%  select(all_of(forestgeo_columns)),
  by="row_id"
  )


```
Found forestgeonames for `r nrow(bci_binomialmatch_orig)` by the wcvp accepted binomial name.



## Add woody column
```{r addwoody field}

bcilist <- bcilist %>%
  mutate(
    # If it has an sp6 code then it is woody or if the binomial is in woody too. 
    woody = !is.na(sp6) | orig_binomial %in% forestgeo_binomial_lookup$match_key | wcvp_accepted_binomial %in% forestgeo_binomial_lookup$match_key, 
    garwood_diff_forestgeo = !is.na(forestgeo_current_name) & (
      orig_name != forestgeo_current_name & 
      wcvp_accepted_name != forestgeo_current_name &
      orig_binomial != forestgeo_current_name &
      wcvp_accepted_binomial != forestgeo_current_name)
  ) %>% select(-pref_forestgeo)

colnames(bcilist)

# Species in the woody list
datatable(bcilist %>%
  filter(woody) %>%
  select(sp6, sp4, forestgeo_current_name, orig_name, wcvp_matched_name, wcvp_accepted_name, wcvp_accepted_family, 
         wcvp_accepted_lifeform), 
  caption="BCI species that are in the woody list",
  options = list(scrollX = TRUE, pageLength = 10))


# Species that have a sp6 code but the forestgeo current name is different from the garwood name. 
datatable(bcilist %>%
  filter(garwood_diff_forestgeo) %>%
  select(sp6, sp4, forestgeo_current_name, orig_name, wcvp_matched_name, wcvp_accepted_name), caption="BCI species where the forestgeo name is different from the garwood name. This is likely due to a misidentification in the species. The current forestgeoname is the most recently accepted name.")

          
```

The forestgeo current name is the most accepted name for the species, even above wcvp accepted name. 

Pachira quinata	for example, has wcvp accepted name Bombax ceiba, 

Pochota fendleri is the forestgeo current name and also appears as a synonym of Pachira quinata in wcvp. 

I think wcvp is wrong here, Bombax ceiba even has a different distribution from Pochota fendleri, so how come they both have Pachira quinata as synonyms?




## Save Botanist high quality list
```{r bcilistwoody, echo=TRUE}

# BCI BOTANISTS HIGH QUALITY LIST
write_xlsx(bcilist, FILECOMBINEDLIST)


```


### Possibly woody species

```{r posiblywoody}

#bcilist <- bcilist %>% left_join(garwood_list %>% select(`BCI cult`, orig_name), by="orig_name") 

bcilist <- bcilist %>% mutate(garwood_lifeform = LF_RF)

lifeform_woody = c(
  "tree",
  "shrub",
  "liana",
  "climbing shrub",
  "shrub or tree",
  "climbing shrub or tree",
  "scrambling shrub or tree",
  "subshrub or shrub",
  "subshrub, shrub or tree",
  "scrambling shrub",
  "perennial or shrub",
  "scrambling subshrub or shrub",
  "epiphytic shrub",
  "scrambling tree",
  "climbing shrub or liana",
  "scrambling shrub or liana",
  "tuberous shrub or tree",
  "herbaceous tree",
  "climbing herbaceous tree",
  "climbing herbaceous tree or liana",
  "climbing perennial or herbaceous tree",
  "subshrub or climbing shrub",
  "tuberous subshrub",
  "hemiparasitic shrub",
  "hemiepiphytic shrub or tree",
  "annual or shrub",
  "annual, subshrub or shrub"
)

# List of target abbreviations for filtering
#L Liana (to canopy)
#M Medium-sized tree: 10-30 m
#S Shrub: 1-5 m
#T Tree: >30 m
#U Treelet: 5-10 m
  
  
target_codes <- "[LMSTU]"
garwoodlf <- unique(bcilist$garwood_lifeform)
garwood_woody_lifeforms <- garwoodlf[grepl(target_codes, garwoodlf) & !grepl("a", garwoodlf, ignore.case = TRUE)]

print(garwood_woody_lifeforms)

bcilist_possibly_woody <- bcilist %>%
  filter(!woody & is.na(`BCI cult`) &
          (wcvp_accepted_lifeform %in% lifeform_woody |
           garwood_lifeform %in% garwood_woody_lifeforms 
             )  ) %>%
  select(orig_name, wcvp_matched_name, wcvp_accepted_name, wcvp_accepted_family, 
         wcvp_accepted_lifeform, garwood_lifeform)

datatable(bcilist_possibly_woody, 
          caption="BCI species that might be woody according to the wcvp lifeform",
          options = list(scrollX = TRUE, pageLength = 10))

write_xlsx(bcilist_possibly_woody, FILEBCIWOODY)


```


## Panamabiota species that are not in garwood

```{r pbiotanogarwood}

pbiota_list$wcvp_accepted_binomial <- paste0(
  pbiota_list$wcvp_accepted_name_genus, " ",
  pbiota_list$wcvp_accepted_name_species
)

# 1. direct name/binomial match
pbiota_garwood_name_match <- pbiota_list %>%
  filter(

    orig_name %in% bcilist$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist$wcvp_accepted_name) |

    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist$orig_binomial 
    )) |

    (!is.na(wcvp_accepted_binomial) & (
        wcvp_accepted_binomial %in% bcilist$wcvp_accepted_binomial 
    ))
  )

# 2. synonym_sp for remaining
pbiota_garwood_synonym_match <- pbiota_list %>%
  filter(
    !orig_name %in% pbiota_garwood_name_match$orig_name &
    (
      orig_name %in% garwood_synonyms$synonym_sp |
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(orig_binomial) & orig_binomial %in% garwood_synonyms$synonym_binomial) |
      (!is.na(wcvp_accepted_binomial) & wcvp_accepted_binomial %in% garwood_synonyms$synonym_binomial)
    )
  )

pbiota_garwood_synonym_match$matched_garwood_synonym <- TRUE


# see species that matched by a synonym 
pbiota_garwood_synonym_match_print <- pbiota_garwood_synonym_match %>%  left_join(
    garwood_synonyms %>% select(sp_orig = sp_name, synonym_sp, syn_orig = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_match = sp_name, synonym_sp, syn_match = SynonymTextALL) %>% distinct(),
    by = c("wcvp_matched_name" = "synonym_sp"), relationship = "many-to-many"
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_wcvp = sp_name, synonym_sp, syn_wcvp = SynonymTextALL) %>% distinct(),
    by = c("wcvp_accepted_name" = "synonym_sp"),  relationship = "many-to-many") %>%
  mutate(
    SynonymMatchAll = coalesce(syn_orig, syn_match, syn_wcvp),
    orig_garwood = coalesce(sp_orig, sp_match, sp_wcvp),
  ) %>%
  select(-syn_orig, -syn_match, -syn_wcvp,  -sp_orig, -sp_match, -sp_wcvp) %>%
  distinct(orig_name, .keep_all = TRUE)
  

datatable(pbiota_garwood_synonym_match_print %>% 
            select(orig_name, wcvp_matched_name, wcvp_accepted_name, SynonymMatchAll, orig_garwood), 
          caption="Panamabiota species that matched Garwood by synonym",
          options = list(scrollX = TRUE, pageLength = 10))

pbiota_garwood_match <- bind_rows(
  pbiota_garwood_name_match,
  pbiota_garwood_synonym_match
) %>%
  distinct(orig_name, .keep_all = TRUE)

pbiota_no_garwood_at_all <- pbiota_list %>%
  filter(!orig_name %in% pbiota_garwood_match$orig_name)



datatable(pbiota_no_garwood_at_all, 
          caption="Panamabiota species that don't match anything in Garwood",
          options = list(scrollX = TRUE, pageLength = 10))

write_xlsx(pbiota_no_garwood_at_all, FILECHECKPANAMABIOTA)

```


