---
title: "Assemble plant species list for Barro Colorado Island, Panama"
author: "Paula Uzcategui and Helene Muller-Landau"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

This Rmarkdown merges information from multiple high-quality plant species lists for Barro Colorado Island (BCI), Panama, checks species names against the World Checklist of Vascular Plants (WCVP) version 13, and outputs the combined high-quality list. The contributing lists are Garwood et al. 2009, Schnitzer et al. 2024, Zotz 2024, and a file of tree species found in forest inventory plots on BCI.  These lists are considered high-quality because they were each assembled by one or more botanists familiar with the local flora, and build on the work of Croat 1978. Combining the lists involves checking overlap not only in the original names on each list, but also on two lists of "synonyms" for BCI plant species, as well as in the current accepted name.  The synonym lists include one from Garwood 2009 and another from Smithsonian ForestGEO.  In both cases, these include not only synonyms in the botanical sense, but also prior names used for the same BCI morphospecies.  

This Rmarkdown additionally outputs a list of plant species that appear in panamabiota.org's list of plants of Barro Colorado, that are *not* in the combined high-quality list. This list is generated for the purpose of later review.  

The component and output lists are all available in the public GitHub site https://github.com/PanamaForestGEO/Panama_plant_species_lists 


## References 

Croat, Thomas B. 1978. *Flora of Barro Colorado Island*. Stanford University Press.

Garwood, Nancy C., Nancy Hensold, and Robin B. Foster. 2009. Appendix 1.  Revised Species List for Barro Colorado Island, with Synonyms. In *Seedlings of Barro Colorado Island and the Neotropics*. Cornell University Press.

Schnitzer, Stefan, David DeFilippis, Rigoberto Rivera-Camaña, et al. 2024. Liana Species of the Barro Colorado Island 50-Ha Plot 2007 and 2017 Censuses. Version 7. *Dryad* https://doi.org/10.5061/DRYAD.Q2BVQ83V4.

Zotz, Gerhard.  2024.  Table S1. Updated list of epiphyte species on BCI. Online Supplementary Material for Zotz, G. 2024. Epiphyte Research on BCI. In *The First 100 Years of Research on Barro Colorado: Plant and Ecosystem Science*, ed. H. C. Muller-Landau and S. J. Wright. Smithsonian Institution Scholarly Press. https://doi.org/10.5479/si.26880865


## Setup

### Load libraries
```{r setup2,}

rm(list=ls())

library(readxl)
library(writexl)
library(dplyr)
library(tidyr)
library(stringi)
library(stringr)
library(knitr)
library(DT)
```

### Read input files and initial data prep

```{r readprepdata, echo=TRUE, results="hide"}

garwood_list <- garwood_list_orig <- 
  read_excel("splists_out/Garwood_2009_Appendix1HM2_WCVP_2026-02-26.xlsx", col_types = "text")

zotz_list <- read_excel("splists_out/83_Zotz_Epiphytes_TableS1_WCVP_2026-02-26.xlsx", col_types = "text") %>%
  rename_with(~ gsub("[()]", "", gsub(" ", "_", .x)))

lianas_list <- read_excel("splists_out/Lianas_Species_complete_list_WCVP_2026-03-03.xlsx", col_types = "text")

pbiota_list <- read_excel("splists_out/PanamaBiotaBCI_WCVP_2026-02-26.xlsx", col_types = "text")

panwoody_list <- read_excel("splists_out/CurrentPanamaWoody_2026-03-04.xlsx", col_types = "text")

woody_synonyms <- read_excel("splists_out/CurrentPanamaWoody_SYNONYMS_2026-03-04.xlsx", col_types="text")

alternate_names <- read_excel("splists_out/AltNames_2026-03-03.xlsx")

garwood_synonyms <- read_excel("splists_raw/Garwood/Garwood_2009_Appendix1_Synonyms.xlsx") 

garwood_synonyms <- garwood_synonyms %>% mutate(
  sp_name = trimws(paste0(Genus, " ", Species)),
  synonym_sp = trimws(paste0(SynonymGenusSingle, " ", SynonymSpeciesSingle)),
  synonym_binomial = paste0(SynonymGenusSingle, " ", SynonymSpeciesSingle)) 
# note that synonym_sp and synonym_binomial are identical; retaining for clarity of completeness

garwood_list <- garwood_list %>%
  mutate(wcvp_accepted_binomial = 
           paste0(wcvp_accepted_name_genus, " ", wcvp_accepted_name_species),
         wcvp_matched_binomial = word(wcvp_matched_name,))
```

### Output file names 

MAIN OUTPUTS: 

- Combined, high-quality list of BCI plant species.  

- List of BCI species that might be woody according to WCVP lifeform. 

ADDITIONAL OUTPUTS OF FILES TO CHECK: 

- List of Zotz species not in Garwood

- List of BCI plot species not in Garwood

- List of BCI liana list species not in Garwood

- List of PanamaBiota BCI species that are not in the combined high-quality list

```{r outfilenames}
save_plantlists_dir = "splists_out/Plantnet/"

FILECOMBINEDLIST <- paste0(save_plantlists_dir, 
                               "BCI_SPECIES_BOTANISTS_LIST_",Sys.Date(),".xlsx")
FILEBCIWOODY <- paste0(save_plantlists_dir, 
                           "BCI_SPECIES_POSSIBLY_WOODY_",Sys.Date(),".xlsx")

FILECHECKZOTZ <- "tocheck/zotz_species_not_in_garwood.xlsx"

FILECHECKPLOTS <-  "tocheck/plotsbci_not_garwood.xlsx"

FILECHECKLIANAS <- "tocheck/lianasbci_not_garwood.xlsx"

FILECHECKPANAMABIOTA <- "tocheck/panamabiota_not_garwood.xlsx"

```



## Assembling a combined BCI high-quality botanist-determined plant species list

### Garwood 2009 list

Remove species that are listed as being in the Barro Colorado Nature Monument but not on BCI itself (chg=="BCNM").
```{r garwood}

# filter out unidentified species
#garwood_list <- garwood_list %>% filter(!is.na(word(orig_binomial, 2))) 
garwoodnobinomial <- subset(garwood_list_orig,is.na(word(orig_binomial,2)))

#garwood_synonyms <- garwood_synonyms %>% filter(!is.na(Species) & 
#                                                  substr(Species,1,3) !="sp.")

# filter out species that are in the mainland peninsulas but not BCI
garwood_list <- garwood_list %>% filter(chg != "BCNM" | is.na(chg))
garwoodtoremove <- subset(garwood_list_orig,chg=="BCNM")

kable(garwoodtoremove %>% select(orig_name,SYNONYMS),
      caption = "Species dropped from Garwood because they are in the mainland peninsulas but not on BCI itself.")
```

This leaves `r nrow(garwood_list)` plant species on the Garwood list.  Note that in `r nrow(garwoodnobinomial)` cases these species lack proper binomials, indicating their species identity remains uncertain (including possible new species).  

### Zotz 2024 epiphyte species list

There are `r nrow(zotz_list)` species on the Zotz 2024 epiphyte list.  

Match Zotz species to the Garwood list as far as possible, including by synonynms in the broad sense, and identify Zotz species not on the Garwood list.  
```{r zotznogarwood}
# removing morphospecies identified only at the genus level (there is only one)
zotz_list <- zotz_list %>% filter(is.na(orig_binomial) | !is.na(word(orig_binomial,2) ))

# when the name listed is the same as in Croat 1978, the column Sp_Croat_1978 is blank (NA)
# so create a new name_Croat field that is the name of the species in Croat 1978
zotz_list <- zotz_list %>%
  mutate(wcvp_accepted_binomial = 
           paste0(wcvp_accepted_name_genus," ",wcvp_accepted_name_species),
         wcvp_matched_binomial = word(wcvp_matched_name,1,2),
         croat_name = ifelse(!is.na(Sp_Croat_1978),Sp_Croat_1978,orig_name)) %>%
  mutate(croat_binomial = word(croat_name,1,2))

# make one manual correction so matches work right
zotz_list <- zotz_list %>%
  mutate(croat_binomial =ifelse(croat_binomial=="Bolbiltis nicotianifolia",
         "Bolbitis nicotianifolia",croat_binomial))

# matching is done on the binomials (not full names including subspecies)

zotz_list <- zotz_list %>%
  mutate(matchorig = 
           orig_binomial %in% garwood_list$orig_binomial,
         matchcroat = 
           croat_binomial %in% garwood_list$orig_binomial,
         matchorigsynonym = 
           orig_binomial %in% garwood_synonyms$synonym_binomial,
         matchcroatsynonym =
           croat_binomial %in% garwood_synonyms$synonym_binomial,
         matchwcvpmatch = 
           !is.na(wcvp_matched_name) & 
              (wcvp_matched_binomial %in% c(garwood_list$wcvp_matched_binomial,
                                        garwood_list$orig_binomial,
                                        garwood_synonyms$synonym_binomial)),
         matchwcvpaccept = 
           !is.na(wcvp_accepted_name) & 
              wcvp_accepted_binomial %in% c(garwood_list$wcvp_matched_binomial,
                                        garwood_list$orig_binomial,
                                        garwood_synonyms$synonym_binomial)
  ) %>%
  mutate(matchany = matchorig | matchcroat | matchorigsynonym |matchcroatsynonym | 
           matchwcvpmatch | matchwcvpaccept) %>% 
  mutate(firstmatchtype=ifelse(matchorig,"orig-orig",
                               ifelse(matchcroat,"croat-orig",
                                      ifelse(matchorigsynonym,"orig-syn",
                                             ifelse(matchcroatsynonym,"croat-syn",
                                                    ifelse(matchwcvpmatch,"wcvpmatch",
                                                           ifelse(matchwcvpaccept,"wcvpaccpt",
                                                                  NA)))))))
           
zotz_nomatchgarwood <- zotz_list %>%  filter(!matchany)

write_xlsx(zotz_nomatchgarwood, FILECHECKZOTZ)

# check the species that match only on the WCVP names
checkmatch1 <- subset(zotz_list, matchany & 
                        !matchcroat & !matchorigsynonym & !matchorig & !matchcroatsynonym) %>%
  select(orig_name,match_similarity,wcvp_matched_name,wcvp_accepted_name,Sp_Croat_1978)
kable(checkmatch1,
      caption="Species in Zotz that match Garwood only via WCVP matched or accepted names.  These are all cases in which the name is misspelled in Zotz.")

# see species that matched by a synonym and not by original or Croat names
checkmatch2a <- zotz_list %>% filter(matchorigsynonym & !matchorig & !matchcroat) %>%  
  left_join(
    garwood_synonyms %>% select(garwood_orig = sp_name, synonym_binomial, syn_text = SynonymTextALL) %>% distinct(),
    by = c("orig_binomial" = "synonym_binomial")
  )%>% 
  select(orig_binomial,croat_binomial,garwood_orig,firstmatchtype,syn_text)

checkmatch2b <- zotz_list %>% filter(matchcroatsynonym & !matchorig & !matchcroat) %>%  
  left_join(
    garwood_synonyms %>% select(garwood_orig = sp_name, synonym_binomial, syn_text = SynonymTextALL) %>% distinct(),
    by = c("croat_binomial" = "synonym_binomial")
  ) %>% mutate(matchtype="croat-to-synonym") %>%
  select(orig_binomial,croat_binomial,garwood_orig,firstmatchtype,syn_text)

checkmatch2 <- rbind(checkmatch2a,
                     subset(checkmatch2b,!orig_binomial %in% checkmatch2a$orig_binomial))

datatable(checkmatch2, 
          caption="Zotz species that matched Garwood only by synonym. Note that in some cases there was more than one match by synonym",
          options = list(scrollX = TRUE, pageLength = 10))

zotznocroatbutmatched <- zotz_list %>%
  filter(Sp_Croat_1978=="**" & matchany==T) %>%
  arrange(firstmatchtype,orig_name) %>%
  select(orig_name,wcvp_matched_name,wcvp_accepted_name,firstmatchtype,Sp_Croat_1978)

kable(zotznocroatbutmatched,
      caption="Zotz species that Zotz says do not appear in Croat 1978 that nonetheless
      matched in Garwood.  ")
  
kable(zotz_nomatchgarwood  %>% select(orig_name, wcvp_matched_name, wcvp_accepted_name, Sp_Croat_1978), 
          caption="Zotz species that don't match anything in Garwood.  Note that in all cases Sp_Croat_1978 has value **, indicating that the species does not appear in Croat 1978 (according to Zotz)")

```

In total there are `r nrow(zotz_nomatchgarwood)` species on the Zotz list not already on the Garwood list.  

Add these Zotz species to the combined BCI list, which starts with Garwood.
```{r}
# using garwood as a base
garwood_list$source_list = "Garwood2009"
zotz_nomatchgarwood$source_list = "Zotz"

listscol <- c("orig_name",
              "orig_binomial",
              "wcvp_accepted_binomial",
              "wcvp_matched_name",
              "wcvp_matched_plant_name_id",
              "wcvp_matched_ipni_id",
              "wcvp_matched_status",
              "wcvp_accepted_name",
              "wcvp_accepted_family", 
              "wcvp_accepted_author",
              "wcvp_accepted_plant_name_id", 
              "wcvp_accepted_powo_id",
              "wcvp_accepted_ipni_id",
              "wcvp_lifeform_description",
              "source_list")

bcilist <- bind_rows(
  garwood_list %>% 
    select(all_of(c(listscol, "SYNONYMS"))),
  zotz_nomatchgarwood %>% select(all_of(listscol))
)

```


### List of tree species recorded in censuses of forest plots on BCI 

Check that all the tree species recorded from forest plots on BCI are also in the combined species list.  

```{r plotslist}
# Preparing plotlist

# 1. Read data
bci_plots <- read_excel("splists_raw/Plots/abundance_plots_HM.xlsx")

bci_plots <- bci_plots %>% select(sp, binomial, bci, `bci 10 ha plot`, P11, P14)

bci_other_plots <- read.delim(
  "splists_raw/Plots/doi_10_5061_dryad_1g1jwsvc3__v20260226/mainfile.txt"
)

# 2. Summarise other plots safely
bci_other_plots <- bci_other_plots %>%
  group_by(sp6) %>%
  summarise(
    bigplot = n_distinct(tag),
    .groups = "drop"
  )

# 3. Clean and rename BEFORE joining
bci_plots <- bci_plots %>%
  rename(
    sp6 = sp,
    sp_name_plot = binomial,
    bci50ha = bci,
    bci10ha = `bci 10 ha plot`
  )

# 4. Join safely (keeps all bci_plots species)
bci_plots <- bci_plots %>%
  left_join(bci_other_plots, by = "sp6") %>%
  mutate(
    bigplot = coalesce(bigplot, 0)   # replace NA with 0
  )

# 5. Filter safely (handle NA just in case)
bci_plots <- bci_plots %>%
  mutate(across(c(bci50ha, bci10ha, P14, P11), ~coalesce(., 0))) %>%
  filter(
    bci50ha > 0 |
    bci10ha > 0 |
    P14 > 0 |
    P11 > 0 |
    bigplot > 0
  )

# 6. Create plot_name column safely
bci_plots <- bci_plots %>%
  rowwise() %>%
  mutate(
    plot_name = paste(
      c(
        if (bci50ha > 0) "bci50ha",
        if (bigplot > 0) "bcimultiplots",
        if (bci10ha > 0) "bci10ha",
        if (P14 > 0) "P14",
        if (P11 > 0) "P11"
      ),
      collapse = ", "
    )
  ) %>%
  ungroup()

# the plots files don't have wcvp accepted name, merge them by 6-letter species code in the Panama woody species list
match21 <- match(bci_plots$sp6, panwoody_list$sp6)            
nomatchview <- bci_plots[is.na(match21),] %>% select(sp6, sp_name_plot) # Those that don't match are all unidentified

if (nrow(nomatchview)>0)
knitr::kable(nomatchview, caption="Warning: Species recorded in plots on BCI whose 6-letter codes do not have a match in the Panama ForestGEO woody plant species database.")

bci_plots <- bci_plots %>% left_join(
  panwoody_list, by="sp6"
)

# if no orig name add the binomial
bci_plots <- bci_plots %>% mutate(orig_name = ifelse(is.na(orig_name), sp_name_plot, orig_name))

# remove the row for "unidentified", if present
bci_plots <- bci_plots %>% filter(sp6!="uniden")

bcidups <- bci_plots %>% filter(duplicated(sp6)) %>% select(sp6, sp_name_plot)

if (nrow(bcidups)>0)
knitr::kable(bcidups, caption="Warning: Some 6-letter species codes are repeated in the bci_plots data.frame.")

# create binomial first if not already
bci_plots$wcvp_accepted_binomial <- paste0(
  bci_plots$wcvp_accepted_name_genus, " ",
  bci_plots$wcvp_accepted_name_species
)

```
In total there are `nrow(bci_plots)` tree morphospecies recorded for forest plot censuses on BCI.


#### Preparing alternate names

```{r altnames}

# Alternate names combines species names in the synonyms columns of currentwoody, viewtaxa and garwood. 

# These are not necessarily synonyms in the botanical sense; they are any names formerly used for a given BCI morphospecies (e.g., due to misidentification) 

# Here we create a list of the synonyms, and if they are somewhere in the woody list, we add the name of origin y woody and the synonym they have in Garwood.
alternate_names_garwood <- alternate_names %>% 
  # DIRECT MATCH — prefname
  left_join(
    garwood_list %>% select(garwood_direct_pref_join=orig_name, garwood_direct_pref=orig_name),
    by = c("prefname" = "garwood_direct_pref_join")
  ) %>%
  left_join(
    garwood_list %>% select(garwood_direct_prefbi_join=orig_binomial, garwood_direct_prefbi=orig_name),
    by = c("prefname" = "garwood_direct_prefbi_join"), relationship = "many-to-many"
  ) %>%
  
  # DIRECT MATCH — altname
  left_join(
    garwood_list %>% select(garwood_direct_alt_join=orig_name, garwood_direct_alt=orig_name),
    by = c("altname" = "garwood_direct_alt_join")
  ) %>%
  left_join(
    garwood_list %>% select(garwood_direct_altbi_join=orig_binomial, garwood_direct_altbi=orig_name),
    by = c("altname" = "garwood_direct_altbi_join"), relationship = "many-to-many"
  ) %>%
  
  # SYNONYM MATCH — prefname
  left_join(
    garwood_synonyms %>%
      select(
        SynonymTextSingle,
        garwood_syn_orig_pref = sp_name,
        synonym_full_pref = SynonymTextALL,
      ),
    by = c("prefname" = "SynonymTextSingle")
  ) 
  
alternate_names_garwood <- alternate_names_garwood %>%
mutate(
    
    # DIRECT MATCH (either pref or alt)
    garwood_direct_match = coalesce(
      garwood_direct_pref,
      garwood_direct_prefbi,
      garwood_direct_alt,
      garwood_direct_altbi
      
    ))  %>%   
  left_join(
    garwood_list %>% select(orig_name, garwood_synonymall = SYNONYMS),
    by = c("garwood_direct_match" = "orig_name")
  ) 


alternate_names_garwood <- alternate_names_garwood %>% filter(!is.na(garwood_direct_match))

datatable(alternate_names_garwood %>% select(
  prefname, altname, garwood_direct_match, garwood_synonymall
),
caption="Species that changed their identification. Pref name is their preferred name, altname is the name it used to have. Note that these are not all synonyms in the botanical sense."
)

```

#### Matching  plot species to bcilist so far

```{r bciplotmatch}
# 1. direct name/binomial match
bci_plots_garwood_name_match <- bci_plots %>%
  filter(
    orig_name %in% bcilist$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist$wcvp_accepted_name) |

    # match binomials
    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist$orig_binomial 
    )) |

    (!is.na(wcvp_accepted_binomial) & (
        wcvp_accepted_binomial %in% bcilist$wcvp_accepted_binomial 
    ))
  )


# Everything that did not match directly, match with synonyms

bci_plots_garwood_synonym_match1 <- bci_plots %>% 
  filter(
    !orig_name %in% bci_plots_garwood_name_match$orig_name &
    (
      orig_name %in% garwood_synonyms$synonym_sp |
      (!is.na(wcvp_matched_name) & wcvp_matched_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(wcvp_accepted_name) & wcvp_accepted_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(orig_binomial) & orig_binomial %in% garwood_synonyms$synonym_binomial) |
      (!is.na(wcvp_accepted_binomial) & wcvp_accepted_binomial %in% garwood_synonyms$synonym_binomial)
    )
  )

# add  the garwood synonym it matched to
bci_plots_garwood_synonym_print1 <- bci_plots_garwood_synonym_match1 %>%  left_join(
    garwood_synonyms %>% select(sp_orig = sp_name, synonym_sp, syn_orig = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_binomial = sp_name, synonym_sp, syn_binomial = SynonymTextALL) %>% distinct(),
    by = c("orig_binomial" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_match = sp_name, synonym_sp, syn_match = SynonymTextALL) %>% distinct(),
    by = c("wcvp_matched_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_wcvp = sp_name, synonym_sp, syn_wcvp = SynonymTextALL) %>% distinct(),
    by = c("wcvp_accepted_name" = "synonym_sp")) %>%
  mutate(
    SynonymMatchAll = coalesce(syn_orig, syn_match, syn_wcvp, syn_binomial),
    orig_garwood = coalesce(sp_orig, sp_match, sp_wcvp, sp_binomial),
  ) %>%
  select(-syn_orig, -syn_match, -syn_wcvp, -syn_binomial, -sp_orig, -sp_match, -sp_wcvp, -sp_binomial) %>%
  distinct(orig_name, orig_garwood, .keep_all = TRUE)


# everything that did not match to garwood synonyms, match to alternate names
bci_plots_garwood_synonym_match2 <- bci_plots %>% 
  filter(
    !orig_name %in% bci_plots_garwood_synonym_match1$orig_name &
      !orig_name %in% bci_plots_garwood_name_match$orig_name &
    (
      orig_name %in% alternate_names_garwood$prefname |
      orig_binomial %in% alternate_names_garwood$prefname |
      
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% alternate_names_garwood$prefname) |
      
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% alternate_names_garwood$prefname) |
        
      orig_name %in% alternate_names_garwood$altname |
      orig_binomial %in% alternate_names_garwood$altname |
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% alternate_names_garwood$altname) |
      
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% alternate_names_garwood$altname)
    )
  )

datatable(
  bci_plots_garwood_synonym_print1 %>% 
  select(orig_name, wcvp_matched_name, wcvp_accepted_name, orig_garwood ,SynonymMatchAll), 
          caption="BCI plot species that matched a garwood synonym",
          options = list(scrollX = TRUE, pageLength = 10))

```


```{r matchsyn2}
## add the altname it matched to and why that name is in garwood
bci_plots_garwood_synonym_print2 <- bci_plots_garwood_synonym_match2 %>%
  # A synonym or alternate name of the species is in the garwood list. 
  left_join(
    alternate_names_garwood %>% 
      select(prefname, altname, garwood_direct_match, garwood_synonymall),
    by = c("orig_name" = "prefname")
  ) %>%
  left_join(
    alternate_names_garwood %>% 
      select(prefname, altname, garwood_direct_match, garwood_synonymall),
    by = c("orig_binomial" = "prefname"),
    suffix = c("", "_bi")
  ) %>%
  left_join(
    alternate_names_garwood %>% 
      select(prefname, altname, garwood_direct_match, garwood_synonymall),
    by = c("wcvp_matched_name" = "prefname"),
    suffix = c("", "_match")
  ) %>%  
  left_join(
    alternate_names_garwood %>% 
      select(prefname, altname, garwood_direct_match, garwood_synonymall),
    by = c("wcvp_accepted_name" = "prefname"),
    suffix = c("", "_acc")
  ) %>%  
  
  # now the same but for altname
  left_join(
    alternate_names_garwood %>% 
      select(prefname, altname, altname, garwood_direct_match, garwood_synonymall),
    by = c("orig_name" = "altname"),
    suffix = c("", "_altorig")
  ) %>%
  left_join(
    alternate_names_garwood %>% 
      select(prefname, altname, garwood_direct_match, garwood_synonymall),
    by = c("orig_binomial" = "altname"),
    suffix = c("", "_altbi")
  ) %>%
  left_join(
    alternate_names_garwood %>% 
      select(prefname, altname, garwood_direct_match, garwood_synonymall),
    by = c("wcvp_matched_name" = "altname"),
    suffix = c("", "_altmatch")
  ) %>%  
  left_join(
    alternate_names_garwood %>% 
      select(prefname, altname, garwood_direct_match, garwood_synonymall),
    by = c("wcvp_accepted_name" = "altname"),
    suffix = c("", "_altacc")
  ) 

bci_plots_garwood_synonym_print2 <- bci_plots_garwood_synonym_print2 %>%
  mutate(
    # the garwood name it matched to or of the synonym it matched to
    orig_garwood = coalesce(
      garwood_direct_match,
      garwood_direct_match_bi,
      garwood_direct_match_match,
      garwood_direct_match_acc,
      garwood_direct_match_altorig,
      garwood_direct_match_altbi,
      garwood_direct_match_altmatch,
      garwood_direct_match_altacc
    ),
    
    alternate_name =  altname,
    prefered_name = prefname
    
  )%>%
  distinct(orig_name, orig_garwood, .keep_all = TRUE)



datatable(
  bci_plots_garwood_synonym_print2 %>% 
  select(orig_name, wcvp_matched_name, wcvp_accepted_name, orig_garwood,
         alternate_name, prefered_name, garwood_synonymall), 
          caption="BCI plot species where an alternate name of the species matched to garwood",
          options = list(scrollX = TRUE, pageLength = 10))


```

In this cases, the alternate name in forestgeo is more reliable than the name the garwood list. 

Therefore, we change the names in garwood to the names in forestgeo. 
```{r changegarwoodname}

# replace values in the bcilist to the values in the 
bci_plots_garwood_synonym_print2$source_list <- "GarwoodForestGeo"

corrections <- bci_plots_garwood_synonym_print2 %>%
    select(orig_garwood, all_of(listscol))

bcilist <- bcilist  %>%
  mutate(orig_garwood = orig_name)

# Update bcilist based on the matching 'orig_name'
bcilist <- bcilist %>%
  rows_update(corrections, by = "orig_garwood")



bci_plots_garwood_synonym_match <- bind_rows(bci_plots_garwood_synonym_match1, bci_plots_garwood_synonym_match2)
bci_plots_garwood_synonym_match <- bci_plots_garwood_synonym_match %>% mutate(matched_garwood_synonym = TRUE)

# combine
bci_plots_garwood_match <- bind_rows(
  bci_plots_garwood_name_match,
  bci_plots_garwood_synonym_match
) %>%
  distinct(sp6, .keep_all = TRUE)

```



```{r allnomatchbciplot}
# 4. unmatched
bci_plots_no_garwood_at_all <- bci_plots %>%
  filter(!sp6 %in% bci_plots_garwood_match$sp6)

datatable(bci_plots_no_garwood_at_all  %>% select(sp6, orig_name, wcvp_matched_name, wcvp_accepted_name, plot_name), 
          caption="Species recorded in BCI plots census that don't match anything in Garwood",
          options = list(scrollX = TRUE, pageLength = 10))



datatable(garwood_list_orig %>% 
  filter(orig_name %in% bci_plots_no_garwood_at_all$orig_name) %>%
  select(orig_name, chg), 
          caption="Some of these species we removed earlier...",
          options = list(scrollX = TRUE, pageLength = 10))


write_xlsx(bci_plots_no_garwood_at_all, FILECHECKPLOTS)

```

According to the garwood notes Damburneya purpurea is Nectandra purpurea. But maybe garwood was not sure?

Add to bci list
```{r}

bci_plots_no_garwood_at_all$source_list = "PlotCensusBCI"


bcilist <- bind_rows(
  bcilist,
  bci_plots_no_garwood_at_all %>% select(all_of(c(listscol, "plot_name", "bci50ha", "bci10ha", "P11", "P14"))
))

```


### Lianas BCI list
```{r lianasnobcilist}

lianas_list <- lianas_list %>% filter(isbci=="TRUE")
lianas_list <- lianas_list %>% distinct(orig_name, .keep_all = TRUE)
lianas_list$wcvp_accepted_binomial <- paste0(
  lianas_list$wcvp_accepted_name_genus, " ",
  lianas_list$wcvp_accepted_name_species
)

# 1. direct name/binomial match
lianas_garwood_name_match <- lianas_list %>%
  filter(

    orig_name %in% bcilist$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist$wcvp_accepted_name) |

    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist$orig_binomial 
    )) |

    (!is.na(wcvp_accepted_binomial) & (
        wcvp_accepted_binomial %in% bcilist$wcvp_accepted_binomial 
    ))
  )

# 2. synonym_sp for remaining
lianas_garwood_synonym_match <- lianas_list %>%
  filter(
    !orig_name %in% lianas_garwood_name_match$orig_name &
    (
      orig_name %in% garwood_synonyms$synonym_sp |
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(orig_binomial) & orig_binomial %in% garwood_synonyms$synonym_binomial) |
      (!is.na(wcvp_accepted_binomial) & wcvp_accepted_binomial %in% garwood_synonyms$synonym_binomial)
    )
  )

lianas_garwood_synonym_match$matched_garwood_synonym <- TRUE


# see species that matched by a synonym 
lianas_garwood_synonym_match_print <- lianas_garwood_synonym_match %>%  left_join(
    garwood_synonyms %>% select(sp_orig = sp_name, synonym_sp, syn_orig = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_match = sp_name, synonym_sp, syn_match = SynonymTextALL) %>% distinct(),
    by = c("wcvp_matched_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_wcvp = sp_name, synonym_sp, syn_wcvp = SynonymTextALL) %>% distinct(),
    by = c("wcvp_accepted_name" = "synonym_sp")) %>%
  mutate(
    SynonymMatchAll = coalesce(syn_orig, syn_match, syn_wcvp),
    orig_garwood = coalesce(sp_orig, sp_match, sp_wcvp),
  ) %>%
  select(-syn_orig, -syn_match, -syn_wcvp,  -sp_orig, -sp_match, -sp_wcvp) %>%
  distinct(orig_name, .keep_all = TRUE)
  

datatable(lianas_garwood_synonym_match_print %>% 
            select(orig_name, wcvp_matched_name, wcvp_accepted_name, SynonymMatchAll, orig_garwood), 
          caption="Liana species that matched garwood by synonym",
          options = list(scrollX = TRUE, pageLength = 10))


lianas_garwood_match <- bind_rows(
  lianas_garwood_name_match,
  lianas_garwood_synonym_match
) %>%
  distinct(orig_name, .keep_all = TRUE)

lianas_no_garwood_at_all <- lianas_list %>%
  filter(!orig_name %in% lianas_garwood_match$orig_name)



# 5. Output and Display
datatable(lianas_no_garwood_at_all %>%
  select(orig_name, wcvp_matched_name, wcvp_accepted_name), 
          caption = "Lianas species that don't match to anything in Garwood",
          options = list(scrollX = TRUE, pageLength = 10))

write_xlsx(lianas_no_garwood_at_all, FILECHECKLIANAS)

```

Add to bci list
```{r lianasbciadd}

lianas_no_garwood_at_all$source_list = "lianas"

bcilist <- bind_rows(
  bcilist,
  lianas_no_garwood_at_all %>% select(all_of(listscol))
)

```

### Check consistency 
```{r savebotanists}

# Check consistency
# removing rows that are not identified at species level
datatable(bcilist %>% filter(is.na(word(orig_binomial, 2))) %>% select(orig_name, wcvp_accepted_name), caption="Species to be removed from final list because they are not fully identified" )

bcilist <- bcilist %>% filter(!is.na(word(orig_binomial, 2))) 


datatable(bcilist %>% group_by(orig_name) %>% filter(n()> 1) %>%
          select(orig_name, wcvp_accepted_name) %>% arrange(orig_name),
          caption="Species duplicated original name",
  options = list(scrollX = TRUE, pageLength = 10)
          )

datatable(bcilist %>% group_by(wcvp_accepted_name) %>% 
            filter(n()> 1) %>% select(orig_name, wcvp_accepted_name)%>% 
            arrange(wcvp_accepted_name), 
          caption="Species duplicated wcvp_accepted name",
  options = list(scrollX = TRUE, pageLength = 10)
          )



datatable(
  bcilist %>%
    select(wcvp_accepted_name, orig_name, source_list) %>% 
    arrange(source_list),
    caption="Species list in BCI and their source",
  options = list(scrollX = TRUE, pageLength = 10))

```


### Add woody species column
 
If the species is in the woody list

```{r bcilistwoody, echo=TRUE}
# set woody if:
# original, matched or accepted name is anywhere in woody list

bcilist <- bcilist %>% mutate(
  woody_name =  wcvp_accepted_name %in% panwoody_list$current_name | 
                wcvp_accepted_name %in% panwoody_list$orig_name | 
                wcvp_accepted_binomial %in% panwoody_list$orig_binomial |
                wcvp_accepted_binomial %in% panwoody_list$current_name |
                wcvp_matched_name %in% panwoody_list$current_name |
                wcvp_matched_name %in% panwoody_list$orig_name |
                orig_name %in% panwoody_list$orig_name |
                orig_name %in% panwoody_list$current_name |
                orig_binomial %in% panwoody_list$orig_binomial,
  
  woody_synonym =  wcvp_accepted_name %in% woody_synonyms$SynonymTextSingle |
                   orig_name %in% woody_synonyms$SynonymTextSingle |
                   wcvp_matched_name %in% woody_synonyms$SynonymTextSingle,
  woody = woody_name | woody_synonym
) 

# BCI BOTANISTS HIGH QUALITY LIST
write_xlsx(bcilist, FILECOMBINEDLIST)


bcilist_woody <- bcilist %>%
  filter(woody) %>%
  select(orig_name, wcvp_matched_name, wcvp_accepted_name, wcvp_accepted_family, wcvp_lifeform_description)

datatable(bcilist_woody, caption="BCI species that are in the woody list")

```

### Possibly woody species

```{r posiblywoody}

lifeform_woody = c(
  "tree",
  "shrub",
  "liana",
  "climbing shrub",
  "shrub or tree",
  "climbing shrub or tree",
  "scrambling shrub or tree",
  "subshrub or shrub",
  "subshrub, shrub or tree",
  "scrambling shrub",
  "perennial or shrub",
  "scrambling subshrub or shrub",
  "epiphytic shrub",
  "scrambling tree",
  "climbing shrub or liana",
  "scrambling shrub or liana",
  "tuberous shrub or tree",
  "herbaceous tree",
  "climbing herbaceous tree",
  "climbing herbaceous tree or liana",
  "climbing perennial or herbaceous tree",
  "subshrub or climbing shrub",
  "tuberous subshrub",
  "hemiparasitic shrub",
  "hemiepiphytic shrub or tree",
  "annual or shrub",
  "annual, subshrub or shrub"
)

bcilist_possibly_woody <- bcilist %>%
  filter(!woody &
          wcvp_lifeform_description %in% lifeform_woody ) %>%
  select(orig_name, wcvp_matched_name, wcvp_accepted_name, wcvp_accepted_family, wcvp_lifeform_description)

datatable(bcilist_possibly_woody, caption="BCI species that might be woody according to the wcvp lifeform")

write_xlsx(bcilist_possibly_woody, FILEBCIWOODY)


```


## Panamabiota species that are not in garwood

```{r pbiotanogarwood}

pbiota_list$wcvp_accepted_binomial <- paste0(
  pbiota_list$wcvp_accepted_name_genus, " ",
  pbiota_list$wcvp_accepted_name_species
)

# 1. direct name/binomial match
pbiota_garwood_name_match <- pbiota_list %>%
  filter(

    orig_name %in% bcilist$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist$wcvp_accepted_name) |

    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist$orig_binomial 
    )) |

    (!is.na(wcvp_accepted_binomial) & (
        wcvp_accepted_binomial %in% bcilist$wcvp_accepted_binomial 
    ))
  )

# 2. synonym_sp for remaining
pbiota_garwood_synonym_match <- pbiota_list %>%
  filter(
    !orig_name %in% pbiota_garwood_name_match$orig_name &
    (
      orig_name %in% garwood_synonyms$synonym_sp |
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(orig_binomial) & orig_binomial %in% garwood_synonyms$synonym_binomial) |
      (!is.na(wcvp_accepted_binomial) & wcvp_accepted_binomial %in% garwood_synonyms$synonym_binomial)
    )
  )

pbiota_garwood_synonym_match$matched_garwood_synonym <- TRUE


# see species that matched by a synonym 
pbiota_garwood_synonym_match_print <- pbiota_garwood_synonym_match %>%  left_join(
    garwood_synonyms %>% select(sp_orig = sp_name, synonym_sp, syn_orig = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_match = sp_name, synonym_sp, syn_match = SynonymTextALL) %>% distinct(),
    by = c("wcvp_matched_name" = "synonym_sp"), relationship = "many-to-many"
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_wcvp = sp_name, synonym_sp, syn_wcvp = SynonymTextALL) %>% distinct(),
    by = c("wcvp_accepted_name" = "synonym_sp"),  relationship = "many-to-many") %>%
  mutate(
    SynonymMatchAll = coalesce(syn_orig, syn_match, syn_wcvp),
    orig_garwood = coalesce(sp_orig, sp_match, sp_wcvp),
  ) %>%
  select(-syn_orig, -syn_match, -syn_wcvp,  -sp_orig, -sp_match, -sp_wcvp) %>%
  distinct(orig_name, .keep_all = TRUE)
  

datatable(pbiota_garwood_synonym_match_print %>% 
            select(orig_name, wcvp_matched_name, wcvp_accepted_name, SynonymMatchAll, orig_garwood), 
          caption="Panama biota species that matched garwood by synonym",
          options = list(scrollX = TRUE, pageLength = 10))

pbiota_garwood_match <- bind_rows(
  pbiota_garwood_name_match,
  pbiota_garwood_synonym_match
) %>%
  distinct(orig_name, .keep_all = TRUE)

pbiota_no_garwood_at_all <- pbiota_list %>%
  filter(!orig_name %in% pbiota_garwood_match$orig_name)



datatable(pbiota_no_garwood_at_all, 
          caption="Panama biota species that don't match to anything in Garwood",
          options = list(scrollX = TRUE, pageLength = 10))

write_xlsx(pbiota_no_garwood_at_all, FILECHECKPANAMABIOTA)

```


