---
title: "Assemble plant species list for Barro Colorado Island, Panama"
author: "Paula Uzcategui and Helene Muller-Landau"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

This Rmarkdown merges information from multiple high-quality plant species lists for Barro Colorado Island (BCI), Panama, updates to the best current woody plant species names based on the current Smithsonian ForestGEO woody species lists, utilizes prior checks of species names against the World Checklist of Vascular Plants (WCVP) version 13, and outputs the combined high-quality list. The contributing lists are Garwood et al. 2009, Schnitzer et al. 2024, Zotz 2024, the Smithsonian ForestGEO Panama woody plant species and ViewTaxonomy lists (Perez et al. 2025), and a file of tree species found in forest inventory plots on BCI.  These lists are considered high-quality because they were each assembled by one or more botanists familiar with the local flora, and build on the work of Croat 1978. Combining the lists involves checking overlap not only in the original names on each list, but also on two lists of "synonyms" for BCI plant species, as well as in the current accepted name.  The synonym lists include one from Garwood 2009 and another from Smithsonian ForestGEO.  In both cases, these include not only synonyms in the botanical sense, but also prior names used for the same BCI morphospecies.  

This Rmarkdown additionally outputs a list of plant species that appear in panamabiota.org's list of plants of Barro Colorado, that are *not* in the combined high-quality list. This list is generated for the purpose of later review.  

The component and output lists are all available in the public GitHub site https://github.com/PanamaForestGEO/Panama_plant_species_lists 


## References 

Croat, Thomas B. 1978. *Flora of Barro Colorado Island*. Stanford University Press.

Garwood, Nancy C., Nancy Hensold, and Robin B. Foster. 2009. Appendix 1.  Revised Species List for Barro Colorado Island, with Synonyms. In *Seedlings of Barro Colorado Island and the Neotropics*. Cornell University Press.

Pérez, Rolando, Salomón Aguilar and David Mitre.  2025. Lista de árboles, arbustos, palmas, lianas y enredaderas de la flora de Panamá.  Version 2025-12-04.  

Schnitzer, Stefan, David DeFilippis, Rigoberto Rivera-Camaña, et al. 2024. Liana Species of the Barro Colorado Island 50-Ha Plot 2007 and 2017 Censuses. Version 7. *Dryad* https://doi.org/10.5061/DRYAD.Q2BVQ83V4.

Zotz, Gerhard.  2024.  Table S1. Updated list of epiphyte species on BCI. Online Supplementary Material for Zotz, G. 2024. Epiphyte Research on BCI. In *The First 100 Years of Research on Barro Colorado: Plant and Ecosystem Science*, ed. H. C. Muller-Landau and S. J. Wright. Smithsonian Institution Scholarly Press. https://doi.org/10.5479/si.26880865

### To do 

Address cases with Prefix %in% c("X", "sp. aff.") in the Garwood list.  This means the morphospecies is a hybrid.  Our current workflow ignores this info and is constructing "orig_name" and orig_binomial without including the prefix, leading to matches on the name without the prefix, but that is incorrect.  We should keep the prefix in the orig_name and the orig_binomial, and remove these from the taxa that are matched under WCVP.  The Garwood name should be kept (these aren't very important as they are mostly cultivated species and there aren't many of them). 

When reading in files, it would be good to rename columns right away if necessary to get rid of blanks, /’s, and other problematic characters, so subsequent references can be simplified.


## Setup

### Load libraries
```{r setup2,}

rm(list=ls())

library(readxl)
library(writexl)
library(dplyr)
library(tidyr)
library(stringi)
library(stringr)
library(knitr)
library(DT)
```

### Read input files and initial data prep

```{r readprepdata, echo=TRUE, results="hide"}

garwood_list <- garwood_list_orig <- 
  read_excel("splists_out/Garwood_2009_Appendix1HM2_WCVP_2026-02-26.xlsx", col_types = "text")

zotz_list <- read_excel("splists_out/83_Zotz_Epiphytes_TableS1_WCVP_2026-02-26.xlsx", col_types = "text") %>%
  rename_with(~ gsub("[()]", "", gsub(" ", "_", .x)))

lianas_list <- read_excel("splists_out/Lianas_Species_complete_list_WCVP_2026-03-03.xlsx", col_types = "text") %>%
  mutate(sp6 = tolower(sp6)) %>%
  rename(wcvp_accepted_lifeform=wcvp_lifeform_description)

pbiota_list <- read_excel("splists_out/PanamaBiotaBCI_WCVP_2026-02-26.xlsx", col_types = "text")

panwoody_list <- read_excel("splists_out/CurrentPanamaWoody_2026-03-09.xlsx", col_types = "text")
panwoody_synonyms <- read_excel("splists_out/CurrentPanamaWoody_SYNONYMS_2026-03-09.xlsx", col_types="text")

viewtaxa_list <- read_excel("splists_raw/Forestgeo/2026-02-25FromSuzanne/ViewTaxonomy_dec4_25_HM2.xlsx")
viewtaxa_synonyms <- read_excel("splists_out/ViewTaxonomy_dec4_25_SYNONYMS.xlsx")

alternate_names <- read_excel("splists_out/AltNames_2026-03-06.xlsx")

garwood_synonyms <- read_excel("splists_raw/Garwood/Garwood_2009_Appendix1_Synonyms.xlsx") 

garwood_synonyms <- garwood_synonyms %>% mutate(
  sp_name = trimws(paste0(Genus, " ", Species)),
  synonym_sp = trimws(paste0(SynonymGenusSingle, " ", SynonymSpeciesSingle)),
  synonym_binomial = paste0(SynonymGenusSingle, " ", SynonymSpeciesSingle)) 
# note that synonym_sp and synonym_binomial are identical; retaining for clarity of completeness

garwood_list <- garwood_list %>%
  mutate(wcvp_accepted_binomial = 
           paste0(wcvp_accepted_name_genus, " ", wcvp_accepted_name_species),
         wcvp_matched_binomial = word(wcvp_matched_name,)) %>%
  rename(wcvp_accepted_lifeform = wcvp_lifeform_description)
```

### Output file names 

MAIN OUTPUTS: 

- Combined, high-quality list of BCI plant species.  

- List of BCI species that might be woody according to WCVP lifeform. 

ADDITIONAL OUTPUTS OF FILES TO CHECK: 

- List of Zotz species not in Garwood

- List of BCI plot species not in Garwood

- List of BCI liana list species not in Garwood

- List of PanamaBiota BCI species that are not in the combined high-quality list

```{r outfilenames}
save_plantlists_dir = "splists_out/Plantnet/"

FILECOMBINEDLIST <- paste0(save_plantlists_dir, 
                               "BCI_SPECIES_BOTANISTS_LIST_",Sys.Date(),".xlsx")
FILEBCIWOODY <- paste0(save_plantlists_dir, 
                           "BCI_SPECIES_POSSIBLY_WOODY_",Sys.Date(),".xlsx")

FILECHECKZOTZ <- "tocheck/zotz_species_not_in_garwood.xlsx"

FILECHECKPLOTS <-  "tocheck/plotsbci_not_garwood.xlsx"

FILECHECKLIANAS <- "tocheck/lianasbci_not_garwood.xlsx"

FILECHECKPANAMABIOTA <- "tocheck/panamabiota_not_garwood.xlsx"

```



## Assembling a combined BCI high-quality botanist-determined plant species list

### Garwood 2009 list



### Remove non-BCI species from Garwood list

Remove species that are listed as being in the Barro Colorado Nature Monument but not on BCI itself (chg=="BCNM").

```{r garwoodnobcnm}


# filter out species that are in the mainland peninsulas but not BCI, except for those that turn out to be on BCI after all...

yesonbcibinomial <- c("Matayba oppositifolia", "Myrcia zetekiana")

garwood_list <- garwood_list %>% filter(chg != "BCNM" | is.na(chg) | orig_name %in% yesonbcibinomial)
garwoodtoremove <- subset(garwood_list_orig, chg=="BCNM" & 
                            !orig_binomial %in% yesonbcibinomial)

kable(garwoodtoremove %>% select(orig_name,SYNONYMS),
      caption = "Species dropped from Garwood because they are in the mainland peninsulas but not on BCI itself (at least according to Garwood).")

kable(data.frame(binomial=yesonbcibinomial),caption="Species that Garwood lists as only in the mainland peninsulas that are nonetheless kept because they turn out to be in other lists for BCI")
```

```{r garwoodnoid}

# view unidentified species
garwoodnobinomial <- subset(garwood_list_orig,is.na(word(orig_binomial,2)))

kable(garwoodnobinomial %>% select(orig_name,Syn_notes) %>% rename(garwood_name=orig_name),
      caption="Garwood species lacking a complete binomial")

```

This leaves `r nrow(garwood_list)` plant species on the Garwood list.  Note that in `r nrow(garwoodnobinomial)` cases these species lack proper binomials, indicating their species identity remains uncertain (including possible new species).  These are left in anyway because they are known morphospecies.  


### Zotz 2024 epiphyte species list

There are `r nrow(zotz_list)` species on the Zotz 2024 epiphyte list.  

Match Zotz species to the Garwood list as far as possible, including by synonynms in the broad sense, and identify Zotz species not on the Garwood list.  

Note that we make one manual correction to a Zotz misspelling of a Croat name to enable matching with the Garwood list for that name.  Other misspellings to the Zotz names themselves are addressed by the matching with WCVP.  
```{r zotznogarwood}
# removing morphospecies identified only at the genus level (there is only one)
zotz_list <- zotz_list %>% filter(is.na(orig_binomial) | !is.na(word(orig_binomial,2) ))

# when the name listed is the same as in Croat 1978, the column Sp_Croat_1978 is blank (NA)
# so create a new name_Croat field that is the name of the species in Croat 1978
zotz_list <- zotz_list %>%
  mutate(wcvp_accepted_binomial = 
           paste0(wcvp_accepted_name_genus," ",wcvp_accepted_name_species),
         wcvp_matched_binomial = word(wcvp_matched_name,1,2),
         croat_name = ifelse(!is.na(Sp_Croat_1978),Sp_Croat_1978,orig_name)) %>%
  mutate(croat_binomial = word(croat_name,1,2)) %>%
  rename(wcvp_accepted_lifeform=wcvp_lifeform_description)

# make one manual correction so matches work right
zotz_list <- zotz_list %>%
  mutate(croat_binomial =ifelse(croat_binomial=="Bolbiltis nicotianifolia",
         "Bolbitis nicotianifolia",croat_binomial))

# matching is done on the binomials (not full names including subspecies)

zotz_list <- zotz_list %>%
  mutate(matchorig = 
           orig_binomial %in% garwood_list$orig_binomial,
         matchcroat = 
           croat_binomial %in% garwood_list$orig_binomial,
         matchorigsynonym = 
           orig_binomial %in% garwood_synonyms$synonym_binomial,
         matchcroatsynonym =
           croat_binomial %in% garwood_synonyms$synonym_binomial,
         matchwcvpmatch = 
           !is.na(wcvp_matched_name) & 
              (wcvp_matched_binomial %in% c(garwood_list$wcvp_matched_binomial,
                                        garwood_list$orig_binomial,
                                        garwood_synonyms$synonym_binomial)),
         matchwcvpaccept = 
           !is.na(wcvp_accepted_name) & 
              wcvp_accepted_binomial %in% c(garwood_list$wcvp_matched_binomial,
                                        garwood_list$orig_binomial,
                                        garwood_synonyms$synonym_binomial)
  ) %>%
  mutate(matchany = matchorig | matchcroat | matchorigsynonym |matchcroatsynonym | 
           matchwcvpmatch | matchwcvpaccept) %>% 
  mutate(firstmatchtype=ifelse(matchorig,"orig-orig",
                               ifelse(matchcroat,"croat-orig",
                                      ifelse(matchorigsynonym,"orig-syn",
                                             ifelse(matchcroatsynonym,"croat-syn",
                                                    ifelse(matchwcvpmatch,"wcvpmatch",
                                                           ifelse(matchwcvpaccept,"wcvpaccpt",
                                                                  NA)))))))
           
zotz_nomatchgarwood <- zotz_list %>%  filter(!matchany)

write_xlsx(zotz_nomatchgarwood, FILECHECKZOTZ)

# check the species that match only on the WCVP names
checkmatch1 <- subset(zotz_list, matchany & 
                        !matchcroat & !matchorigsynonym & !matchorig & !matchcroatsynonym) %>%
  select(orig_name,match_similarity,wcvp_matched_name,wcvp_accepted_name,Sp_Croat_1978)
kable(checkmatch1,
      caption="Species in Zotz that match Garwood only via WCVP matched or accepted names.  These are all cases in which the name is misspelled in Zotz.")

# see species that matched by a synonym and not by original or Croat names
checkmatch2a <- zotz_list %>% filter(matchorigsynonym & !matchorig & !matchcroat) %>%  
  left_join(
    garwood_synonyms %>% select(garwood_orig = sp_name, synonym_binomial, syn_text = SynonymTextALL) %>% distinct(),
    by = c("orig_binomial" = "synonym_binomial")
  )%>% 
  select(orig_binomial,croat_binomial,garwood_orig,firstmatchtype,syn_text)

checkmatch2b <- zotz_list %>% filter(matchcroatsynonym & !matchorig & !matchcroat) %>%  
  left_join(
    garwood_synonyms %>% select(garwood_orig = sp_name, synonym_binomial, syn_text = SynonymTextALL) %>% distinct(),
    by = c("croat_binomial" = "synonym_binomial")
  ) %>% mutate(matchtype="croat-to-synonym") %>%
  select(orig_binomial,croat_binomial,garwood_orig,firstmatchtype,syn_text)

checkmatch2 <- rbind(checkmatch2a,
                     subset(checkmatch2b,!orig_binomial %in% checkmatch2a$orig_binomial))

kable(checkmatch2, 
          caption="Zotz species that matched Garwood only by synonym. Note that in some cases there was more than one match by synonym",
          options = list(scrollX = TRUE, pageLength = 10))

zotznocroatbutmatched <- zotz_list %>%
  filter(Sp_Croat_1978=="**" & matchany==T) %>%
  arrange(firstmatchtype,orig_name) %>%
  select(orig_name,wcvp_matched_name,wcvp_accepted_name,firstmatchtype,Sp_Croat_1978)

kable(zotznocroatbutmatched,
      caption="Zotz species that Zotz says do not appear in Croat 1978 that nonetheless
      matched in Garwood.  ")
  
kable(zotz_nomatchgarwood  %>% select(orig_name, wcvp_matched_name, wcvp_accepted_name, Sp_Croat_1978), 
          caption="Zotz species that don't match anything in Garwood.  Note that in all cases Sp_Croat_1978 has value **, indicating that Zotz 2024 states that the species does not appear in Croat 1978.")

```

In total there are `r nrow(zotz_nomatchgarwood)` species on the Zotz list not already on the Garwood list.  

Add these Zotz species to the combined BCI list, which starts with Garwood.
```{r}
# using garwood as a base
garwood_list$source_list = "Garwood2009"
zotz_nomatchgarwood$source_list = "Zotz"

listscol <- c("orig_name",
              "orig_binomial",
              "wcvp_accepted_binomial",
              "wcvp_matched_name",
              "wcvp_matched_plant_name_id",
              "wcvp_matched_ipni_id",
              "wcvp_matched_status",
              "wcvp_accepted_name",
              "wcvp_accepted_family", 
              "wcvp_accepted_authority",
              "wcvp_accepted_plant_name_id", 
              "wcvp_accepted_powo_id",
              "wcvp_accepted_ipni_id",
              "wcvp_accepted_lifeform",
              "source_list")

bcilist <- bind_rows(
  garwood_list %>% 
    select(all_of(c(listscol, "SYNONYMS"))),
  zotz_nomatchgarwood %>% select(all_of(listscol))
)

```


# Preparing alternate names
The pre-existing "altnerate_names" file combines species names in the synonyms or "old names" columns of currentwoody, viewtaxa and garwood. 

These are not necessarily synonyms in the botanical sense; they are any names formerly used for a given BCI morphospecies (e.g., due to misidentification) 

If a species in the BCI plots or in the lianas list does not match the garwood directly, an alternate name related to the species could match. 

Here we convert the alternate names list filtering by matches with garwood. We will use this list to check for garwood presence in the BCI plot species and the lianas species list that dont match directly with garwood. 

```{r altnames}

# Here we create a list of the synonyms, and if they are somewhere in the woody list, we add the name of origin y woody and the synonym they have in Garwood.
alternate_names_garwood <- alternate_names %>% 
  # DIRECT MATCH — prefname to Garwood orig_name
  left_join(
    garwood_list %>% select(garwood_direct_pref_join=orig_name, 
                            garwood_direct_pref=orig_name),
    by = c("prefname" = "garwood_direct_pref_join")
  ) %>%
  # DIRECT MATCH — prefname to Garwood orig_binomial
  left_join(
    garwood_list %>% select(garwood_direct_prefbi_join=orig_binomial, 
                            garwood_direct_prefbi=orig_name),
    by = c("prefname" = "garwood_direct_prefbi_join"), relationship = "many-to-many"
  ) %>%
  # DIRECT MATCH — altname to Garwood orig_name
  left_join(
    garwood_list %>% select(garwood_direct_alt_join=orig_name, 
                            garwood_direct_alt=orig_name),
    by = c("altname" = "garwood_direct_alt_join"), relationship = "many-to-many"
  ) %>%
  # DIRECT MATCH — altname to Garwood orig_binomial
  left_join(
    garwood_list %>% select(garwood_direct_altbi_join=orig_binomial, 
                            garwood_direct_altbi=orig_name),
    by = c("altname" = "garwood_direct_altbi_join"), relationship = "many-to-many"
  ) %>%
  
  # SYNONYM MATCH — prefname to Garwood synonym full 
  left_join(
    garwood_synonyms %>%
      select(
        synonym_full_join = synonym_sp,  # note - change in variable selected by Helene
        garwood_syn_orig_pref = sp_name,
        synonym_pref_full = synonym_sp,
      ),
    by = c("prefname" = "synonym_full_join"), relationship = "many-to-many"
  ) %>%
  # SYNONYM MATCH — prefname to Garwood synonym binomial [added by Helene]
  left_join(
    garwood_synonyms %>%
      select(
        synonym_bi_join = synonym_binomial,
        garwood_syn_bi_pref = sp_name,
        synonym_pref_bi = synonym_binomial,
      ),
    by = c("prefname" = "synonym_bi_join"), relationship = "many-to-many"
  ) %>%
   # SYNONYM MATCH — prefname to Garwood synonym full 
  left_join(
    garwood_synonyms %>%
      select(
        synonym_full_join = synonym_sp,  # note - change in variable selected by Helene
        garwood_syn_orig_alt = sp_name,
        synonym_alt_full = synonym_sp,
      ),
    by = c("altname" = "synonym_full_join"), relationship = "many-to-many"
  ) %>%
  # SYNONYM MATCH — prefname to Garwood synonym binomial [added by Helene]
  left_join(
    garwood_synonyms %>%
      select(
        synonym_bi_join = synonym_binomial,
        garwood_syn_bi_alt = sp_name,
        synonym_alt_bi = synonym_binomial,
      ),
    by = c("altname" = "synonym_bi_join"), relationship = "many-to-many"
  ) 

alternate_names_garwood <- alternate_names_garwood %>%
  mutate(
    # Garwood original name (the accepted/canonical name)
    garwood_orig_match = coalesce(
      garwood_direct_pref,
      garwood_direct_prefbi,
      garwood_direct_alt,
      garwood_direct_altbi,
      garwood_syn_orig_pref,
      garwood_syn_bi_pref,
      garwood_syn_orig_alt,
      garwood_syn_bi_alt
    ),
    
    # Garwood synonym (the specific synonym used to match)
    garwood_synonym_match = coalesce(
      synonym_pref_full,
      synonym_pref_bi,
      synonym_alt_full,
      synonym_alt_bi
    ),
    
    # Source / reason of the match following the hierarchy above
    firstmatchtype = case_when(
      in_garwood2009 ~ "direct synonym",
      !is.na(garwood_direct_pref)    ~ "pref-orig",
      !is.na(garwood_direct_prefbi)  ~ "pref-origbi",
      !is.na(garwood_direct_alt)     ~ "alt-orig",
      !is.na(garwood_direct_altbi)   ~ "alt-origbi",
      !is.na(garwood_syn_orig_pref)  ~ "pref-syn",
      !is.na(garwood_syn_bi_pref)    ~ "pref-synbi",
      !is.na(garwood_syn_orig_alt)   ~ "alt-syn",
      !is.na(garwood_syn_bi_alt)     ~ "alt-synbi",
      TRUE                           ~ NA_character_
    ),
  
  )

alternate_names_garwood <- alternate_names_garwood %>% filter(!is.na(garwood_orig_match) | in_garwood2009)

datatable(alternate_names_garwood %>% select(
  Current_name = prefname,
  Former_name = altname, 
  Garwood_original_name = garwood_orig_match, 
  Garwood_synonym = garwood_synonym_match, firstmatchtype, in_garwood2009
),
caption="List of alternate names that are in Garwood."
)

# summarize match types
match_summary <- alternate_names_garwood %>%
  group_by(firstmatchtype) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

knitr::kable(match_summary, caption="Summary of matches")

# Create a single mapping table that captures all match directions
alt_match_lookup <- bind_rows(
  alternate_names_garwood %>% select(match_key = prefname, garwood_orig_match, garwood_synonym_match, firstmatchtype, alt_name = altname, pref_name = prefname),
  alternate_names_garwood %>% select(match_key = altname,  garwood_orig_match, garwood_synonym_match,firstmatchtype, alt_name = altname, pref_name = prefname)
) %>% 
  filter(!is.na(match_key)) %>%
  distinct(match_key, .keep_all = TRUE)


```


### List of tree species recorded in censuses of forest plots on BCI 

Check that all the tree species recorded from forest plots on BCI are also in the combined species list.  

```{r plotslist}
# Preparing plotlist

# 1. Read data
bci_plots <- read_excel("splists_raw/Plots/abundance_plots_HM.xlsx")

bci_plots <- bci_plots %>% select(sp, binomial, bci, `bci 10 ha plot`, P11, P14)

bci_other_plots <- read.delim(
  "splists_raw/Plots/doi_10_5061_dryad_1g1jwsvc3__v20260226/mainfile.txt"
)

# 2. Summarise other plots safely
bci_other_plots <- bci_other_plots %>%
  group_by(sp6) %>%
  summarise(
    bigplot = n_distinct(tag),
    .groups = "drop"
  )

# 3. Clean and rename BEFORE joining
bci_plots <- bci_plots %>%
  rename(
    sp6 = sp,
    sp_name_plot = binomial,
    bci50ha = bci,
    bci10ha = `bci 10 ha plot`
  )

# 4. Join safely (keeps all bci_plots species)
bci_plots <- bci_plots %>%
  left_join(bci_other_plots, by = "sp6") %>%
  mutate(
    bigplot = coalesce(bigplot, 0)   # replace NA with 0
  )

# 5. Filter safely (handle NA just in case)
bci_plots <- bci_plots %>%
  mutate(across(c(bci50ha, bci10ha, P14, P11), ~coalesce(., 0))) %>%
  filter(
    bci50ha > 0 |
    bci10ha > 0 |
    P14 > 0 |
    P11 > 0 |
    bigplot > 0
  )

# 6. Create plot_name column safely
bci_plots <- bci_plots %>%
  rowwise() %>%
  mutate(
    plot_name = paste(
      c(
        if (bci50ha > 0) "bci50ha",
        if (bigplot > 0) "bcimultiplots",
        if (bci10ha > 0) "bci10ha",
        if (P14 > 0) "P14",
        if (P11 > 0) "P11"
      ),
      collapse = ", "
    )
  ) %>%
  ungroup()

# the plots files don't have wcvp accepted name, merge them by 6-letter species code in the Panama woody species list
match21 <- match(bci_plots$sp6, panwoody_list$sp6)            
nomatchview <- bci_plots[is.na(match21),] %>% select(sp6, sp_name_plot) # Those that don't match are all unidentified

if (nrow(nomatchview)>0)
knitr::kable(nomatchview, caption="Warning: Species recorded in plots on BCI whose 6-letter codes do not have a match in the Panama ForestGEO woody plant species database.")

bci_plots <- bci_plots %>% left_join(
  panwoody_list, by="sp6"
)

# if no orig name add the binomial
bci_plots <- bci_plots %>% mutate(orig_name = ifelse(is.na(orig_name), sp_name_plot, orig_name))

# remove the row for "unidentified", if present
bci_plots <- bci_plots %>% filter(sp6!="uniden")

bcidups <- bci_plots %>% filter(duplicated(sp6)) %>% select(sp6, sp_name_plot)

if (nrow(bcidups)>0)
knitr::kable(bcidups, caption="Warning: Some 6-letter species codes are repeated in the bci_plots data.frame.")

# create binomial first if not already
bci_plots$wcvp_accepted_binomial <- paste0(
  bci_plots$wcvp_accepted_genus, " ",
  bci_plots$wcvp_accepted_species
)

```
In total there are `r nrow(bci_plots)` tree morphospecies recorded for forest plot censuses on BCI.



Matching to original garwood names (direct match) or to synonyms or alternate names that match to garwood.
```{r bciplotmatch}
# 1. direct name/binomial match
bci_plots_garwood_name_match <- bci_plots %>%
  filter(
    orig_name %in% bcilist$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist$wcvp_accepted_name) |

    # match binomials
    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist$orig_binomial 
    )) |

    (!is.na(wcvp_accepted_binomial) & (
        wcvp_accepted_binomial %in% bcilist$wcvp_accepted_binomial 
    ))
  )


# 2. everything that did not match directly to garwood, match to alternate names
bci_plots_garwood_synonym_match <- bci_plots %>% 
  filter(
      !orig_name %in% bci_plots_garwood_name_match$orig_name &
    (
      orig_name %in% alt_match_lookup$match_key |
      orig_binomial %in% alt_match_lookup$match_key |
      
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% alt_match_lookup$match_key) |
        
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% alt_match_lookup$match_key)
    )
  )


# 3. pront the altname it matched to and why that name is in garwood

bci_plots_garwood_synonym_print <- bci_plots_garwood_synonym_match %>%
  # Join using a chain of coalesce to check the most reliable fields first
  left_join(alt_match_lookup, by = c("orig_name" = "match_key")) %>%
  left_join(alt_match_lookup, by = c("orig_binomial" = "match_key"), suffix = c("", "_bi")) %>%
  left_join(alt_match_lookup, by = c("wcvp_matched_name" = "match_key"), suffix = c("", "_match")) %>%
  left_join(alt_match_lookup, by = c("wcvp_accepted_name" = "match_key"), suffix = c("", "_acc")) %>%
  
  # Now, condense everything into your final fields
  mutate(
    orig_garwood    = coalesce(garwood_orig_match, garwood_orig_match_bi, garwood_orig_match_match, garwood_orig_match_acc),
    orig_garwood_synonym = coalesce(garwood_synonym_match, garwood_synonym_match_bi, garwood_synonym_match_match, garwood_synonym_match_acc),
    firstmatchtype  = coalesce(firstmatchtype, firstmatchtype_bi, firstmatchtype_match, firstmatchtype_acc),
    alternate_name  = coalesce(alt_name, alt_name_bi, alt_name_match, alt_name_acc),
    prefered_name   = coalesce(pref_name, pref_name_bi, pref_name_match, pref_name_acc)
  ) %>%
  select(-ends_with("_bi"), -ends_with("_match"), -ends_with("_acc")) # Clean up extra columns


datatable(
  bci_plots_garwood_synonym_print %>% 
  select(orig_name, wcvp_matched_name, wcvp_accepted_name, firstmatchtype, orig_garwood,
         alternate_name, prefered_name, orig_garwood_synonym), 
          caption="BCI plot species where an alternate name of the species matched to garwood",
          options = list(scrollX = TRUE, pageLength = 10))


```

Printing species in the plots that are not in Garwood. 
```{r allnomatchbciplot}

# combine direct matches and matches by alternate names or synonyms
bci_plots_garwood_match <- bind_rows(
  bci_plots_garwood_name_match,
  bci_plots_garwood_synonym_match
) %>%
  distinct(sp6, .keep_all = TRUE)

# filter unmatched rows
bci_plots_no_garwood_at_all <- bci_plots %>%
  filter(!sp6 %in% bci_plots_garwood_match$sp6)

kable(bci_plots_no_garwood_at_all  %>% select(sp6, orig_name, wcvp_matched_name, 
                                              wcvp_accepted_name, plot_name), 
          caption="Species recorded in BCI plots census that don't match anything in Garwood")


plotsingarwoodorig <- garwood_list_orig %>% 
  filter(orig_name %in% bci_plots_no_garwood_at_all$orig_name)

if (nrow(bci_plots_no_garwood_at_all)>0 & nrow(plotsingarwoodorig)>0) {
  datatable(plotsingarwoodorig %>%
    select(orig_name, chg), 
            caption="In some cases, these species were on the original Garwood list but recorded there as being only in the BCNM, not on BCI.")
    
}

write_xlsx(bci_plots_no_garwood_at_all, FILECHECKPLOTS)

```

Add the completely new rows to the BCI list


```{r addplottobci}
# bci_plots_no_garwood_at_all$source_list = "PlotCensusBCI"

bcilist <- bind_rows(
  bcilist,
  bci_plots_no_garwood_at_all %>% 
    mutate(source_list="plotcensuses") %>%
    select(all_of(c(listscol, "plot_name", "bci50ha", "bci10ha", "P11", "P14"))
))

```


### Lianas BCI list

Check whether they are in Garwood directly or by alternate names.  
```{r lianasnobcilist}

lianas_list <- lianas_list %>% filter(isbci=="TRUE")
lianas_list <- lianas_list %>% distinct(orig_name, .keep_all = TRUE)
lianas_list <- lianas_list %>% 
  mutate(wcvp_accepted_binomial = paste(wcvp_accepted_name_genus, wcvp_accepted_name_species),
         former_binomial = ifelse(is.na(GenusFormer),NA,paste(GenusFormer,SpeciesOnlyFormer))) %>%
  mutate(former_name = paste(ifelse(is.na(SubspeciesFormer),former_binomial,
                                    paste(former_binomial,SubspeciesFormer))))

# 1. direct name/binomial match
lianas_garwood_name_match <- lianas_list %>%
  filter(

    orig_name %in% bcilist$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist$wcvp_accepted_name) |

    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist$orig_binomial 
    )) |

    (!is.na(wcvp_accepted_binomial) & (
        wcvp_accepted_binomial %in% bcilist$wcvp_accepted_binomial 
    ))
  )


# 2. synonym_sp for remaining
lianas_garwood_synonym_match <- lianas_list %>%
  filter(
    !orig_name %in% lianas_garwood_name_match$orig_name &
    (
      orig_name %in% garwood_synonyms$synonym_sp |
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(orig_binomial) & orig_binomial %in% garwood_synonyms$synonym_binomial) |
      (!is.na(wcvp_accepted_binomial) & wcvp_accepted_binomial %in% garwood_synonyms$synonym_binomial)
    )
  )

lianas_garwood_synonym_match$matched_garwood_synonym <- TRUE


# see species that matched by a synonym 
lianas_garwood_synonym_match_print <- lianas_garwood_synonym_match %>%  left_join(
    garwood_synonyms %>% select(sp_orig = sp_name, synonym_sp, syn_orig = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_match = sp_name, synonym_sp, syn_match = SynonymTextALL) %>% distinct(),
    by = c("wcvp_matched_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_wcvp = sp_name, synonym_sp, syn_wcvp = SynonymTextALL) %>% distinct(),
    by = c("wcvp_accepted_name" = "synonym_sp")) %>%
  mutate(
    SynonymMatchAll = coalesce(syn_orig, syn_match, syn_wcvp),
    orig_garwood = coalesce(sp_orig, sp_match, sp_wcvp),
  ) %>%
  select(-syn_orig, -syn_match, -syn_wcvp,  -sp_orig, -sp_match, -sp_wcvp) %>%
  distinct(orig_name, .keep_all = TRUE)
  

kable(lianas_garwood_synonym_match_print %>% 
            select(orig_name, wcvp_matched_name, wcvp_accepted_name, SynonymMatchAll, orig_garwood), 
          caption="Liana species that matched garwood by synonym",
          options = list(scrollX = TRUE, pageLength = 10))



#3 match of former liana name to current Garwood name 
lianas_formername_match <- lianas_list %>%
  filter(
    !orig_name %in% c(lianas_garwood_name_match$orig_name,lianas_garwood_synonym_match$orig_name)
                      &
    (former_name %in% bcilist$orig_name | 
       former_binomial %in% bcilist$orig_binomial)  )

kable(lianas_formername_match %>% select(former_name,orig_name),
      caption="Liana species in which former names matched to Garwood original name.")


#4 match of former liana name to Garwood synonym
lianas_formertosynonym_match <- lianas_list %>%
  filter(
    !orig_name %in% c(lianas_garwood_name_match$orig_name,lianas_garwood_synonym_match$orig_name)
                      &
    (former_name %in% garwood_synonyms$synonym_sp |
       former_binomial %in% garwood_synonyms$synonym_binomial))

# see species that matched by a synonym 
lianas_formertosynonym_print <- lianas_formertosynonym_match %>%  left_join(
    garwood_synonyms %>% select(garwood_name = sp_name, synonym_sp, syn_text = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) 

kable(lianas_formertosynonym_print %>% 
            select(orig_name, former_name, garwood_name, wcvp_accepted_name,syn_text), 
          caption="Liana species whose former names matched a Garwood synonym.",
          options = list(scrollX = TRUE, pageLength = 10))



# 5. everything that did not match directly to garwood, match to alternate names
lianas_garwood_alternate_match <- lianas_list %>% 
  filter(
       !orig_name %in% c(lianas_garwood_name_match$orig_name,lianas_garwood_synonym_match$orig_name, lianas_formertosynonym_match$orig_name) &
    (
      orig_name %in% alt_match_lookup$match_key |
      orig_binomial %in% alt_match_lookup$match_key |
      
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% alt_match_lookup$match_key) |
        
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% alt_match_lookup$match_key)
    )
  )


# print the altname it matched to and why that name is in garwood

lianas_garwood_alternate_print <- lianas_garwood_alternate_match %>%
  # Join using a chain of coalesce to check the most reliable fields first
  left_join(alt_match_lookup, by = c("orig_name" = "match_key")) %>%
  left_join(alt_match_lookup, by = c("orig_binomial" = "match_key"), suffix = c("", "_bi")) %>%
  left_join(alt_match_lookup, by = c("wcvp_matched_name" = "match_key"), suffix = c("", "_match")) %>%
  left_join(alt_match_lookup, by = c("wcvp_accepted_name" = "match_key"), suffix = c("", "_acc")) %>%
  
  # Now, condense everything into your final fields
  mutate(
    orig_garwood    = coalesce(garwood_orig_match, garwood_orig_match_bi, garwood_orig_match_match, garwood_orig_match_acc),
    orig_garwood_synonym = coalesce(garwood_synonym_match, garwood_synonym_match_bi, garwood_synonym_match_match, garwood_synonym_match_acc),
    firstmatchtype  = coalesce(firstmatchtype, firstmatchtype_bi, firstmatchtype_match, firstmatchtype_acc),
    alternate_name  = coalesce(alt_name, alt_name_bi, alt_name_match, alt_name_acc),
    prefered_name   = coalesce(pref_name, pref_name_bi, pref_name_match, pref_name_acc)
  ) %>%
  select(-ends_with("_bi"), -ends_with("_match"), -ends_with("_acc")) # Clean up extra columns


datatable(
  lianas_garwood_alternate_print %>% 
  select(orig_name, wcvp_matched_name, wcvp_accepted_name, firstmatchtype, orig_garwood,
         alternate_name, prefered_name, orig_garwood_synonym), 
          caption="Liana species where an alternate name of the species matched to garwood",
          options = list(scrollX = TRUE, pageLength = 10))


# additional species still unmatched

lianas_garwood_match <- bind_rows(
  lianas_garwood_name_match,
  lianas_garwood_synonym_match,
  lianas_formername_match,
  lianas_formertosynonym_match,
  lianas_garwood_alternate_match
) %>%
  distinct(orig_name, .keep_all = TRUE)

lianas_no_garwood_at_all <- lianas_list %>%
  filter(!orig_name %in% lianas_garwood_match$orig_name)


# 6. Output and Display
kable(lianas_no_garwood_at_all %>%
  select(former_name, orig_name, wcvp_matched_name, wcvp_accepted_name), 
          caption = "Lianas species recorded in the 50-ha plot that don't have a match in Garwood",
          options = list(scrollX = TRUE, pageLength = 10))

write_xlsx(lianas_no_garwood_at_all, FILECHECKLIANAS)

```

Add to bci list
```{r lianasbciadd}

lianas_no_garwood_at_all$source_list = "lianas"

bcilist <- bind_rows(
  bcilist,
  lianas_no_garwood_at_all %>% select(all_of(listscol))
)

# Add the forestegeo names of the species that matched by alternate names
bcilist <- bcilist %>% 
  rows_update(
    lianas_garwood_alternate_print %>% 
      # Rename 'orig_garwood' to 'orig_name' to match bcilist exactly
      select(orig_name = orig_garwood, pref_forestgeo = prefered_name),
    by = "orig_name"
  )

```

### Check consistency 
```{r savebotanists}

# Check consistency
# removing rows that are not identified at species level
kable(bcilist %>% filter(is.na(word(orig_binomial, 2))) %>% select(orig_name, wcvp_accepted_name), caption="Species to be removed from final list because they are not fully identified" )

bcilist <- bcilist %>% filter(!is.na(word(orig_binomial, 2))) 

bcidupnames <- bcilist %>% group_by(orig_name) %>% filter(n()> 1) 
if (nrow(bcidupnames)>0) {
  datatable(bcidupnames%>%
          select(orig_name, wcvp_accepted_name) %>% arrange(orig_name),
          caption="Species with duplicated original name",
  options = list(scrollX = TRUE, pageLength = 10)
          )
  
}

bcidupwcvp <- bcilist %>% group_by(wcvp_accepted_name) %>% 
            filter(n()> 1)
if (nrow(bcidupwcvp)>0) {
datatable(bcidupwcvp %>% select(orig_name, wcvp_accepted_name)%>% 
            arrange(wcvp_accepted_name), 
          caption="Species duplicated wcvp_accepted name",
  options = list(scrollX = TRUE, pageLength = 10)
          )
}

bcilist %>%
  count(source_list) %>%
  arrange(desc(n)) %>%
  kable(col.names = c("source_list", "count"),caption="Sources of morphospecies in the BCI plant species list.")

datatable(
  bcilist %>%
    select(wcvp_accepted_name, orig_name, source_list) %>% 
    arrange(source_list),
    caption="Species list for BCI, with the source of each entry",
  options = list(scrollX = TRUE, pageLength = 10))

```


### Use the Panama woody plant morphospecies list to link in currento ForestGEO names and 6-letter and 4-letter species codes for all species.

And add column indicating if the species is in the woody list
```{r addforestgeocurrentname}

# For all the species in the bcilist
# check if they are directly on the forestgeo /woody list
# check if they have an alternate name on that list

# Add a column with the forestgeo name
# Flag when the forestgeo_current_name != orig_garwood name

# 1. Create lookup table for woody species
forestgeo_names_lookup <- bind_rows(
  panwoody_list %>% select(match_key = orig_name, forestgeo_current_name = current_name, sp6, sp4),
  panwoody_list %>% select(match_key = orig_binomial, forestgeo_current_name = current_name, sp6, sp4),
  panwoody_list %>% select(match_key = current_name, forestgeo_current_name = current_name, sp6, sp4),
  panwoody_list %>% select(match_key = wcvp_accepted_name, forestgeo_current_name = current_name, sp6, sp4),
  panwoody_list %>% select(match_key = wcvp_matched_name, forestgeo_current_name = current_name, sp6, sp4),
  panwoody_synonyms %>% select(match_key = SynonymTextSingle, forestgeo_current_name = current_name, sp6, sp4),
) %>% 
  filter(!is.na(match_key)) %>%
  distinct(match_key, .keep_all = TRUE)

# check for any cases where one forestgeo name corresponds with multiple garwood names


# check for cases where one garwood name corresponds to multiple forestgeo names.  


# 2. Match forestgeo names and codes from the woody list
bcilist <- bcilist %>% 
  left_join(forestgeo_names_lookup, by = c("pref_forestgeo"= "match_key")) %>%
  left_join(forestgeo_names_lookup, by = c("orig_name"= "match_key"), suffix = c("", "_orig")) %>%
  left_join(forestgeo_names_lookup, by = c("orig_binomial"= "match_key"), suffix = c("", "_origbi")) %>%
  left_join(forestgeo_names_lookup, by = c("wcvp_accepted_name"= "match_key"), suffix = c("", "_wcvpacc")) %>%
  left_join(forestgeo_names_lookup, by = c("wcvp_accepted_binomial"= "match_key"), suffix = c("", "_wcvpaccbi")) %>%
  left_join(forestgeo_names_lookup, by = c("wcvp_matched_name"= "match_key"), suffix = c("", "_wcvpmatch"))
  
bcilist <- bcilist %>%  mutate(
    forestgeo_current_name = coalesce(
      forestgeo_current_name,
      forestgeo_current_name_orig,
      forestgeo_current_name_origbi,
      forestgeo_current_name_wcvpacc,
      forestgeo_current_name_wcvpaccbi,
      forestgeo_current_name_wcvpmatch
    ),
    sp6 = coalesce(
      sp6,
      sp6_orig,
      sp6_origbi,
      sp6_wcvpacc,
      sp6_wcvpaccbi,
      sp6_wcvpmatch
    ),
    sp4 = coalesce(
      sp4,
      sp4_orig,
      sp4_origbi,
      sp4_wcvpacc,
      sp4_wcvpaccbi,
      sp4_wcvpmatch
    )) %>%
  select(-ends_with(c("_orig", "_origbi", "_wcvpacc", "_wcvpaccbi", "_wcvpmatch"))) 
  
bcilist <- bcilist %>%
  mutate(
    woody = !is.na(sp6), # If they matched it also means it is a woody species 
    garwood_diff_forestgeo = orig_name != forestgeo_current_name & 
      wcvp_accepted_name != forestgeo_current_name &
      orig_binomial != forestgeo_current_name &
      wcvp_accepted_binomial != forestgeo_current_name
  ) 

# see the species with code
datatable(bcilist %>% filter(!is.na(sp6)) %>%select(sp6, sp4, forestgeo_current_name, orig_name,wcvp_accepted_name, garwood_diff_forestgeo, source_list),
          caption="BCI species list with sp6 code from the woody list")
          
```


## Save Botanist high quality list
```{r bcilistwoody, echo=TRUE}


# BCI BOTANISTS HIGH QUALITY LIST
write_xlsx(bcilist, FILECOMBINEDLIST)

datatable(bcilist %>%
  filter(woody) %>%
  select(orig_name, wcvp_matched_name, wcvp_accepted_name, wcvp_accepted_family, 
         wcvp_accepted_lifeform), caption="BCI species that are in the woody list")

```


### Possibly woody species

```{r posiblywoody}

lifeform_woody = c(
  "tree",
  "shrub",
  "liana",
  "climbing shrub",
  "shrub or tree",
  "climbing shrub or tree",
  "scrambling shrub or tree",
  "subshrub or shrub",
  "subshrub, shrub or tree",
  "scrambling shrub",
  "perennial or shrub",
  "scrambling subshrub or shrub",
  "epiphytic shrub",
  "scrambling tree",
  "climbing shrub or liana",
  "scrambling shrub or liana",
  "tuberous shrub or tree",
  "herbaceous tree",
  "climbing herbaceous tree",
  "climbing herbaceous tree or liana",
  "climbing perennial or herbaceous tree",
  "subshrub or climbing shrub",
  "tuberous subshrub",
  "hemiparasitic shrub",
  "hemiepiphytic shrub or tree",
  "annual or shrub",
  "annual, subshrub or shrub"
)

bcilist_possibly_woody <- bcilist %>%
  filter(!woody &
          wcvp_accepted_lifeform %in% lifeform_woody ) %>%
  select(orig_name, wcvp_matched_name, wcvp_accepted_name, wcvp_accepted_family, 
         wcvp_accepted_lifeform)

datatable(bcilist_possibly_woody, caption="BCI species that might be woody according to the wcvp lifeform")

write_xlsx(bcilist_possibly_woody, FILEBCIWOODY)


```


## Panamabiota species that are not in garwood

```{r pbiotanogarwood}

pbiota_list$wcvp_accepted_binomial <- paste0(
  pbiota_list$wcvp_accepted_name_genus, " ",
  pbiota_list$wcvp_accepted_name_species
)

# 1. direct name/binomial match
pbiota_garwood_name_match <- pbiota_list %>%
  filter(

    orig_name %in% bcilist$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist$wcvp_accepted_name) |

    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist$orig_binomial 
    )) |

    (!is.na(wcvp_accepted_binomial) & (
        wcvp_accepted_binomial %in% bcilist$wcvp_accepted_binomial 
    ))
  )

# 2. synonym_sp for remaining
pbiota_garwood_synonym_match <- pbiota_list %>%
  filter(
    !orig_name %in% pbiota_garwood_name_match$orig_name &
    (
      orig_name %in% garwood_synonyms$synonym_sp |
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(orig_binomial) & orig_binomial %in% garwood_synonyms$synonym_binomial) |
      (!is.na(wcvp_accepted_binomial) & wcvp_accepted_binomial %in% garwood_synonyms$synonym_binomial)
    )
  )

pbiota_garwood_synonym_match$matched_garwood_synonym <- TRUE


# see species that matched by a synonym 
pbiota_garwood_synonym_match_print <- pbiota_garwood_synonym_match %>%  left_join(
    garwood_synonyms %>% select(sp_orig = sp_name, synonym_sp, syn_orig = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_match = sp_name, synonym_sp, syn_match = SynonymTextALL) %>% distinct(),
    by = c("wcvp_matched_name" = "synonym_sp"), relationship = "many-to-many"
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_wcvp = sp_name, synonym_sp, syn_wcvp = SynonymTextALL) %>% distinct(),
    by = c("wcvp_accepted_name" = "synonym_sp"),  relationship = "many-to-many") %>%
  mutate(
    SynonymMatchAll = coalesce(syn_orig, syn_match, syn_wcvp),
    orig_garwood = coalesce(sp_orig, sp_match, sp_wcvp),
  ) %>%
  select(-syn_orig, -syn_match, -syn_wcvp,  -sp_orig, -sp_match, -sp_wcvp) %>%
  distinct(orig_name, .keep_all = TRUE)
  

datatable(pbiota_garwood_synonym_match_print %>% 
            select(orig_name, wcvp_matched_name, wcvp_accepted_name, SynonymMatchAll, orig_garwood), 
          caption="Panamabiota species that matched Garwood by synonym",
          options = list(scrollX = TRUE, pageLength = 10))

pbiota_garwood_match <- bind_rows(
  pbiota_garwood_name_match,
  pbiota_garwood_synonym_match
) %>%
  distinct(orig_name, .keep_all = TRUE)

pbiota_no_garwood_at_all <- pbiota_list %>%
  filter(!orig_name %in% pbiota_garwood_match$orig_name)



datatable(pbiota_no_garwood_at_all, 
          caption="Panamabiota species that don't match anything in Garwood",
          options = list(scrollX = TRUE, pageLength = 10))

write_xlsx(pbiota_no_garwood_at_all, FILECHECKPANAMABIOTA)

```


