---
title: "Assembling a combined list of alternate names for Panamanian plant morphospecies"
author: "Paula Uzcategui and Helene Muller-Landau"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##  Create a combined alternate name file

```{r cars}
rm(list=ls())

library(readxl)
library(writexl)
library(dplyr)
library(tidyr)
library(stringi)
library(stringr)
library(DT)


viewtaxonomy <-  read_excel("splists_raw/Forestgeo/2025-12-04FromSuzanne/ViewTaxonomy_dec4_25.xlsx")

woody_synonyms <- read_excel("splists_out/CurrentPanamaWoody_SYNONYMS_2026-03-04.xlsx")
garwood_synonyms <- read_excel("splists_raw/Garwood/Garwood_2009_Appendix1_Synonyms.xlsx") 


garwood_synonyms <- garwood_synonyms %>% mutate(
  sp_name = trimws(paste0(Genus, " ", Species)),
  synonym_sp = trimws(paste0(SynonymGenusSingle, " ", SynonymSpeciesSingle)))


FILEVIEWTAXSYN <- "splists_out/ViewTaxonomy_dec4_25_SYNONYMS.xlsx"

FILEALLSYN <- paste0("splists_out/AltNames_",Sys.Date(), ".xlsx")

```

## Create alternate name list for view taxonomy

```{r synonyms}


viewtaxonomy <- viewtaxonomy %>%
mutate(across(everything(), ~ {
  if(is.character(.)) {
    na_if(., "NULL") # Solo aplica na_if si la columna es texto
  } else {
    . # Si es numérica, la deja igual
  }
}))

 #set the code to the subspecies code if it is a subspecies
viewtaxonomy <- viewtaxonomy %>% 
  mutate(
         sp6 = ifelse(is.na(Subspecies), Mnemonic, subspMnemonic), 
         binomial=ifelse(Genus=="Unidentified",paste(Genus,sp6),paste(Genus,SpeciesName)),
         species_name=ifelse(is.na(Subspecies),
                                paste0(Genus," ",SpeciesName),
                                paste0(Genus," ",SpeciesName," ",Rank, " ", Subspecies))
       )
  

viewtaxonomy$binomial <- stri_trans_general(viewtaxonomy$binomial,"Latin-ASCII")

viewtaxonomy_synonyms <- viewtaxonomy %>%

  mutate(OriginalOrder = row_number()) %>%
  
  select(
    OriginalOrder, 
    sp6, species_name, 
    SynonymTextALL = ListOfOldNames
  ) %>%
  
  # Filter out empty synonym rows
  filter(!is.na(SynonymTextALL) & SynonymTextALL != "") %>%

  mutate(SynonymTextSingle = SynonymTextALL) %>%
  separate_longer_delim(SynonymTextSingle, delim = ",") %>%
  mutate(SynonymTextSingle = trimws(SynonymTextSingle)) %>%
  
  # Order of the synonym in the original text
  group_by(OriginalOrder) %>%
    mutate(SynonymNumber = row_number()) %>%
    ungroup() %>%
  
  # Extract Genus and Species from the single synonym
  # Using too_many="merge" in case there are infraspecific names
  separate_wider_delim(
    SynonymTextSingle, 
    delim = " ", 
    names = c("SynonymGenusSingle", "SynonymSpeciesSingle"),
    too_many = "merge",
    cols_remove = FALSE
  ) %>% arrange(species_name)

# View the result
datatable(viewtaxonomy_synonyms, 
             caption = "Mapping: Original Name to Individual Synonym")


write_xlsx(viewtaxonomy_synonyms, FILEVIEWTAXSYN)

```
Writing the synonyms from the ViewTaxonomy file to `r FILEVIEWTAXSYN`


## Combine these into a single table of synonyms and alternate names

```{r combinedsyn}

v_tax <- viewtaxonomy_synonyms %>%
  select(prefname = species_name, altname = SynonymTextSingle) %>%
  mutate(in_viewtaxcurr = TRUE)

w_syn <- woody_synonyms %>%
  select(prefname = orig_name, altname = SynonymTextSingle) %>%
  mutate(in_panwoodycurr = TRUE)

g_syn <- garwood_synonyms %>%
  select(prefname = sp_name, altname = synonym_sp, GarwoodNotes = SynonymTextSingle) %>%
  mutate(in_garwood2009 = TRUE)

# 2. Join them all together
altnames <- w_syn %>%
  full_join(v_tax, by = c("prefname", "altname")) %>%
  full_join(g_syn, by = c("prefname", "altname")) %>%
  # 3. Handle NAs and finalize column order
  mutate(across(starts_with("in_"), ~ coalesce(.x, FALSE))) %>%
  select(prefname, altname, in_panwoodycurr, in_viewtaxcurr, in_garwood2009, GarwoodNotes)

datatable(altnames)

```

Find and examine problematic entries.  

```{r checkaltnames}

# find cases in which the two names are the same
identnames <- altnames %>% filter(altname==prefname)

# show these cases
datatable(identnames,caption="Altname and prefname are the same")

# remove them
altnames <- altnames %>%
  filter(altname!=prefname)

# find cases in which two rows have the same altname and prefname
dup_altpref <- altnames %>%
  arrange(prefname,altname,GarwoodNotes) %>%
  group_by(altname, prefname) %>%
  filter(n() > 1) %>%
  ungroup()

# check these cases
datatable(dup_altpref,caption="Cases with the same combination of altname and prefname binomials (likely different full names including subspecies)")

# consolidate these duplicate rows
altnames <- altnames %>%
  group_by(altname, prefname) %>%
  summarise(
    in_panwoodycurr = any(in_panwoodycurr, na.rm = TRUE),
    in_viewtaxcurr  = any(in_viewtaxcurr, na.rm = TRUE),
    in_garwood2009  = any(in_garwood2009, na.rm = TRUE),
    GarwoodNotes = {
      x <- unique(na.omit(GarwoodNotes))
      if (length(x) == 0) NA_character_ else paste(x, collapse = ". ")
    },
    .groups = "drop"
  )


# detect cases where the same combination of names appears both directions:
recipnames <- altnames %>%
  rowwise() %>%
  mutate(
    name1 = sort(c(altname, prefname))[1],
    name2 = sort(c(altname, prefname))[2]
  ) %>%
  ungroup() %>%
  group_by(name1, name2) %>%
  filter(n() > 1) %>%
  ungroup() %>%
  arrange(name1,in_garwood2009) %>%
  select(-name1,-name2)

# check these cases
datatable(recipnames,caption="Cases in which the same pair of names appears in opposite positions.")


```

These are all cases in which Garwood 2009 has one set of changes, and the more recent ForestGEO species lists (Panama woody and/or ViewTaxonomy) have the opposite. In these cases, the ForestGEO preference is kept, because it is more recent.


```{r writecombined}

# drop the Garwood variants of the reciprocol changes
altnames <- altnames %>%
  anti_join(
    recipnames %>% filter(in_garwood2009 == TRUE) %>% select(altname, prefname),
    by = c("altname", "prefname")
  )


write_xlsx(altnames, FILEALLSYN)
```


Writing all remaining sets of alternate names to the file `r FILEALLSYN`
