---
title: "CheckForestgeoSpListChanges"
author: "Helene Muller-Landau"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Overview

This document is intended to run checks of changes in the Panama woody species lists assembled by ForestGEO, including species codes and species names.  

## How to run

Put this document in the main folder to run it.

When running this following changes, save the html output with the name and date in the reports folder.

```{r libraries}
library(readxl)
library(writexl)
library(dplyr)
library(tidyr)
library(knitr)
```

Read in the old and new species lists, and align the column names.
```{r readfiles}
FNOLD <- "splists_raw/PanamaWoodySpLists/2025-08-28FromSuzanne/FloraPanama_28Aug25.xlsx"
FNNEW <- "splists_raw/PanamaWoodySpLists/2025-12-04FromSuzanne/FloradePanama_dec4_2025_HM.xlsx"

oldlist <- read_xlsx(FNOLD,guess_max=5000) %>%
  rename_with(tolower)%>%
  rename(taxorder=order,
         mnemonic=spcode,
         speciesname=species) %>%
  mutate(binomial=paste(genus,speciesname)) %>%
  mutate(fullname=ifelse(is.na(subspecies),binomial,paste(binomial,subspecies)))
newlist <- read_xlsx(FNNEW,guess_max=5000)  %>%
  rename_with(tolower) %>%
  mutate(binomial=paste(genus,speciesname))  %>%
  mutate(fullname=ifelse(is.na(subspecies),binomial,paste(binomial,subspecies)))
```

Check for differences in species names included.  
```{r aligncolnames}
droppedcases <- oldlist %>%
  filter(!mnemonic %in% newlist$mnemonic & !fullname %in% newlist$fullname) %>%
  select(mnemonic,fullname,authority,synonyms)
kable(droppedcases,caption="Dropped cases:  Species name and code are on old list and are not on new list.")

droppedcases <- newlist %>%
  filter(!mnemonic %in% oldlist$mnemonic & !fullname %in% oldlist$fullname) %>%
  select(mnemonic,fullname,authority,synonyms)
kable(droppedcases,caption="Added cases:  Species name and code are on new list and are not on old.")

changedcodes <- oldlist %>%
  filter(!mnemonic %in% newlist$mnemonic & fullname %in% newlist$fullname) %>%
  rename(oldcode = mnemonic) %>%
  mutate(newcode = newlist$mnemonic[match(fullname,newlist$fullname)]) %>%
  select(oldcode,newcode,fullname,authority,synonyms)

kable(changedcodes,caption = "Species that changed codes from the old list to the new list")

changednames <- oldlist %>%
  filter(mnemonic %in% newlist$mnemonic & !fullname %in% newlist$fullname) %>%
  rename(oldname = fullname,
         oldauthority=authority) %>%
  mutate(newname = newlist$fullname[match(mnemonic,newlist$mnemonic)],
         newauthority = newlist$authority[match(mnemonic,newlist$mnemonic)]) %>%
  select(mnemonic,oldname,oldauthority,newname,newauthority)

kable(changednames,caption = "Codes with name changes from the old list to the new list")

```


