---
title: "Check and merge Panama woody species lists"
author: "Helene Muller-Landau and Paula Uzcategui"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This document is intended to check woody plant species lists for Panama, and merge them together to produce a single document with a complete set of codes and up-to-date names.

## How to run

Put this document in the main folder to run it.

When running this following changes, save the html output with the name and date in the reports folder.

## Source files

1. *Panama woody plant species table*, specifically the most recent version from the folder "PanamaWoodySpLists" (filenames typically start with "FloraPanama" or similar).  This includes only fully recognized botanical species, omitting morphospecies lacking complete taxonomic information.  It is assembled and maintained by Smithsonian ForestGEO botanists in Panama.  It includes a 6-letter species code (mnemonic). 

 
2. *Panana plots woody species table*, specifically the most recent version from the folder "ForestGEO" (filenames typically start "ViewTaxonomy").  This includes all recognized morphospecies that appear in any of the ForestGEO plots. It differs from the Panama woody plant species list in that it omits species that are not in the plots, and adds morphospecies that do not yet have a full current taxonomic name (whether because their species ID has not yet been determined or they are new species).  It is assembled and maintained by Smithsonian ForestGEO botanists in Panama.  It includes a 6-letter species code (mnemonic). 

 
3. *Wright plant nomenclature table*, specifically the most recent version from the folder "Wright" (filenames start with "nomenclature").  This includes all plant morphospecies (including non-woody morphospecies) that are recorded in seed and seedling datasets collected at STRI under the supervision of S. Joseph Wright.  This is included in order to link the 4-letter codes used in these datasets. 


## Workflow

*The workflow here has the following steps: *

- Check each list individually.  This involves checking for duplicate entries and other errors, and checking species names against WCVP (World Checklist of Vascular Plants) version 13.  Note that prior versions of this workflow checked names against TNRS. 

- Merge the Panama woody plant species and Panama plots tree species lists, making sure 6-letter codes are consistent.  Output any mismatches.

- Merge in sp4 codes from the Wright list. 

- Output a combined Panama woody species list including sp4 codes.

- Output a combined list of "synonyms" in the broad sense of the word (not the technical botanical definition): alternative names used for the same Panama woody plant morphospecies in the past.  These are pulled from the other lists.  

- Report any apparent discrepancies / issues within this RMarkdown and in output files to check. 


## Future to do

To be added in the future - also include data from the following file, and check for consistency of codes, etc. 

Panama plot liana species table, specifically from the folder Schnitzer.  These are lists of liana species that are included in censuses of plots in central Panama conducted under the supervision of Stefan Schnitzer.  It includes a 6-letter species code that should align with the ForestGEO codes.  


# Initial setup

```{r libraries, echo=TRUE, results="hide", message=FALSE, warning=FALSE,}
library(readxl)
library(writexl)
library(dplyr)
library(tidyr)
library(stringi)
library(stringr)
suppressPackageStartupMessages({
  library(rWCVP)
  library(rWCVPdata)
})
library(knitr)
library(DT)

```


```{r clearworkspace, include=FALSE}
rm(list=ls())
```


```{r globalvars, echo=TRUE, results="hide"}
redoWCVP = FALSE  # if TRUE, always rerun the time-consuming WCVP checks for all names
# if false, will try to use saved versions of files instead

FIX_SPECIES = TRUE # a new file will be saved with the fixed species names. 

DIRINSP <- "splists_raw/"
DIRMIDSP <- "splists_mid/"
DIROUTSP <- "splists_out/"
DIRCHECK <- "tocheck/"

# input files # all in directory DIRINSP
FNWRIGHTSPLISTIN <- "splists_out/wright_nomenclature_R_20210224_WCVP_2026-02-27.xlsx"
FNFGEOSPLISTIN <- "Forestgeo/2026-02-25FromSuzanne/ViewTaxonomy_dec4_25_HM2.xlsx"
FNPANAMASPLISTIN <- "PanamaWoodySpLists/2025-12-04FromSuzanne/FloradePanama_dec4_2025_HM.xlsx" 


FNPANAMASPLISTWCVP <- paste0(DIRMIDSP,tools::file_path_sans_ext(basename(FNPANAMASPLISTIN)),"_WCVP_MATCH.rds")
FNPANAMASPLISTWCVP2 <- paste0(DIRMIDSP,tools::file_path_sans_ext(basename(FNPANAMASPLISTIN)),"_WCVP_MATCH2.rds") # binomial matches


PANAMA_COMBINED_OUT_PATH <- paste0(DIROUTSP,"CurrentPanamaWoody_",Sys.Date(),".xlsx")

PANAMA_COMBINED_CHANGE_SUMMARY_PATH <- paste0(DIROUTSP,"CurrentPanamaWoody_CHANGES_",Sys.Date(),".xlsx")

WRIGHT_MISSING_PANWOODY_PATH <- paste0(DIROUTSP,"Wright_sp6_not_CurrentPanamaWoody_",Sys.Date(),".xlsx")
WRIGHT_MISSING_PANWOODY_PATH2 <- paste0(DIROUTSP,"Wright_no_sp6_not_CurrentPanamaWoody_",Sys.Date(),".xlsx")
  
  
FNPANAMATAXA <- paste0(DIROUTSP,"AllTaxa_CurrentPanamaWoody_",Sys.Date(),".xlsx")

FNSYNONYMS <- paste0(DIROUTSP,"CurrentPanamaWoody_SYNONYMS_",Sys.Date(),".xlsx")

# the last is the list of all unique accepted species, genera, and families for photo labeling

```

## Load the Panama woody species list from ForestGEO

This includes only fully recognized botanical species, omitting morphospecies lacking complete taxonomic information.  It is assembled and maintained by Smithsonian ForestGEO botanists in Panama.  It includes a 6-letter species code (mnemonic). 

```{r loadsppanama, echo=TRUE}
panwoodtaxaorig <- read_excel(paste0(DIRINSP,FNPANAMASPLISTIN),col_types="text") 

panwoodtaxa <- panwoodtaxaorig %>%
  mutate(
    across(
      c(family, genus, speciesname, authority, subspecies),
      str_squish
    )
  ) %>%
  rename(
    orig_family       = family,
    orig_genus        = genus,
    orig_species      = speciesname,
    orig_authority  = authority,
    orig_infraspecies = subspecies,
    orig_synonyms = synonyms,
    sp6=mnemonic,
    vouchers=herbarium,
    habit=liana) %>%
  mutate(sp6=tolower(sp6),
         habit=ifelse(habit=="l","Climbing","Freestanding"),
         orig_binomial=paste0(orig_genus," ",orig_species)) %>%
  mutate(orig_name=str_trim(ifelse(is.na(orig_infraspecies),orig_binomial,
                                paste0(orig_binomial," ",orig_infraspecies))),
         habit=ifelse(is.na(habit),"Freestanding",habit))%>%
  mutate(orig_infraspecies_rank = word(orig_infraspecies,1),
         orig_infraspecies = word(orig_infraspecies,2))



```

Number of species in Panama woody plant species list: `r paste(nrow(panwoodtaxa))`

Manual corrections
```{r correctpanwood}
panwoodtaxa <- panwoodtaxa %>%
  filter(sp6!="swarsu")  # remove a morphospecies lacking full species ID

```

Perform some basic checks on the Panama woody species list. 

```{r checkpanwood}
sp6dupspan <- panwoodtaxa %>% group_by(sp6) %>% filter(n()>1) %>% ungroup()
  
if(nrow(sp6dupspan) > 0){
  knitr::kable(sp6dupspan %>% select(sp6, orig_name), caption="Duplicated sp6 codes in Panama woody species list")
} 

# check duplicate species names
spnamedupspan <- panwoodtaxa %>% group_by(orig_name) %>% filter(n()>1) %>% ungroup() %>% arrange(orig_name)
  
if(nrow(spnamedupspan) > 0){
    datatable(spnamedupspan %>% select(sp6, species_name, IDLevel), caption="Duplicated species names in Panama woody species list",options = list(scrollX = TRUE, pageLength = 10))
}

```



## Load the ForestGEO viewtaxonomy dataset

This includes all recognized morphospecies that appear in any of the ForestGEO plots. It differs from the Panama woody plant species list in that it omits species that are not in the plots, and adds morphospecies that do not yet have a full current taxonomic name (whether because their species ID has not yet been determined or they are new species).  It is assembled and maintained by Smithsonian ForestGEO botanists in Panama.  It includes a 6-letter species code (mnemonic). 

```{r loadsptreeplottaxa, echo=TRUE}
treeplottaxaorig <- read_excel(paste0(DIRINSP,FNFGEOSPLISTIN)) 

treeplottaxa <- treeplottaxaorig %>%
mutate(across(everything(), ~ {
  if(is.character(.)) {
    na_if(., "NULL") # Solo aplica na_if si la columna es texto
  } else {
    . # Si es numérica, la deja igual
  }
}))

 #set the code to the subspecies code if it is a subspecies
treeplottaxa <- treeplottaxa %>% 
  mutate(
         sp6 = ifelse(is.na(subspMnemonic), Mnemonic, subspMnemonic), 
         authority= ifelse(is.na(subspAuthority), Authority, 
                           paste0(Authority," ",subspAuthority)), 
         Family=ifelse(Family=="Unknown" & !is.na(FieldFamily),FieldFamily,Family)) %>%
  mutate(binomial=ifelse(IDLevel=="family",paste(Family,SpeciesName),
                         paste(Genus,SpeciesName))) %>%
  mutate(binomial=stri_trans_general(binomial,"Latin-ASCII")) %>% # get rid of nonstandard characters
  mutate(species_name=ifelse(is.na(Subspecies),binomial,
                                paste0(binomial," ",Rank, " ", Subspecies)))
  
# forestgeo stats
kable(as.data.frame(table(treeplottaxa$IDLevel)) %>% rename(ID_Level=Var1, N_MorphoSpecies=Freq),
      caption="Number of morphospecies identified to different levels in the ForestGEO plot tree taxa list (ViewTaxonomy file)")

```
Total number of morphospecies in ForestGEO tree plot taxa list: `r paste(nrow(treeplottaxa))`


Run basic checks on this species list

```{r checkplottaxa}
sp6dups <- treeplottaxa %>% group_by(sp6) %>% filter(n()>1) %>% ungroup()
  
if(nrow(sp6dups) > 0){
  knitr::kable(sp6dups %>% select(sp6, species_name), caption="Duplicated sp6 codes in forestgeo viewtaxa")
} 

# check duplicates in cases with idlevel species
spnamedups <- treeplottaxa %>% filter(IDLevel =="species") %>% group_by(species_name) %>% filter(n()>1) %>% ungroup() %>% arrange(species_name)
  
if(nrow(spnamedups) > 0){
    datatable(spnamedups %>% select(sp6, species_name, IDLevel), caption="Duplicated species names in forestgeo viewtaxa",options = list(scrollX = TRUE, pageLength = 10))
}
  

```



### Check for consistency between the Panama woody plant species list and the ForestGEO plots taxa list.  

These are both assembled by ForestGEO, and generally updated in tandem, so should be fully consistent.  


Check that species with the same code in the two datasets also have the same name.
```{r checkconsistencyfgeo2}
checknamesbycode <- treeplottaxa %>%
  filter(sp6 %in% panwoodtaxa$sp6) %>%
  select(sp6,Mnemonic,subspMnemonic,species_name,binomial) %>%
  rename(viewtaxaname=species_name,viewtaxabinomial=binomial,
         vtsp6=Mnemonic,vtsubsp6=subspMnemonic) %>%
  left_join(panwoodtaxa[,c("sp6","orig_name","orig_binomial")],by=c("sp6"="sp6")) %>%
  rename(panflorname=orig_name,panflorbinomial=orig_binomial) %>%
  filter(viewtaxaname!=panflorname | viewtaxabinomial!=panflorbinomial) %>%
  select(sp6,vtsp6,vtsubsp6,panflorname,viewtaxaname,panflorbinomial,viewtaxabinomial)
  
if (nrow(checknamesbycode)>0) {
  kable(checknamesbycode,
      caption = "Cases where the same sp6 code corresponds to different species names in the ViewTaxonomy and Panama woody plant species lists")
} else print("For every shared sp6 code, names are exactly the same in the ViewTaxonomy and Panama woody plant species lists.")

```
Both of these are cases where the binomial is the same, but the two databases differ in whether there is infraspecific information or not.  The Panama woody plant species list has the subspecies information included.  Go with the Panama woody plant species list in these cases, which is the default.  (That is, later on we only add records from viewtaxa if their 6-letter codes are not already in the Panama woody taxa list.)     



Check that all taxa in the ViewTaxonomy list that are identified to species are also in the Panama woody plants list.  

```{r checkconsistencyfgeo1}
spvtnotpan <- treeplottaxa %>%
  filter(IDLevel=="species"& is.na(match(sp6,panwoodtaxa$sp6)))

if (nrow(spvtnotpan)>0) {
  kable(spvtnotpan %>% select(sp6,Mnemonic,subspMnemonic,species_name),
      caption = "Species in the ViewTaxonomy dataset whose sp6 codes are missing from the Panama woody plant species list")
} else (print("All species with IDLevel species in the ViewTaxonomy list are also in the Panama woody plant species list."))
```


```{r checkconsistencyfgeo1b}
spvtnotpansp6binom <- spvtnotpan %>%
    select(Mnemonic,binomial,species_name) %>%
    left_join(panwoodtaxa %>% select(sp6,orig_binomial,orig_name), 
  by=c("Mnemonic"="sp6")) %>%
    rename(viewtaxabinom=binomial,
           viewtaxaname=species_name,
           panwoodbinom=orig_binomial,
           panwoodname=orig_name)
  if (nrow(spvtnotpansp6binom)>0)
    kable(spvtnotpansp6binom,
           caption="Of those species in the ViewTaxonomy dataset whose subspecies sp6 codes are missing from the Panama woody plant species list, ones that have a match on their species-level (rather than subspecies-level) sp6 codes")
```

```{r checkconsistencyfgeo1c}
 spvtnotpanbinom <- spvtnotpan %>%
    select(Mnemonic,subspMnemonic,binomial,species_name) %>%
    left_join(panwoodtaxa %>% select(sp6,orig_binomial,orig_name), 
  by=c("binomial"="orig_binomial"),multiple="all") %>%
    rename(vtsp6=Mnemonic,
           vtsubsp6=subspMnemonic,
           panwoodsp6=sp6,
           viewtaxaname=species_name,
           panwoodname=orig_name) %>%
    select(vtsp6,vtsubsp6,panwoodsp6,binomial,viewtaxaname,panwoodname)
  if (nrow(spvtnotpanbinom)>0)
    kable(spvtnotpanbinom,
           caption="Of those species in the ViewTaxonomy dataset whose subspecies sp6 codes are missing from the Panama woody plant species list, ones that have a match on their binomials")


```
In two of the four species, the Panama woody name has a subspecies that differs from the listing in the ViewTaxonomy database.  What's going on here?


```{r checkconsistencyfgeo1d}
spvtnotpanrelated <- treeplottaxa %>%
  filter(binomial %in% spvtnotpan$binomial) %>%
  select(Mnemonic,subspMnemonic,binomial,species_name) %>%
  left_join(panwoodtaxa %>% select(sp6,orig_binomial,orig_name), 
  by=c("Mnemonic"="sp6"),multiple="all") %>%
    rename(vtsp6=Mnemonic,
           vtsubsp6=subspMnemonic,
           viewtaxabinom=binomial,
           panwoodbinom=orig_binomial,
           viewtaxaname=species_name,
           panwoodname=orig_name) %>%
    select(vtsp6,vtsubsp6,viewtaxabinom,panwoodbinom,viewtaxaname,panwoodname)
  
if (nrow(spvtnotpanrelated)>0)
    kable(spvtnotpanrelated,
           caption="Of those species in the ViewTaxonomy dataset whose subspecies sp6 codes are missing from the Panama woody plant species list, all cases with the same binomial in viewtaxa and associated matches based on the species-level Mnemonic.")
```
It appears that in all these cases, the main species or subspecies is in the Panama woody taxa list, but the subspecies or secondary subspecies is not.  Thus the default behavior will work appropriately (we add any records in ViewTaxa that correspond with sp6 codes not already in the full Panama woody species list.)

 

Adding species in forestgeo Viewtaxonomy that are not in the woodylist
```{r addsp, echo=TRUE, results="hide"}

# adding sp6 prior column for sp6 codes in forestgeolist
panwoodtaxa <- panwoodtaxa %>% mutate(
sp6prior=treeplottaxa$Mnemonic[match(orig_name,treeplottaxa$binomial)]
)


# add species in forestgeo that are not in panama woody
panwoodtaxafull <- bind_rows(
  panwoodtaxa %>% mutate(IDLevel = "species"),
  treeplottaxa %>% filter(IDLevel == "species" & !sp6 %in% panwoodtaxa$sp6) %>%  
              select(sp6 = sp6, orig_family = Family, orig_genus = Genus, 
                        orig_name = species_name, orig_species = SpeciesName,
                        orig_infraspecies = Subspecies, orig_infraspecies_rank = Rank,
                        orig_binomial = binomial,
                        orig_authority = authority, orig_synonyms = ListOfOldNames,
                        IDLevel
                        ) 
)


```

Working with: `r nrow(panwoodtaxafull)` fully identified species.  

# Match species names with WCVP v13:

Many names in the original file will change after using WCVP but mainly because of spelling differences. The workflow to find the accepted wcvp name for each species is the following: 

**1. Match species names against WCVP (World Checklist of Vascular Plants)**

**2. Identify and review potential mismatches:**

- A mismatch occurs when WCVP returns a name different from the original lookup name. This can lead to incorrect accepted names, requiring manual verification. Most mismatches arise from minor spelling differences between the original names and WCVP entries. Others occur because some infraspecies are not in WCVP. And very weird mismatches occur when WCVP fuzzy matching points to the wrong name.
      
**3. Correct identified mismatches.**

**4. Set a current_name column following this rules: **

Set current_name to wcvp_name:

1. If species spelling differs in the WCVP matched name and that matched name is accepted
2. If WCVP says the original name is a synonym or illegitimate

Keep the orig_name as current name if:

1. WCVP did not return an accepted name. (The taxon is not in wcvp)
2. The WCVP accepted match means that 2 or more forestgeo species would have the same WCVP name.
3. The wcvp name is missing the subspecies in the original name

At the end we have the same number of species. And we have columns:

current_name -> the name according to the criteria defined above
orig_name -> the name as it appears in the forestgeo table
wcvp_name -> the accepted name according to wcvp
wcvp_matched_name -> the name matched by wcvp 

Among other useful columns returned by wcvp...

### Match WCVP names

```{r mergewcvp}

#########################  WCVP names ##################################

# Helper function to merge and format accepted names from WCVP backbone
matchwcvp_accepted_names <- function(matcheswcvp, wcvp_backbone){
  matcheswcvp %>% 
    left_join(
      wcvp_backbone %>% 
        select(
          plant_name_id,
          # rename immediately
          wcvp_accepted_ipni_id  = ipni_id,
          wcvp_accepted_powo_id  = powo_id,
          wcvp_accepted_family   = family,
          wcvp_accepted_rank = taxon_rank,
          wcvp_accepted_status = taxon_status,
          wcvp_accepted_lifeform = lifeform_description,

          # rename after the mutate for readability 
          genus, species, infraspecific_rank, infraspecies,
          parenthetical_author, primary_author
        ),
      by = c("wcvp_accepted_id" = "plant_name_id")
    ) %>% 
    mutate( 
      wcvp_accepted_binomial = ifelse(is.na(species) | species == "",
                                      genus, paste(genus, species)),
      wcvp_accepted_authority =
        paste0(
          ifelse(!is.na(parenthetical_author), paste0("(", parenthetical_author, ") "),""),
          coalesce(primary_author, "")
        )) %>%
    select(-parenthetical_author,-primary_author) %>%
    mutate(wcvp_accepted_name = ifelse(!is.na(infraspecies) & infraspecies != "",
       paste(wcvp_accepted_binomial,infraspecific_rank,infraspecies), 
       wcvp_accepted_binomial)) %>% 
    rename(
      wcvp_accepted_genus = genus,
      wcvp_accepted_species = species,
      wcvp_accepted_infrarank = infraspecific_rank,
      wcvp_accepted_infraspecies = infraspecies
    )
}

# --- 1. Run or Load Match ---
if(file.exists(FNPANAMASPLISTWCVP) & !redoWCVP){
  matchresult <- readRDS(FNPANAMASPLISTWCVP)
} else {
  # Run the matching process
  matchresult <- wcvp_match_names(panwoodtaxafull, wcvp_names, name_col = "orig_name")

  # Save as RDS for speed
  saveRDS(matchresult, FNPANAMASPLISTWCVP)
}

# do the same for the orig_binomial in cases where this is not equal to orig_name
if(file.exists(FNPANAMASPLISTWCVP2) & !redoWCVP){
  matchresult2 <- readRDS(FNPANAMASPLISTWCVP2)
} else {
  matchresult2 <- wcvp_match_names(subset(panwoodtaxafull,orig_name!=orig_binomial), wcvp_names, 
                                   name_col = "orig_binomial")

  # Save as RDS for speed
  saveRDS(matchresult2, FNPANAMASPLISTWCVP2)
}


# --- 2. Sanity Checks & Duplicate Resolution ---

# Identify names that mapped to multiple 'Accepted' records
manyaccepted <- matchresult %>%
  group_by(orig_name) %>%
  filter(sum(wcvp_status == "Accepted", na.rm = TRUE) > 1) %>%
  ungroup()

# If duplicates exist, force use of the original ID to avoid grouping explosion
if (nrow(manyaccepted) > 0) {
  matchresult <- matchresult %>%
    mutate(wcvp_accepted_id = ifelse(orig_name %in% manyaccepted$orig_name, wcvp_id, wcvp_accepted_id))
}

# --- 3. Clean Match Results (1 row per species) ---
matchresult_clean <- matchresult %>%
  mutate(
    status_priority = case_when(
      wcvp_status == "Accepted"     ~ 1,
      wcvp_status == "Synonym"      ~ 2,
      wcvp_status == "Illegitimate" ~ 3,
      wcvp_status == "Unplaced"     ~ 4,
      TRUE                          ~ 5
    )
  ) %>%
  group_by(orig_name) %>%
  slice_min(status_priority, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  matchwcvp_accepted_names(wcvp_names)

matchresult2_clean <- matchresult2 %>%
  mutate(
    status_priority = case_when(
      wcvp_status == "Accepted"     ~ 1,
      wcvp_status == "Synonym"      ~ 2,
      wcvp_status == "Illegitimate" ~ 3,
      wcvp_status == "Unplaced"     ~ 4,
      TRUE                          ~ 5
    )
  ) %>%
  group_by(orig_name) %>%
  slice_min(status_priority, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  matchwcvp_accepted_names(wcvp_names)

```


Next we will detect inconsistencies in the matched names and the accepted names and correct errors as we go. A record of the changes will be kept in a separate file. 


### Fix mismatches

The default is to assume that the WCVP matched name for the full name is the one to use going forward.  This will be overwritten in some cases below.  

```{r spot_mismatch}

mismatches <- matchresult_clean %>% 
  filter(orig_name != wcvp_name | 
           is.na(wcvp_name) | 
           is.na(wcvp_accepted_name)) %>%
  left_join(matchresult2_clean %>% 
              select(-orig_binomial) %>% 
              rename_with(~paste0(., "_binomial"), -orig_name),
            by = "orig_name") %>%
  mutate(
    # Extract wcvp genus, species, infraspecies rank and infraspecies
    wcvp_genus = word(wcvp_name, 1),
    wcvp_species = word(wcvp_name, 2),
    wcvp_infraspecies_rank = word(wcvp_name, 3),
    wcvp_infraspecies = word(wcvp_name, 4),
    
   # Add booleans for mismatches
    genus_mismatch = !is.na(orig_genus) & 
                      (orig_genus != wcvp_genus | is.na(wcvp_genus)),
    species_mismatch = !is.na(orig_species) & 
                        (orig_species != wcvp_species | is.na(wcvp_species)),
    infraspecies_rank_mismatch = !is.na(orig_infraspecies_rank) & 
                                  (orig_infraspecies_rank != wcvp_infraspecies_rank | 
                                   is.na(wcvp_infraspecies_rank)),
    infraspecies_mismatch = !is.na(orig_infraspecies) & 
                             (orig_infraspecies != wcvp_infraspecies | 
                              is.na(wcvp_infraspecies))
  )


printmismatches <- mismatches %>% 
  select(
    orig_name, wcvp_name, wcvp_accepted_name, match_similarity, 
    orig_binomial, wcvp_name_binomial, wcvp_accepted_name_binomial, match_similarity_binomial,
    genus_mismatch, species_mismatch, infraspecies_rank_mismatch, infraspecies_mismatch
  ) %>% 
  arrange(match_similarity, match_similarity_binomial)

datatable(printmismatches,options = list(scrollX = TRUE, pageLength = 10),
          caption="The name matched in WCVP differs from the original name. Sorted by lowest match_similarity.")


colsprint = c("orig_name", "wcvp_name","wcvp_accepted_name" ,"orig_binomial", "wcvp_name_binomial", "wcvp_accepted_name_binomial")


mismatches$preferred = "full"
```


#### Case 1: both full name and binomial name show a mismatch
```{r printmisses}
mismatchcase1 <- mismatches %>% 
  filter( orig_name != wcvp_name   &
            ( (orig_binomial != wcvp_name_binomial) | 
                is.na(wcvp_name_binomial)) & !is.na(orig_binomial) )

datatable( mismatchcase1 %>% select(all_of(colsprint)),
           caption="Both binomial and full name mismatched",
options = list(scrollX = TRUE, pageLength = 10))

```

In all these cases the original spelling differs slightly from the name matched in WCVP.  We go with the matched name in WCVP (the default).


#### Case 2: The full name matched except for the infraspecies rank
```{r case3}
mismatchcase3 <- mismatches %>% filter( infraspecies_rank_mismatch & !genus_mismatch & !species_mismatch & !infraspecies_mismatch) 
  
datatable( mismatchcase3 %>% select(all_of(colsprint)),
           caption="full name matched except for infraspecies rank",options = list(scrollX = TRUE, pageLength = 10))

```
We go with the matched name in WCVP (the default).

#### Case 3: Binomial matched perfectly, but no match on full name.
```{r findingcase2}
mismatchcase2 <- mismatches %>% filter( (orig_binomial == wcvp_name_binomial) &
                                          is.na(wcvp_name) | (
                      (orig_name != wcvp_name &
                      (genus_mismatch | species_mismatch) &
                       !is.na(orig_infraspecies) & orig_infraspecies!= ""))
                         ) 
datatable( mismatchcase2 %>% select(all_of(colsprint)),
           caption="Perfect match on binomial but no match on full name.",
           options = list(scrollX = TRUE, pageLength = 10)
           )


```
In this cases the match to binomial is better, so use that (overwriting the default).

```{r fixingcase3}

inc <- with(mismatches, orig_name %in% mismatchcase2$orig_name)
if (any(inc, na.rm = TRUE)) {
  tmp <- mismatches[inc, ]
  tmp$preferred = "binomial"
  mismatches[inc, ] <- tmp
}

```

#### Case 4: genus, species and rank matched but infraspecies didn't
```{r case4}
mismatchcase4 <- mismatches %>% filter( !genus_mismatch & !species_mismatch & infraspecies_mismatch & !infraspecies_rank_mismatch)

datatable( mismatchcase4 %>% select(all_of(colsprint)), caption="genus, species and infraspecific rank matched but infraspecies name did not ",options = list(scrollX = TRUE, pageLength = 10))


```

This is just a slight spelling difference in the subspecies name.  Go with the WCVP match on the full name (the default).

#### Other cases:
```{r case5}
# species not considered in the previous cases:
missrest <- mismatches %>% 
  filter(!orig_name %in% c(mismatchcase1$orig_name, mismatchcase2$orig_name, mismatchcase3$orig_name, mismatchcase4$orig_name))

datatable( missrest %>% select(all_of(colsprint)), caption="Remaining mismatched species",options = list(scrollX = TRUE, pageLength = 10))

```
In four of the five cases, WCVP matches the original name, but has no current accepted name.  The fifth case, Witheringia knappiae is not recognized by WCVP or by TNRS (separate check).  It has a forestgeo list synonym of Brachistus knappieae, which is recognized.  





#### Summary of mismatches after fixing:
List of wcvp columns to keep
```{r wcvpcolumns}

# list of wcvp columns we are interested to keep at the end.
wcvp_columns <- c(
        # columns of matched name
        "wcvp_name", "wcvp_authors", "wcvp_rank", 
        "wcvp_status", "wcvp_id", "wcvp_ipni_id", "match_similarity",
        
        # columns of accepted name
        "wcvp_accepted_id","wcvp_accepted_powo_id", "wcvp_accepted_ipni_id",
        "wcvp_accepted_binomial","wcvp_accepted_name", "wcvp_accepted_authority",
        "wcvp_accepted_family", "wcvp_accepted_genus",  "wcvp_accepted_species",
        "wcvp_accepted_infrarank", "wcvp_accepted_infraspecies",
        "wcvp_accepted_rank", 
        "wcvp_accepted_lifeform"
        )
```

Now all the mismatches should make more sense
```{r fixingmismatches}

# summary of mismatches
knitr::kable(mismatches %>% group_by(preferred) %>% summarise(species = n()))

# correct the mismatches with the binomial  matches when necessary 
mismatches <- mismatches %>% 
  mutate(
    across(
      all_of(c(wcvp_columns)),
      ~ifelse(
        preferred == "binomial",
        get(paste0(cur_column(), "_binomial")),
        .x
      )
    )
  )

datatable(mismatches %>% select(orig_name, wcvp_name, wcvp_accepted_name),
          options = list(scrollX = TRUE, pageLength = 10),
          caption="Species after setting the wcvp matched name to a closer match")

```



#### Joining mismatch fixes to the main list

```{r matcheshandling}

# --- 4. Handle  matches ---

# joins the corrected mismatches back to the original wcvp matched list
inc <- with(matchresult_clean, orig_name %in% mismatches$orig_name)
if (any(inc, na.rm = TRUE)) {
  corrections <- mismatches %>%
    select(all_of(c("orig_name", wcvp_columns)))
  
  matchresult_clean <- rows_update(matchresult_clean, corrections, 
                                   by = "orig_name")
}

# --- 5. Final Join and Rename ---
# This merges the WCVP column with the original columns of the list
processlist <- panwoodtaxafull %>%
  left_join(matchresult_clean %>% 
              select(all_of(c("orig_name", wcvp_columns))),
            by = "orig_name") %>%
  rename(
    # rename matched columns
    wcvp_matched_name = wcvp_name,
    wcvp_matched_authors = wcvp_authors,
    wcvp_matched_rank = wcvp_rank,
    wcvp_matched_status=wcvp_status,
    wcvp_matched_plant_name_id = wcvp_id,
    wcvp_matched_ipni_id = wcvp_ipni_id,
    
    wcvp_accepted_plant_name_id =  wcvp_accepted_id,
  ) %>%
  mutate(
  origmatches = orig_name == wcvp_matched_name
  )

# Review Results
knitr::kable(processlist %>% count(wcvp_matched_status))

datatable(subset(processlist,
             is.na(wcvp_accepted_name) | !wcvp_matched_status %in% c("Accepted", "Synonym") ) %>% 
        select(orig_family,orig_name,wcvp_matched_status,wcvp_accepted_name),
      options = list(scrollX = TRUE, pageLength = 10),
      caption = "Species not listed as Accepted or Synonym in WCVP")
```


# Make decision for current name

We will now put together a list with current_name, current_family, and current_authority, which provide the name that should be preferred after this exercise.  

At the end we have the same number of species. And we have columns:

current_name -> the name according to the criteria defined above
orig_name -> the name as it appears in the forestgeo table
wcvp_matched_name -> the name matched by wcvp 
wcvp_accepted name -> the accepted name according to wcvp

We also add a variable for whether the matched name is the same as the original ForestGEO name, and whether the accepted name is the same as the original ForestGEO name. 

The current name is set based on this rules: 

- The current name is the wcvp_accepted_name if it is not NA. Otherwise the current name is the original name. (Previously we made sure the wcvp_accepted_name are all from good matches, otherwise we keep the wcvp_accepted_name as NA.)

- EXCEPT when two or more species have the same wcvp_accepted_name, in that case we keep the original name

We have to be specially carefull with species such as the two varieties of Swartzia simplex, which the botanist are confident are two different varieties but they are not yet in WCVP.


1. First find the duplicate wcvp_accepted_names
```{r finddups}

acceptednamesdup <- processlist %>% 
  filter(!is.na(wcvp_accepted_name)) %>%
  group_by(wcvp_accepted_name) %>% 
  arrange(wcvp_accepted_name) %>% filter(n()>1) %>% ungroup() %>%
  mutate(mismatch = (orig_name != wcvp_matched_name)) %>% 
  arrange(wcvp_accepted_name, mismatch, match_similarity) %>%
  select(sp6, wcvp_accepted_name, orig_name, wcvp_matched_name, wcvp_matched_status, mismatch, match_similarity )

if( nrow(acceptednamesdup) > 0){
  datatable(acceptednamesdup,
            caption="Cases where two morphospecies are associated with the same WCVP accepted name",
            options = list(scrollX = TRUE, pageLength = 10),)
}



```

See all the species names changes:

```{r seechanges}

changes_names <- processlist %>% 
  filter( !is.na(wcvp_accepted_name) &  
            wcvp_accepted_name != orig_name & 
            !orig_name %in% acceptednamesdup$orig_name) %>%
  select(sp6,  orig_name, wcvp_accepted_name, wcvp_matched_name, wcvp_matched_status)


datatable(changes_names,
          caption="Changes to be applied, the original name is different from the wcvp_accepted_name and this is not a case in which two different morphospecies have the same WCVP accepted name.",
          options = list(scrollX = TRUE, pageLength = 10),)


```

2. Set current name, family and author based on the rules mentioned above.

```{r currentname choice}

processlist <- processlist %>% 
  mutate(
    # Use accepted name if available, otherwise use original name
    current_name = ifelse(is.na(wcvp_accepted_name) | orig_name %in% acceptednamesdup$orig_name, orig_name, wcvp_accepted_name),
    
    # Use accepted authority if available, otherwise use original authority
    current_authority = ifelse(is.na(wcvp_accepted_authority) | orig_name %in% acceptednamesdup$orig_name, orig_authority, wcvp_accepted_authority),
    
    # Use accepted family if available, otherwise use original family
    current_family = ifelse(is.na(wcvp_accepted_family) | orig_name %in% acceptednamesdup$orig_name, orig_family, wcvp_accepted_family),
    
    # Flag whether the name changed from original to WCVP accepted
    orig_name_changed_wcvp =  orig_name != current_name,
    
    # Flag whether the accepted name is duplicated
    wcvp_accepted_duplicated = orig_name %in% acceptednamesdup$orig_name
  )


```


## Check for missing information and consistency

The Unplaced  status can have missing information. 

```{r checkmissing, echo=TRUE}

processlist <- processlist %>% mutate(
  current_genus = word(current_name, 1),
  current_species = word(current_name, 2)
) %>% rename(orig_habit = habit)

missing_info <- processlist %>% filter(
   is.na(current_name) | is.na(current_family) | is.na(current_authority))   

if (nrow(missing_info)>0)
datatable(
  missing_info %>% select(sp6, current_name, orig_name,
         current_family, orig_family, current_authority, 
         orig_authority, orig_name_changed_wcvp, wcvp_matched_status),
  caption = "Species with missing taxonomic information in wcvp",
  options = list(scrollX = TRUE, pageLength = 10),
  rownames = FALSE
)

```


Now add in all the morphospecies lacking complete species IDs from the ViewTaxa list.  (There would be no point in running these through WCVP.)

```{r addpartialid}
treeplottaxapartialid <- treeplottaxa %>% filter(!sp6 %in% panwoodtaxafull$sp6) %>%  
              select(sp6 = sp6, orig_family = Family, orig_genus = Genus, 
                        orig_name = species_name, orig_species = SpeciesName,
                        orig_infraspecies = Subspecies, orig_infraspecies_rank = Rank,
                        orig_binomial = binomial,
                        orig_authority = authority, orig_synonyms = ListOfOldNames,
                        IDLevel
                        ) %>%
  mutate(current_name=orig_name,
         current_authority=orig_authority,
         current_family=orig_family,
         current_genus=orig_genus,
         current_species=orig_species,
         current_binomial=orig_binomial)

allpanwoodtaxa <- bind_rows(processlist,treeplottaxapartialid)

```

The woody species list now has `r nrow(allpanwoodtaxa)` species. 

# Add sp4 codes from Wright list

## Load Wright species list dataset: 

We will use the sp6 codes and species names to match records and merge in sp4 codes. 

Note that the current version of the Wright list that is loaded includes only taxa with IDlevel=="species", even though there may be some cases where morphospecies have sp6 and sp4 codes and the sp4 codes could otherwise be mapped over.  This affects only a handful of morphospecies, so ignored for now.  

```{r loadspwright, echo=TRUE}
# Joe Wright's taxonomy dataset (includes 4-letter codes and some non-woody species as well as trees and lianas)

# this list was already processed with the wcvp v13 names
allwrighttaxa <- read_excel(FNWRIGHTSPLISTIN)

# given that we are only interested in linking sp4, and in woody species, remove other cases
usewrighttaxa <- allwrighttaxa %>%
  filter(!is.na(sp4) & 
           (herb==0 | is.na(herb)) & 
           (epiphyte==0 | is.na(epiphyte)) & 
           (parasite==0 | is.na(parasite)))

# some species in wright are not proper binomials but instead morphospecies names
unidentified_species <- c(
  "Xylopia roubik", 
  "Nectandra fuzzy", 
  "Phyllanthus chagres",
  "Crossopetalum coba",
  "Leandra roja",
  "Mouriri coba", 
  "Guarea arbusto", 
  "Neea roja", 
  "Licania coba",
  "Drypetes coba",
  "Faramea coba",
  "Cassipourea hoja grande",
  "Guatteria flor roja", 
  "Licaria arco seco", 
  "Solanum enves purpura", 
  "Unonopsis peludo", 
  "Apeiba hybrid", 
  "Anthurium NA", 
  "Couepia chocolate", 
  "Crossopetalum coba", 
  "Guapira campochagres", 
  "Guarea sherman", 
  "Leandra roja", 
  "Licania coba", 
  "Plinia cerrojefe", 
  "Randia campo chagres", 
  "Sapium broadleaf", 
  "Cojoba hoja grande",
  "Meliosma hoja grande")


usewrighttaxa <- usewrighttaxa %>% 
  mutate(Idlevel = ifelse(orig_name %in% unidentified_species, "genus", Idlevel))

### check for consistency

# duplicated sp6 codes
sp6dups <- usewrighttaxa %>% filter(!is.na(sp6)) %>% group_by(sp6) %>% filter(n()>1) %>% ungroup()
  
if(nrow(sp6dups) > 0){
  knitr::kable(sp6dups %>% select(sp4, sp6, wcvp_accepted_name), caption="Duplicated sp6 codes in wright list")
} else{
  print("No duplicated sp6 codes in Wright nomenclature file.")
}

#duplicated sp4
sp4dups <- usewrighttaxa %>% filter(!is.na(sp4)) %>% group_by(sp4) %>% filter(n()>1) %>% ungroup()
  
if(nrow(sp4dups) > 0){
  knitr::kable(sp4dups %>% select(sp4, sp6, wcvp_accepted_name), caption="Duplicated sp6 codes in wright list")
} else{
  print("No duplicated sp4 codes in Wright nomenclature file.")
}
  

# duplicated names
spnamedups <- usewrighttaxa %>% group_by(orig_name) %>% filter(n()>1) %>% ungroup() %>% arrange(orig_name)
  
if(nrow(spnamedups) > 0){
    datatable(spnamedups %>% select(sp6, orig_name, wcvp_accepted_name, current_name), caption="Duplicated original species names in wright list",options = list(scrollX = TRUE, pageLength = 10))
} else{
    print("No duplicated original species names in Wright nomenclature file.")
}

# duplicated current names
spcurrnamedups <- usewrighttaxa %>% group_by(current_name) %>% filter(n()>1) %>% ungroup() %>% arrange(current_name)
  
if(nrow(spcurrnamedups) > 0){
    datatable(spcurrnamedups %>% select(sp6, wcvp_accepted_name, current_name), caption="Duplicated current species names in wright list",options = list(scrollX = TRUE, pageLength = 10))
} else{
    print("No duplicated current species names in Wright nomenclature file.")
}



```



There are `r nrow(usewrighttaxa)` species in the Wright list and `r nrow(allpanwoodtaxa)` in the Panama woody species list. So naturally. We expect some species won't have a sp4 code. 
Now we add the sp4 codes from the Wright list to our species list. 

We match the sp4 code by the following 
(1) on sp6 code if it works, otherwise
(2) on original names in the two lists if it works, otherwise
(3) on original binomials in the two lists if it works, otherwise
(4) on current names (post WCVP) in the two lists 

```{r sp4check, echo=TRUE}

# add sp4 codes from Wright list

m1 <- match(usewrighttaxa$sp6, allpanwoodtaxa$sp6)
m2 <- match(usewrighttaxa$orig_name,allpanwoodtaxa$orig_name)
m3 <- match(usewrighttaxa$orig_binomial,allpanwoodtaxa$orig_binomial)
m4 <- match(usewrighttaxa$current_name,allpanwoodtaxa$current_name)

m <- ifelse(!is.na(m1),m1,
               ifelse(!is.na(m2),m2,
                      ifelse(!is.na(m3),m3,m4)))

usewrighttaxa <- usewrighttaxa %>%
  mutate(sp6match = allpanwoodtaxa$sp6[m])

allpanwoodtaxa <- allpanwoodtaxa %>%
  mutate(sp4 = usewrighttaxa$sp4[match(allpanwoodtaxa$sp6,usewrighttaxa$sp6match)])


summary_matches <- data.frame(
    M1_sp6 = sum(!is.na(m1)),
    M2_orig_name = sum(!is.na(m2)),
    M2_notM1 = sum(!is.na(m2) & is.na(m1)),
    M3_orig_binomial = sum(!is.na(m3)),
    M3_notM1M2 = sum(!is.na(m3) & is.na(m1) & is.na(m2)),
    M4_current_name = sum(!is.na(m4)),
    M4_notM1M2M3 = sum(!is.na(m4) & is.na(m1) & is.na(m2) & is.na(m3)),
    No_Match = sum(is.na(m))
  )

# print table
knitr::kable(t(summary_matches), 
      caption = "Summary of species matching results for non-herbaceous Wright species with sp4 codes")

wrightnomatch <- usewrighttaxa[is.na(m),]

datatable(wrightnomatch %>% select(sp4,sp6,orig_name,current_name,free,herb, epiphyte) %>%
            arrange(free,orig_name),
          caption="Remaining Wright taxa without a match (among species that have a sp4 code and are not listed as herbs, epiphytes, or parasites.")
```

Check for duplicate sp4 after match
```{r checksp4match,  echo=TRUE}

sp4dups <- allpanwoodtaxa %>% filter(!is.na(sp4)) %>% group_by(sp4) %>% filter(n()>1) %>% ungroup()
  
if(nrow(sp4dups) > 0){
  knitr::kable(sp4dups %>% select(sp4, sp6, wcvp_accepted_name), caption="Duplicated sp4 codes after linking with Wright list")
} 

```


`r paste(sum(!is.na(allpanwoodtaxa$sp4)), "/", nrow(allpanwoodtaxa))` woody taxa were successfully assigned a sp4 code.



# Save files

Saving combined panama woody, changes in panama woody, sp4 codes that need revision:

```{r finaloutput}

list_wcvp_changes <- allpanwoodtaxa %>% filter(orig_name != wcvp_accepted_name)


write_xlsx(allpanwoodtaxa, PANAMA_COMBINED_OUT_PATH)
  
write_xlsx(list_wcvp_changes, PANAMA_COMBINED_CHANGE_SUMMARY_PATH)
  


```

# Alternate names file

```{r alternates}

alternate_synonym_table <- allpanwoodtaxa %>%

  mutate(OriginalOrder = row_number()) %>%
  
  select(
    OriginalOrder, sp6, sp4, current_name, orig_name, wcvp_matched_name, wcvp_accepted_name,
    SynonymTextALL = orig_synonyms
  ) %>%
  
  # Filter out empty synonym rows
  filter(!is.na(SynonymTextALL) & SynonymTextALL != "") %>%

  mutate(SynonymTextSingle = SynonymTextALL) %>%
  separate_longer_delim(SynonymTextSingle, delim = ",") %>%
  mutate(SynonymTextSingle = trimws(SynonymTextSingle)) %>%
  
  # Order of the synonym in the original text
  group_by(OriginalOrder) %>%
    mutate(SynonymNumber = row_number()) %>%
    ungroup() %>%
  
  # 6. Extract Genus and Species from the single synonym
  # Using too_many="merge" in case there are infraspecific names
  separate_wider_delim(
    SynonymTextSingle, 
    delim = " ", 
    names = c("SynonymGenusSingle", "SynonymSpeciesSingle"),
    too_many = "merge",
    cols_remove = FALSE
  ) %>% arrange(current_name) %>%
  select(-OriginalOrder)

# View the result
datatable(alternate_synonym_table, 
             caption = "Mapping: Original Name to Individual Synonym",
          options = list(scrollX = TRUE, pageLength = 10))

# make a file that gives the synonyms as 1-to-1 list
write_xlsx(alternate_synonym_table, FNSYNONYMS)


```

# Wright species missing from woody
```{r sp4missing}
# 4. Find codes in Wright list NOT present in the woody list AND 
# and that DO have #1 in any of the following columns: tree, midstory, understory, shrub, liana.  Include a column for their lifeform according to Wright (based on these columns)



wright_no_woody_match <- usewrighttaxa %>%
  mutate(
    # field if there is no match at all with the woody list
    no_woody_match = !(
      sp6 %in% na.omit(allpanwoodtaxa$sp6) |
      sp6 %in% na.omit(allpanwoodtaxa$sp6prior) |
      sp4 %in% na.omit(allpanwoodtaxa$sp4) |
      current_name %in% allpanwoodtaxa$current_name |
      orig_name %in% allpanwoodtaxa$current_name |
      orig_binomial %in% allpanwoodtaxa$orig_binomial |
      orig_name %in% alternate_synonym_table$SynonymTextSingle |
      current_name %in% alternate_synonym_table$SynonymTextSingle
    ),
    lifeform_wright = paste(
      if_else(!is.na(tree) & tree != 0, "TREE", "_"),
      if_else(!is.na(midstory) & midstory != 0, "MIDSTORY", "_"),
      if_else(!is.na(understory) & understory != 0, "UNDERSTORY", "_"),
      if_else(!is.na(shrub) & shrub != 0, "SHRUB", "_"),
      if_else(!is.na(liana) & liana != 0, "LIANA", "_"), # No longer breaks!
      if_else(!is.na(climber) & climber != 0, "CLIMBER", "_"),
      if_else(!is.na(free) & free != 0, "FREE", "_"),
      if_else(!is.na(vine) & vine != 0, "VINE", "_"),
      if_else(!is.na(herb) & herb != 0, "HERB", "_"),
      if_else(!is.na(epiphyte) & epiphyte != 0, "EPIPHYTE", "_"),
      if_else(!is.na(hemiepiphyte) & hemiepiphyte != 0, "HEMIEPIPHYTE", "_"),
      if_else(!is.na(parasite) & parasite != 0, "PARASITE", "_"),
      sep = ", "
    ),
    
    # Remove the underscores and extra commas
    # This regex looks for "_" followed by a comma or a comma followed by "_"
    lifeform_wright = str_remove_all(lifeform_wright, "_, | ,_|_"),
    
    # 3. Final polish to remove any leading/trailing commas or spaces
    lifeform_wright = str_trim(str_remove_all(lifeform_wright, "^, |, $")),
    
    # 4. If nothing is left, make it a true NA
    lifeform_wright = if_else(lifeform_wright == "", NA_character_, lifeform_wright)
           )

# Find codes in Wright NOT in woody list AND have #1 in woody-relevant columns
wright_sp6_no_woody_list <- wright_no_woody_match %>% 
  filter(no_woody_match, !is.na(sp6))

# Table of Wright woody species lacking 6-letter codes
# Note: I used a regex check on lifeform_wright to see if it contains 
# any of your 5 target woody types.
missing_wright_nosp6_probably_woody <- wright_no_woody_match %>%
  filter(
    no_woody_match,
    is.na(sp6),
    str_detect(lifeform_wright, "TREE|MIDSTORY|UNDERSTORY|SHRUB|LIANA")
  )

# Table of Wright unmatched species with 6-letter codes 
datatable(wright_sp6_no_woody_list %>%
            select(sp4, sp6, orig_name, current_name, lifeform_wright),
          caption = "Wright species with sp6 codes that are not in the woody list",
          options = list(scrollX = TRUE, pageLength = 10))

datatable(missing_wright_nosp6_probably_woody %>% select(sp4, sp6, orig_name, current_name,lifeform_wright),
          caption = "Wright species that might be woody and dont have a sp6 code and the name doesnt match anything in the woody list",
          options = list(scrollX = TRUE, pageLength = 10))


write_xlsx(wright_sp6_no_woody_list, WRIGHT_MISSING_PANWOODY_PATH)
write_xlsx(missing_wright_nosp6_probably_woody, WRIGHT_MISSING_PANWOODY_PATH2)
```

# Some checks on consistency of codes for higher level taxa:


```{r checkhigher, echo=TRUE, results="hide"}

allsp <- allpanwoodtaxa # the fixed list

# currwcvp_name is now the correct name, based on the fixes previously done. 
allsp <- allsp %>% mutate(
       genus = current_genus,
       family = current_family,
       genuscode=ifelse(is.na(sp6),NA,substr(sp6,1,4)))

# check that a single genus always has the same family:
genus_check1 <- allsp %>%
  group_by(genus) %>%
  summarise(n_families = n_distinct(family), .groups = "drop") %>%
  filter(n_families > 1)
genus_check1

# check whether a single genus always has the same lifeform:
genus_check2 <- allsp %>%
  group_by(genus) %>%
  summarise(
    habit_class = case_when(
      n_distinct(orig_habit) == 1 & first(orig_habit) == "Climbing"     ~ "Climbing only",
      n_distinct(orig_habit) == 1 & first(orig_habit) == "Freestanding" ~ "Freestanding only",
      n_distinct(orig_habit)  > 1                                  ~ "Both"
    ),
    .groups = "drop"
  ) %>%
  count(habit_class, name = "n_genera")
genus_check2

# to print genus and species with different habits
genera_both <- allsp %>%
  group_by(genus) %>%
  filter(n_distinct(orig_habit) > 1) %>%
  ungroup()

table_2rows <- genera_both %>%
  group_by(genus, orig_habit) %>%
  summarise(
    Species = paste(sort(unique(current_name)), collapse = ", "),
    .groups = "drop"
  ) %>%
  arrange(genus, orig_habit)


# check that a single genus always has the same first 4 letters of species code:
genus_check3 <- allsp %>%
  group_by(genus) %>%
  summarise(n_genuscode = n_distinct(genuscode), 
            genuscode_list = paste(sort(unique(genuscode)), collapse = ", "),
            .groups = "drop") %>%
  filter(n_genuscode > 1)
genus_check3

# find cases where same genuscode corresponds to more than 1 Accepted genus
genus_check4 <- allsp %>%
  group_by(genuscode) %>%
  summarise(
    Accepted_gen4_list = paste(sort(unique(substr(genus,1,4))),collapse=", "),
    Accepted_gen3_list = paste(sort(unique(substr(genus,1,3))),collapse=", "),
    n_Accepted_genus = n_distinct(genus),
    Accepted_genus_list = paste(sort(unique(genus)), collapse = ", "),
    .groups = "drop"
  ) %>%
  filter(n_Accepted_genus > 1)
genus_check4


# just get cases in which there are multiple genus names that have the same first 3 letters as the genus code:  
genus_check5 <- allsp %>%
  filter(!is.na(genuscode)) %>%
  mutate(prefix3 = substr(genuscode, 1, 3),
         genus_prefix3 = tolower(substr(genus, 1, 3))) %>%
  # keep only genus names matching the first 3 letters of genuscode
  filter(genus_prefix3 == prefix3) %>%
  group_by(genuscode) %>%
  summarise(
    n_distinct_genus = n_distinct(genus),
    Accepted_genus_list = paste(sort(unique(genus)), collapse = ", "),
    .groups = "drop"
  ) %>%
  filter(n_distinct_genus > 1) %>%
  arrange(genuscode)
genus_check5

# i changed this, the previous version could count species twice in different genus
genusdfall <- allsp %>%
  group_by(genus) %>%
  summarize(nsptree=sum(orig_habit=="Freestanding"),
            nspliana=sum(orig_habit=="Climbing")) %>%
  ungroup() %>% 
  mutate(Accepted_name=genus,
         Accepted_name_rank="genus")  %>%
  dplyr::select(Accepted_name,Accepted_name_rank,nsptree,nspliana)

#sanity check
sum(genusdfall$nspliana + genusdfall$nsptree)
nrow(allsp)

familydfall <- allsp %>%
  group_by(family) %>%
  summarize(nsptree=sum(orig_habit=="Freestanding"),
            nspliana=sum(orig_habit=="Climbing")) %>%
  ungroup() %>% 
  mutate(Accepted_name=family,
         Accepted_name_rank="family")  %>%
  dplyr::select(Accepted_name,Accepted_name_rank,nsptree,nspliana)

#sanity check
sum(familydfall$nspliana + familydfall$nsptree)
nrow(allsp)

# Merge genus and family in higher taxa
allhighertaxa <- rbind(genusdfall,familydfall) %>%
  mutate(lifeform=paste0(ifelse(nsptree>0,"A",""),ifelse(nspliana>0,"L",""))) %>%
  dplyr::select(Accepted_name,Accepted_name_rank,lifeform)

```
Checking consistency in higher taxa: 

1. Genus names related to more than 1 family: `r nrow(genus_check1)`

```{r genuscheck1}
if (nrow(genus_check1) > 0) { 
  datatable(genus_check1, caption="Genus with more than 1 family",
            options = list(scrollX = TRUE, pageLength = 10))
} 
```

2. Genus with species with different life forms: `r nrow(genus_check2$n_genera[genus_check2$habit_class == "Both"])` 

```{r genuscheck2}
if (nrow(genus_check2) > 0) { 
  datatable(genus_check2, caption="Genus lifeform. Check wether a single genus always has the same life form", options = list(scrollX = TRUE, pageLength = 10))
} 

numbergenus <- genus_check2 %>% filter(habit_class == "Both") %>% select(n_genera)
if(nrow(numbergenus) > 0){
  datatable(table_2rows, caption="Genus with species with two life forms",options = list(scrollX = TRUE, pageLength = 10))
}
```

3. Codes that don't match genus (probably because of name changes): `r nrow(genus_check3)`

```{r genuscheck3}
if (nrow(genus_check3) > 0) { 
  datatable(head(genus_check3, 10), caption="Genus with codes that dont match genus name",options = list(scrollX = TRUE, pageLength = 10))
} 
```

4. Codes that match to more than one genus (probably because of name changes): `r nrow(genus_check4)`

```{r genuscheck4}
if (nrow(genus_check4) > 0) { 
  datatable(head(genus_check4, 10), caption="Genus with codes that match more than one genus",options = list(scrollX = TRUE, pageLength = 10))
} 
```


5. Genus code matches genus but is present in more than one genus (probably because of name changes): `r nrow(genus_check5)`

```{r genuscheck5}
if (nrow(genus_check5) > 0) { 
  datatable(head(genus_check5,10), caption="Genus with the same 3 letters of code",options = list(scrollX = TRUE, pageLength = 10))
} 
```

## Write taxa summary

```{r taxasummary, echo=TRUE, results="hide"}

spdfall <- allsp %>%
  mutate(lifeform=ifelse(orig_habit=="Freestanding","A","L"),
         Accepted_name_rank = "species") %>%
  select(current_name,sp6,sp4,Accepted_name_rank,lifeform) %>%
  rename(Accepted_name=current_name,sp6=sp6)

outtaxa <- bind_rows(spdfall,allhighertaxa) %>%
  rename(namerank=Accepted_name_rank,taxaname=Accepted_name) %>%
  arrange(taxaname)

table(outtaxa$namerank)

if(FIX_SPECIES){
  write_xlsx(outtaxa,FNPANAMATAXA)
}
```

