---
title: "Overview - Check and merge species lists Markdown"
author: "Helene Muller-Landau, Paula Uzcategui"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Put this document in the main folder to run it.

When running this following changes, save the html output with the name and date in the reports folder.

This document is intended to check woody plant species lists for Panama, and merge them together to produce a single document with a complete set of codes and up-to-date names.

Source files:

1. *Panama woody plant species table*, specifically the most recent version from the folder "PanamaWoodySpLists" (filenames typically start with "FloraPanama" or similar).  This includes only fully recognized botanical species, omitting morphospecies lacking complete taxonomic information.  It is assembled and maintained by Smithsonian ForestGEO botanists in Panama.  It includes a 6-letter species code (mnemonic). 

 
2. *Panana plots woody species table*, specifically the most recent version from the folder "ForestGEO" (filenames typically start "ViewTaxonomy").  This includes all recognized morphospecies that appear in any of the ForestGEO plots. It differs from the Panama woody plant species list in that it omits species that are not in the plots, and adds morphospecies that do not yet have a full current taxonomic name (whether because their species ID has not yet been determined or they are new species).  It is assembled and maintained by Smithsonian ForestGEO botanists in Panama.  It includes a 6-letter species code (mnemonic). 

 
3. *Wright plant nomenclature table*, specifically the most recent version from the folder "Wright" (filenames start with "nomenclature").  This includes all plant morphospecies (including non-woody morphospecies) that are recorded in seed and seedling datasets collected at STRI under the supervision of S. Joseph Wright.  This is included in order to link the 4-letter codes used in these datasets. 


*The workflow here has the following steps: *


- Check each list individually.  Check for duplicate entries.  Check species names against WCVP (World Checklist of Vascular Plants).  Prior versions of this workflow checked names against TNRS. 
 

- Merge the Panama woody plant species and Panama plots tree species lists, making sure 6-letter codes are consistent.  Output any mismatches.


- Merge in sp4 codes from the Wright list. 


- Report on any apparent discrepancies / issues within this RMarkdown and in output files to check. 


To be added in the future - also include data from the following file, and check for consistency of codes, etc. 

Panama plot liana species table, specifically from the folder Schnitzer.  These are lists of liana species that are included in censuses of plots in central Panama conducted under the supervision of Stefan Schnitzer.  It includes a 6-letter species code that should align with the ForestGEO codes.  

## Settings
```{r setpath, include=FALSE}
rm(list=ls())
```

```{r libraries, echo=TRUE, results="hide", message=FALSE, warning=FALSE,}
library("readxl")
library("writexl")
library("dplyr")
library(tidyr)
library("stringi")
library("stringr")
suppressPackageStartupMessages({
  library(rWCVP)
  library(rWCVPdata)
})
library(knitr)
library(DT)

```

```{r initialsetup, echo=TRUE, results="hide"}
#rm(list=ls())

redoWCVP = FALSE

FIX_SPECIES = TRUE # a new file will be saved with the fixed species names. 

DIRINSP <- "splists_raw/"
DIRMIDSP <- "splists_mid/"
DIROUTSP <- "splists_out/"
DIRCHECK <- "tocheck/"

# input files # all in directory DIRINSP
FNWRIGHTSPLISTIN <- "splists_out/wright_nomenclature_R_20210224_WCVP_2026-02-27.xlsx"
FNFGEOSPLISTIN <- "Forestgeo/2025-12-04FromSuzanne/ViewTaxonomy_dec4_25.xlsx"
FNPANAMASPLISTIN <- "PanamaWoodySpLists/2025-12-04FromSuzanne/FloradePanama_dec4_2025_HM.xlsx" 


FNPANAMASPLISTWCVP <- paste0(DIRMIDSP,tools::file_path_sans_ext(basename(FNPANAMASPLISTIN)),"_WCVP_MATCH.rds")
FNPANAMASPLISTWCVP2 <- paste0(DIRMIDSP,tools::file_path_sans_ext(basename(FNPANAMASPLISTIN)),"_WCVP_MATCH2.rds") # binomial matches


PANAMA_COMBINED_OUT_PATH <- paste0(DIROUTSP,"CurrentPanamaWoody_",Sys.Date(),".xlsx")

PANAMA_COMBINED_CHANGE_SUMMARY_PATH <- paste0(DIROUTSP,"CurrentPanamaWoody_CHANGES_",Sys.Date(),".xlsx")

WRIGHT_MISSING_PANWOODY_PATH = paste0(DIROUTSP,"Wright_sp6_not_CurrentPanamaWoody_",Sys.Date(),".xlsx")
WRIGHT_MISSING_PANWOODY_PATH2 = paste0(DIROUTSP,"Wright_no_sp6_not_CurrentPanamaWoody_",Sys.Date(),".xlsx")
  
  
FNPANAMATAXA <- paste0(DIROUTSP,"AllTaxa_CurrentPanamaWoody_",Sys.Date(),".xlsx")

FNSYNONYMS <- paste0(DIROUTSP,"CurrentPanamaWoody_SYNONYMS_",Sys.Date(),".xlsx")

# the last is the list of all unique accepted species, genera, and families for photo labeling

```

We will: 

1. Make sure sp6 codes in the woody list are consistent with forestgeo codes

2. Match woody species names to wcvp

3. Match sp4 codes from wright list


## Load the ForestGEO viewtaxonomy dataset

We use this dataset to match species names with species sp6 codes

```{r loadsptreeplottaxa, echo=TRUE}
# ForestGEO dataset of taxonomy for species codes used in Panama plot censuses - includes morphospecies

treeplottaxa <- read_excel(paste0(DIRINSP,FNFGEOSPLISTIN))

treeplottaxa <- treeplottaxa %>%
mutate(across(everything(), ~ {
  if(is.character(.)) {
    na_if(., "NULL") # Solo aplica na_if si la columna es texto
  } else {
    . # Si es numérica, la deja igual
  }
}))

 #set the code to the subspecies code if it is a subspecies
treeplottaxa <- treeplottaxa %>% 
  mutate(
         sp6 = ifelse(is.na(Subspecies), Mnemonic, subspMnemonic), 
         author_name= ifelse(is.na(subspAuthority), Authority, subspAuthority), 
         binomial=ifelse(Genus=="Unidentified",paste(Genus,sp6),paste(Genus,SpeciesName)),
         species_name=ifelse(is.na(Subspecies),
                                paste0(Genus," ",SpeciesName),
                                paste0(Genus," ",SpeciesName," ",Rank, " ", Subspecies))
       )
  

treeplottaxa$binomial <- stri_trans_general(treeplottaxa$binomial,"Latin-ASCII")

sp6dups <- treeplottaxa %>% group_by(sp6) %>% filter(n()>1) %>% ungroup()
  
if(nrow(sp6dups) > 0){
  knitr::kable(sp6dups %>% select(sp6, species_name), caption="Duplicated sp6 codes in forestgeo viewtaxa")
} else{
  print("No duplicated codes")
}
  

# check duplicates with idlevel species
spnamedups <- treeplottaxa %>% filter(IDLevel =="species") %>% group_by(species_name) %>% filter(n()>1) %>% ungroup() %>% arrange(species_name)
  
if(nrow(spnamedups) > 0){
    datatable(spnamedups %>% select(sp6, species_name, IDLevel), caption="Duplicated species names in forestgeo viewtaxa",options = list(scrollX = TRUE, pageLength = 10))
} else{
    print("No duplicated species names")
}
  
# leave because they have codes

# forestgeo stats
kable(as.data.frame(table(treeplottaxa$IDLevel)) %>% rename(ID_Level=Var1, N_MorphoSpecies=Freq),
      caption="ForestGEO plot taxa list")


```

## Load Wright species list dataset: 

We will use the sp6 codes and species names to find matches for sp4 codes. 

```{r loadspwright, echo=TRUE}
# Joe Wright's taxonomy dataset (includes 4-letter codes and some non-woody species as well as trees and lianas)

# this list was already processed with the wcvp v13 names
usewrighttaxa <- read_excel(FNWRIGHTSPLISTIN)

### check for consistency

# duplicated sp6 codes
sp6dups <- usewrighttaxa %>% filter(!is.na(sp6)) %>% group_by(sp6) %>% filter(n()>1) %>% ungroup()
  
if(nrow(sp6dups) > 0){
  knitr::kable(sp6dups %>% select(sp4, sp6, wcvp_accepted_name), caption="Duplicated sp6 codes in wright list")
} else{
  print("No duplicated codes")
}

#duplicated sp4
sp4dups <- usewrighttaxa %>% filter(!is.na(sp4)) %>% group_by(sp4) %>% filter(n()>1) %>% ungroup()
  
if(nrow(sp4dups) > 0){
  knitr::kable(sp4dups %>% select(sp4, sp6, wcvp_accepted_name), caption="Duplicated sp6 codes in wright list")
} else{
  print("No duplicated codes")
}
  

# duplicated names
spnamedups <- usewrighttaxa %>% group_by(orig_name, wcvp_accepted_name) %>% filter(n()>1) %>% ungroup() %>% arrange(wcvp_accepted_name)
  
if(nrow(spnamedups) > 0){
    datatable(spnamedups %>% select(sp6, wcvp_accepted_name, current_name), caption="Duplicated species names in wright list",options = list(scrollX = TRUE, pageLength = 10))
} else{
    print("No duplicated species names")
}


# some species in wright are unidentified
unidentified_species <- c(
  "Xylopia roubik", 
  "Nectandra fuzzy", 
  "Phyllanthus chagres",
  "Crossopetalum coba",
  "Leandra roja",
  "Mouriri coba", 
  "Guarea arbusto", 
  "Neea roja", 
  "Licania coba",
  "Drypetes coba",
  "Faramea coba",
  "Cassipourea hoja grande",
  "Guatteria flor roja", 
  "Licaria arco seco", 
  "Solanum enves purpura", 
  "Unonopsis peludo", 
  "Apeiba hybrid", 
  "Anthurium NA", 
  "Couepia chocolate", 
  "Crossopetalum coba", 
  "Guapira campochagres", 
  "Guarea sherman", 
  "Leandra roja", 
  "Licania coba", 
  "Plinia cerrojefe", 
  "Randia campo chagres", 
  "Sapium broadleaf", 
  "Cojoba hoja grande",
  "Meliosma hoja grande")


usewrighttaxa <- usewrighttaxa %>% mutate(Idlevel = ifelse(orig_name %in% unidentified_species, "genus", Idlevel))

```
 
## Load woody species list to be processed

```{r loadsppanama, echo=TRUE}

# ForestGEO Panama woody plant species list 
panwoodtaxa <- read_excel(paste0(DIRINSP,FNPANAMASPLISTIN),col_types="text")  %>%
  mutate(
    across(
      c(family, genus, speciesname, authority, subspecies),
      str_squish
    )
  ) %>%
  rename(
    orig_family       = family,
    orig_genus        = genus,
    orig_species      = speciesname,
    orig_name_author  = authority,
    orig_infraspecies = subspecies,
    orig_synonyms = synonyms,
    sp6curr=mnemonic,
    vouchers=herbarium,
    habit=liana) %>%
  mutate(sp6curr=tolower(sp6curr),
         habit=ifelse(habit=="l","Climbing","Freestanding"),
         orig_name=ifelse(is.na(orig_infraspecies),
                                paste0(orig_genus," ",orig_species),
                                paste0(orig_genus," ",orig_species," ",orig_infraspecies)),
         orig_name = str_trim(orig_name),
         orig_infraspecies_rank = word(orig_infraspecies,1),
         orig_infraspecies = word(orig_infraspecies,2),
         habit=ifelse(is.na(habit),"Freestanding",habit),
         sp6prior=treeplottaxa$sp6[match(orig_name,treeplottaxa$species_name)])


```

Number of species in Panama woody plant species list: `r paste(nrow(panwoodtaxa))`

### Check for consistency of sp6 codes

We initially match sp6 prior using the raw species names

```{r checkmalformed, echo=TRUE}
# after matching sp6 prior by species name:

panwoodtaxa <- panwoodtaxa %>% mutate(
  sp6 = coalesce(sp6curr, sp6prior),
  binomial = paste0(orig_genus, " ", orig_species))

#  check for duplicated codes
sp6dups <- panwoodtaxa %>% filter(!is.na(sp6)) %>% group_by(sp6) %>% 
  arrange(sp6) %>% filter(n()>1) %>% ungroup()
 
sp6problems <- panwoodtaxa %>%
  filter(
    is.na(sp6) | 
      (!is.na(sp6) & !is.na(sp6prior) & sp6 != sp6prior) | 
      (!is.na(sp6) & nchar(sp6) != 6) | 
      (sp6 %in% sp6dups)
  ) %>%
  mutate(problem_reason = case_when(
    is.na(sp6)                ~ "Code is missing (NULL)",
    sp6 != sp6prior           ~ "Code mismatch with prior record",
    nchar(sp6) != 6           ~ "Code is not 6 characters",
    sp6 %in% sp6dups          ~ "Duplicate species code",
    TRUE                      ~ "Unknown error"
  ))

if(nrow(sp6problems) > 0){
  knitr::kable(sp6problems %>% select(sp6, sp6curr, sp6prior, orig_name), caption="Species code problems")
} else { print("No inconsistency with the codes")}


```

```{r}
panwoodtaxa <- panwoodtaxa %>%
  filter(sp6!="swarsu")  # remove a morphospecies lacking full species ID

```


### Check duplicated names
```{r checkduplicate, echo=TRUE, results="hide"}
# check for duplicate binomials and address if needed
bindups <- panwoodtaxa %>% group_by(orig_name) %>% arrange(orig_name) %>% filter(n()>1) %>% ungroup()

nrow(bindups)
```

- Duplicated species names: `r paste(nrow(bindups))`

```{r showbindups}
if (nrow(bindups) > 0) { 
    knitr::kable(bindups %>% select(orig_name, sp6curr), captio="Duplicated species names")
    write_xlsx(bindups,paste0(DIRCHECK,tools::file_path_sans_ext(basename(FNPANAMASPLISTIN)),"_nombreduplicado.xlsx"))
}
``` 

⚠️ If there are duplicates, manually check and leave the correct species in the "_touse" file:
```{r manuadupcheck}
if (nrow(bindups)>0 & file.exists(paste0(DIRCHECK,tools::file_path_sans_ext(basename(FNPANAMASPLISTIN)),"_touse.xlsx"))) 
  {
  # currently no duplicates!
  # previously, I manually went through and picked one of the entries as correct for each duplicated orig_name
  # exception is Swartzia simplex, where both are good
  panwoodtaxa <- filter(panwoodtaxa, !orig_name %in% bindups$orig_name)
  gooddups <- read_excel(paste0(DIRCHECK,tools::file_path_sans_ext(basename(FNPANAMASPLISTIN)),"_touse.xlsx"))
  panwoodtaxa <- rbind(panwoodtaxa,gooddups)
} else{
  print("Found no file to repair duplicates")
}
```


### Check for consistency between the Panama woody plant species list and the ForestGEO plots taxa list.  

These are both assembled by ForestGEO, and generally updated in tandem, so should be fully consistent.  

Check that all taxa in the ViewTaxonomy list that are identified to species are also in the Panama woody plants list. 

```{r checkforestgeoconsistency}
# check for 6-letter codes in the Fgeo codes list that are not in the Panama woody plant species list (does not matter if they dont have full species id)
sp6dropped <- subset(treeplottaxa,!sp6 %in% panwoodtaxa$sp6) %>%
      mutate(binomial_in_flora=binomial %in% panwoodtaxa$binomial) %>%
      arrange(sp6)

if(nrow(sp6dropped) > 0){
  datatable(
    sp6dropped %>% select(sp6, species_name, binomial_in_flora),
    caption="Species in forestgeo list not in tha panama woody list. binomial_in_flora = TRUE means the binomial name does show in   forestgeo list",
    options = list(scrollX = TRUE, pageLength = 10),
    rownames = FALSE
  )

} else{ print("All identified species in forestgeo are in the panama woody list")}

# check fo na prior
sp6match_namenomatch <- panwoodtaxa %>% filter(is.na(sp6prior)) %>% left_join(treeplottaxa %>% select(sp6, species_name, Subspecies), by=c("sp6curr" = "sp6")) %>% filter(!is.na(species_name)) %>% 
  select(sp6curr, sp6prior, orig_name, species_name, Subspecies) %>%
  rename(sp6= sp6curr, species_name_panwood = orig_name, species_name_treeplottaxa = species_name, subspecies_treeplottaxa = Subspecies  )
  
if(nrow(sp6match_namenomatch) > 0){
  knitr::kable(sp6match_namenomatch, caption="Species code in treeplottaxa matches with panama wood but the species name dont match")
} else { print("No inconsistency with the same code pointing to different names in fgeo and panama woody")}

```

Adding species in forestgeo Viewtaxonomy that are not in the woodylist
```{r addsp, echo=TRUE, results="hide"}

# add species in forestgeo that are not in panama woody
panwoodtaxa <- bind_rows(
  panwoodtaxa %>% mutate(IDLevel = "species"),
  sp6dropped %>% select(sp6 = sp6, sp6curr = sp6, orig_family = Family, orig_genus = Genus, 
                        orig_name = species_name, orig_species = SpeciesName,
                        orig_infraspecies = Subspecies, orig_infraspecies_rank = Rank,
                        binomial = binomial,
                        orig_name_author = author_name, orig_synonyms = ListOfOldNames,
                        IDLevel
                        ) 
)

```

Working with: `r nrow(panwoodtaxa)` morphospecies

# Match species names with WCVP v13:

Many names in the original file will change after using WCVP but mainly because of spelling differences. The workflow to find the accepted wcvp name for each species is the following: 

**1. Select only the rows identified at the species level **

**2. Match species names against WCVP (World Checklist of Vascular Plants)**

**3. Identify and review potential mismatches:**

- A mismatch occurs when WCVP returns a name different from the original lookup name. This can lead to incorrect accepted names, requiring manual verification. Most mismatches arise from minor spelling differences between the original names and WCVP entries. Others occur because some infraspecies are not in WCVP. And very weird mismatches occur when WCVP fuzzy matching points to the wrong name.
      
**4. Correct identified mismatches.**

**5. Set a current_name column following this rules: **

Set current_name to wcvp_name:

1. If species spelling differs, we keep the wcvp spelling
2. If species name is a synonym or illegitimate we keep the wcvp name

Otherwise keep the orig_name as current name if:

1. We couldnt find a wcvp accepted name. (The taxon is not in wcvp)
2. If the WCVP accepted match means that 2 or more forestgeo species would have the same WCVP name.
4. The wcvp name doesnt have the subspecies in the original name

At the end we have the same number of species. And we have columns:

current_name -> the name according to the criteria defined above
orig_name -> the name as it appears in the forestgeo table
wcvp_name -> the accepted name according to wcvp
wcvp_matched_name -> the name matched by wcvp 

Among other useful columns returned by wcvp...


```{r create_processlist}

# we will make changes in this list and then merge back to the original list

origlist <- panwoodtaxa %>% mutate(
      orig_binomial = str_squish(paste0(orig_genus, " ", 
                               ifelse(substr(orig_species, 1, 3) == "sp." | IDLevel !="species", "", orig_species))))

```

Working with `r nrow(origlist)` species. 
 
### Match WCVP names

```{r mergewcvp}

#########################  WCVP names ##################################

# Helper function to merge and format accepted names from WCVP backbone
matchwcvp_accepted_names <- function(matcheswcvp, wcvp_backbone){
  matcheswcvp %>% 
    left_join(
      wcvp_backbone %>% 
        select(
          plant_name_id,
          # rename immediately
          wcvp_accepted_ipni_id  = ipni_id,
          wcvp_accepted_powo_id  = powo_id,
          wcvp_accepted_family   = family,
          wcvp_accepted_name_rank = taxon_rank,
          wcvp_wcvp_accepted_name_status = taxon_status,
          wcvp_lifeform_description = lifeform_description,

          # rename after the mutate for readability 
          genus, species, infraspecific_rank, infraspecies,
          parenthetical_author, primary_author
        ),
      by = c("wcvp_accepted_id" = "plant_name_id")
    ) %>% 
    mutate( 
      wcvp_accepted_name = case_when(
        is.na(species) | species == "" ~ genus,
        !is.na(infraspecies) & infraspecies != "" ~ paste(genus, species, infraspecific_rank, infraspecies),
        TRUE ~ paste(genus, species)
      ),
      
      wcvp_accepted_authority =
        paste0(
          ifelse(!is.na(parenthetical_author), paste0("(", parenthetical_author, ") "),""),
          coalesce(primary_author, "")
        )
    ) %>% 
    rename(
      wcvp_accepted_author = primary_author,
      wcvp_accepted_name_genus = genus,
      wcvp_accepted_name_species = species,
      wcvp_accepted_name_infraspecific_rank = infraspecific_rank,
      wcvp_accepted_name_infraspecies = infraspecies
    )
}

# --- 1. Run or Load Match ---
if(file.exists(FNPANAMASPLISTWCVP) & !redoWCVP){
  matchresult <- readRDS(FNPANAMASPLISTWCVP)
} else {
  # Run the matching process
  matchresult <- wcvp_match_names(subset(origlist,IDLevel == "species"), wcvp_names, name_col = "orig_name")

  # Save as RDS for speed
  saveRDS(matchresult, FNPANAMASPLISTWCVP)
}

# do the same for the orig_binomial in cases where this is not equal to orig_name
if(file.exists(FNPANAMASPLISTWCVP2) & !redoWCVP){
  matchresult2 <- readRDS(FNPANAMASPLISTWCVP2)
} else {
  matchresult2 <- wcvp_match_names(subset(origlist,orig_name!=orig_binomial), wcvp_names, 
                                   name_col = "orig_binomial")

  # Save as RDS for speed
  saveRDS(matchresult2, FNPANAMASPLISTWCVP2)
}


# --- 2. Sanity Checks & Duplicate Resolution ---

# Identify names that mapped to multiple 'Accepted' records
manyaccepted <- matchresult %>%
  group_by(orig_name) %>%
  filter(sum(wcvp_status == "Accepted", na.rm = TRUE) > 1) %>%
  ungroup()

# If duplicates exist, force use of the original ID to avoid grouping explosion
if (nrow(manyaccepted) > 0) {
  matchresult <- matchresult %>%
    mutate(wcvp_accepted_id = ifelse(orig_name %in% manyaccepted$orig_name, wcvp_id, wcvp_accepted_id))
}

# --- 3. Clean Match Results (1 row per species) ---
matchresult_clean <- matchresult %>%
  mutate(
    status_priority = case_when(
      wcvp_status == "Accepted"     ~ 1,
      wcvp_status == "Synonym"      ~ 2,
      wcvp_status == "Illegitimate" ~ 3,
      wcvp_status == "Unplaced"     ~ 4,
      TRUE                          ~ 5
    )
  ) %>%
  group_by(orig_name) %>%
  slice_min(status_priority, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  matchwcvp_accepted_names(wcvp_names)

matchresult2_clean <- matchresult2 %>%
  mutate(
    status_priority = case_when(
      wcvp_status == "Accepted"     ~ 1,
      wcvp_status == "Synonym"      ~ 2,
      wcvp_status == "Illegitimate" ~ 3,
      wcvp_status == "Unplaced"     ~ 4,
      TRUE                          ~ 5
    )
  ) %>%
  group_by(orig_name) %>%
  slice_min(status_priority, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  matchwcvp_accepted_names(wcvp_names)

```


Next we will detect inconsistencies in the matched names and the accepted names and correct errors as we go. A record of the changes will be kept in a separate file. 


### Fix mismatches

```{r spot_mismatch}

mismatches <- matchresult_clean %>% 
  filter(orig_name != wcvp_name | is.na(wcvp_name) | is.na(wcvp_accepted_name)) %>%
  left_join(matchresult2_clean %>% select(-orig_binomial) %>% rename_with(~paste0(., "_binomial"), -orig_name),
            by = "orig_name") %>%
  mutate(
    # Extract wcvp genus, species, infraspecies rank and infraspecies
    wcvp_genus = word(wcvp_name, 1),
    wcvp_species = word(wcvp_name, 2),
    wcvp_infraspecies_rank = word(wcvp_name, 3),
    wcvp_infraspecies = word(wcvp_name, 4),
    
   # Add booleans for mismatches
    genus_mismatch = !is.na(orig_genus) & 
                      (orig_genus != wcvp_genus | is.na(wcvp_genus)),
    species_mismatch = !is.na(orig_species) & 
                        (orig_species != wcvp_species | is.na(wcvp_species)),
    infraspecies_rank_mismatch = !is.na(orig_infraspecies_rank) & 
                                  (orig_infraspecies_rank != wcvp_infraspecies_rank | 
                                   is.na(wcvp_infraspecies_rank)),
    infraspecies_mismatch = !is.na(orig_infraspecies) & 
                             (orig_infraspecies != wcvp_infraspecies | 
                              is.na(wcvp_infraspecies))
  )


printmismatches <- mismatches %>% 
  select(
    orig_name, wcvp_name, wcvp_accepted_name, match_similarity, 
    orig_binomial, wcvp_name_binomial, wcvp_accepted_name_binomial, match_similarity_binomial,
    genus_mismatch, species_mismatch, infraspecies_rank_mismatch, infraspecies_mismatch
  ) %>% 
  arrange(match_similarity, match_similarity_binomial)

datatable(printmismatches,options = list(scrollX = TRUE, pageLength = 10))


colsprint = c("orig_name", "wcvp_name","wcvp_accepted_name" ,"orig_binomial", "wcvp_name_binomial", "wcvp_accepted_name_binomial")


mismatches$preferred = "full"
```

#### Case 1:
```{r printmisses}
# both binomial and full mismatched
mismatchcase1 <- mismatches %>% filter( orig_name != wcvp_name   &
                        ( (orig_binomial != wcvp_name_binomial) | is.na(wcvp_name_binomial)) & !is.na(orig_binomial) )

datatable( mismatchcase1 %>% select(all_of(colsprint)),
           caption="both binomial and full name mismatched",
options = list(scrollX = TRUE, pageLength = 10))

```

In this cases the original spelling differs slightly from the name matched in WCVP. 


#### Case 2:
```{r case3}
# full name matched except infraspecies rank
mismatchcase3 <- mismatches %>% filter( infraspecies_rank_mismatch & !genus_mismatch & !species_mismatch & !infraspecies_mismatch) 
  
datatable( mismatchcase3 %>% select(all_of(colsprint)),
           caption="full name matched except infraspecies rank",options = list(scrollX = TRUE, pageLength = 10))

```

The match is different only in infraspecific rank. Use the WCVP accepted name for the full match. 

#### Case 3:
```{r findingcase2}
# full name mismatched or didint match to anything, binomial didin't. And genus and species did not mismatch
# this is because the infraspecies is messing the fuzzy matching
mismatchcase2 <- mismatches %>% filter( (orig_binomial == wcvp_name_binomial) &
                                          is.na(wcvp_name) | (
                      (orig_name != wcvp_name &
                      (genus_mismatch | species_mismatch) &
                       !is.na(orig_infraspecies) & orig_infraspecies!= ""))
                         ) 
datatable( mismatchcase2 %>% select(all_of(colsprint)),
           caption="Mismatch for full name involving a mismatch on genus or species, and at the same time perfect match when just on binomial.",
           options = list(scrollX = TRUE, pageLength = 10)
           )


```
In this cases the match to binomial is better, so use that. 

```{r fixingcase3}

inc <- with(mismatches, orig_name %in% mismatchcase2$orig_name)
if (any(inc, na.rm = TRUE)) {
  tmp <- mismatches[inc, ]
  tmp$preferred = "binomial"
  mismatches[inc, ] <- tmp
}

```

#### Case 4:
```{r case4}
# genus, species and rank matched but infraspecies didint 
mismatchcase4 <- mismatches %>% filter( !genus_mismatch & !species_mismatch & infraspecies_mismatch & !infraspecies_rank_mismatch)

datatable( mismatchcase4 %>% select(all_of(colsprint)), caption="genus, species and infraspecific rank matched but infraspecies name did not ",options = list(scrollX = TRUE, pageLength = 10))


```

In all of this cases the subspecies did not match in wcvp, but by a slight spelling difference.  Thus using the full match

#### Other cases:
```{r case5}
# species not considered in the previous cases:
missrest <- mismatches %>% 
  filter(!orig_name %in% c(mismatchcase1$orig_name, mismatchcase2$orig_name, mismatchcase3$orig_name, mismatchcase4$orig_name))

datatable( missrest %>% select(all_of(colsprint)), caption="Species not considered in previous cases",options = list(scrollX = TRUE, pageLength = 10))

```
For all of these, there is no WCVP match to the full name. And it is not better to use the binomial. 

Iin the cases where there is no match, it is because of a lack of taxonomic resolution in the original ID (just to family).  


#### Summary of mismatches after fixing:
List of wcvp columns to keep
```{r wcvpcolumns}

# list of wcvp columns we are interested to keep at the end.
wcvp_columns <- c(
        # columns of matched name
        "wcvp_name", "wcvp_authors", "wcvp_rank", "wcvp_status", "wcvp_id", "wcvp_ipni_id", "match_similarity",
        
        # columns of accepted name
        "wcvp_accepted_id","wcvp_accepted_powo_id", "wcvp_accepted_ipni_id",
        "wcvp_accepted_name", "wcvp_accepted_name_genus",  "wcvp_accepted_name_species",
        "wcvp_accepted_name_rank", "wcvp_accepted_family", 
        "wcvp_accepted_name_infraspecific_rank", "wcvp_accepted_name_infraspecies",
        "wcvp_accepted_author", "wcvp_accepted_authority",
        "wcvp_lifeform_description"
        )
```

Now all the mismatches should make more sense
```{r fixingmismatches}

# summary of mismatches
knitr::kable(mismatches %>% group_by(preferred) %>% summarise(species = n()))

# correct the mismatches with the binomial  matches when necessary 
mismatches <- mismatches %>% 
  mutate(
    across(
      all_of(c(wcvp_columns)),
      ~ifelse(
        preferred == "binomial",
        get(paste0(cur_column(), "_binomial")),
        .x
      )
    )
  )

datatable(mismatches %>% select(orig_name, wcvp_name, wcvp_accepted_name),
          options = list(scrollX = TRUE, pageLength = 10),
          caption="Species after setting the wcvp matched name to a closer match")

```



#### Joining mismatch fixes to the main list

```{r matcheshandling}

# --- 4. Handle  matches ---

# joins the corrected mismatches back to the original wcvp matched list
inc <- with(matchresult_clean, orig_name %in% mismatches$orig_name)
if (any(inc, na.rm = TRUE)) {
  corrections <- mismatches %>%
    select(all_of(c("orig_name", wcvp_columns)))
  
  matchresult_clean <- rows_update(matchresult_clean, corrections, 
                                   by = "orig_name")
}

# --- 5. Final Join and Rename ---
# This merges the WCVP column with the original columns of the list
processlist <- origlist %>%
  left_join(matchresult_clean %>% 
              select(all_of(c("orig_name", wcvp_columns))),
            by = "orig_name") %>%
  rename(
    # rename matched columns
    wcvp_matched_name = wcvp_name,
    wcvp_matched_authors = wcvp_authors,
    wcvp_matched_rank = wcvp_rank,
    wcvp_matched_status=wcvp_status,
    wcvp_matched_plant_name_id = wcvp_id,
    wcvp_matched_ipni_id = wcvp_ipni_id,
    
    wcvp_accepted_plant_name_id =  wcvp_accepted_id,
  ) %>%
  mutate(
  origmatches = orig_name == wcvp_matched_name
  )

# Review Results
knitr::kable(processlist %>% count(wcvp_matched_status))

datatable(subset(processlist,
             is.na(wcvp_accepted_name) | !wcvp_matched_status %in% c("Accepted", "Synonym") ) %>% 
        select(orig_family,orig_name,wcvp_matched_status,wcvp_accepted_name),
      options = list(scrollX = TRUE, pageLength = 10),
      caption = "Species not listed as Accepted or Synonym in WCVP")
```


# Make decision for current name

We will now put together a list with current_name, current_family, and current_authority, which provide the name that should be preferred after this exercise.  

At the end we have the same number of species. And we have columns:

current_name -> the name according to the criteria defined above
orig_name -> the name as it appears in the forestgeo table
wcvp_matched_name -> the name matched by wcvp 
wcvp_accepted name -> the accepted name according to wcvp

We also add a variable for whether the matched name is the same as the original ForestGEO name, and whether the accepted name is the same as the original ForestGEO name. 

The current name is set based on this rules: 

- The current name is the wcvp_accepted_name if it is not NA. Otherwise the current name is the original name. (Previously we made sure the wcvp_accepted_name are all from good matches, otherwise we keep the wcvp_accepted_name as NA.)

- EXCEPT when two or more species have the same wcvp_accepted_name, in that case we keep the original name

We have to be specially carefull with species such as the two varieties of Swartzia simplex, which the botanist are confident are two different varieties but they are not yet in WCVP.


1. First find the duplicate wcvp_accepted_names
```{r finddups}

acceptednamesdup <- processlist %>% 
  filter(!is.na(wcvp_accepted_name)) %>%
  group_by(wcvp_accepted_name) %>% 
  arrange(wcvp_accepted_name) %>% filter(n()>1) %>% ungroup() %>%
  mutate(mismatch = (orig_name != wcvp_matched_name)) %>% 
  arrange(wcvp_accepted_name, mismatch, match_similarity) %>%
  select(sp6, wcvp_accepted_name, orig_name, wcvp_matched_name, wcvp_matched_status, mismatch, match_similarity )

if( nrow(acceptednamesdup) > 0){
  datatable(acceptednamesdup,
            caption="Two species have now the same accepted name",
            options = list(scrollX = TRUE, pageLength = 10),)
}



```

See all the species names changes:

```{r seechanges}

changes_names <- processlist %>% 
  filter( !is.na(wcvp_accepted_name) &  wcvp_accepted_name != orig_name & !orig_name %in% acceptednamesdup$orig_name) %>%
  select(sp6,  orig_name, wcvp_accepted_name, wcvp_matched_name, wcvp_matched_status)


datatable(changes_names,
          caption="Changes to be applied, the original name is different from the wcvp_name",
          options = list(scrollX = TRUE, pageLength = 10),)


```

2. Set current name, family and author based on the rules mentioned above.

```{r currentname choice}

processlist <- processlist %>% 
  mutate(
    # Use accepted name if available, otherwise use original name
    current_name = ifelse(is.na(wcvp_accepted_name) | orig_name %in% acceptednamesdup$orig_name, orig_name, wcvp_accepted_name),
    
    # Use accepted author if available, otherwise use original author
    current_author = ifelse(is.na(wcvp_accepted_author) | orig_name %in% acceptednamesdup$orig_name, orig_name_author, wcvp_accepted_author),
    
    current_authority = ifelse(is.na(wcvp_accepted_authority) | orig_name %in% acceptednamesdup$orig_name, orig_name_author, wcvp_accepted_authority),
    
    # Use accepted family if available, otherwise use original family
    current_family = ifelse(is.na(wcvp_accepted_family) | orig_name %in% acceptednamesdup$orig_name, orig_family, wcvp_accepted_family),
    
    # Flag whether the name changed from original to WCVP accepted
    orig_name_changed_wcvp =  orig_name != current_name,
    
    # Flag whether the accepted name is duplicated
    wcvp_accepted_duplicated = orig_name %in% acceptednamesdup$orig_name
  )


```


## Check for missing information and consistency

The Unplaced  status can have missing information. 

```{r checkmissing, echo=TRUE}

processlist <- processlist %>% mutate(
  current_genus = word(current_name, 1),
  current_species = word(current_name, 2)
) %>% rename(orig_habit = habit)

missing_info <- processlist %>% filter(
   is.na(current_name) | is.na(current_family) | is.na(current_author))   


datatable(
  missing_info %>% select(sp6, current_name, orig_name,
         current_family, orig_family, current_author, 
         orig_name_author, orig_name_changed_wcvp, wcvp_matched_status),
  caption = "Species with missing taxonomic information in wcvp",
  options = list(scrollX = TRUE, pageLength = 10),
  rownames = FALSE
)

```


# Add sp4 codes from Wright list

There are `r nrow(usewrighttaxa)` species in the Wright list. And `r nrow(processlist)` in the woody species list. So naturally. We expect some species won't have a sp4 code. 
Now we add the sp4 codes to our species list. This codes are in the Wright list. 

We match the sp4 code by:
(1) on sp6 code
(2) on forestgeo / original name 
(3) on forestgeo/ original name binomial
(4) on accepted name (from WCVP) 

Output the number of species matched on 1 alone, on 2 but not 1, and on 3 but not 2 or 1. 

```{r sp4check, echo=TRUE}

# add sp4 codes from Wright list

wright_no_sp6 <- usewrighttaxa %>% 
  anti_join(processlist, by = "sp6")

wright_no_sp6_noorigname <- wright_no_sp6 %>% 
  anti_join(processlist, by = "orig_name")

wright_no_sp6_nowcvpname <- wright_no_sp6_noorigname %>% 
  anti_join(processlist, by = c("current_name"))


# 2. Trying to match to wright
processlist <- processlist %>%
  mutate(
    m1_idx = match(sp6, usewrighttaxa$sp6),
    m2_idx = ifelse(is.na(m1_idx), match(orig_name, wright_no_sp6$orig_name), NA),
    m3_idx = ifelse(is.na(m1_idx) & is.na(m2_idx), match(current_name, wright_no_sp6_noorigname$current_name), NA),
    m4_idx = ifelse(is.na(m1_idx) & is.na(m2_idx) &  is.na(m3_idx), match(orig_binomial, wright_no_sp6_nowcvpname$orig_binomial), NA),
    
    # Use the specific dataframes for each index
    sp4 = case_when(
      !is.na(m1_idx) ~ usewrighttaxa$sp4[m1_idx],
      !is.na(m2_idx) ~ wright_no_sp6$sp4[m2_idx],
      !is.na(m3_idx) ~ wright_no_sp6_noorigname$sp4[m3_idx], # Fixed reference
      !is.na(m4_idx) ~ wright_no_sp6_nowcvpname$sp4[m3_idx],
      TRUE ~ NA_character_
    )
  )

resumen_matches <- processlist %>%
  summarise(
    M1_sp6 = sum(!is.na(m1_idx)),
    M2_orig_name = sum(!is.na(m2_idx)),
    M3_current_name = sum(!is.na(m3_idx)),
    M4_orig_binomial = sum(!is.na(m4_idx)),
    No_Match = sum(is.na(sp4))
  ) %>%
  tidyr::pivot_longer(everything(), names_to = "Matching Step", values_to = "Count")

# print table
knitr::kable(resumen_matches, 
      caption = "Summary of Species Matching Results (M1, M2, M3)")

processlist <- processlist %>% select(-m1_idx, -m2_idx, -m3_idx, -m4_idx)
```

Identify inconsistencies
```{r fixspecialsp4,  echo=TRUE}

# 1. Identify duplicated sp4 codes
usage_frequency <- processlist %>%
  filter(!is.na(sp4)) %>%
  count(sp4, name = "occurrence_count") %>%
  count(occurrence_count, name = "number_of_sp4_codes")

print(usage_frequency)

# 2. Identify, Join, and Display Duplicates
processlist %>%
  # Filter only for rows where the sp4 is duplicated
  group_by(sp4) %>%
  filter(n() > 1 & !is.na(sp4)) %>%
  ungroup() %>%
  # Join with Wright list 
  left_join(
    usewrighttaxa %>% select(sp4, wcvp_name_wright = current_name, sp6_wright = sp6),
    by = "sp4"
  ) %>%
  # Organize and Select columns for the final table
  select(sp4, sp6, current_name, orig_name, wcvp_matched_name, wcvp_name_wright, sp6_wright) %>%
  arrange(sp4) %>%
  knitr::kable(caption = "Species sharing the same sp4 code (compared with Wright List)") 

# 3. Check for NAs (Species that failed all 3 match steps)
na_count <- sum(is.na(processlist$sp4))
matched_count <- sum(!is.na(processlist$sp4))



```


`r paste(sum(!is.na(processlist$sp4)), "/", nrow(processlist))` woody taxa were successfully assigned a sp4 code.



# Save files

Saving combined panama woody, changes in panama woody, sp4 codes that need revision:

```{r finaloutput}

list_wcvp_changes <- processlist %>% filter(orig_name != wcvp_accepted_name)


write_xlsx(processlist, PANAMA_COMBINED_OUT_PATH)
  
write_xlsx(list_wcvp_changes, PANAMA_COMBINED_CHANGE_SUMMARY_PATH)
  


```

# Alternate names file

```{r alternates}

alternate_synonym_table <- processlist %>%

  mutate(OriginalOrder = row_number()) %>%
  
  select(
    OriginalOrder, 
    sp6, sp4, current_name, orig_name, wcvp_matched_name, 
    SynonymTextALL = orig_synonyms
  ) %>%
  
  # Filter out empty synonym rows
  filter(!is.na(SynonymTextALL) & SynonymTextALL != "") %>%

  mutate(SynonymTextSingle = SynonymTextALL) %>%
  separate_longer_delim(SynonymTextSingle, delim = ",") %>%
  mutate(SynonymTextSingle = trimws(SynonymTextSingle)) %>%
  
  # Order of the synonym in the original text
  group_by(OriginalOrder) %>%
    mutate(SynonymNumber = row_number()) %>%
    ungroup() %>%
  
  # 6. Extract Genus and Species from the single synonym
  # Using too_many="merge" in case there are infraspecific names
  separate_wider_delim(
    SynonymTextSingle, 
    delim = " ", 
    names = c("SynonymGenusSingle", "SynonymSpeciesSingle"),
    too_many = "merge",
    cols_remove = FALSE
  ) %>% arrange(current_name)

# View the result
datatable(alternate_synonym_table, 
             caption = "Mapping: Original Name to Individual Synonym",
          options = list(scrollX = TRUE, pageLength = 10))

# make a file that gives the synonyms as 1-to-1 list
write_xlsx(alternate_synonym_table, FNSYNONYMS)


```

# Wright species missing from woody
```{r sp4missing}
# 4. Find codes in Wright list NOT present in the woody list AND 
# and that DO have #1 in any of the following columns: tree, midstory, understory, shrub, liana.  Include a column for their lifeform according to Wright (based on these columns)



wright_no_woody_match <- usewrighttaxa %>%
  mutate(
    # field if there is no match at all with the woody list
    no_woody_match = !(
      sp6 %in% na.omit(processlist$sp6) |
      sp6 %in% na.omit(processlist$sp6prior) |
      sp4 %in% na.omit(processlist$sp4) |
      current_name %in% processlist$current_name |
      orig_name %in% processlist$current_name |
      orig_binomial %in% processlist$orig_binomial |
      orig_name %in% alternate_synonym_table$SynonymTextSingle |
      current_name %in% alternate_synonym_table$SynonymTextSingle
    ),
    lifeform_wright = paste(
      if_else(!is.na(tree) & tree != 0, "TREE", "_"),
      if_else(!is.na(midstory) & midstory != 0, "MIDSTORY", "_"),
      if_else(!is.na(understory) & understory != 0, "UNDERSTORY", "_"),
      if_else(!is.na(shrub) & shrub != 0, "SHRUB", "_"),
      if_else(!is.na(liana) & liana != 0, "LIANA", "_"), # No longer breaks!
      if_else(!is.na(climber) & climber != 0, "CLIMBER", "_"),
      if_else(!is.na(free) & free != 0, "FREE", "_"),
      if_else(!is.na(vine) & vine != 0, "VINE", "_"),
      if_else(!is.na(herb) & herb != 0, "HERB", "_"),
      if_else(!is.na(epiphyte) & epiphyte != 0, "EPIPHYTE", "_"),
      if_else(!is.na(hemiepiphyte) & hemiepiphyte != 0, "HEMIEPIPHYTE", "_"),
      if_else(!is.na(parasite) & parasite != 0, "PARASITE", "_"),
      sep = ", "
    ),
    
    # Remove the underscores and extra commas
    # This regex looks for "_" followed by a comma or a comma followed by "_"
    lifeform_wright = str_remove_all(lifeform_wright, "_, | ,_|_"),
    
    # 3. Final polish to remove any leading/trailing commas or spaces
    lifeform_wright = str_trim(str_remove_all(lifeform_wright, "^, |, $")),
    
    # 4. If nothing is left, make it a true NA
    lifeform_wright = if_else(lifeform_wright == "", NA_character_, lifeform_wright)
           )

# Find codes in Wright NOT in woody list AND have #1 in woody-relevant columns
wright_sp6_no_woody_list <- wright_no_woody_match %>% 
  filter(no_woody_match, !is.na(sp6))

# Table of Wright woody species lacking 6-letter codes
# Note: I used a regex check on lifeform_wright to see if it contains 
# any of your 5 target woody types.
missing_wright_nosp6_probably_woody <- wright_no_woody_match %>%
  filter(
    no_woody_match,
    is.na(sp6),
    Idlevel == "species",
    str_detect(lifeform_wright, "TREE|MIDSTORY|UNDERSTORY|SHRUB|LIANA")
  )

# Table of Wright unmatched species with 6-letter codes 
datatable(wright_sp6_no_woody_list %>%
            select(sp4, sp6, orig_name, current_name, lifeform_wright),
          caption = "Wright species with sp6 codes that are not in the woody list",
          options = list(scrollX = TRUE, pageLength = 10))

datatable(missing_wright_nosp6_probably_woody %>% select(sp4, sp6, orig_name, current_name,lifeform_wright),
          caption = "Wright species that might be woody and dont have a sp6 code and the name doesnt match anything in the woody list",
          options = list(scrollX = TRUE, pageLength = 10))


write_xlsx(wright_sp6_no_woody_list, WRIGHT_MISSING_PANWOODY_PATH)
write_xlsx(missing_wright_nosp6_probably_woody, WRIGHT_MISSING_PANWOODY_PATH2)
```

# Higher level taxa:


```{r checkhigher, echo=TRUE, results="hide"}

allsp <- processlist # the fixed list

# currwcvp_name is now the correct name, based on the fixes previously done. 
allsp <- allsp %>% mutate(
       genus = current_genus,
       family = current_family,
       genuscode=ifelse(is.na(sp6),NA,substr(sp6,1,4)))

# check that a single genus always has the same family:
genus_check1 <- allsp %>%
  group_by(genus) %>%
  summarise(n_families = n_distinct(family), .groups = "drop") %>%
  filter(n_families > 1)
genus_check1

# check whether a single genus always has the same lifeform:
genus_check2 <- allsp %>%
  group_by(genus) %>%
  summarise(
    habit_class = case_when(
      n_distinct(orig_habit) == 1 & first(orig_habit) == "Climbing"     ~ "Climbing only",
      n_distinct(orig_habit) == 1 & first(orig_habit) == "Freestanding" ~ "Freestanding only",
      n_distinct(orig_habit)  > 1                                  ~ "Both"
    ),
    .groups = "drop"
  ) %>%
  count(habit_class, name = "n_genera")
genus_check2

# to print genus and species with different habits
genera_both <- allsp %>%
  group_by(genus) %>%
  filter(n_distinct(orig_habit) > 1) %>%
  ungroup()

table_2rows <- genera_both %>%
  group_by(genus, orig_habit) %>%
  summarise(
    Species = paste(sort(unique(current_name)), collapse = ", "),
    .groups = "drop"
  ) %>%
  arrange(genus, orig_habit)


# check that a single genus always has the same first 4 letters of species code:
genus_check3 <- allsp %>%
  group_by(genus) %>%
  summarise(n_genuscode = n_distinct(genuscode), 
            genuscode_list = paste(sort(unique(genuscode)), collapse = ", "),
            .groups = "drop") %>%
  filter(n_genuscode > 1)
genus_check3

# find cases where same genuscode corresponds to more than 1 Accepted genus
genus_check4 <- allsp %>%
  group_by(genuscode) %>%
  summarise(
    Accepted_gen4_list = paste(sort(unique(substr(genus,1,4))),collapse=", "),
    Accepted_gen3_list = paste(sort(unique(substr(genus,1,3))),collapse=", "),
    n_Accepted_genus = n_distinct(genus),
    Accepted_genus_list = paste(sort(unique(genus)), collapse = ", "),
    .groups = "drop"
  ) %>%
  filter(n_Accepted_genus > 1)
genus_check4


# just get cases in which there are multiple genus names that have the same first 3 letters as the genus code:  
genus_check5 <- allsp %>%
  filter(!is.na(genuscode)) %>%
  mutate(prefix3 = substr(genuscode, 1, 3),
         genus_prefix3 = tolower(substr(genus, 1, 3))) %>%
  # keep only genus names matching the first 3 letters of genuscode
  filter(genus_prefix3 == prefix3) %>%
  group_by(genuscode) %>%
  summarise(
    n_distinct_genus = n_distinct(genus),
    Accepted_genus_list = paste(sort(unique(genus)), collapse = ", "),
    .groups = "drop"
  ) %>%
  filter(n_distinct_genus > 1) %>%
  arrange(genuscode)
genus_check5

# i changed this, the previous version could count species twice in different genus
genusdfall <- allsp %>%
  group_by(genus) %>%
  summarize(nsptree=sum(orig_habit=="Freestanding"),
            nspliana=sum(orig_habit=="Climbing")) %>%
  ungroup() %>% 
  mutate(Accepted_name=genus,
         Accepted_name_rank="genus")  %>%
  dplyr::select(Accepted_name,Accepted_name_rank,nsptree,nspliana)

#sanity check
sum(genusdfall$nspliana + genusdfall$nsptree)
nrow(allsp)

familydfall <- allsp %>%
  group_by(family) %>%
  summarize(nsptree=sum(orig_habit=="Freestanding"),
            nspliana=sum(orig_habit=="Climbing")) %>%
  ungroup() %>% 
  mutate(Accepted_name=family,
         Accepted_name_rank="family")  %>%
  dplyr::select(Accepted_name,Accepted_name_rank,nsptree,nspliana)

#sanity check
sum(familydfall$nspliana + familydfall$nsptree)
nrow(allsp)

# Merge genus and family in higher taxa
allhighertaxa <- rbind(genusdfall,familydfall) %>%
  mutate(lifeform=paste0(ifelse(nsptree>0,"A",""),ifelse(nspliana>0,"L",""))) %>%
  dplyr::select(Accepted_name,Accepted_name_rank,lifeform)

```
Checking consistency in higher taxa: 

1. Genus names related to more than 1 family: `r nrow(genus_check1)`

```{r genuscheck1}
if (nrow(genus_check1) > 0) { 
  datatable(genus_check1, caption="Genus with more than 1 family",
            options = list(scrollX = TRUE, pageLength = 10))
} 
```

2. Genus with species with different life forms: `r nrow(genus_check2$n_genera[genus_check2$habit_class == "Both"])` 

```{r genuscheck2}
if (nrow(genus_check2) > 0) { 
  datatable(genus_check2, caption="Genus lifeform. Check wether a single genus always has the same life form", options = list(scrollX = TRUE, pageLength = 10))
} 

numbergenus <- genus_check2 %>% filter(habit_class == "Both") %>% select(n_genera)
if(nrow(numbergenus) > 0){
  datatable(table_2rows, caption="Genus with species with two life forms",options = list(scrollX = TRUE, pageLength = 10))
}
```

3. Codes that don't match genus (probably because of name changes): `r nrow(genus_check3)`

```{r genuscheck3}
if (nrow(genus_check3) > 0) { 
  datatable(head(genus_check3, 10), caption="Genus with codes that dont match genus name",options = list(scrollX = TRUE, pageLength = 10))
} 
```

4. Codes that match to more than one genus (probably because of name changes): `r nrow(genus_check4)`

```{r genuscheck4}
if (nrow(genus_check4) > 0) { 
  datatable(head(genus_check4, 10), caption="Genus with codes that match more than one genus",options = list(scrollX = TRUE, pageLength = 10))
} 
```


5. Genus code matches genus but is present in more than one genus (probably because of name changes): `r nrow(genus_check5)`

```{r genuscheck5}
if (nrow(genus_check5) > 0) { 
  datatable(head(genus_check5,10), caption="Genus with the same 3 letters of code",options = list(scrollX = TRUE, pageLength = 10))
} 
```

## Write taxa summary

```{r taxasummary, echo=TRUE, results="hide"}

spdfall <- allsp %>%
  mutate(lifeform=ifelse(orig_habit=="Freestanding","A","L"),
         Accepted_name_rank = "species") %>%
  select(current_name,sp6,sp4,Accepted_name_rank,lifeform) %>%
  rename(Accepted_name=current_name,sp6=sp6)

outtaxa <- bind_rows(spdfall,allhighertaxa) %>%
  rename(namerank=Accepted_name_rank,taxaname=Accepted_name) %>%
  arrange(taxaname)

table(outtaxa$namerank)

if(FIX_SPECIES){
  write_xlsx(outtaxa,FNPANAMATAXA)
}
```

