---
title: "Assemble BCI PlantNet list"
author: "Helene Muller-Landau, Paula Uzcategui"
date: "2026-02-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Using three sources: garwood list (2009), zotz list, and panama biota list. 

We match the names by the wcvp accepted name, if its not in any of the other lists it is added as a new row. The original name of the species in each list is preserved, as well as the wcvp_matched_name. 

OUTPUTS: 

- list of BCI panama plant species

- list of Zotz species that are not in any way in Garwood (synonym, Croat name, original name, matched name)

- list of PanamaBiota species that are not in any way in Garwood (synonym, original name, matched name)

- list of BCI species that might be woody according to the wcvp lifeform (needs revision)

```{r setup2, echo=TRUE,message=FALSE, warning=FALSE, results="hide"}

rm(list=ls())

library("readxl")
library("writexl")
library("dplyr")
library(tidyr)
library("stringi")
library("stringr")
library(DT)
```

# Preparing the datasets

```{r prep, echo=TRUE, results="hide"}

save_plantlists_dir = "splists_out/Plantnet species/"


garwood_list <- read_excel("splists_out/Garwood_2009_Appendix1HM2_WCVP_2026-02-26.xlsx", col_types = "text")

zotz_list <- read_excel("splists_out/83_Zotz_Epiphytes_TableS1_WCVP_2026-02-26.xlsx", col_types = "text")


lianas_list <- read_excel("splists_out/Lianas_Species_complete_list_WCVP_2026-03-03.xlsx", col_types = "text")

pbiota_list <- read_excel("splists_out/PanamaBiotaBCI_WCVP_2026-02-26.xlsx", col_types = "text")

woody_list <- read_excel("splists_out/CurrentPanamaWoody_2026-03-03.xlsx", col_types = "text")
woody_synonyms <- read_excel("splists_out/CurrentPanamaWoody_SYNONYMS_2026-03-03.xlsx", col_types="text")

alternate_names <- read_excel("splists_out/AltNames_2026-03-03.xlsx")

garwood_synonyms <- read_excel("splists_raw/Garwood/Garwood_2009_Appendix1_Synonyms.xlsx") 


garwood_synonyms <- garwood_synonyms %>% mutate(
  sp_name = trimws(paste0(Genus, " ", Species)),
  synonym_sp = trimws(paste0(SynonymGenusSingle, " ", SynonymSpeciesSingle)),
  synonym_binomial = paste0(garwood_synonyms$SynonymGenusSingle, " ", garwood_synonyms$SynonymSpeciesSingle))

garwood_list$wcvp_binomial <- paste0(garwood_list$wcvp_accepted_name_genus, " ", garwood_list$wcvp_accepted_name_species)


```


## BCI high-quality botanist-determined list

### Garwood list
```{r garwood}

#removing BCNM
garwood_list_orig <- garwood_list

# filter out unidentified species
garwood_list <- garwood_list %>% filter(!is.na(word(orig_binomial, 2))) 

garwood_synonyms <- garwood_synonyms %>% filter(!is.na(Species) & 
                                                  Species != "sp." &
                                                  Species != "sp. ined.") 

garwood_toremove <- garwood_list %>% filter(chg == "BCNM")

datatable(garwood_toremove %>% select(orig_name,wcvp_accepted_name ))

garwood_list <- garwood_list %>% filter(chg != "BCNM" | is.na(chg))
```

### Zotz list

Zotz species that are not in garwood
```{r zotznogarwood}

zotz_list$wcvp_binomial = paste0(zotz_list$wcvp_accepted_name_genus, " ", zotz_list$wcvp_accepted_name_species)

# removing species at the genus level
zotz_list <- zotz_list %>% filter(is.na(orig_binomial) | !is.na(word(orig_binomial,2) ))
  
# see species that match by any means to garwood
zotz_garwood_name_match <-  zotz_list %>% filter(
  
  orig_name %in% garwood_list$orig_name | 
  (!is.na(wcvp_matched_name) & wcvp_matched_name %in% garwood_list$wcvp_matched_name) | 
  (!is.na(wcvp_accepted_name) & wcvp_accepted_name %in% garwood_list$wcvp_accepted_name) |
  
  # Croat names are the original names used to build garwood
  (!is.na(`Sp (Croat 1978)`) & (
    `Sp (Croat 1978)` %in% garwood_list$orig_name |
    `Sp (Croat 1978)` %in% garwood_list$wcvp_matched_name |
    `Sp (Croat 1978)` %in% garwood_list$wcvp_accepted_name
  )) |
   
  #match binomials 
  (!is.na(orig_binomial) & (
     orig_binomial %in% garwood_list$orig_binomial
  )) |
  
  (!is.na(wcvp_binomial) & (
     wcvp_binomial %in% garwood_list$wcvp_binomial)
  ))

    
zotz_garwood_synonym_match <- zotz_list %>% 
  filter(
    !orig_name %in% zotz_garwood_name_match$orig_name &
    (
      orig_name %in% garwood_synonyms$synonym_sp |
      (!is.na(wcvp_matched_name) & wcvp_matched_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(wcvp_accepted_name) & wcvp_accepted_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(`Sp (Croat 1978)`) & `Sp (Croat 1978)` %in% garwood_synonyms$synonym_sp) |
      (!is.na(orig_binomial) & orig_binomial %in% garwood_synonyms$synonym_binomial) |
      (!is.na(wcvp_binomial) & wcvp_binomial %in% garwood_synonyms$synonym_binomial)
    )
  )

zotz_garwood_synonym_match$matched_garwood_synonym = TRUE

# see species that matched by a synonym 
zotz_garwood_synonym_match_print <- zotz_garwood_synonym_match %>%  left_join(
    garwood_synonyms %>% select(sp_orig = sp_name, synonym_sp, syn_orig = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_match = sp_name, synonym_sp, syn_match = SynonymTextALL) %>% distinct(),
    by = c("wcvp_matched_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_wcvp = sp_name, synonym_sp, syn_wcvp = SynonymTextALL) %>% distinct(),
    by = c("wcvp_accepted_name" = "synonym_sp")) %>%
  left_join(
    garwood_synonyms %>% select(sp_croat = sp_name, synonym_sp, syn_croat = SynonymTextALL) %>% distinct(),
    by = c("Sp (Croat 1978)" = "synonym_sp")) %>%
  mutate(
    SynonymMatchAll = coalesce(syn_orig, syn_match, syn_wcvp, syn_croat),
    orig_garwood = coalesce(sp_orig, sp_match, sp_wcvp, sp_croat),
  ) %>%
  select(-syn_orig, -syn_match, -syn_wcvp, -syn_croat, -sp_orig, -sp_match, -sp_wcvp, -sp_croat) %>%
  distinct(orig_name, orig_garwood, .keep_all = TRUE)
  

datatable(zotz_garwood_synonym_match_print %>% 
            select(orig_name, wcvp_matched_name, wcvp_accepted_name, `Sp (Croat 1978)`, SynonymMatchAll, orig_garwood), 
          caption="Zotz species that matched garwood by synonym",
          options = list(scrollX = TRUE, pageLength = 10))
  
  
zotz_garwood_match <- bind_rows(
  zotz_garwood_name_match,
  zotz_garwood_synonym_match
) %>%
  distinct(orig_name, .keep_all = TRUE)

zotz_no_garwood_at_all <- zotz_list %>%
  filter(!orig_name %in% zotz_garwood_match$orig_name)

datatable(zotz_no_garwood_at_all  %>% select(orig_name, wcvp_matched_name, wcvp_accepted_name, `Sp (Croat 1978)`), 
          caption="Zotz species that dont match to anything in garwood",
          options = list(scrollX = TRUE, pageLength = 10))

write_xlsx(zotz_no_garwood_at_all, "splists_out/zotz_species_not_in_garwood.xlsx")

```

Add to bci list
```{r}
# using garwood as a base
garwood_list$source_list = "Garwood2009"
zotz_no_garwood_at_all$source_list = "Zotz"

listscol <- c("orig_name",
              "orig_binomial",
              "wcvp_binomial",
              "wcvp_matched_name",
              "wcvp_matched_plant_name_id",
              "wcvp_matched_ipni_id",
              "wcvp_matched_status",
              "wcvp_accepted_name",
              "wcvp_accepted_family", 
              "wcvp_accepted_author",
              "wcvp_accepted_plant_name_id", 
              "wcvp_accepted_powo_id",
              "wcvp_accepted_ipni_id",
              "wcvp_lifeform_description",
              "source_list")

bcilist <- bind_rows(
  garwood_list %>% 
    select(all_of(c(listscol, "SYNONYMS"))),
  zotz_no_garwood_at_all %>% select(all_of(listscol))
)

```


### Species in BCI plots census

```{r plotslist}
# Preparing plotlist

# 1. Read data
bci_plots <- read_excel("splists_raw/Plots/abundance_plots_HM.xlsx")

bci_plots <- bci_plots %>% select(sp, binomial, bci, `bci 10 ha plot`, P11, P14)

bci_other_plots <- read.delim(
  "splists_raw/Plots/doi_10_5061_dryad_1g1jwsvc3__v20260226/mainfile.txt"
)

# 2. Summarise other plots safely
bci_other_plots <- bci_other_plots %>%
  group_by(sp6) %>%
  summarise(
    bigplot = n_distinct(tag),
    .groups = "drop"
  )

# 3. Clean and rename BEFORE joining
bci_plots <- bci_plots %>%
  rename(
    sp6 = sp,
    sp_name_plot = binomial,
    bci50ha = bci,
    bci10ha = `bci 10 ha plot`
  )

# 4. Join safely (keeps all bci_plots species)
bci_plots <- bci_plots %>%
  left_join(bci_other_plots, by = "sp6") %>%
  mutate(
    bigplot = coalesce(bigplot, 0)   # replace NA with 0
  )

# 5. Filter safely (handle NA just in case)
bci_plots <- bci_plots %>%
  mutate(across(c(bci50ha, bci10ha, P14, P11), ~coalesce(., 0))) %>%
  filter(
    bci50ha > 0 |
    bci10ha > 0 |
    P14 > 0 |
    P11 > 0 |
    bigplot > 0
  )

# 6. Create plot_name column safely
bci_plots <- bci_plots %>%
  rowwise() %>%
  mutate(
    plot_name = paste(
      c(
        if (bci50ha > 0) "bci50ha",
        if (bigplot > 0) "bcimultiplots",
        if (bci10ha > 0) "bci10ha",
        if (P14 > 0) "P14",
        if (P11 > 0) "P11"
      ),
      collapse = ", "
    )
  ) %>%
  ungroup()

# the plots dont have wcvp accepted name, merge them by woody list
match21 <- match(bci_plots$sp6, woody_list$sp6)            
nomatchview <- bci_plots[is.na(match21),] %>% select(sp6, sp_name_plot) # All that dont match are unidentified

knitr::kable(nomatchview, caption="Species in plot that dont match with the woody bci list")

bci_plots <- bci_plots %>% left_join(
  woody_list, by="sp6"
)

# if no orig name add the binomial
bci_plots <- bci_plots %>% mutate(orig_name = ifelse(is.na(orig_name), sp_name_plot, orig_name))

knitr::kable(bci_plots %>% filter(duplicated(sp6)) %>% select(sp6, sp_name_plot))

# create binomial first if not already
bci_plots$wcvp_binomial <- paste0(
  bci_plots$wcvp_accepted_name_genus, " ",
  bci_plots$wcvp_accepted_name_species
)




```

#### Preparing alternate names

```{r altnames}

# Alternate names combines species names in the synonyms columns of currentwoody, viewtaxa and garwood. 

# These are not necesarily synonyms but can also be names the species used to be given until it was revised again

# Here we create a list of the synonyms, and if they are somewhere in the woody list, we add the name of origin y woody and the synonym they have in garwood.
alternate_names_garwood <- alternate_names %>% 
  # DIRECT MATCH — prefname
  left_join(
    garwood_list %>% select(garwood_direct_pref_=orig_name, garwood_direct_pref=orig_name),
    by = c("prefname" = "garwood_direct_pref_")
  ) %>%
  
  left_join(
    garwood_list %>% select(garwood_direct_prefbi_=orig_binomial, garwood_direct_prefbi=orig_name),
    by = c("prefname" = "garwood_direct_prefbi_"), relationship = "many-to-many"
  ) %>%
  
  # DIRECT MATCH — altname
  left_join(
    garwood_list %>% select(garwood_direct_alt_=orig_name, garwood_direct_alt=orig_name),
    by = c("altname" = "garwood_direct_alt_")
  ) %>%
  left_join(
    garwood_list %>% select(garwood_direct_altbi_=orig_binomial, garwood_direct_altbi=orig_name),
    by = c("altname" = "garwood_direct_altbi_"), relationship = "many-to-many"
  ) %>%
  
  # SYNONYM MATCH — prefname
  left_join(
    garwood_synonyms %>%
      select(
        SynonymTextSingle,
        garwood_syn_orig_pref = sp_name,
        synonym_full_pref = SynonymTextALL,
      ),
    by = c("prefname" = "SynonymTextSingle")
  ) 
  
alternate_names_garwood <- alternate_names_garwood %>%
mutate(
    
    # DIRECT MATCH (either pref or alt)
    garwood_direct_match = coalesce(
      garwood_direct_pref,
      garwood_direct_prefbi,
      garwood_direct_alt,
      garwood_direct_altbi
      
    ))  %>%   
  left_join(
    garwood_list %>% select(orig_name, garwood_synonymall = SYNONYMS),
    by = c("garwood_direct_match" = "orig_name")
  ) 


alternate_names_garwood <- alternate_names_garwood %>% filter(!is.na(garwood_direct_match))

datatable(alternate_names_garwood %>% select(
  prefname, altname, garwood_direct_match, garwood_synonymall
),
caption="Species that changed their identification. Pref name is their prefeerred name, altname is the name it used to have. Note that these are not all synonyms in the botanical sense."
)

```

#### Matching  plot species to bcilist so far

```{r bciplotmatch}
# 1. direct name/binomial match
bci_plots_garwood_name_match <- bci_plots %>%
  filter(
    orig_name %in% bcilist$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist$wcvp_accepted_name) |

    # match binomials
    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist$orig_binomial 
    )) |

    (!is.na(wcvp_binomial) & (
        wcvp_binomial %in% bcilist$wcvp_binomial 
    ))
  )


# Everything that did not match directly, match with synonyms

bci_plots_garwood_synonym_match1 <- bci_plots %>% 
  filter(
    !orig_name %in% bci_plots_garwood_name_match$orig_name &
    (
      orig_name %in% garwood_synonyms$synonym_sp |
      (!is.na(wcvp_matched_name) & wcvp_matched_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(wcvp_accepted_name) & wcvp_accepted_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(orig_binomial) & orig_binomial %in% garwood_synonyms$synonym_binomial) |
      (!is.na(wcvp_binomial) & wcvp_binomial %in% garwood_synonyms$synonym_binomial)
    )
  )

# add  the garwood synonym it matched to
bci_plots_garwood_synonym_print1 <- bci_plots_garwood_synonym_match1 %>%  left_join(
    garwood_synonyms %>% select(sp_orig = sp_name, synonym_sp, syn_orig = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_binomial = sp_name, synonym_sp, syn_binomial = SynonymTextALL) %>% distinct(),
    by = c("orig_binomial" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_match = sp_name, synonym_sp, syn_match = SynonymTextALL) %>% distinct(),
    by = c("wcvp_matched_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_wcvp = sp_name, synonym_sp, syn_wcvp = SynonymTextALL) %>% distinct(),
    by = c("wcvp_accepted_name" = "synonym_sp")) %>%
  mutate(
    SynonymMatchAll = coalesce(syn_orig, syn_match, syn_wcvp, syn_binomial),
    orig_garwood = coalesce(sp_orig, sp_match, sp_wcvp, sp_binomial),
  ) %>%
  select(-syn_orig, -syn_match, -syn_wcvp, -syn_binomial, -sp_orig, -sp_match, -sp_wcvp, -sp_binomial) %>%
  distinct(orig_name, orig_garwood, .keep_all = TRUE)


# everything that did not match to garwood synonyms, match to alternate names
bci_plots_garwood_synonym_match2 <- bci_plots %>% 
  filter(
    !orig_name %in% bci_plots_garwood_synonym_match1$orig_name &
      !orig_name %in% bci_plots_garwood_name_match$orig_name &
    (
      orig_name %in% alternate_names_garwood$prefname |
      orig_binomial %in% alternate_names_garwood$prefname |
      
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% alternate_names_garwood$prefname) |
      
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% alternate_names_garwood$prefname) |
        
      orig_name %in% alternate_names_garwood$altname |
      orig_binomial %in% alternate_names_garwood$altname |
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% alternate_names_garwood$altname) |
      
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% alternate_names_garwood$altname)
    )
  )

datatable(
  bci_plots_garwood_synonym_print1 %>% 
  select(orig_name, wcvp_matched_name, wcvp_accepted_name, orig_garwood ,SynonymMatchAll), 
          caption="BCI plot species that matched a garwood synonym",
          options = list(scrollX = TRUE, pageLength = 10))

```


```{r matchsyn2}
## add the altname it matched to and why that name is in garwood
bci_plots_garwood_synonym_print2 <- bci_plots_garwood_synonym_match2 %>%
  # A synonym or alternate name of the species is in the garwood list. 
  left_join(
    alternate_names_garwood %>% 
      select(prefname, altname, garwood_direct_match, garwood_synonymall),
    by = c("orig_name" = "prefname")
  ) %>%
  left_join(
    alternate_names_garwood %>% 
      select(prefname, altname, garwood_direct_match, garwood_synonymall),
    by = c("orig_binomial" = "prefname"),
    suffix = c("", "_bi")
  ) %>%
  left_join(
    alternate_names_garwood %>% 
      select(prefname, altname, garwood_direct_match, garwood_synonymall),
    by = c("wcvp_matched_name" = "prefname"),
    suffix = c("", "_match")
  ) %>%  
  left_join(
    alternate_names_garwood %>% 
      select(prefname, altname, garwood_direct_match, garwood_synonymall),
    by = c("wcvp_accepted_name" = "prefname"),
    suffix = c("", "_acc")
  ) %>%  
  
  # now the same but for altname
  left_join(
    alternate_names_garwood %>% 
      select(prefname, altname, altname, garwood_direct_match, garwood_synonymall),
    by = c("orig_name" = "altname"),
    suffix = c("", "_altorig")
  ) %>%
  left_join(
    alternate_names_garwood %>% 
      select(prefname, altname, garwood_direct_match, garwood_synonymall),
    by = c("orig_binomial" = "altname"),
    suffix = c("", "_altbi")
  ) %>%
  left_join(
    alternate_names_garwood %>% 
      select(prefname, altname, garwood_direct_match, garwood_synonymall),
    by = c("wcvp_matched_name" = "altname"),
    suffix = c("", "_altmatch")
  ) %>%  
  left_join(
    alternate_names_garwood %>% 
      select(prefname, altname, garwood_direct_match, garwood_synonymall),
    by = c("wcvp_accepted_name" = "altname"),
    suffix = c("", "_altacc")
  ) 

bci_plots_garwood_synonym_print2 <- bci_plots_garwood_synonym_print2 %>%
  mutate(
    # the garwood name it matched to or of the synonym it matched to
    orig_garwood = coalesce(
      garwood_direct_match,
      garwood_direct_match_bi,
      garwood_direct_match_match,
      garwood_direct_match_acc,
      garwood_direct_match_altorig,
      garwood_direct_match_altbi,
      garwood_direct_match_altmatch,
      garwood_direct_match_altacc
    ),
    
    alternate_name =  altname,
    prefered_name = prefname
    
  )%>%
  distinct(orig_name, orig_garwood, .keep_all = TRUE)



datatable(
  bci_plots_garwood_synonym_print2 %>% 
  select(orig_name, wcvp_matched_name, wcvp_accepted_name, orig_garwood,
         alternate_name, prefered_name, garwood_synonymall), 
          caption="BCI plot species where an alternate name of the species matched to garwood",
          options = list(scrollX = TRUE, pageLength = 10))


```

In this cases, the alternate name in forestgeo is more reliable than the name the garwood list. 

Therefore, we change the names in garwood to the names in forestgeo. 
```{r changegarwoodname}

# replace values in the bcilist to the values in the 
bci_plots_garwood_synonym_print2$source_list <- "GarwoodForestGeo"

corrections <- bci_plots_garwood_synonym_print2 %>%
    select(orig_garwood, all_of(listscol))

bcilist <- bcilist  %>%
  mutate(orig_garwood = orig_name)

# Update bcilist based on the matching 'orig_name'
bcilist <- bcilist %>%
  rows_update(corrections, by = "orig_garwood")



bci_plots_garwood_synonym_match <- bind_rows(bci_plots_garwood_synonym_match1, bci_plots_garwood_synonym_match2)
bci_plots_garwood_synonym_match <- bci_plots_garwood_synonym_match %>% mutate(matched_garwood_synonym = TRUE)

# combine
bci_plots_garwood_match <- bind_rows(
  bci_plots_garwood_name_match,
  bci_plots_garwood_synonym_match
) %>%
  distinct(sp6, .keep_all = TRUE)

```



```{r allnomatchbciplot}
# 4. unmatched
bci_plots_no_garwood_at_all <- bci_plots %>%
  filter(!sp6 %in% bci_plots_garwood_match$sp6)

datatable(bci_plots_no_garwood_at_all  %>% select(sp6, orig_name, wcvp_matched_name, wcvp_accepted_name, plot_name), 
          caption="Species recorded in BCI plots census that dont match to anything in garwood",
          options = list(scrollX = TRUE, pageLength = 10))



datatable(garwood_list_orig %>% 
  filter(orig_name %in% bci_plots_no_garwood_at_all$orig_name) %>%
  select(orig_name, chg), 
          caption="Some of these species we removed earlier...",
          options = list(scrollX = TRUE, pageLength = 10))


write_xlsx(bci_plots_no_garwood_at_all, "splists_out/Plantnet species/plots_bci_not_garwood.xlsx")

```

According to the garwood notes Damburneya purpurea is Nectandra purpurea. But maybe garwood was not sure?

Add to bci list
```{r}

bci_plots_no_garwood_at_all$source_list = "PlotCensusBCI"


bcilist <- bind_rows(
  bcilist,
  bci_plots_no_garwood_at_all %>% select(all_of(c(listscol, "plot_name", "bci50ha", "bci10ha", "P11", "P14"))
))

```


### Lianas BCI list
```{r lianasnobcilist}

lianas_list <- lianas_list %>% filter(isbci=="TRUE")
lianas_list <- lianas_list %>% distinct(orig_name, .keep_all = TRUE)
lianas_list$wcvp_binomial <- paste0(
  lianas_list$wcvp_accepted_name_genus, " ",
  lianas_list$wcvp_accepted_name_species
)

# 1. direct name/binomial match
lianas_garwood_name_match <- lianas_list %>%
  filter(

    orig_name %in% bcilist$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist$wcvp_accepted_name) |

    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist$orig_binomial 
    )) |

    (!is.na(wcvp_binomial) & (
        wcvp_binomial %in% bcilist$wcvp_binomial 
    ))
  )

# 2. synonym_sp for remaining
lianas_garwood_synonym_match <- lianas_list %>%
  filter(
    !orig_name %in% lianas_garwood_name_match$orig_name &
    (
      orig_name %in% garwood_synonyms$synonym_sp |
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(orig_binomial) & orig_binomial %in% garwood_synonyms$synonym_binomial) |
      (!is.na(wcvp_binomial) & wcvp_binomial %in% garwood_synonyms$synonym_binomial)
    )
  )

lianas_garwood_synonym_match$matched_garwood_synonym <- TRUE


# see species that matched by a synonym 
lianas_garwood_synonym_match_print <- lianas_garwood_synonym_match %>%  left_join(
    garwood_synonyms %>% select(sp_orig = sp_name, synonym_sp, syn_orig = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_match = sp_name, synonym_sp, syn_match = SynonymTextALL) %>% distinct(),
    by = c("wcvp_matched_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_wcvp = sp_name, synonym_sp, syn_wcvp = SynonymTextALL) %>% distinct(),
    by = c("wcvp_accepted_name" = "synonym_sp")) %>%
  mutate(
    SynonymMatchAll = coalesce(syn_orig, syn_match, syn_wcvp),
    orig_garwood = coalesce(sp_orig, sp_match, sp_wcvp),
  ) %>%
  select(-syn_orig, -syn_match, -syn_wcvp,  -sp_orig, -sp_match, -sp_wcvp) %>%
  distinct(orig_name, .keep_all = TRUE)
  

datatable(lianas_garwood_synonym_match_print %>% 
            select(orig_name, wcvp_matched_name, wcvp_accepted_name, SynonymMatchAll, orig_garwood), 
          caption="Liana species that matched garwood by synonym",
          options = list(scrollX = TRUE, pageLength = 10))


lianas_garwood_match <- bind_rows(
  lianas_garwood_name_match,
  lianas_garwood_synonym_match
) %>%
  distinct(orig_name, .keep_all = TRUE)

lianas_no_garwood_at_all <- lianas_list %>%
  filter(!orig_name %in% lianas_garwood_match$orig_name)



# 5. Output and Display
datatable(lianas_no_garwood_at_all %>%
  select(orig_name, wcvp_matched_name, wcvp_accepted_name), 
          caption = "Lianas species that don't match to anything in Garwood",
          options = list(scrollX = TRUE, pageLength = 10))

write_xlsx(lianas_no_garwood_at_all, "splists_out/lianas_species_not_in_garwood.xlsx")

```

Add to bci list
```{r lianasbciadd}

lianas_no_garwood_at_all$source_list = "lianas"

bcilist <- bind_rows(
  bcilist,
  lianas_no_garwood_at_all %>% select(all_of(listscol))
)

```

### Check consistency 
```{r savebotanists}

# Check consistency
# removing rows that are not identified at species level
datatable(bcilist %>% filter(is.na(word(orig_binomial, 2))) %>% select(orig_name, wcvp_accepted_name), caption="Species to be removed from final list because they are not fully identified" )

bcilist <- bcilist %>% filter(!is.na(word(orig_binomial, 2))) 


datatable(bcilist %>% group_by(orig_name) %>% filter(n()> 1) %>%
          select(orig_name, wcvp_accepted_name) %>% arrange(orig_name),
          caption="Species duplicated original name",
  options = list(scrollX = TRUE, pageLength = 10)
          )

datatable(bcilist %>% group_by(wcvp_accepted_name) %>% 
            filter(n()> 1) %>% select(orig_name, wcvp_accepted_name)%>% 
            arrange(wcvp_accepted_name), 
          caption="Species duplicated wcvp_accepted name",
  options = list(scrollX = TRUE, pageLength = 10)
          )



datatable(
  bcilist %>%
    select(wcvp_accepted_name, orig_name, source_list) %>% 
    arrange(source_list),
    caption="Species list in BCI and their source",
  options = list(scrollX = TRUE, pageLength = 10))

```


### Add woody species column
 
If the species is in the woody list

```{r bcilistwoody, echo=TRUE}
# set woody if:
# original, matched or accepted name is anywhere in woody list

bcilist <- bcilist %>% mutate(
  woody_name =  wcvp_accepted_name %in% woody_list$current_name | 
                wcvp_accepted_name %in% woody_list$orig_name | 
                wcvp_binomial %in% woody_list$orig_binomial |
                wcvp_binomial %in% woody_list$current_name |
                wcvp_matched_name %in% woody_list$current_name |
                wcvp_matched_name %in% woody_list$orig_name |
                orig_name %in% woody_list$orig_name |
                orig_name %in% woody_list$current_name |
                orig_binomial %in% woody_list$orig_binomial,
  
  woody_synonym =  wcvp_accepted_name %in% woody_synonyms$SynonymTextSingle |
                   orig_name %in% woody_synonyms$SynonymTextSingle |
                   wcvp_matched_name %in% woody_synonyms$SynonymTextSingle,
  woody = woody_name | woody_synonym
) 

# BCI BOTANISTS HIGH QUALITY LIST
write_xlsx(bcilist, paste0(save_plantlists_dir, "BCI_SPECIES_BOTANISTS_LIST_",Sys.Date(),".xlsx"))


bcilist_woody <- bcilist %>%
  filter(woody) %>%
  select(orig_name, wcvp_matched_name, wcvp_accepted_name, wcvp_accepted_family, wcvp_lifeform_description)

datatable(bcilist_woody, caption="BCI species that are in the woody list")

```

### Possibly woody species

```{r posiblywoody}

lifeform_woody = c(
  "tree",
  "shrub",
  "liana",
  "climbing shrub",
  "shrub or tree",
  "climbing shrub or tree",
  "scrambling shrub or tree",
  "subshrub or shrub",
  "subshrub, shrub or tree",
  "scrambling shrub",
  "perennial or shrub",
  "scrambling subshrub or shrub",
  "epiphytic shrub",
  "scrambling tree",
  "climbing shrub or liana",
  "scrambling shrub or liana",
  "tuberous shrub or tree",
  "herbaceous tree",
  "climbing herbaceous tree",
  "climbing herbaceous tree or liana",
  "climbing perennial or herbaceous tree",
  "subshrub or climbing shrub",
  "tuberous subshrub",
  "hemiparasitic shrub",
  "hemiepiphytic shrub or tree",
  "annual or shrub",
  "annual, subshrub or shrub"
)

bcilist_possibly_woody <- bcilist %>%
  filter(!woody &
          wcvp_lifeform_description %in% lifeform_woody ) %>%
  select(orig_name, wcvp_matched_name, wcvp_accepted_name, wcvp_accepted_family, wcvp_lifeform_description)

datatable(bcilist_possibly_woody, caption="BCI species that might be woody according to the wcvp lifeform")

write_xlsx(bcilist_possibly_woody, paste0(save_plantlists_dir, "BCI_SPECIES_LIST_POSSIBLY_WOODY_",Sys.Date(),".xlsx"))


```


## Panamabiota species that are not in garwood

```{r pbiotanogarwood}

pbiota_list$wcvp_binomial <- paste0(
  pbiota_list$wcvp_accepted_name_genus, " ",
  pbiota_list$wcvp_accepted_name_species
)

# 1. direct name/binomial match
pbiota_garwood_name_match <- pbiota_list %>%
  filter(

    orig_name %in% bcilist$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist$wcvp_accepted_name) |

    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist$orig_binomial 
    )) |

    (!is.na(wcvp_binomial) & (
        wcvp_binomial %in% bcilist$wcvp_binomial 
    ))
  )

# 2. synonym_sp for remaining
pbiota_garwood_synonym_match <- pbiota_list %>%
  filter(
    !orig_name %in% pbiota_garwood_name_match$orig_name &
    (
      orig_name %in% garwood_synonyms$synonym_sp |
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(orig_binomial) & orig_binomial %in% garwood_synonyms$synonym_binomial) |
      (!is.na(wcvp_binomial) & wcvp_binomial %in% garwood_synonyms$synonym_binomial)
    )
  )

pbiota_garwood_synonym_match$matched_garwood_synonym <- TRUE


# see species that matched by a synonym 
pbiota_garwood_synonym_match_print <- pbiota_garwood_synonym_match %>%  left_join(
    garwood_synonyms %>% select(sp_orig = sp_name, synonym_sp, syn_orig = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_match = sp_name, synonym_sp, syn_match = SynonymTextALL) %>% distinct(),
    by = c("wcvp_matched_name" = "synonym_sp"), relationship = "many-to-many"
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_wcvp = sp_name, synonym_sp, syn_wcvp = SynonymTextALL) %>% distinct(),
    by = c("wcvp_accepted_name" = "synonym_sp"),  relationship = "many-to-many") %>%
  mutate(
    SynonymMatchAll = coalesce(syn_orig, syn_match, syn_wcvp),
    orig_garwood = coalesce(sp_orig, sp_match, sp_wcvp),
  ) %>%
  select(-syn_orig, -syn_match, -syn_wcvp,  -sp_orig, -sp_match, -sp_wcvp) %>%
  distinct(orig_name, .keep_all = TRUE)
  

datatable(pbiota_garwood_synonym_match_print %>% 
            select(orig_name, wcvp_matched_name, wcvp_accepted_name, SynonymMatchAll, orig_garwood), 
          caption="Panama biota species that matched garwood by synonym",
          options = list(scrollX = TRUE, pageLength = 10))

pbiota_garwood_match <- bind_rows(
  pbiota_garwood_name_match,
  pbiota_garwood_synonym_match
) %>%
  distinct(orig_name, .keep_all = TRUE)

pbiota_no_garwood_at_all <- pbiota_list %>%
  filter(!orig_name %in% pbiota_garwood_match$orig_name)



datatable(pbiota_no_garwood_at_all, 
          caption="Panama biota species that dont match to anything in garwood",
          options = list(scrollX = TRUE, pageLength = 10))

write_xlsx(pbiota_no_garwood_at_all, "splists_out/panamabiota_species_not_in_garwood.xlsx")

```


