---
title: "species_plantnet"
author: "Helene Muller-Landau, Paula Uzcategui"
date: "2026-02-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Create a combined BCI species list

Using three sources: garwood list (2009), zotz list, and panama biota list. 

We match the names by the wcvp accepted name, if its not in any of the other lists it is added as a new row. The original name of the species in each list is preserved, as well as the wcvp_matched_name. 

OUTPUTS: 

- list of BCI panama plant species

- list of Zotz species that are not in any way in Garwood (synonym, Croat name, original name, matched name)

- list of PanamaBiota species that are not in any way in Garwood (synonym, original name, matched name)

- list of BCI species that might be woody according to the wcvp lifeform (needs revision)

```{r setup2, echo=TRUE,message=FALSE, warning=FALSE, results="hide"}

rm(list=ls())

library("readxl")
library("writexl")
library("dplyr")
library(tidyr)
library("stringi")
library("stringr")
library(DT)
```

# Preparing the datasets

```{r prep, echo=TRUE, results="hide"}

save_plantlists_dir = "splists_out/Plantnet species/"


garwood_list <- read_excel("splists_out/Garwood_2009_Appendix1HM2_WCVP_2026-02-26.xlsx", col_types = "text")

zotz_list <- read_excel("splists_out/83_Zotz_Epiphytes_TableS1_WCVP_2026-02-26.xlsx", col_types = "text")

woody_list <- read_excel("splists_out/CurrentPanamaWoody_2026-03-03.xlsx",
                         col_types = "text")



lianas_list <- read_excel("splists_out/Lianas_Species_complete_list_WCVP_2026-03-03.xlsx", col_types = "text")

pbiota_list <- read_excel("splists_out/PanamaBiotaBCI_WCVP_2026-02-26.xlsx", col_types = "text")



woody_synonyms <- read_excel("splists_out/CurrentPanamaWoody_SYNONYMS_2026-03-03.xlsx")
garwood_synonyms <- read_excel("splists_raw/Garwood/Garwood_2009_Appendix1_Synonyms.xlsx") 


garwood_synonyms <- garwood_synonyms %>% mutate(
  sp_name = trimws(paste0(Genus, " ", Species)),
  synonym_sp = trimws(paste0(SynonymGenusSingle, " ", SynonymSpeciesSingle)))

garwood_list$wcvp_binomial <- paste0(garwood_list$wcvp_accepted_name_genus, " ", garwood_list$wcvp_accepted_name_species)

garwood_synonyms$synonym_binomial <- paste0(garwood_synonyms$SynonymGenusSingle, " ", garwood_synonyms$SynonymSpeciesSingle)

```

## BCI high-quality botanist-determined list

### Garwood list
```{r garwood}

#removing BCNM

garwood_toremove <- garwood_list %>% filter(chg == "BCNM")

datatable(garwood_toremove %>% select(orig_name,wcvp_accepted_name ))

garwood_list <- garwood_list %>% filter((!is.na(chg) & chg != "BCNM") | is.na(chg))
```

### Zotz list

Zotz species that are not in garwood
```{r zotznogarwood}

zotz_list$wcvp_binomial = paste0(zotz_list$wcvp_accepted_name_genus, " ", zotz_list$wcvp_accepted_name_species)

# see species that match by any means to garwood
zotz_garwood_name_match <-  zotz_list %>% filter(
  
  orig_name %in% garwood_list$orig_name | 
  (!is.na(wcvp_matched_name) & wcvp_matched_name %in% garwood_list$wcvp_matched_name) | 
  (!is.na(wcvp_accepted_name) & wcvp_accepted_name %in% garwood_list$wcvp_accepted_name) |
  
  # Croat names are the original names used to build garwood
  (!is.na(`Sp (Croat 1978)`) & (
    `Sp (Croat 1978)` %in% garwood_list$orig_name |
    `Sp (Croat 1978)` %in% garwood_list$wcvp_matched_name |
    `Sp (Croat 1978)` %in% garwood_list$wcvp_accepted_name
  )) |
   
  #match binomials 
  (!is.na(orig_binomial) & (
     orig_binomial %in% garwood_list$orig_binomial |
     orig_binomial %in% garwood_synonyms$synonym_binomial
  )) |
  
  (!is.na(wcvp_binomial) & (
     wcvp_binomial %in% garwood_list$wcvp_binomial |
     wcvp_binomial %in% garwood_synonyms$synonym_binomial
  ))
  )

    
zotz_garwood_synonym_match <- zotz_list %>% 
  filter(
    !orig_name %in% zotz_garwood_name_match$orig_name &
    (
      orig_name %in% garwood_synonyms$synonym_sp |
      (!is.na(wcvp_matched_name) & wcvp_matched_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(wcvp_accepted_name) & wcvp_accepted_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(`Sp (Croat 1978)`) & `Sp (Croat 1978)` %in% garwood_synonyms$synonym_sp)
    )
  )

zotz_garwood_synonym_match$matched_garwood_synonym = TRUE

# see species that matched by a synonym 
zotz_garwood_synonym_match_print <- zotz_garwood_synonym_match %>%  left_join(
    garwood_synonyms %>% select(sp_orig = sp_name, synonym_sp, syn_orig = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_match = sp_name, synonym_sp, syn_match = SynonymTextALL) %>% distinct(),
    by = c("wcvp_matched_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_wcvp = sp_name, synonym_sp, syn_wcvp = SynonymTextALL) %>% distinct(),
    by = c("wcvp_accepted_name" = "synonym_sp")) %>%
  left_join(
    garwood_synonyms %>% select(sp_croat = sp_name, synonym_sp, syn_croat = SynonymTextALL) %>% distinct(),
    by = c("Sp (Croat 1978)" = "synonym_sp")) %>%
  mutate(
    SynonymMatchAll = coalesce(syn_orig, syn_match, syn_wcvp, syn_croat),
    orig_garwood = coalesce(sp_orig, sp_match, sp_wcvp, sp_croat),
  ) %>%
  select(-syn_orig, -syn_match, -syn_wcvp, -syn_croat, -sp_orig, -sp_match, -sp_wcvp, -sp_croat) %>%
  distinct(orig_name, .keep_all = TRUE)
  

datatable(zotz_garwood_synonym_match_print %>% 
            select(orig_name, wcvp_matched_name, wcvp_accepted_name, `Sp (Croat 1978)`, SynonymMatchAll, orig_garwood), 
          caption="Zotz species that matched garwood by synonym",
          options = list(scrollX = TRUE, pageLength = 10))
  
  
zotz_garwood_match <- bind_rows(
  zotz_garwood_name_match,
  zotz_garwood_synonym_match
) %>%
  distinct(orig_name, .keep_all = TRUE)

zotz_no_garwood_at_all <- zotz_list %>%
  filter(!orig_name %in% zotz_garwood_match$orig_name)

datatable(zotz_no_garwood_at_all  %>% select(orig_name, wcvp_matched_name, wcvp_accepted_name, `Sp (Croat 1978)`), 
          caption="Zotz species that dont match to anything in garwood",
          options = list(scrollX = TRUE, pageLength = 10))

write_xlsx(zotz_no_garwood_at_all, "splists_out/zotz_species_not_in_garwood.xlsx")

```

Add to bci list
```{r}
# using garwood as a base
garwood_list$source_list = "Garwood2009"
zotz_no_garwood_at_all$source_list = "Zotz"

listscol <- c("orig_name",
              "orig_binomial",
              "wcvp_binomial",
              "wcvp_matched_name",
              "wcvp_matched_plant_name_id",
              "wcvp_matched_ipni_id",
              "wcvp_matched_status",
              "wcvp_accepted_name",
              "wcvp_accepted_family", 
              "wcvp_accepted_author",
              "wcvp_accepted_plant_name_id", 
              "wcvp_accepted_powo_id",
              "wcvp_accepted_ipni_id",
              "wcvp_lifeform_description",
              "source_list")

bcilist <- bind_rows(
  garwood_list %>% 
    select(all_of(c(listscol, "SYNONYMS"))),
  zotz_no_garwood_at_all %>% select(all_of(listscol))
)

```


### Species in BCI plots census

```{r plotslist}
# Preparing plotlist

# 1. Read data
bci_plots <- read_excel("splists_raw/Plots/abundance_plots_HM.xlsx")

bci_other_plots <- read.delim(
  "splists_raw/Plots/doi_10_5061_dryad_1g1jwsvc3__v20260226/mainfile.txt"
)

# 2. Summarise other plots safely
bci_other_plots <- bci_other_plots %>%
  group_by(sp6) %>%
  summarise(
    bigplot = n_distinct(tag),
    .groups = "drop"
  )

# 3. Clean and rename BEFORE joining
bci_plots <- bci_plots %>%
  rename(
    sp6 = sp,
    sp_name_plot = binomial,
    bci50ha = bci,
    bci10ha = `bci 10 ha plot`
  )

# 4. Join safely (keeps all bci_plots species)
bci_plots <- bci_plots %>%
  left_join(bci_other_plots, by = "sp6") %>%
  mutate(
    bigplot = coalesce(bigplot, 0)   # replace NA with 0
  )

# 5. Filter safely (handle NA just in case)
bci_plots <- bci_plots %>%
  mutate(across(c(bci50ha, bci10ha, P14, P11), ~coalesce(., 0))) %>%
  filter(
    bci50ha > 0 |
    bci10ha > 0 |
    P14 > 0 |
    P11 > 0 |
    bigplot > 0
  )

# 6. Create plot_name column safely
bci_plots <- bci_plots %>%
  rowwise() %>%
  mutate(
    plot_name = paste(
      c(
        if (bci50ha > 0) "bci50ha",
        if (bigplot > 0) "bcimultiplots",
        if (bci10ha > 0) "bci10ha",
        if (P14 > 0) "P14",
        if (P11 > 0) "P11"
      ),
      collapse = ", "
    )
  ) %>%
  ungroup()

# the plots dont have wcvp accepted name, merge them by woody list
match21 <- match(bci_plots$sp6, woody_list$sp6)            
nomatchview <- bci_plots[is.na(match21),] %>% select(sp6, sp_name_plot) # All that dont match are unidentified

knitr::kable(nomatchview, caption="Species in plot that dont match with the woody bci list")

bci_plots <- bci_plots %>% left_join(
  woody_list, by="sp6"
)

knitr::kable(bci_plots %>% filter(duplicated(sp6)))

# create binomial first if not already
bci_plots$wcvp_binomial <- paste0(
  bci_plots$wcvp_accepted_name_genus, " ",
  bci_plots$wcvp_accepted_name_species
)

```


```{r plotbci}
# see the matches by the woody synonyms
woody_synonyms_garwood <- woody_synonyms %>% filter(
  SynonymTextSingle %in% garwood_list$orig_name |
  SynonymTextSingle %in% garwood_list$wcvp_accepted_name |
  SynonymTextSingle %in% garwood_list$wcvp_matched_name |
  SynonymTextSingle %in% garwood_synonyms$SynonymTextSingle
) 


# 1. direct name/binomial match
bci_plots_garwood_name_match <- bci_plots %>%
  filter(

    orig_name %in% bcilist$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist$wcvp_accepted_name) |

    # match binomials
    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist$orig_binomial |
        orig_binomial %in% garwood_synonyms$synonym_binomial
    )) |

    (!is.na(wcvp_binomial) & (
        wcvp_binomial %in% bcilist$wcvp_binomial |
        wcvp_binomial %in% garwood_synonyms$synonym_binomial
    ))
  )

# 2. synonym_sp match for remaining
bci_plots_garwood_synonym_match <- bci_plots %>%
  filter(
    !orig_name %in% bci_plots_garwood_name_match$orig_name &
    (
      orig_name %in% garwood_synonyms$synonym_sp |
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% garwood_synonyms$synonym_sp) |
        
        orig_name %in% woody_synonyms_garwood$orig_name |
        wcvp_accepted_name %in%  woody_synonyms_garwood$orig_name 
      )
  )

bci_plots_garwood_synonym_match$matched_garwood_synonym <- TRUE

# see species that matched by a synonym 
bci_plots_garwood_synonym_print <- bci_plots_garwood_synonym_match %>%  left_join(
    garwood_synonyms %>% select(sp_orig = sp_name, synonym_sp, syn_orig = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_match = sp_name, synonym_sp, syn_match = SynonymTextALL) %>% distinct(),
    by = c("wcvp_matched_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_wcvp = sp_name, synonym_sp, syn_wcvp = SynonymTextALL) %>% distinct(),
    by = c("wcvp_accepted_name" = "synonym_sp")) %>%
  left_join(
    woody_synonyms_garwood %>% 
      select(sp_origwoody_syn = orig_name, syn_origwoody_syn = SynonymTextALL, synonym_sp =SynonymTextSingle) %>% distinct(),
    by = c("orig_name" = "sp_origwoody_syn")) %>%
  left_join(
      woody_synonyms_garwood %>% 
        select(sp_matchwoody_syn = orig_name, syn_matchwoody_syn = SynonymTextALL, synonym_sp =SynonymTextSingle) %>% distinct(),
    by = c("wcvp_matched_name" = "sp_matchwoody_syn")) %>%
  left_join(
    woody_synonyms_garwood %>% select(sp_wcvpwoody_syn = orig_name, syn_wcvpwoody_syn = SynonymTextALL, synonym_sp =SynonymTextSingle) %>% distinct(),
    by = c("wcvp_accepted_name" = "sp_wcvpwoody_syn")) %>%
  mutate(
    GarwoodSynonymMatchAll = coalesce(syn_orig, syn_match, syn_wcvp),
    WoodyGarwoodSynonymMatch = coalesce(syn_origwoody_syn, syn_matchwoody_syn, syn_wcvpwoody_syn),
    orig_garwood = coalesce(sp_orig, sp_match, sp_wcvp),
  ) %>%
  select(-syn_orig, -syn_match, -syn_wcvp, 
           -sp_orig, -sp_match, -sp_wcvp, 
         -syn_origwoody_syn, -syn_matchwoody_syn, -syn_wcvpwoody_syn
         ) %>%
  distinct(orig_name, .keep_all = TRUE)
  

datatable(bci_plots_garwood_synonym_print %>% 
            select(orig_name, wcvp_matched_name, wcvp_accepted_name, GarwoodSynonymMatchAll, orig_garwood, WoodyGarwoodSynonymMatch), 
          caption="BCI plot species that matched garwood by synonym",
          options = list(scrollX = TRUE, pageLength = 10))

# 3. combine
bci_plots_garwood_match <- bind_rows(
  bci_plots_garwood_name_match,
  bci_plots_garwood_synonym_match
) %>%
  distinct(sp6, .keep_all = TRUE)

# 4. unmatched
bci_plots_no_garwood_at_all <- bci_plots %>%
  filter(!sp6 %in% bci_plots_garwood_match$sp6)

datatable(bci_plots_no_garwood_at_all  %>% select(sp6, orig_name, wcvp_matched_name, wcvp_accepted_name, plot_name), 
          caption="Species recorded in BCI plots census that dont match to anything in garwood",
          options = list(scrollX = TRUE, pageLength = 10))



write_xlsx(bci_plots_no_garwood_at_all, "splists_out/Plantnet species/plots_bci_not_garwood.xlsx")

```

Add to bci list
```{r}

bci_plots_no_garwood_at_all$source_list = "PlotCensusBCI"


bcilist <- bind_rows(
  bcilist,
  bci_plots_no_garwood_at_all %>% select(all_of(c(listscol, "plot_name", "bci50ha", "bci10ha", "P11", "P14"))
))

```


### Lianas BCI list
```{r lianasnobcilist}

lianas_list$wcvp_binomial <- paste0(
  lianas_list$wcvp_accepted_name_genus, " ",
  lianas_list$wcvp_accepted_name_species
)

# 1. direct name/binomial match
lianas_garwood_name_match <- lianas_list %>%
  filter(

    orig_name %in% bcilist$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist$wcvp_accepted_name) |

    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist$orig_binomial |
        orig_binomial %in% garwood_synonyms$synonym_binomial
    )) |

    (!is.na(wcvp_binomial) & (
        wcvp_binomial %in% bcilist$wcvp_binomial |
        wcvp_binomial %in% garwood_synonyms$synonym_binomial
    ))
  )

# 2. synonym_sp for remaining
lianas_garwood_synonym_match <- lianas_list %>%
  filter(
    !orig_name %in% lianas_garwood_name_match$orig_name &
    (
      orig_name %in% garwood_synonyms$synonym_sp |
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% garwood_synonyms$synonym_sp)
    )
  )

lianas_garwood_synonym_match$matched_garwood_synonym <- TRUE


# see species that matched by a synonym 
lianas_garwood_synonym_match_print <- lianas_garwood_synonym_match %>%  left_join(
    garwood_synonyms %>% select(sp_orig = sp_name, synonym_sp, syn_orig = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_match = sp_name, synonym_sp, syn_match = SynonymTextALL) %>% distinct(),
    by = c("wcvp_matched_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_wcvp = sp_name, synonym_sp, syn_wcvp = SynonymTextALL) %>% distinct(),
    by = c("wcvp_accepted_name" = "synonym_sp")) %>%
  mutate(
    SynonymMatchAll = coalesce(syn_orig, syn_match, syn_wcvp),
    orig_garwood = coalesce(sp_orig, sp_match, sp_wcvp),
  ) %>%
  select(-syn_orig, -syn_match, -syn_wcvp,  -sp_orig, -sp_match, -sp_wcvp) %>%
  distinct(orig_name, .keep_all = TRUE)
  

datatable(lianas_garwood_synonym_match_print %>% 
            select(orig_name, wcvp_matched_name, wcvp_accepted_name, SynonymMatchAll, orig_garwood), 
          caption="Liana species that matched garwood by synonym",
          options = list(scrollX = TRUE, pageLength = 10))


lianas_garwood_match <- bind_rows(
  lianas_garwood_name_match,
  lianas_garwood_synonym_match
) %>%
  distinct(orig_name, .keep_all = TRUE)

lianas_no_garwood_at_all <- lianas_list %>%
  filter(!orig_name %in% lianas_garwood_match$orig_name)



# 5. Output and Display
datatable(lianas_no_garwood_at_all %>%
  select(orig_name, wcvp_matched_name, wcvp_accepted_name), 
          caption = "Lianas species that don't match to anything in Garwood",
          options = list(scrollX = TRUE, pageLength = 10))

write_xlsx(lianas_no_garwood_at_all, "splists_out/lianas_species_not_in_garwood.xlsx")

```

Add to bci list
```{r lianasbciadd}

lianas_no_garwood_at_all$source_list = "lianas"

bcilist <- bind_rows(
  bcilist,
  lianas_no_garwood_at_all %>% select(all_of(listscol))
)

```

### Save BCI BOTANISTS HIGH QUALITY LIST
```{r}

write_xlsx(bcilist, paste0(save_plantlists_dir, "BCI_SPECIES_BOTANISTS_LIST_",Sys.Date(),".xlsx"))

datatable(
  bcilist %>%
    select(wcvp_accepted_name, orig_name, source_list) %>% 
    arrange(source_list),
    caption="Species list in BCI and their source",
  options = list(scrollX = TRUE, pageLength = 10))

```


## BCI "everything list"

Here we add the panama biota species.

### Panamabiota species that are not in garwood

```{r pbiotanogarwood}

pbiota_list$wcvp_binomial <- paste0(
  pbiota_list$wcvp_accepted_name_genus, " ",
  pbiota_list$wcvp_accepted_name_species
)

# 1. direct name/binomial match
pbiota_garwood_name_match <- pbiota_list %>%
  filter(

    orig_name %in% bcilist$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist$wcvp_accepted_name) |

    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist$orig_binomial |
        orig_binomial %in% garwood_synonyms$synonym_binomial
    )) |

    (!is.na(wcvp_binomial) & (
        wcvp_binomial %in% bcilist$wcvp_binomial |
        wcvp_binomial %in% garwood_synonyms$synonym_binomial
    ))
  )

# 2. synonym_sp for remaining
pbiota_garwood_synonym_match <- pbiota_list %>%
  filter(
    !orig_name %in% pbiota_garwood_name_match$orig_name &
    (
      orig_name %in% garwood_synonyms$synonym_sp |
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% garwood_synonyms$synonym_sp)
    )
  )

pbiota_garwood_synonym_match$matched_garwood_synonym <- TRUE


# see species that matched by a synonym 
pbiota_garwood_synonym_match_print <- pbiota_garwood_synonym_match %>%  left_join(
    garwood_synonyms %>% select(sp_orig = sp_name, synonym_sp, syn_orig = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_match = sp_name, synonym_sp, syn_match = SynonymTextALL) %>% distinct(),
    by = c("wcvp_matched_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_wcvp = sp_name, synonym_sp, syn_wcvp = SynonymTextALL) %>% distinct(),
    by = c("wcvp_accepted_name" = "synonym_sp")) %>%
  mutate(
    SynonymMatchAll = coalesce(syn_orig, syn_match, syn_wcvp),
    orig_garwood = coalesce(sp_orig, sp_match, sp_wcvp),
  ) %>%
  select(-syn_orig, -syn_match, -syn_wcvp,  -sp_orig, -sp_match, -sp_wcvp) %>%
  distinct(orig_name, .keep_all = TRUE)
  

datatable(pbiota_garwood_synonym_match_print %>% 
            select(orig_name, wcvp_matched_name, wcvp_accepted_name, SynonymMatchAll, orig_garwood), 
          caption="Panama biota species that matched garwood by synonym",
          options = list(scrollX = TRUE, pageLength = 10))

pbiota_garwood_match <- bind_rows(
  pbiota_garwood_name_match,
  pbiota_garwood_synonym_match
) %>%
  distinct(orig_name, .keep_all = TRUE)

pbiota_no_garwood_at_all <- pbiota_list %>%
  filter(!orig_name %in% pbiota_garwood_match$orig_name)



datatable(pbiota_no_garwood_at_all, 
          caption="Panama biota species that dont match to anything in garwood",
          options = list(scrollX = TRUE, pageLength = 10))

write_xlsx(pbiota_no_garwood_at_all, "splists_out/panamabiota_species_not_in_garwood.xlsx")

```

Add to bci list
```{r mergebci}

pbiota_no_garwood_at_all$source_list = "PanamaBiota"

bcilist <- bind_rows(bcilist, pbiota_no_garwood_at_all)

```



### Save BCI everything list
```{r}

write_xlsx(bcilist, paste0(save_plantlists_dir, "BCI_SPECIES_EVERYTHING_LIST_",Sys.Date(),".xlsx"))


```

```{r printresult}
datatable(
  bcilist %>%
    select(wcvp_accepted_name, orig_name, source_list) %>% 
    arrange(source_list),
    caption="Species list in BCI and their source",
  options = list(scrollX = TRUE, pageLength = 10))

```


# See species that matched by a garwood synonyms
```{r matchsynonims}
bcilist<- bcilist %>%
  left_join(
    garwood_synonyms %>% select(sp_orig = sp_name, synonym_sp, syn_orig = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_curr = sp_name, synonym_sp, syn_curr = SynonymTextALL) %>% distinct(),
    by = c("wcvp_accepted_name" = "synonym_sp"), relationship = "many-to-many") %>%
  mutate(
    SynonymTextALL = coalesce(syn_orig, syn_curr),
    orig_garwood = coalesce(sp_orig, sp_curr),
  ) %>%
  select(-syn_orig, -syn_curr, -sp_orig, -sp_curr) %>%
  distinct(wcvp_accepted_name, orig_name, .keep_all = TRUE)
 
```


## Add woody species column
 
If the species is in the woody list

```{r bcilistwoody, echo=TRUE}
# set woody if:
# original, matched or accepted name is anywhere in woody list

bcilist <- bcilist %>% mutate(
  woody_name =  wcvp_accepted_name %in% woody_list$current_name | 
                wcvp_accepted_name %in% woody_list$orig_name | 
                wcvp_matched_name %in% woody_list$current_name |
                wcvp_matched_name %in% woody_list$orig_name |
                orig_name %in% woody_list$orig_name |
                orig_name %in% woody_list$current_name,
  woody_synonym =  wcvp_accepted_name %in% woody_synonyms$SynonymTextSingle |
                   orig_name %in% woody_synonyms$SynonymTextSingle |
                   wcvp_matched_name %in% woody_synonyms$SynonymTextSingle,
  woody = woody_name | woody_synonym
) 

#write_xlsx(bcilist, paste0(save_plantlists_dir, "BCI_SPECIES_LIST_",Sys.Date(),".xlsx"))


```

### Possibly woody species

```{r posiblywoody, echo=TRUE, results="hide"}

lifeform_woody = c(
  "tree",
  "shrub",
  "liana",
  "climbing shrub",
  "shrub or tree",
  "climbing shrub or tree",
  "scrambling shrub or tree",
  "subshrub or shrub",
  "subshrub, shrub or tree",
  "scrambling shrub",
  "perennial or shrub",
  "scrambling subshrub or shrub",
  "epiphytic shrub",
  "scrambling tree",
  "climbing shrub or liana",
  "scrambling shrub or liana",
  "tuberous shrub or tree",
  "herbaceous tree",
  "climbing herbaceous tree",
  "climbing herbaceous tree or liana",
  "climbing perennial or herbaceous tree",
  "subshrub or climbing shrub",
  "tuberous subshrub",
  "hemiparasitic shrub",
  "hemiepiphytic shrub or tree",
  "annual or shrub",
  "annual, subshrub or shrub"
)

bcilist_possibly_woody <- bcilist %>%
  filter(!woody &
          wcvp_lifeform_description %in% lifeform_woody ) %>%
  select(orig_name, wcvp_matched_name, wcvp_accepted_name, wcvp_accepted_family, wcvp_lifeform_description)

datatable(bcilist_possibly_woody, caption="BCI species that might be woody according to the wcvp lifeform")

write_xlsx(bcilist_possibly_woody, paste0(save_plantlists_dir, "BCI_SPECIES_LIST_POSSIBLY_WOODY_",Sys.Date(),".xlsx"))


```
