---
title: "Assemble plant species list for Barro Colorado Island, Panama"
author: "Paula Uzcategui and Helene Muller-Landau"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

This Rmarkdown merges information from multiple high-quality plant species lists for Barro Colorado Island (BCI), Panama, updates to the best current woody plant species names based on the current Smithsonian ForestGEO woody species lists, utilizes prior checks of species names against the World Checklist of Vascular Plants (WCVP) version 13, and outputs the combined high-quality list. The reason for using WCVP version 13 is because that is the version currently in use by PlantNet.  The contributing lists are Garwood et al. 2009, Schnitzer et al. 2024, Zotz 2024, the Smithsonian ForestGEO Panama woody plant species and ViewTaxonomy lists (Perez et al. 2025), and a file of tree species found in forest inventory plots on BCI.  These lists are considered high-quality because they were each assembled by one or more botanists familiar with the local flora, and build on the work of Croat 1978. Combining the lists involves checking overlap not only in the original names on each list, but also on two lists of "synonyms" for BCI plant species, as well as in the current accepted name.  The synonym lists include one from Garwood 2009 and two from Smithsonian ForestGEO.  In all cases, these include not only synonyms in the botanical sense, but also prior names used for the same BCI morphospecies (including in cases where species determination was revised).  

This Rmarkdown additionally outputs a list of plant species that appear in panamabiota.org's list of plants of Barro Colorado, that are *not* in the combined high-quality list. This list is generated for the purpose of later review.  

The component and output lists are all available in the public GitHub site https://github.com/PanamaForestGEO/Panama_plant_species_lists 


## References 

Croat, Thomas B. 1978. *Flora of Barro Colorado Island*. Stanford University Press.

Garwood, Nancy C., Nancy Hensold, and Robin B. Foster. 2009. Appendix 1.  Revised Species List for Barro Colorado Island, with Synonyms. In *Seedlings of Barro Colorado Island and the Neotropics*. Cornell University Press.

Pérez, Rolando, Salomón Aguilar and David Mitre.  2025. Lista de árboles, arbustos, palmas, lianas y enredaderas de la flora de Panamá.  Version 2025-12-04.  

Schnitzer, Stefan, David DeFilippis, Rigoberto Rivera-Camaña, et al. 2024. Liana Species of the Barro Colorado Island 50-Ha Plot 2007 and 2017 Censuses. Version 7. *Dryad* https://doi.org/10.5061/DRYAD.Q2BVQ83V4.

Zotz, Gerhard.  2024.  Table S1. Updated list of epiphyte species on BCI. Online Supplementary Material for Zotz, G. 2024. Epiphyte Research on BCI. In *The First 100 Years of Research on Barro Colorado: Plant and Ecosystem Science*, ed. H. C. Muller-Landau and S. J. Wright. Smithsonian Institution Scholarly Press. https://doi.org/10.5479/si.26880865

### To do 

When reading in files, it would be good to rename columns right away if necessary to get rid of blanks, /’s, and other problematic characters, so subsequent references can be simplified.

for kable(zotznocroatbutmatched %>% filter(firstmatchtype=="orig-syn"),
      caption="Zotz species that Zotz says do not appear in Croat 1978 that nonetheless
      matched to a Garwood synonym. ")
      
      change to giving the name or names that show up in Garwood for this taxon.  
      
    
## Setup

### Load libraries
```{r setup2,}

rm(list=ls())

library(readxl)
library(writexl)
library(dplyr)
library(tidyr)
library(stringi)
library(stringr)
library(knitr)
library(DT)
```

### Read input files and initial data prep

```{r readprepdata, echo=TRUE, results="hide"}


garwood_list_orig  <- read_excel("splists_out/Garwood_2009_Appendix1HM2_WCVP_2026-03-15.xlsx", col_types = "text")

zotz_list <- read_excel("splists_out/83_Zotz_Epiphytes_TableS1_WCVP_2026-03-25.xlsx", col_types = "text") %>%
  rename_with(~ gsub("[()]", "", gsub(" ", "_", .x)))

lianas_list <- read_excel("splists_out/Lianas_Species_complete_list_WCVP_2026-03-16.xlsx", col_types = "text") %>%
  mutate(sp6 = tolower(sp6)) %>%
  rename(wcvp_accepted_lifeform=wcvp_lifeform_description)

pbiota_list <- read_excel("splists_out/PanamaBiotaBCI_WCVP_2026-02-26.xlsx", col_types = "text")

panwoody_list <- read_excel("splists_out/CurrentPanamaWoody_2026-03-26.xlsx", col_types = "text")
panwoody_synonyms <- read_excel("splists_out/alternates_and_synonyms/CurrentPanamaWoody_SYNONYMS_2026-03-26.xlsx", col_types="text")

viewtaxa_list <- read_excel("splists_raw/Forestgeo/2026-02-25FromSuzanne/ViewTaxonomy_dec4_25_HM2.xlsx")
viewtaxa_synonyms <- read_excel("splists_out/alternates_and_synonyms/ViewTaxonomy_dec4_25_SYNONYMS.xlsx")

alternate_names <- read_excel("splists_out/alternates_and_synonyms/AltNames_2026-03-24.xlsx")
alternate_names_garwood <- read_excel("splists_out/alternates_and_synonyms/alternate_names_garwood_2026-03-24.xlsx")


allwrighttaxa <- read_excel("splists_out/wright_nomenclature_R_20210224_WCVP_2026-03-26.xlsx", na = c("", "NA"))

garwood_synonyms <- read_excel("splists_raw/Garwood/Garwood_2009_Appendix1_Synonyms.xlsx") 

garwood_synonyms <- garwood_synonyms %>% mutate(
  sp_name = trimws(paste0(Genus, " ", Species)),
  synonym_sp = trimws(paste0(SynonymGenusSingle, " ", SynonymSpeciesSingle)),
  synonym_binomial = paste0(SynonymGenusSingle, " ", SynonymSpeciesSingle)) 
# note that synonym_sp and synonym_binomial are identical; retaining for clarity of completeness

garwood_list <- garwood_list_orig %>%
  mutate(wcvp_accepted_binomial = 
           paste0(wcvp_accepted_name_genus, " ", wcvp_accepted_name_species),
         wcvp_matched_binomial = word(wcvp_matched_name,)) %>%
  rename(wcvp_accepted_lifeform = wcvp_lifeform_description)





```

We have a list of alternate_names_garwood is a list of alternate names that match to garwood. It is generated in the alternate_species_names.Rmd to be used in this script.

```{r buildaltmatchlookup}
# ─────────────────────────────────────────────────────────────────────────────
# Build a flat lookup table used downstream to resolve BCI plot species and
# lianas list entries that did not match Garwood directly.
# Both the preferred name and the alternate name are indexed as possible keys
# so that either form can be used to retrieve the Garwood match.
# ─────────────────────────────────────────────────────────────────────────────
alt_match_lookup <- bind_rows(
  alternate_names_garwood %>%
    select(match_key = prefname, garwood_orig_match, garwood_synonym_match,
           firstmatchtype, alt_name = altname, pref_name = prefname),
  alternate_names_garwood %>%
    select(match_key = altname,  garwood_orig_match, garwood_synonym_match,
           firstmatchtype, alt_name = altname, pref_name = prefname)
) %>%
  filter(!is.na(match_key)) %>%
  distinct(match_key, .keep_all = TRUE)

```

### Output file names 

MAIN OUTPUTS: 

- Combined, high-quality list of BCI plant species.  

- List of BCI species that might be woody according to WCVP lifeform. 

ADDITIONAL OUTPUTS OF FILES TO CHECK: 

- List of Zotz species not in Garwood

- List of BCI plot species not in Garwood

- List of BCI liana list species not in Garwood

- List of PanamaBiota BCI species that are not in the combined high-quality list

```{r outfilenames}
save_plantlists_dir = "splists_out/Plantnet/"

FILECOMBINEDLIST <- paste0(save_plantlists_dir, 
                               "BCI_SPECIES_BOTANISTS_LIST_",Sys.Date(),".xlsx")
FILEBCIWOODY <- paste0("splists_out/needs_revision/BCI_SPECIES_POSSIBLY_WOODY_",Sys.Date(),".xlsx")

FILECHECKZOTZ <- "splists_out/needs_revision/zotz_species_not_in_garwood.xlsx"

FILECHECKPLOTS <-  "splists_out/needs_revision/plotsbci_not_garwood.xlsx"

FILECHECKLIANAS <- "splists_out/needs_revision/lianasbci_not_garwood.xlsx"

FILECHECKPANAMABIOTA <- "splists_out/needs_revision/panamabiota_not_garwood.xlsx"

WRIGHT_MISSING_PANWOODY_PATH <- paste0("splists_out/needs_revision/Wright_not_CurrentPanamaWoody_",Sys.Date(),".xlsx")

```



## Assembling a combined BCI high-quality botanist-determined plant species list

### Garwood 2009 list

Some species in Garwood are listed as being only in the Barro Colorado Nature Monument but not on BCI itself (chg=="BCNM"). However, after matching with species recorded in the plots in BCI these species show up. Because of that we decide not to remove these species listed as only BCNM.

```{r garwoodnobcnm}


# filter out species that are in the mainland peninsulas but not BCI, except for those that turn out to be on BCI after all...

#Buchenavia tetraphylla is an alternate name of Terminalia tetraphylla which is found in the plots list

yesonbcibinomial <- c("Matayba oppositifolia", "Myrcia zetekiana", "Buchenavia tetraphylla")

#garwood_list <- garwood_list %>% filter(chg != "BCNM" | is.na(chg) | #orig_name %in% yesonbcibinomial)
garwoodtoremove <- subset(garwood_list_orig, chg=="BCNM")

kable(garwoodtoremove %>% select(orig_name,SYNONYMS),
      caption = "Species that according to Garwood are in the mainland peninsulas but not on BCI itself.")
```



```{r garwoodnoid}

# view unidentified species
garwoodnobinomial <- subset(garwood_list_orig,is.na(word(orig_binomial,2)))

kable(garwoodnobinomial %>% select(orig_name,Syn_notes) %>% rename(garwood_name=orig_name),
      caption="Garwood species lacking a complete binomial")

```

This leaves `r nrow(garwood_list)` plant species on the Garwood list.  Note that in `r nrow(garwoodnobinomial)` cases these species lack proper binomials, indicating their species identity remains uncertain (including possible new species).  These are left in anyway because they are known morphospecies.  

### Zotz 2024 epiphyte species list

There are `r nrow(zotz_list)` species on the Zotz 2024 epiphyte list.  

Match Zotz species to the Garwood list as far as possible, including by synonynms in the broad sense, and identify Zotz species not on the Garwood list.  

Note that we make one manual correction to a Zotz misspelling of a Croat name to enable matching with the Garwood list for that name.  Other misspellings to the Zotz names themselves are addressed by the matching with WCVP.  
```{r zotznogarwood}
# removing morphospecies identified only at the genus level (there is only one)
zotz_list <- zotz_list %>% filter(is.na(orig_binomial) | !is.na(word(orig_binomial,2) ))

# when the name listed is the same as in Croat 1978, the column Sp_Croat_1978 is blank (NA)
# so create a new name_Croat field that is the name of the species in Croat 1978
zotz_list <- zotz_list %>%
  mutate(wcvp_accepted_binomial = 
           paste0(wcvp_accepted_name_genus," ",wcvp_accepted_name_species),
         wcvp_matched_binomial = word(wcvp_matched_name,1,2),
         croat_name = ifelse(!is.na(Sp_Croat_1978),Sp_Croat_1978,orig_name)) %>%
  mutate(croat_binomial = word(croat_name,1,2)) %>%
  rename(wcvp_accepted_lifeform=wcvp_lifeform_description)

# make one manual correction so matches work right
zotz_list <- zotz_list %>%
  mutate(croat_binomial =ifelse(croat_binomial=="Bolbiltis nicotianifolia",
         "Bolbitis nicotianifolia",croat_binomial))

# matching is done on the binomials (not full names including subspecies)

zotz_list <- zotz_list %>%
  mutate(matchorig = 
           orig_binomial %in% garwood_list$orig_binomial,
         matchcroat = 
           croat_binomial %in% garwood_list$orig_binomial,
         matchorigsynonym = 
           orig_binomial %in% garwood_synonyms$synonym_binomial,
         matchcroatsynonym =
           croat_binomial %in% garwood_synonyms$synonym_binomial,
         matchwcvpmatch = 
           !is.na(wcvp_matched_name) & 
              (wcvp_matched_binomial %in% c(garwood_list$wcvp_matched_binomial,
                                        garwood_list$orig_binomial,
                                        garwood_synonyms$synonym_binomial)),
         matchwcvpaccept = 
           !is.na(wcvp_accepted_name) & 
              wcvp_accepted_binomial %in% c(garwood_list$wcvp_matched_binomial,
                                        garwood_list$orig_binomial,
                                        garwood_synonyms$synonym_binomial)
  ) %>%
  mutate(matchany = matchorig | matchcroat | matchorigsynonym |matchcroatsynonym | 
           matchwcvpmatch | matchwcvpaccept) %>% 
  mutate(firstmatchtype=ifelse(matchorig,"orig-orig",
                               ifelse(matchcroat,"croat-orig",
                                      ifelse(matchorigsynonym,"orig-syn",
                                             ifelse(matchcroatsynonym,"croat-syn",
                                                    ifelse(matchwcvpmatch,"wcvpmatch",
                                                           ifelse(matchwcvpaccept,"wcvpaccpt",
                                                                  NA)))))))
           
zotz_nomatchgarwood <- zotz_list %>%  filter(!matchany)

write_xlsx(zotz_nomatchgarwood, FILECHECKZOTZ)

# check the species that match only on the WCVP names
checkmatch1 <- subset(zotz_list, matchany & 
                        !matchcroat & !matchorigsynonym & !matchorig & !matchcroatsynonym) %>%
  select(orig_name,match_similarity,wcvp_matched_name,wcvp_accepted_name,Sp_Croat_1978)
kable(checkmatch1,
      caption="Species in Zotz that match Garwood only via WCVP matched or accepted names.  These are all cases in which the name is misspelled in Zotz.")

# see species that matched by a synonym and not by original or Croat names
checkmatch2a <- zotz_list %>% filter(matchorigsynonym & !matchorig & !matchcroat) %>%  
  left_join(
    garwood_synonyms %>% select(garwood_orig = sp_name, synonym_binomial, syn_text = SynonymTextALL) %>% distinct(),
    by = c("orig_binomial" = "synonym_binomial")
  )%>% 
  select(orig_binomial,croat_binomial,garwood_orig,firstmatchtype,syn_text)

checkmatch2b <- zotz_list %>% filter(matchcroatsynonym & !matchorig & !matchcroat) %>%  
  left_join(
    garwood_synonyms %>% select(garwood_orig = sp_name, synonym_binomial, syn_text = SynonymTextALL) %>% distinct(),
    by = c("croat_binomial" = "synonym_binomial")
  ) %>% mutate(matchtype="croat-to-synonym") %>%
  select(orig_binomial,croat_binomial,garwood_orig,firstmatchtype,syn_text)

checkmatch2 <- rbind(checkmatch2a,
                     subset(checkmatch2b,!orig_binomial %in% checkmatch2a$orig_binomial))

kable(checkmatch2, 
          caption="Zotz species that matched Garwood only by synonym. Note that in some cases there was more than one match by synonym.  Here orig_binomial means the current name in Zotz.",
          options = list(scrollX = TRUE, pageLength = 10))

zotznocroatbutmatched <- zotz_list %>%
  filter(Sp_Croat_1978=="**" & matchany==T) %>%
  arrange(firstmatchtype,orig_name) %>%
  select(orig_name,wcvp_matched_name,wcvp_accepted_name,firstmatchtype,Sp_Croat_1978)

kable(zotznocroatbutmatched %>% filter(firstmatchtype=="orig-orig"),
      caption="Zotz species that Zotz says do not appear in Croat 1978 that nonetheless
      matched in Garwood, and appear with the same name in both")
  

kable(zotznocroatbutmatched %>% filter(firstmatchtype=="orig-syn"),
      caption="Zotz species that Zotz says do not appear in Croat 1978 that nonetheless
      matched to a Garwood synonym. ")

kable(zotz_nomatchgarwood  %>% select(orig_name, wcvp_matched_name, wcvp_accepted_name, Sp_Croat_1978), 
          caption="Zotz species that don't match anything in Garwood.  Note that in all cases Sp_Croat_1978 has value **, indicating that Zotz 2024 states that the species does not appear in Croat 1978.")

```

In total there are `r nrow(zotz_nomatchgarwood)` species on the Zotz list not already on the Garwood list.  

Add these Zotz species to the combined BCI list, which starts with Garwood.
```{r}
# using garwood as a base
garwood_list$source_list = "Garwood2009"
zotz_nomatchgarwood$source_list = "Zotz"

listscol <- c("orig_name",
              "orig_binomial",
              "wcvp_accepted_binomial",
              "wcvp_matched_name",
              "wcvp_matched_plant_name_id",
              "wcvp_matched_ipni_id",
              "wcvp_matched_status",
              "wcvp_accepted_name",
              "wcvp_accepted_family", 
              "wcvp_accepted_authority",
              "wcvp_accepted_plant_name_id", 
              "wcvp_accepted_powo_id",
              "wcvp_accepted_ipni_id",
              "wcvp_accepted_lifeform",
              "source_list"
              )

bcilist <- bind_rows(
  garwood_list %>% 
    select(all_of(c(listscol, "SYNONYMS", "LF_RF", "BCI cult"))),
  zotz_nomatchgarwood %>% select(all_of(listscol))
)

```

The pre-existing "altnerate_names" file combines species names in the synonyms or "old names" columns of currentwoody, viewtaxa and garwood. 

These are not necessarily synonyms in the botanical sense; they are any names formerly used for a given BCI morphospecies (e.g., due to misidentification) 

If a species in the BCI plots or in the lianas list does not match the garwood directly, an alternate name related to the species could match. 


### List of tree species recorded in censuses of forest plots on BCI 

Check that all the tree species recorded from forest plots on BCI are also in the combined species list.  

```{r plotslist}
# Preparing plotlist

# 1. Read data
bci_plots <- read_excel("splists_raw/Plots/abundance_plots_HM.xlsx")

bci_plots <- bci_plots %>% select(sp, binomial, bci, `bci 10 ha plot`, P11, P14)

bci_other_plots <- read.delim(
  "splists_raw/Plots/doi_10_5061_dryad_1g1jwsvc3__v20260226/mainfile.txt"
)

# 2. Summarise other plots safely
bci_other_plots <- bci_other_plots %>%
  group_by(sp6) %>%
  summarise(
    bigplot = n_distinct(tag),
    .groups = "drop"
  )

# 3. Clean and rename BEFORE joining
bci_plots <- bci_plots %>%
  rename(
    sp6 = sp,
    sp_name_plot = binomial,
    bci50ha = bci,
    bci10ha = `bci 10 ha plot`
  )

# 4. Join safely (keeps all bci_plots species)
bci_plots <- bci_plots %>%
  left_join(bci_other_plots, by = "sp6") %>%
  mutate(
    bigplot = coalesce(bigplot, 0)   # replace NA with 0
  )

# 5. Filter safely (handle NA just in case)
bci_plots <- bci_plots %>%
  mutate(across(c(bci50ha, bci10ha, P14, P11), ~coalesce(., 0))) %>%
  filter(
    bci50ha > 0 |
    bci10ha > 0 |
    P14 > 0 |
    P11 > 0 |
    bigplot > 0
  )

# 6. Create plot_name column safely
bci_plots <- bci_plots %>%
  rowwise() %>%
  mutate(
    plot_name = paste(
      c(
        if (bci50ha > 0) "bci50ha",
        if (bigplot > 0) "bcimultiplots",
        if (bci10ha > 0) "bci10ha",
        if (P14 > 0) "P14",
        if (P11 > 0) "P11"
      ),
      collapse = ", "
    )
  ) %>%
  ungroup()

# the plots files don't have wcvp accepted name, merge them by 6-letter species code in the Panama woody species list
match21 <- match(bci_plots$sp6, panwoody_list$sp6)            
nomatchview <- bci_plots[is.na(match21),] %>% select(sp6, sp_name_plot) # Those that don't match are all unidentified

if (nrow(nomatchview)>0){
knitr::kable(nomatchview, caption="Warning: Species recorded in plots on BCI whose 6-letter codes do not have a match in the Panama ForestGEO woody plant species database.")} else{
  print("All species in the plot have a match to a sp6 code in the current woody list.")
}

bci_plots <- bci_plots %>% left_join(
  panwoody_list, by="sp6"
) %>%
  rename(forestgeo_current_name = orig_name)

# Some names in the plots do not match the names in forestgeo
datatable(
bci_plots %>% 
  filter(is.na(sp_name_plot)|sp_name_plot != forestgeo_current_name) %>% 
  select(sp6, sp4, sp_name_plot, forestgeo_current_name = forestgeo_current_name),
caption="Sp6 codes in the plots match to forestgeo codes, but some of the species names used in the plots do not match the forestgeo name associated with that code. "
)

# We will use the forestgeo current name to make all the comparisons with this list
bci_plots <- bci_plots %>% mutate(orig_name = forestgeo_current_name)

# remove the row for "unidentified", if present
bci_plots <- bci_plots %>% filter(sp6!="uniden")

bcidups <- bci_plots %>% filter(duplicated(sp6)) %>% select(sp6, sp_name_plot)

if (nrow(bcidups)>0)
knitr::kable(bcidups, caption="Warning: Some 6-letter species codes are repeated in the bci_plots data.frame.")

# create binomial first if not already
bci_plots$wcvp_accepted_binomial <- paste0(
  bci_plots$wcvp_accepted_genus, " ",
  bci_plots$wcvp_accepted_species
)

datatable(
bci_plots %>%
  select(sp6, sp4, sp_name_plot, forestgeo_current_name,),
caption="BCI plots species list"
)

```

*In total there are `r nrow(bci_plots)` tree morphospecies recorded for forest plot censuses on BCI.*

Matching to original garwood names (direct match) or to synonyms or alternate names that match to garwood.
 
#### 1. Direct match to garwood names

```{r bciplotdirectmatch}
# 1. direct name/binomial match
bci_plots_garwood_name_match <- bci_plots %>%
  filter(
    orig_name %in% bcilist$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist$wcvp_accepted_name) |

    # match binomials
    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist$orig_binomial 
    )) |

    (!is.na(wcvp_accepted_binomial) & (
        wcvp_accepted_binomial %in% bcilist$wcvp_accepted_binomial 
    ))
  )

datatable(
  bci_plots_garwood_name_match %>% 
  select(sp6, forestgeo_current_name, wcvp_matched_name, wcvp_accepted_name), 
          caption="BCI plot species matching directly to garwood names",
          options = list(scrollX = TRUE, pageLength = 10))
```

          
          
*Matched `r nrow(bci_plots_garwood_name_match)` species in the plots directly to garwood.*

#### 2. Alternate name match 
```{r plotalternatematchtry}


# 2. everything that did not match directly to garwood, match to alternate names
bci_plots_garwood_synonym_match <- bci_plots %>% 
  filter(
      !orig_name %in% bci_plots_garwood_name_match$orig_name &
    (
      orig_name %in% alt_match_lookup$match_key |
      orig_binomial %in% alt_match_lookup$match_key |
      
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% alt_match_lookup$match_key) |
        
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% alt_match_lookup$match_key)
    )
  )


# 3. print the altname it matched to and why that name is in garwood

bci_plots_garwood_synonym_print <- bci_plots_garwood_synonym_match %>%
  # Join using a chain of coalesce to check the most reliable fields first
  left_join(alt_match_lookup, by = c("orig_name" = "match_key")) %>%
  left_join(alt_match_lookup, by = c("orig_binomial" = "match_key"), suffix = c("", "_bi")) %>%
  left_join(alt_match_lookup, by = c("wcvp_matched_name" = "match_key"), suffix = c("", "_match")) %>%
  left_join(alt_match_lookup, by = c("wcvp_accepted_name" = "match_key"), suffix = c("", "_acc")) %>%
  
  # Now, condense everything into final fields
  mutate(
    orig_garwood    = coalesce(garwood_orig_match, garwood_orig_match_bi, garwood_orig_match_match, garwood_orig_match_acc),
    orig_garwood_synonym = coalesce(garwood_synonym_match, garwood_synonym_match_bi, garwood_synonym_match_match, garwood_synonym_match_acc),
    firstmatchtype  = coalesce(firstmatchtype, firstmatchtype_bi, firstmatchtype_match, firstmatchtype_acc),
    alternate_name  = coalesce(alt_name, alt_name_bi, alt_name_match, alt_name_acc),
    prefered_name   = coalesce(pref_name, pref_name_bi, pref_name_match, pref_name_acc)
  ) %>%
  select(-ends_with("_bi"), -ends_with("_match"), -ends_with("_acc")) # Clean up extra columns

```

*Matched `r nrow(bci_plots_garwood_synonym_match)` species in the plots by synonyms or alternate names to garwood.*

The table below shows BCI plot species that could not be matched directly to Garwood (2009) but were linked through an alternate name. The `firstmatchtype` column indicates which name pairing produced the link:

Definitions:

- F=ForestGEO

- G=Garwood

- Fcur = ForestGEO current name from most recent list

- Falt = ForestGEO alternate name from most recent list for Panama woody plants in general

- Gorig = Garwood 2009 original name

- Galt = Garwood 2009 alternate name

- **direct synonym** — the name appears directly in the Garwood list as a synonym

 Definitions:

F=ForestGEO

G=Garwood

Fcur = ForestGEO current name from most recent list

Falt = ForestGEO alternate name from most recent list for Panama woody plants in general

Gorig = Garwood 2009 original name

Galt = Garwood 2009 alternate name

 
- **Fcur-Gorig** — the current ForestGEO name matches a name listed by Garwood

- **Fcur-Gorigbi** — the current ForestGEO name matches the binomial of a Garwood original name.

- **Falt-Gorig** — a ForestGEO alternate name matches a name originally listed by Garwood

- **Falt-Gorigbi** — a ForestGEO alternate name matches the binomial of a Garwood original name.

- **Fcur-Galt** — the current ForestGEO name matches a synonym documented by Garwood that is conflicting with a forestgeo alternate-prefered pair

- **Fcur-Galtbi** — the current ForestGEO name matches the binomial of a Garwood synonym

- **Falt-Galt** — a historical ForestGEO alternate name matches a synonym documented by Garwood

- **Falt-Galtbi** — a historical ForestGEO alternate name matches the binomial of a Garwood synonym

 

Note that "Garwood original name" refers to the name as it appears in the Garwood (2009) revision, not necessarily the currently accepted name.

```{r printaltmatches}
datatable(
  bci_plots_garwood_synonym_print %>% 
  select(sp6, forestgeo_current_name, wcvp_matched_name, wcvp_accepted_name, firstmatchtype, orig_garwood,
         alternate_name, prefered_name, orig_garwood_synonym), 
          caption="BCI plot species where an alternate name of the species matched to garwood",
          options = list(scrollX = TRUE, pageLength = 10))


```

Printing species in the plots that are not in Garwood. 
```{r allnomatchbciplot}

# combine direct matches and matches by alternate names or synonyms
bci_plots_garwood_match <- bind_rows(
  bci_plots_garwood_name_match,
  bci_plots_garwood_synonym_match
) %>%
  distinct(sp6, .keep_all = TRUE)

# filter unmatched rows
bci_plots_no_garwood_at_all <- bci_plots %>%
  filter(!sp6 %in% bci_plots_garwood_match$sp6)

kable(bci_plots_no_garwood_at_all  %>% select(sp6, sp_name_plot, forestgeo_current_name , wcvp_matched_name, 
                                              wcvp_accepted_name, plot_name), 
          caption="Species recorded in BCI plots census that don't match anything in Garwood")


plotsingarwoodorig <- garwood_list_orig %>% 
  filter(orig_name %in% bci_plots_no_garwood_at_all$orig_name)

if (nrow(bci_plots_no_garwood_at_all)>0 & nrow(plotsingarwoodorig)>0) {
  datatable(plotsingarwoodorig %>%
    select(sp6, orig_name, chg), 
            caption="In some cases, these species were on the original Garwood list but recorded there as being only in the BCNM, not on BCI.")
    
}

write_xlsx(bci_plots_no_garwood_at_all, FILECHECKPLOTS)

```

Added `r nrow(bci_plots_no_garwood_at_all)` completely new rows to the BCI list


```{r addplottobci}

bcilist <- bind_rows(
  bcilist,
  bci_plots_no_garwood_at_all %>% 
    mutate(source_list="plotcensuses",
           orig_name = sp_name_plot # the orig_name in bcilist represents the name originally listed in the source list.
           ) %>%
    select(all_of(c(listscol, "plot_name", "bci50ha", "bci10ha", "P11", "P14", "sp6", "sp4", "forestgeo_current_name"))
))


# add a field for the names that matched by alternate names
bcilist <- bcilist %>% mutate(
  pref_forestgeo = NA_character_,
  orig_garwood = NA_character_
)

# Add the prefered name in forestgeo to the species that matched with the alternate names file. 
# This is to facilitate finding the current_forestgeo name later. Since some of this matches are entirely trought alternate names
bcilist <- bcilist %>% 
  rows_update(
    bci_plots_garwood_synonym_print %>% 
      # Rename 'orig_garwood' to 'orig_name' to match bcilist exactly
    select(orig_name = orig_garwood, pref_forestgeo = prefered_name),
    by = "orig_name"
  )

```


### Lianas BCI list

Check whether they are in Garwood directly or by alternate names.  
```{r lianasnobcilist}

lianas_list <- lianas_list %>% filter(isbci=="TRUE")
lianas_list <- lianas_list %>% distinct(orig_name, .keep_all = TRUE)
lianas_list <- lianas_list %>% 
  mutate(wcvp_accepted_binomial = paste(wcvp_accepted_name_genus, wcvp_accepted_name_species),
         former_binomial = ifelse(is.na(GenusFormer),NA,paste(GenusFormer,SpeciesOnlyFormer))) %>%
  mutate(former_name = paste(ifelse(is.na(SubspeciesFormer),former_binomial,
                                    paste(former_binomial,SubspeciesFormer))))
```

Working with `r nrow(lianas_list)` liana species. 

```{r lianasdirect}

# 1. direct name/binomial match
lianas_garwood_name_match <- lianas_list %>%
  filter(

    orig_name %in% bcilist$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist$wcvp_accepted_name) |

    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist$orig_binomial 
    )) |

    (!is.na(wcvp_accepted_binomial) & (
        wcvp_accepted_binomial %in% bcilist$wcvp_accepted_binomial 
    ))
  )


#2 match of former liana name to current Garwood name 
lianas_formername_match <- lianas_list %>%
  filter(
    !orig_name %in% c(lianas_garwood_name_match$orig_name)
                      &
    (former_name %in% bcilist$orig_name | 
       former_binomial %in% bcilist$orig_binomial)  )

kable(lianas_formername_match %>% select(former_name,orig_name),
      caption="Liana species in which former names matched to Garwood original name.")
```

Matched `r nrow(lianas_garwood_name_match) + nrow(lianas_formername_match)` species in the lianas list that are directly in Garwood. 

```{r lianaaltmatch}

#4 match of former liana name to Garwood synonym
lianas_formertosynonym_match <- lianas_list %>%
  filter(
    !orig_name %in% c(lianas_garwood_name_match$orig_name)&
    (former_name %in% garwood_synonyms$synonym_sp |
       former_binomial %in% garwood_synonyms$synonym_binomial))

# see species that matched by a synonym 
lianas_formertosynonym_print <- lianas_formertosynonym_match %>%  left_join(
    garwood_synonyms %>% select(garwood_name = sp_name, synonym_sp, syn_text = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) 

kable(lianas_formertosynonym_print %>% 
            select(orig_name, former_name, garwood_name, wcvp_accepted_name,syn_text), 
          caption="Liana species whose former names matched a Garwood synonym.",
          options = list(scrollX = TRUE, pageLength = 10))

# 5. everything that did not match directly to garwood, match to alternate names
lianas_garwood_alternate_match <- lianas_list %>% 
  filter(
       !orig_name %in% c(lianas_garwood_name_match$orig_name, lianas_formertosynonym_match$orig_name) &
    (
      orig_name %in% alt_match_lookup$match_key |
      orig_binomial %in% alt_match_lookup$match_key |
      
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% alt_match_lookup$match_key) |
        
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% alt_match_lookup$match_key)
    )
  )


# print the altname it matched to and why that name is in garwood

lianas_garwood_alternate_print <- lianas_garwood_alternate_match %>%
  # Join using a chain of coalesce to check the most reliable fields first
  left_join(alt_match_lookup, by = c("orig_name" = "match_key")) %>%
  left_join(alt_match_lookup, by = c("orig_binomial" = "match_key"), suffix = c("", "_bi")) %>%
  left_join(alt_match_lookup, by = c("wcvp_matched_name" = "match_key"), suffix = c("", "_match")) %>%
  left_join(alt_match_lookup, by = c("wcvp_accepted_name" = "match_key"), suffix = c("", "_acc")) %>%
  
  # Now, condense everything into your final fields
  mutate(
    orig_garwood    = coalesce(garwood_orig_match, garwood_orig_match_bi, garwood_orig_match_match, garwood_orig_match_acc),
    orig_garwood_synonym = coalesce(garwood_synonym_match, garwood_synonym_match_bi, garwood_synonym_match_match, garwood_synonym_match_acc),
    firstmatchtype  = coalesce(firstmatchtype, firstmatchtype_bi, firstmatchtype_match, firstmatchtype_acc),
    alternate_name  = coalesce(alt_name, alt_name_bi, alt_name_match, alt_name_acc),
    prefered_name   = coalesce(pref_name, pref_name_bi, pref_name_match, pref_name_acc)
  ) %>%
  select(-ends_with("_bi"), -ends_with("_match"), -ends_with("_acc")) # Clean up extra columns

```

The table below shows liana species that could not be matched directly to Garwood (2009) but were linked through an alternate name. The `firstmatchtype` column indicates which name pairing produced the link:


Note that "Garwood original name" refers to the name as it appears in the Garwood (2009) revision, not necessarily the currently accepted name.

```{r showlianaalternate}
datatable(
  lianas_garwood_alternate_print %>% 
  select(orig_name, wcvp_matched_name, wcvp_accepted_name, firstmatchtype, orig_garwood,
         alternate_name, prefered_name, orig_garwood_synonym), 
          caption="Liana species where an alternate name of the species matched to garwood",
          options = list(scrollX = TRUE, pageLength = 10))
```

Printing liana species that are not in the list so far and need to be added: 
```{r addlianas}
# additional species still unmatched

lianas_garwood_match <- bind_rows(
  lianas_garwood_name_match,
  lianas_formername_match,
  lianas_formertosynonym_match,
  lianas_garwood_alternate_match
) %>%
  distinct(orig_name, .keep_all = TRUE)

lianas_no_garwood_at_all <- lianas_list %>%
  filter(!orig_name %in% lianas_garwood_match$orig_name)


# 6. Output and Display
kable(lianas_no_garwood_at_all %>%
  select(former_name, orig_name, wcvp_matched_name, wcvp_accepted_name), 
          caption = "Lianas species recorded in the 50-ha plot that don't have a match in Garwood",
          options = list(scrollX = TRUE, pageLength = 10))

write_xlsx(lianas_no_garwood_at_all, FILECHECKLIANAS)

```

Add to bci list
```{r lianasbciadd}

lianas_no_garwood_at_all$source_list = "lianas"

bcilist <- bind_rows(
  bcilist,
  lianas_no_garwood_at_all %>% select(all_of(listscol))
)

# Add the forestegeo names of the species that matched by alternate names
bcilist <- bcilist %>% 
  rows_update(
    lianas_garwood_alternate_print %>% 
      # Rename 'orig_garwood' to 'orig_name' to match bcilist exactly
      select(orig_name = orig_garwood, pref_forestgeo = prefered_name),
    by = "orig_name"
  )

```

### Check consistency 
```{r savebotanists}

# Check consistency
# removing rows that are not identified at species level
kable(bcilist %>% filter(is.na(word(orig_binomial, 2))) %>% select(orig_name, wcvp_accepted_name), caption="Species to be removed from final list because they are not fully identified" )

bcilist <- bcilist %>% filter(!is.na(word(orig_binomial, 2))) 

bcidupnames <- bcilist %>% group_by(orig_name) %>% filter(n()> 1) 
if (nrow(bcidupnames)>0) {
  datatable(bcidupnames%>%
          select(orig_name, wcvp_accepted_name) %>% arrange(orig_name),
          caption="Species with duplicated original name",
  options = list(scrollX = TRUE, pageLength = 10)
          )
  
}

bcidupwcvp <- bcilist %>% group_by(wcvp_accepted_name) %>% 
            filter(n()> 1)
if (nrow(bcidupwcvp)>0) {
datatable(bcidupwcvp %>% select(orig_name, wcvp_accepted_name)%>% 
            arrange(wcvp_accepted_name), 
          caption="Species duplicated wcvp_accepted name",
  options = list(scrollX = TRUE, pageLength = 10)
          )
}

bcilist %>%
  count(source_list) %>%
  arrange(desc(n)) %>%
  kable(col.names = c("source_list", "count"),caption="Sources of morphospecies in the BCI plant species list.")

datatable(
  bcilist %>%
    select(wcvp_accepted_name, orig_name, source_list) %>% 
    arrange(source_list),
    caption="Species list for BCI, with the source of each entry",
  options = list(scrollX = TRUE, pageLength = 10))

```


# Use the Panama woody plant morphospecies list to link in current ForestGEO names and 6-letter and 4-letter species codes for all species.

The forestgeo name is the most up-to-date name for these species, as they are the names used by the botanists in Panama. Next we check if the names in the current panama woody list are different from the names in Garwood, in that case the name in forestgeo is preferred. We add a new columns "current name" abd  "current_name_source" with the preferred name and whether it is Garwood, Forestgeo, Zots or the Lianas list. 


```{r renamingsome}
#checkpoint...
bcilist_orig <- bcilist
```


```{r addforestgeocurrentname}

# For all the species in the bcilist
# check if they are directly on the forestgeo /woody list
# check if they have an alternate name on that list

# Add a column with the forestgeo name
# Flag when the forestgeo_current_name != orig_garwood name

# 1. Create lookup table for woody species
forestgeo_names_lookup <- bind_rows(
  panwoody_list %>% mutate(matchby = "Fcurr") %>% select(match_key = orig_name, forestgeo_current_name = orig_name, sp6, sp4, wcvp_accepted_duplicated, matchby),
  panwoody_list %>% mutate(matchby = "wcvpmatch") %>% select(match_key = wcvp_matched_name, forestgeo_current_name = orig_name, sp6, sp4, wcvp_accepted_duplicated, matchby),
    panwoody_list %>% mutate(matchby = "wcvpaccepted") %>% select(match_key = wcvp_accepted_name, forestgeo_current_name = orig_name, sp6, sp4, wcvp_accepted_duplicated, matchby),
) %>% 
  filter(!is.na(match_key)) %>%
  # Group and collapse to keep provenance if a key matched in multiple ways
  group_by(match_key, forestgeo_current_name, sp6, sp4, wcvp_accepted_duplicated) %>%
  summarise(
    matchby = paste(matchby, collapse = ", "),
    .groups = "drop"
  )

forestgeo_names_lookup <- forestgeo_names_lookup %>% mutate(row_id = row_number())

### CHECKS ###
# check if one match_key points to different forestgeo_current_name
# This happens because the wcvp accepted name is different from the forestgeo_current_name. 
# The botanist are confident these species are different. But to wcvp v13 they are the same. 
duplicate_matchkey1 <- forestgeo_names_lookup %>%
  group_by(match_key) %>%
  filter(n_distinct(forestgeo_current_name) > 1 ) %>%
  arrange(match_key)

#datatable(duplicate_matchkey1,
#          options = list(scrollX = TRUE, pageLength = 10),
#  caption="The same key maps to a different current name. For this cases the current #forestgeo name is different from the wcvp accepted name. And the current forestgeo name is #believed to be the accepted name.")

# In this case we remove the species were the match_key is different from the forestgeo_current_name.
alt_names_remove <- duplicate_matchkey1 %>% filter(forestgeo_current_name != match_key)

forestgeo_names_lookup <- forestgeo_names_lookup %>% filter(!row_id %in% alt_names_remove$row_id)

panwoody_synonyms <- panwoody_synonyms %>% select(sp6,
  forestgeo_current_name = orig_name, altname=SynonymTextSingle
)

```

```{r foresgeoadddirect}

# Prepare for the matches
bcilist <- bcilist %>% mutate(row_id = row_number(),
                              current_name = NA_character_, 
                              source_current_name = NA_character_
                              )

```  

### Case 0. The species comes from a forestgeo list

```{r case0matchfg}
# 0. The source of the names is the forestgeo list
case0_fgsource <- bcilist %>% filter(!is.na(sp6)) %>% 
                  mutate(current_name = forestgeo_current_name,
                         source_current_name = "ForestGeo"
                         )

bcilist_fgeomatch <- bcilist %>% filter(is.na(sp6)) %>% select(-sp6, -forestgeo_current_name)

# The 6 species that are uniquely in the plots and not in the garwood list already have their forestgeo name. 
datatable(case0_fgsource %>%
            select(sp6,sp4, orig_name, forestgeo_current_name, source_list, current_name, source_current_name),
          caption="Species that are only in the plot list already have a foresgeo code",
           options = list(scrollX = TRUE, pageLength = 10),
          )

matched_species <- case0_fgsource
```

### Case 1. The species name is a current forestgeoname

```{r case1fgname}

case1_directmatch <- bcilist_fgeomatch %>% 
  left_join(forestgeo_names_lookup, by = c("orig_name"= "match_key"), suffix = c("", "_orig")) %>%
  left_join(forestgeo_names_lookup, by = c("wcvp_matched_name"= "match_key"), suffix = c("", "_wcvpmatch"))

  
case1_directmatch <- case1_directmatch %>%  mutate(
    forestgeo_current_name =     coalesce(
      forestgeo_current_name,
      forestgeo_current_name_wcvpmatch
    ),
    sp6 = coalesce(
      sp6,
      sp6_wcvpmatch
    ),
    sp4 = coalesce(
      sp4,
      sp4_wcvpmatch
    )) %>%   select(-ends_with(c("_orig",  "_wcvpacc", "_wcvpmatch"))) 


case1_directmatch <- case1_directmatch %>%
  mutate(forestgeo_current_name = orig_name,
         current_name = forestgeo_current_name,
         source_current_name = "ForestGeo"
         )  %>%
  filter(!is.na(sp6) & 
           !orig_name %in% matched_species$orig_name &
           !is.na(current_name) & !current_name %in% matched_species$current_name) 

case1_directmatch %>% group_by(forestgeo_current_name) %>% filter(n() > 1)

datatable(case1_directmatch %>%
          select(sp6,sp4, orig_name, wcvp_matched_name, wcvp_accepted_name, forestgeo_current_name, source_list, matchby),
          caption="Species original name is a forestgeo_current_name",
          options = list(scrollX = TRUE, pageLength = 10),
          )

matched_species <- bind_rows(case0_fgsource, case1_directmatch)

bcilist_fgeomatch <- bcilist %>% filter(!orig_name %in% matched_species$orig_name) %>% select(-sp6)

```


### Case 2. The species original name matches to forestgeo alternate name

```{r case2fgname}

case2_forestgeoalternate <- bcilist_fgeomatch %>% 
  left_join(panwoody_synonyms %>% select(altname, forestgeo_current_name , sp6), 
            by=c("orig_name" = "altname"),suffix = c("", "_orig")) %>%
  left_join(panwoody_synonyms %>% select(altname, forestgeo_current_name , sp6), 
            by=c("wcvp_matched_name" = "altname"),suffix = c("", "_wcvpmatch"))

case2_forestgeoalternate <- case2_forestgeoalternate %>%  mutate(
    forestgeo_current_name =     coalesce(
      forestgeo_current_name,
      forestgeo_current_name_wcvpmatch
    ),
    sp6 = coalesce(
      sp6,
      sp6_wcvpmatch
    )) %>%   select(-ends_with(c("_orig",  "_wcvpacc", "_wcvpmatch"))) 



case2_forestgeoalternate <- case2_forestgeoalternate %>%
  mutate(
         current_name = forestgeo_current_name,
         source_current_name = "ForestGeo"
         )%>%
  filter(!is.na(sp6) & 
           !orig_name %in% matched_species$orig_name &
           !is.na(current_name) & !current_name %in% matched_species$current_name
         )


datatable(case2_forestgeoalternate %>%
          select(sp6, orig_name,forestgeo_current_name, source_list),
          caption="Species original name are a forestgeo alternate name",
          options = list(scrollX = TRUE, pageLength = 10),
          )


matched_species <- bind_rows(case0_fgsource, case1_directmatch, case2_forestgeoalternate)

bcilist_fgeomatch <- bcilist %>% filter(!orig_name %in% matched_species$orig_name) %>% select(-sp6)

```

### Case 3. The species alternate name matches to ONLY one forestgeo current name


```{r case3fgname}

# at this point we only have garwood names to worry about
unique(bcilist_fgeomatch$source_list)

# Here we have to be careful, a species can have multiple alt names...
case3_forestgeoalternate <- bcilist_fgeomatch %>% 
  left_join(alternate_names %>% filter(!in_panwoodycurr & !in_viewtaxcurr)  
            %>% select(altname, prefname), 
            by=c("orig_name" = "prefname"), relationship = "many-to-many", suffix=c("","_orig")) %>%
  left_join(alternate_names %>% filter(!in_panwoodycurr & !in_viewtaxcurr)  
            %>% select(altname, prefname), 
            by=c("wcvp_matched_name" = "prefname"), relationship = "many-to-many", suffix=c("","_wcvpmatch"))


case3_forestgeoalternate <- case3_forestgeoalternate %>% 
  mutate(
    altname = coalesce(
      altname,
      altname_wcvpmatch
    )) %>%   select(-ends_with(c("_orig", "_wcvpmatch")))

# Keep only the ones that have alternate names
case3_forestgeoalternate <- case3_forestgeoalternate %>% filter(!is.na(altname)) %>%
  distinct(orig_name, altname, .keep_all = TRUE) # clearing duplicates


# Find the forestgeo name that matches to that alternate name
case3_forestgeoalternate <- case3_forestgeoalternate %>%
  left_join(panwoody_list %>% select(forestgeo_current_name = orig_name, sp6), by = c("altname" = "forestgeo_current_name"))


datatable(case3_forestgeoalternate %>%
            filter(!is.na(sp6)) %>%
            group_by(orig_name) %>% 
            filter(n() > 1) %>%
            select(sp6, orig_name, altname,source_list),
          caption="Two or more alternate names for the species match a current forestgeo name.")


datatable(case3_forestgeoalternate %>%
            filter(!is.na(sp6)) %>%
            group_by(orig_name) %>% 
            filter(n() == 1) %>%
            select(sp6, orig_name, altname, source_list),
          caption="Alternate names for the species match a current forestgeo name.")

case3_forestgeoalternate <- case3_forestgeoalternate %>%
        filter(!is.na(sp6)) %>%
        group_by(orig_name) %>% 
        filter(n() == 1) %>%
        ungroup() %>%
  mutate(current_name = altname,
         source_current_name = "ForestGeo",
         forestgeo_current_name = altname
         ) %>%   
  filter(!orig_name %in% matched_species$orig_name &
           !is.na(current_name) & !current_name %in% matched_species$current_name
         )

matched_species <- bind_rows(case0_fgsource, case1_directmatch, case2_forestgeoalternate, case3_forestgeoalternate)

bcilist_fgeomatch <- bcilist %>% filter(!orig_name %in% matched_species$orig_name) %>% select(-sp6, -forestgeo_current_name)


```

### Case 4. The species alternate name matches to  one forestgeo alternate name
```{r case4fgname}

# Here we have to be careful, a species can have multiple alt names...
case4_forestgeoalternate <- bcilist_fgeomatch %>% 
  left_join(alternate_names %>% filter(!in_panwoodycurr & !in_viewtaxcurr)  
            %>% select(altname, prefname), 
            by=c("orig_name" = "prefname"), relationship = "many-to-many", suffix=c("","_orig")) %>%
  left_join(alternate_names %>% filter(!in_panwoodycurr & !in_viewtaxcurr)  
            %>% select(altname, prefname), 
            by=c("wcvp_matched_name" = "prefname"), relationship = "many-to-many", suffix=c("","_wcvpmatch"))


case4_forestgeoalternate <- case4_forestgeoalternate %>% 
  mutate(
    altname = coalesce(
      altname,
      altname_wcvpmatch
    )) %>%   select(-ends_with(c("_orig", "_wcvpmatch")))


# Keep only the ones that have alternate names
case4_forestgeoalternate <- case4_forestgeoalternate %>% filter(!is.na(altname)) %>%
  distinct(orig_name, altname, .keep_all = TRUE) # clearing duplicates

```

Check the different cases:

```{r checkcase4}
# Find the forestgeo alternate name that matches to that alternate name
case4_forestgeoalternate <- case4_forestgeoalternate %>%
  left_join(panwoody_synonyms %>% select(forestgeo_current_name, altname,  sp6), by = "altname")


datatable(case4_forestgeoalternate %>%
            filter(!is.na(sp6)) %>%
            group_by(orig_name) %>% 
            filter(n() > 1) %>%
            select(sp6, orig_name, altname, forestgeo_current_name, source_list),
          caption="Two or more alternate names for the species match an alternate forestgeo name.")

# keeping distinct orig_name, forestgeo_current_name pairs
# If the forestgeo name is the same as the orig_name also keep that option.
case4_forestgeoalternate <- case4_forestgeoalternate %>%
  distinct(orig_name, forestgeo_current_name, .keep_all = TRUE) %>%# clearing duplicates
  filter(!is.na(sp6)) %>%
  group_by(orig_name) %>% 
  filter(n() == 1 | word(orig_name,2) == word(forestgeo_current_name,2)) %>%
  ungroup()
```

In this case we keep the forestgeo current name if it is the same for both alternate names. For Psychotria hebeclada we keep the one were the species is the same as the orig_name (because it was only a change in genus)

```{r checkcase4_1}
datatable(case4_forestgeoalternate %>%
            filter(!is.na(sp6)) %>%
            group_by(orig_name) %>% 
            filter(n() == 1) %>%
            select(sp6, orig_name, altname, forestgeo_current_name, source_list),
          caption="One alternate name for the species match an alternate forestgeo name.")


case4_forestgeoalternate <- case4_forestgeoalternate %>% 
          filter(!is.na(sp6))  %>%
            group_by(orig_name) %>% 
            filter(n() == 1) %>%
    mutate(current_name = forestgeo_current_name,
         source_current_name = "ForestGeo"
         ) %>%   
  filter(!orig_name %in% matched_species$orig_name &
           !is.na(current_name) & !current_name %in% matched_species$current_name
         )
 
matched_species <- bind_rows(case0_fgsource, case1_directmatch, case2_forestgeoalternate, case3_forestgeoalternate, case4_forestgeoalternate)

bcilist_fgeomatch <- bcilist %>% filter(!orig_name %in% matched_species$orig_name) %>% select(-sp6, -forestgeo_current_name)


```

So far we have the forestgeo names of `r nrow(matched_species)` out of `r nrow(bcilist)`o

### Case 5. The binomial of the species matches a binomial in forestgeo current names
```{r binomialcase}
# Create binomial lookup
forestgeo_binomial_lookup <- bind_rows(
  forestgeo_names_lookup,
  panwoody_list %>%
    mutate(matchby = "origbinomial") %>%
    select(match_key = orig_binomial, forestgeo_current_name = orig_name, sp6, sp4, wcvp_accepted_duplicated, matchby),
  panwoody_list %>% 
    mutate(matchby = "wcvpacceptedbinomial") %>% 
    select(match_key = wcvp_accepted_binomial, forestgeo_current_name = orig_name, sp6, sp4, wcvp_accepted_duplicated, matchby)
) %>% 
  filter(!is.na(match_key)) %>%
  # Group and collapse to keep provenance if a key matched in multiple ways
  group_by(match_key, forestgeo_current_name, sp6, sp4, wcvp_accepted_duplicated) %>%
  summarise(
    matchby = paste(matchby, collapse = ", "),
    .groups = "drop"
  )

case5_binomials <- bcilist_fgeomatch %>%
  left_join(
  forestgeo_binomial_lookup, 
  by=c("orig_binomial" = "match_key"), 
  relationship = "many-to-many") %>% 
  filter(!is.na(sp6))

case5_binomials <- case5_binomials %>%
  # Calculate mapping counts for both sides
  group_by(orig_name) %>%
  mutate(n_orig = n()) %>%
  group_by(forestgeo_current_name) %>%
  mutate(n_fg = n()) %>%
  ungroup() 

datatable(case5_binomials %>%
            filter(n_orig > 1) %>%
            select(sp6, orig_name, forestgeo_current_name, source_list),
          caption="Species matched to two or more different forestgeo binomials. This means forestgeo makes a distinction between two subspecies")

# in this case, since both subspecies are different. Keep the one without the subspecies
case5_binomials <- case5_binomials %>% 
  filter(n_orig == 1 | (n_orig > 1 & is.na(word(forestgeo_current_name,3))))
```

In this case we keep the forestgeo name without the subspecies


```{r checkcase5_2}
datatable(case5_binomials %>%
            filter(n_fg > 1) %>%
            select(sp6, orig_name, forestgeo_current_name, source_list),
          caption="The same forestgeo binomial matched to different species name. This means forestgeo does not make a distinction between two subspecies")

case5_binomials <- case5_binomials %>% 
  filter(n_fg == 1 | 
           (n_fg > 1 & word(orig_name,2) == word(orig_name,4)))
```

In this case, we keep the forestgeo name if in the original name, the subspecies is the same as the species

```{r checkcase5}
datatable(case5_binomials %>%
            group_by(orig_name,forestgeo_current_name) %>% 
            filter(n() == 1) %>%
            select(sp6, orig_name, forestgeo_current_name, source_list),
          caption="The species binomial found a match in forestgeo")

# Keep the distinct pairs of orig_name, forestgeo name
# If they are not distinct, keep only the ones were subspecies is the same as the species. 

case5_binomials <- case5_binomials %>%
          filter(
            !is.na(sp6))  %>%
            group_by(orig_name, forestgeo_current_name) %>% 
            filter(n() == 1) %>%
    mutate(
      current_name = forestgeo_current_name,
      source_current_name = "ForestGeo"
         ) %>%   
      filter(!orig_name %in% matched_species$orig_name &
           !is.na(current_name) & !current_name %in% matched_species$current_name
         )

matched_species <- bind_rows(case0_fgsource, case1_directmatch, case2_forestgeoalternate, case3_forestgeoalternate, case4_forestgeoalternate, case5_binomials)

bcilist_fgeomatch <- bcilist %>% filter(!orig_name %in% matched_species$orig_name) %>% select(-sp6, -forestgeo_current_name)

```

Now we have the forestgeo names of `r nrow(matched_species)` out of `r nrow(bcilist)`o


### Case 6. Lookin at the WCVP accepted name.

So far we have matched with the original name or the wcvp matched name (to prevent mismatches due to spelling differences). 

Next we try to match the species that have not matched yet by the wcvp_accepted_name


```{r case6directwcvp}

case6_directwcvp <- bcilist_fgeomatch %>% 
  left_join(forestgeo_names_lookup, by = c("wcvp_accepted_name"= "match_key"))


case6_directwcvp <- case6_directwcvp %>%
  filter(!is.na(sp6)) %>%
  mutate(forestgeo_current_name = wcvp_accepted_name,
         current_name = forestgeo_current_name,
         source_current_name = "ForestGeo"
         ) %>%   
      filter(!orig_name %in% matched_species$orig_name &
           !is.na(current_name) & !current_name %in% matched_species$current_name
         )

case6_directwcvp %>% group_by(forestgeo_current_name) %>% filter(n() > 1)

datatable(case6_directwcvp %>%
          select(sp6, orig_name, wcvp_matched_name, wcvp_accepted_name, forestgeo_current_name, source_list, matchby),
          caption="Species wcvp accepted name matches a forestgeo name",
          options = list(scrollX = TRUE, pageLength = 10),
          )

matched_species <- bind_rows(case0_fgsource, case1_directmatch,case2_forestgeoalternate, case3_forestgeoalternate, case4_forestgeoalternate, case5_binomials, case6_directwcvp)

bcilist_fgeomatch <- bcilist %>% filter(!orig_name %in% matched_species$orig_name) %>% select(-sp6, -forestgeo_current_name)

```

Matching by WCVP binomial
```{r case6wcvpbinomial}

case6_binomialwcvp <- bcilist_fgeomatch %>%
  left_join(
  forestgeo_binomial_lookup, 
  by=c("wcvp_accepted_binomial" = "match_key"), 
  relationship = "many-to-many") %>% 
  filter(!is.na(sp6))

datatable(case6_binomialwcvp %>%
          select(sp6, orig_name, wcvp_matched_name, wcvp_accepted_name, forestgeo_current_name, source_list, matchby),
          caption="Species original wcvp accepted binomial matches a forestgeo binomial",
          options = list(scrollX = TRUE, pageLength = 10),
          )

case6_binomialwcvp <- case6_binomialwcvp %>%
          filter(
            !is.na(sp6))  %>%
            group_by(orig_name) %>% 
            filter(n() == 1) %>%
            group_by(forestgeo_current_name) %>% 
            filter(n() == 1) %>%
    mutate(
      current_name = forestgeo_current_name,
      source_current_name = "ForestGeo"
         ) %>%   
      filter(!orig_name %in% matched_species$orig_name &
           !is.na(current_name) & !current_name %in% matched_species$current_name
         )
matched_species <- bind_rows(case0_fgsource, case1_directmatch, case2_forestgeoalternate,case3_forestgeoalternate, case4_forestgeoalternate, case5_binomials, case6_directwcvp, case6_binomialwcvp)

bcilist_fgeomatch <- bcilist %>% filter(!orig_name %in% matched_species$orig_name) %>% select(-sp6, -forestgeo_current_name)
```



## Join everything
```{r}
 

cols_to_keep <- colnames(bcilist)

bcilist_fgnames <- bind_rows(
  matched_species %>% select(all_of(cols_to_keep)),
  bcilist %>%
    filter(!orig_name %in% c(matched_species$orig_name)
           )
)

# adding sp4 column
bcilist_fgnames <- bcilist_fgnames %>% select(-sp4) %>%
  left_join(panwoody_list %>% select(sp6, sp4), by="sp6")

nrow(bcilist_fgnames)

# check for duplicated current names
datatable(bcilist_fgnames %>%
            filter(!is.na(current_name)) %>%
          group_by(current_name) %>%
          filter(n() > 1) %>%  
          select(sp6,sp4, orig_name,forestgeo_current_name, source_list) %>% arrange(current_name),
          caption="Current name is duplicated",
          options = list(scrollX = TRUE, pageLength = 10),
          )
```
So far we have found the best matches for forestgeo names. If a species is left unmatched to forestgeo it is because it is in no way in the current panama woody list, or because the match created a conflict with other species in the list that had a more direct match to the same forestgeoname (e.g it was a direct match of name while the other matched through a binomial or synonym)

Now for the species we couldnt find a match for, we set the current name as the name they have in their original list. And check again if this could potentially create a conflict with the forestgeo names. In which case the species needs revision.

```{r currentname}

bcilist_fgnames <- bcilist_fgnames %>% 
  mutate(current_name = ifelse(is.na(current_name),
                               orig_name,
                               current_name
                               ),
         source_current_name = ifelse(is.na(source_current_name),
                               source_list,
                               source_current_name
                               ),
         )

datatable(bcilist_fgnames %>%
            filter(!is.na(current_name)) %>%
          group_by(current_name) %>%
          filter(n() > 1) %>%  
          select(sp6,sp4, current_name, orig_name,forestgeo_current_name, source_current_name, source_list ) %>% arrange(current_name),
          caption="Current name is duplicated",
          options = list(scrollX = TRUE, pageLength = 10),
          )


datatable(bcilist_fgnames %>%
          select(sp6,sp4, current_name, orig_name,forestgeo_current_name, source_current_name, source_list ) %>% arrange(current_name),
          caption="BCI BOTANISTS SPECIES LIST",
          options = list(scrollX = TRUE, pageLength = 10),
          )

```



## Add woody column
```{r addwoody field}

bcilist_fgnames <- bcilist_fgnames %>%
  mutate(
    # If it has an sp6 code then it is woody or if the binomial is in woody too. 
    woody = !is.na(sp6) | orig_binomial %in% forestgeo_binomial_lookup$match_key | wcvp_accepted_binomial %in% forestgeo_binomial_lookup$match_key |
      !is.na(forestgeo_current_name)
  )

colnames(bcilist_fgnames)

# Species in the woody list
datatable(bcilist_fgnames %>%
  filter(woody) %>%
  select(sp6, sp4, forestgeo_current_name, orig_name, wcvp_matched_name, wcvp_accepted_name, wcvp_accepted_family, 
         wcvp_accepted_lifeform), 
  caption="BCI species that are in the woody list",
  options = list(scrollX = TRUE, pageLength = 10))

bcilist_fgnames <- bcilist_fgnames %>% mutate(garwood_lifeform = LF_RF)

wcvp_lifeform_woody_list = c(
  "tree",
  "shrub",
  "liana",
  "climbing shrub",
  "shrub or tree",
  "climbing shrub or tree",
  "scrambling shrub or tree",
  "subshrub or shrub",
  "subshrub, shrub or tree",
  "scrambling shrub",
  "perennial or shrub",
  "scrambling subshrub or shrub",
  "epiphytic shrub",
  "scrambling tree",
  "climbing shrub or liana",
  "scrambling shrub or liana",
  "tuberous shrub or tree",
  "herbaceous tree",
  "climbing herbaceous tree",
  "climbing herbaceous tree or liana",
  "climbing perennial or herbaceous tree",
  "subshrub or climbing shrub",
  "tuberous subshrub",
  "hemiparasitic shrub",
  "hemiepiphytic shrub or tree",
  "annual or shrub",
  "annual, subshrub or shrub"
)

# List of target abbreviations for filtering
#L Liana (to canopy)
#M Medium-sized tree: 10-30 m
#S Shrub: 1-5 m
#T Tree: >30 m
#U Treelet: 5-10 m
  
  
target_codes <- "[LMSTU]"
garwoodlf <- unique(bcilist$garwood_lifeform)
garwood_woody_lifeforms <- garwoodlf[grepl(target_codes, garwoodlf) & !grepl("a", garwoodlf, ignore.case = TRUE)]

print(garwood_woody_lifeforms)



bcilist_possibly_woody <- bcilist_fgnames %>%
  filter(!woody & is.na(`BCI cult`) &
          (wcvp_accepted_lifeform %in% wcvp_lifeform_woody_list |
           garwood_lifeform %in% garwood_woody_lifeforms 
             )  ) %>%
  select(orig_name, wcvp_matched_name, wcvp_accepted_name, wcvp_accepted_family, 
         wcvp_accepted_lifeform, garwood_lifeform)

datatable(bcilist_possibly_woody, 
          caption="BCI species that might be woody according to the wcvp lifeform",
          options = list(scrollX = TRUE, pageLength = 10))




# Save list

write_xlsx(bcilist_fgnames, FILECOMBINEDLIST)

# possibly woody
write_xlsx(bcilist_possibly_woody, FILEBCIWOODY)

```




# Wright revision: Using the BCI Botanist list to find Wright species that need revision.

Here we check what Wright species are not in the panama woody list but might be woody.

```{r sp4missing}

# Lookup table for woody species
woody_names_lookup <- bind_rows(
  panwoody_list %>% mutate(matchby = "Fcurr") %>% 
    select(match_key = orig_name, forestgeo_current_name = orig_name, sp6, sp4, wcvp_accepted_duplicated, matchby),
  
  panwoody_list %>% mutate(matchby = "wcvpmatch") %>% 
    select(match_key = wcvp_matched_name, forestgeo_current_name = orig_name, sp6, sp4, wcvp_accepted_duplicated, matchby),
  
    panwoody_list %>% mutate(matchby = "wcvpaccepted") %>% 
    select(match_key = wcvp_accepted_name, forestgeo_current_name = orig_name, sp6, sp4, wcvp_accepted_duplicated, matchby),
  
    panwoody_list %>%
    mutate(matchby = "origbinomial") %>%
    select(match_key = orig_binomial, forestgeo_current_name = orig_name, sp6, sp4, wcvp_accepted_duplicated, matchby),
  
  panwoody_list %>% 
    mutate(matchby = "wcvpacceptedbinomial") %>% 
    select(match_key = wcvp_accepted_binomial, forestgeo_current_name = orig_name, sp6, sp4, wcvp_accepted_duplicated, matchby),
  
  panwoody_synonyms %>% 
    mutate(matchby = "altname") %>% 
    select(match_key = altname, forestgeo_current_name , matchby)
) %>% 
  filter(!is.na(match_key)) %>%
  # Group and collapse to keep provenance if a key matched in multiple ways
  group_by(match_key, forestgeo_current_name, sp6, sp4, wcvp_accepted_duplicated) %>%
  summarise(
    matchby = paste(matchby, collapse = ", "),
    .groups = "drop"
  )

# Look up table for BCI species
bcilist_lookup <- bind_rows(
  # 1. Match by Original Name
  bcilist_fgnames %>% 
    mutate(matchby = "orig_name") %>% 
    select(match_key = orig_name, sp6, sp4, woody, matchby),
  
  # 2. Match by Original Binomial
  bcilist_fgnames %>% 
    mutate(matchby = "orig_binomial") %>% 
    select(match_key = orig_binomial, sp6, sp4, woody, matchby),
  
  # 3. Match by WCVP Matched Name
  bcilist_fgnames %>% 
    mutate(matchby = "wcvpmatch") %>% 
    select(match_key = wcvp_matched_name, sp6, sp4, woody, matchby),
  
  # 4. Match by WCVP Accepted Name
  bcilist_fgnames %>% 
    mutate(matchby = "wcvpaccepted") %>% 
    select(match_key = wcvp_accepted_name, sp6, sp4, woody, matchby),
  
  # 5. Match by WCVP Accepted Binomial
  bcilist_fgnames %>% 
    mutate(matchby = "wcvpacceptedbinomial") %>% 
    select(match_key = wcvp_accepted_binomial, sp6, sp4, woody, matchby),
    
  # 6. Match by Current Name (from your specific BCI columns)
  bcilist_fgnames %>% 
    mutate(matchby = "current_name") %>% 
    select(match_key = current_name, sp6, sp4, woody, matchby)
) %>% 
  # Drop NAs and apply your regex cleaning from before
  filter(!is.na(match_key)) %>%
  # Group and collapse to keep provenance
  group_by(match_key, sp6, sp4, woody) %>%
  summarise(
    matchby = paste(unique(matchby), collapse = ", "),
    .groups = "drop"
  )

bcilist_lookup_woody <- bcilist_lookup %>% filter(woody)

wright_no_woody_match <- allwrighttaxa %>%
  mutate(
    # field if there is no match at all with the woody list
    no_panwoody_match = !(
      sp6 %in% na.omit(woody_names_lookup$sp6) |
      sp4 %in% na.omit(woody_names_lookup$sp4) |
      orig_name %in% woody_names_lookup$match_key |
      orig_binomial %in% woody_names_lookup$match_key |
      wcvp_accepted_name %in% woody_names_lookup$match_key |
      wcvp_matched_name %in% woody_names_lookup$match_key 
    ),

    no_bci_woody_match <- !(
      sp6 %in% na.omit(bcilist_lookup_woody$sp6) |
      sp4 %in% na.omit(bcilist_lookup_woody$sp4) |
      orig_name %in% bcilist_lookup_woody$match_key |
      orig_binomial %in% bcilist_lookup_woody$match_key |
      wcvp_accepted_name %in% bcilist_lookup_woody$match_key |
      wcvp_matched_name %in% bcilist_lookup_woody$match_key 
    ),
      
    lifeform_wright_is_woody = str_detect(lifeform_wright, "TREE|MIDSTORY|UNDERSTORY|SHRUB|LIANA"),
    congeners_woody = (word(orig_binomial,1) %in% word(woody_names_lookup$match_key,1)),
    climber = str_detect(lifeform_wright, "CLIMBER"),

    # check first by s6 code
    lifeform_garwood = bcilist_fgnames$garwood_lifeform[match(allwrighttaxa$sp6,bcilist_fgnames$sp6)],
    # if still NA check by sp4 code
    lifeform_garwood = ifelse(
             is.na(lifeform_garwood),
             bcilist_fgnames$garwood_lifeform[match(allwrighttaxa$sp4,bcilist_fgnames$sp4)],
             lifeform_garwood
            ),
    
    lifeform_garwood_is_woody =  lifeform_garwood %in% garwood_woody_lifeforms,
    
    lifeform_wcvp_is_woody = wcvp_lifeform_description %in% wcvp_lifeform_woody_list,
    
    might_be_woody = lifeform_wright_is_woody | congeners_woody | lifeform_garwood_is_woody | lifeform_wcvp_is_woody
           )

# Table of Wright unmatched species
datatable(wright_no_woody_match %>% 
            filter(no_panwoody_match & Idlevel != "unidentified") %>%
            select(sp4, sp6, orig_name, orig_name,
                   might_be_woody,          lifeform_wright_is_woody,
         lifeform_wcvp_is_woody,
         lifeform_garwood_is_woody, congeners_woody,
                   lifeform_wright, Idlevel, lifeform_garwood, wcvp_lifeform_description
                   ) %>%
            arrange(desc(might_be_woody), orig_name),
          caption = "Wright morphospecies missing from the woody list. The column might_be_woody indicates if it could be woody based on the wright lifeform",
          options = list(scrollX = TRUE, pageLength = 10))


orig_cols <- colnames(wright_no_woody_match)

wright_no_woody_match <- wright_no_woody_match %>% 
  relocate(might_be_woody,
         lifeform_wright_is_woody,
         lifeform_wcvp_is_woody,
         lifeform_garwood_is_woody,
         congeners_woody,
         Idlevel,
         sp6, sp4, 
         orig_name, 
         wcvp_matched_name, wcvp_accepted_name, wcvp_matched_status,
         orig_family, wcvp_accepted_family, 
         might_be_woody, oldname,
         orig_author, wcvp_accepted_authority,
         lifeform_wright,
         lifeform_garwood,
         wcvp_lifeform_description,
         climber
         ) %>%
  arrange(desc(might_be_woody), Idlevel, orig_name)

write_xlsx(wright_no_woody_match, WRIGHT_MISSING_PANWOODY_PATH)

```

# Panama biota revision
this needs revision...
```{r panamabiotarevision}
pbiota_list$wcvp_accepted_binomial <- paste0(
  pbiota_list$wcvp_accepted_name_genus, " ",
  pbiota_list$wcvp_accepted_name_species
)

# 1. direct name/binomial match
pbiota_garwood_name_match <- pbiota_list %>%
  filter(

    orig_name %in% bcilist_fgnames$orig_name |

    (!is.na(wcvp_matched_name) &
       wcvp_matched_name %in% bcilist_fgnames$wcvp_matched_name) |

    (!is.na(wcvp_accepted_name) &
       wcvp_accepted_name %in% bcilist_fgnames$wcvp_accepted_name) |

    (!is.na(orig_binomial) & (
        orig_binomial %in% bcilist_fgnames$orig_binomial 
    )) |

    (!is.na(wcvp_accepted_binomial) & (
        wcvp_accepted_binomial %in% bcilist_fgnames$wcvp_accepted_binomial 
    )) |
      
    (!is.na(orig_name) & (
        orig_name %in% bcilist_fgnames$current_name 
    ))
  )

# 2. synonym_sp for remaining
pbiota_garwood_synonym_match <- pbiota_list %>%
  filter(
    !orig_name %in% pbiota_garwood_name_match$orig_name &
    (
      orig_name %in% garwood_synonyms$synonym_sp |
      (!is.na(wcvp_matched_name) &
         wcvp_matched_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(wcvp_accepted_name) &
         wcvp_accepted_name %in% garwood_synonyms$synonym_sp) |
      (!is.na(orig_binomial) & orig_binomial %in% garwood_synonyms$synonym_binomial) |
      (!is.na(wcvp_accepted_binomial) & wcvp_accepted_binomial %in% garwood_synonyms$synonym_binomial)
    )
  )

pbiota_garwood_synonym_match$matched_garwood_synonym <- TRUE


# see species that matched by a synonym 
pbiota_garwood_synonym_match_print <- pbiota_garwood_synonym_match %>%  left_join(
    garwood_synonyms %>% select(sp_orig = sp_name, synonym_sp, syn_orig = SynonymTextALL) %>% distinct(),
    by = c("orig_name" = "synonym_sp")
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_match = sp_name, synonym_sp, syn_match = SynonymTextALL) %>% distinct(),
    by = c("wcvp_matched_name" = "synonym_sp"), relationship = "many-to-many"
  ) %>%
  left_join(
    garwood_synonyms %>% select(sp_wcvp = sp_name, synonym_sp, syn_wcvp = SynonymTextALL) %>% distinct(),
    by = c("wcvp_accepted_name" = "synonym_sp"),  relationship = "many-to-many") %>%
  mutate(
    SynonymMatchAll = coalesce(syn_orig, syn_match, syn_wcvp),
    orig_garwood = coalesce(sp_orig, sp_match, sp_wcvp),
  ) %>%
  select(-syn_orig, -syn_match, -syn_wcvp,  -sp_orig, -sp_match, -sp_wcvp) %>%
  distinct(orig_name, .keep_all = TRUE)
  

datatable(pbiota_garwood_synonym_match_print %>% 
            select(orig_name, wcvp_matched_name, wcvp_accepted_name, SynonymMatchAll, orig_garwood), 
          caption="Panamabiota species that matched Garwood by synonym",
          options = list(scrollX = TRUE, pageLength = 10))
```

