---
title: "Matching Liana Species names to WCVP"
author: "Paula Uzcategui and Helene Muller-Landau"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Match WCVP names with species name

Next we do:

Input: list of species

- Match wcvp names
- Tabulate types of changes: spelling differences, Synonyms. 
- After the update two species can have the same wcvp name. 

Output: 
- _WCVP file with the original species list columns and the columns from wcvp
- _WCVP_CHANGES a file containing only the species where the original name differs fromthe wcvp name. 

```{r setup2, echo=TRUE,message=FALSE, warning=FALSE, results="hide"}

# installing wcvp
#install.packages("remotes")
#library(remotes)
#remotes::install_github("matildabrown/rWCVP")
#rm(list=ls())

library(readxl)
library(writexl)
library(dplyr)
library(tidyr)
library(stringr)
library(knitr)
suppressPackageStartupMessages({
  library(rWCVP)
  library(rWCVPdata)
})
library(DT)

citation("rWCVPdata") # according to this, this is using version 13.

# Load WCVP backbone
wcvp_names <- rWCVPdata::wcvp_names 

# INPUT PATHS
original_list_path <- "splists_raw/Schnitzer/Schnitzer_2026-02-13/Lianas_Species_complete_list.xlsx"

pub_list_path <- "splists_raw/Schnitzer/Schnitzer_2024-10-04/BCI_2007_2017__species_names_SEPT27_2024.csv"

# Extract file name without extension 
original_list_name <- tools::file_path_sans_ext(basename(original_list_path))

# Path for intermediate matching results (to skip heavy processing if already done)
matches_mid_path <- paste0("splists_mid/", original_list_name, "_mid_match.rds")
matches_mid_path2 <- paste0("splists_mid/", original_list_name, "_mid_match2.rds")

# OUTPUT PATHS 
out_wcvp_list <- paste0("splists_out/", original_list_name, "_WCVP_", Sys.Date(), ".xlsx")
out_wcvp_changes_list <- paste0("splists_out/", original_list_name, "_WCVP_CHANGES_", Sys.Date(), ".xlsx")


redoWCVP= TRUE



```

# Load species lists, and put them in a consistent format

```{r cleandata,echo=TRUE}
lianas <- read_excel(original_list_path)
lianasbci50 <- data.frame(codename=lianas$`List 1`)
lianasplots <- data.frame(codename=lianas$`List 2`)

lianasbci50 <- lianasbci50 %>% 
  filter(!is.na(codename) & codename!="Desconocida") %>%
  separate(codename, into = c("sp6", "binomial"), sep = " - ", remove = FALSE)%>%
  mutate(sp6 = str_trim(sp6) %>% str_to_upper(),
         binomial = str_squish(binomial) %>% str_to_sentence()) %>%
  arrange(sp6) %>%
  select(-codename)

lianasplots <- lianasplots %>% 
  filter(!is.na(codename) & codename!="Desconocida") %>%
  separate(codename, into = c("sp6", "binomial"), sep = " - ", remove = FALSE)%>%
  mutate(sp6 = str_trim(sp6) %>% str_to_upper(),
         binomial = str_squish(binomial) %>% str_to_sentence())  %>%
  arrange(sp6) %>%
  select(-codename)

lianasbci50pub <- read.csv(pub_list_path)

lianasbci50pub <- lianasbci50pub %>%
  mutate(across(where(is.character), str_squish))  %>% # remove any trailing or leading white space and reduce internal white space to always 1 space
  rename(sp6=Species.Code..Mnemonic.,
         SpeciesFull=Species,
         FamilyFormer=Family..former.,
         GenusFormer=Genus..former.,
         SpeciesFormerFull=Species..former.,
         AuthorityFormer=Authority..former.) %>%
  mutate(SpeciesOnly = word(SpeciesFull, 1),
         Subspecies = str_squish(str_remove(SpeciesFull, "^[^\\s]+\\s*")),
         Subspecies = na_if(Subspecies, ""),
         SpeciesOnlyFormer = word(SpeciesFormerFull,1),
         SubspeciesFormer = str_squish(str_remove(SpeciesFormerFull, "^[^\\s]+\\s*")),
         SubspeciesFormer = na_if(SubspeciesFormer, ""),
         binomial=paste0(Genus," ",SpeciesOnly),
         binomialFormer=paste0(GenusFormer," ",SpeciesOnlyFormer),
         binomialplus=paste0(Genus, " ",SpeciesFull),
         binomialplusFormer=paste0(GenusFormer," ",SpeciesFormerFull)) %>%
  select(sp6,binomial,binomialFormer,binomialplus,binomialplusFormer,
         Family,Genus,SpeciesOnly,Subspecies,Authority,
         FamilyFormer,GenusFormer,SpeciesOnlyFormer,SubspeciesFormer,AuthorityFormer) %>%
  arrange(binomial)
```


## Check for consistency between published list and newer lists for BCI
```{r comparebci50lists}
onlypubbci50 <- lianasbci50pub %>%
  select(sp6, binomial) %>%
  anti_join(lianasbci50 %>% select(sp6), by = "sp6")

onlynewbci50 <- lianasbci50 %>%
  select(sp6, binomial) %>%
  anti_join(lianasbci50pub %>% select(sp6), by = "sp6")

kable(onlypubbci50, caption = "Species in published list but not newer list for BCI 50 ha lianas")
kable(onlynewbci50, caption = "Species in newer list but not published list for BCI 50 ha lianas")

#Find cases where sp6 is the same but binomial differs (potential naming mismatch)
sp6_conflicts2bcilists <- bind_rows(
  lianasbci50 %>% mutate(source = "bci50new") %>% select(source, sp6, binomial),
  lianasbci50pub %>% mutate(source = "bci50pub") %>% select(source, sp6, binomial)
) %>%
  distinct() %>%
  group_by(binomial) %>%
  filter(n_distinct(sp6) > 1) %>%
  arrange(binomial, sp6) %>%
  ungroup()

kable(sp6_conflicts2bcilists, caption = "Conflicts: same sp6, different binomial")

# Find cases where binomial is the same but sp6 differs (potential code mismatch)
binomial_conflicts2bcilists <- bind_rows(
  lianasbci50 %>% mutate(source = "bci50new") %>% select(source, sp6, binomial),
  lianasbci50pub %>% mutate(source = "bci50pub") %>% select(source, sp6, binomial)
) %>%
  distinct() %>%
  group_by(binomial) %>%
  filter(n_distinct(sp6) > 1) %>%
  arrange(binomial, sp6) %>%
  ungroup()

kable(binomial_conflicts2bcilists, caption = "Conflicts: same binomial, different sp6")

```
The checks show that these two lists are the same (including having the same duplicates), so proceed using just the published list.  

## Check for duplicates within lists

### Check for duplicates within BCI 50-ha published list
```{r checkdupbci50}
binomial_conflicts_bci50pub <- lianasbci50pub %>%
  distinct(sp6, binomial) %>%
  group_by(binomial) %>%
  filter(n_distinct(sp6) > 1) %>%
  arrange(binomial, sp6) %>%
  ungroup()

if (nrow(binomial_conflicts_bci50pub)>0) 
  kable(binomial_conflicts_bci50pub,
      caption = "Cases in published BCI 50-ha plot liana species list where the same binomial corresponds to multiple sp6 codes.")

sp6_conflicts_bci50pub <- lianasbci50pub %>%
  distinct(sp6, binomial) %>%
  group_by(sp6) %>%
  filter(n_distinct(binomial) > 1) %>%
  arrange(sp6,binomial) %>%
  ungroup()

if (nrow(sp6_conflicts_bci50pub)>0) 
  kable(sp6_conflicts_bci50pub,
      caption = "Cases in published BCI 50-ha plot liana species list where the same sp6 code corresponds to multiple binomials.")

```
There are 5 cases where the published BCI 50-ha plot liana species lists where two different codes correspond to the same binomial.

There are no cases where a single code corresponds with multiple binomials.  


### Check for duplicates within newer plots list
```{r checkdupplots}
binomial_conflicts_plots <- lianasplots %>%
  distinct(sp6, binomial) %>%
  group_by(binomial) %>%
  filter(n_distinct(sp6) > 1) %>%
  arrange(binomial, sp6) %>%
  ungroup()

if (nrow(binomial_conflicts_plots)>0) 
  kable(binomial_conflicts_plots,
      caption = "Cases in newer non-BCI plots liana species list where the same binomial corresponds to multiple sp6 codes.")

sp6_conflicts_plots <- lianasplots %>%
  distinct(sp6, binomial) %>%
  group_by(sp6) %>%
  filter(n_distinct(binomial) > 1) %>%
  arrange(sp6,binomial) %>%
  ungroup()

if (nrow(sp6_conflicts_plots)>0) 
  kable(sp6_conflicts_plots,
      caption = "Cases in newer non-BCI plots liana species list where the same sp6 code corresponds to multiple binomials.")

```



## Check for overlap and inconsistencies between the published BCI list and the plots list

The following `r length(setdiff(lianasbci50pub$sp6, lianasplots$sp6))` sp6 codes are in the published BCI 50-ha plot liana species list but not in the newer list for non-BCI plots: 

`r paste(setdiff(lianasbci50pub$sp6, lianasplots$sp6), collapse = ", ")`

The following `r length(setdiff(lianasplots$sp6, lianasbci50pub$sp6))` sp6 codes are in the newer lists for non-BCI plots but not in the published BCI 50-ha plot liana species list:  

`r paste(setdiff(lianasplots$sp6, lianasbci50pub$sp6), collapse = ", ")`


```{r comparebci50vplots}

onlybci50 <- lianasbci50pub %>%
  select(sp6, binomial) %>%
  anti_join(lianasplots %>% select(sp6), by = "sp6")

onlyplots <- lianasplots %>%
  select(sp6, binomial) %>%
  anti_join(lianasbci50pub %>% select(sp6), by = "sp6")

kable(onlybci50, caption = "Species in lianasbci50 but not in lianasplots")
kable(onlyplots, caption = "Species in lianasplots but not in lianasbci50")

lianasall <- bind_rows(
  lianasbci50pub %>% mutate(sourcelist = "bci50") %>% select(sourcelist, sp6, binomial),
  lianasplots %>% mutate(sourcelist = "plots") %>% select(sourcelist, sp6, binomial)
) %>%
  distinct(sp6,binomial,.keep_all=TRUE) %>%
  arrange(sp6,binomial)

# Find cases where binomial is the same but sp6 differs (potential code mismatch)
binomial_conflicts2lists <- lianasall %>%
  group_by(binomial) %>%
  filter(n() > 1) %>%               
  arrange(binomial, sp6, sourcelist) %>%
  ungroup()

kable(binomial_conflicts2lists,
      caption = "Cases where the same binomial appears with different codes")

#Find cases where sp6 is the same but binomial differs (potential naming mismatch)
sp6_conflicts2lists <- lianasall %>%
  group_by(sp6) %>%
  filter(n() > 1) %>%                  
  arrange(sp6, binomial, sourcelist) %>%
  ungroup()

kable(sp6_conflicts2lists,
      caption = "Cases where the same sp6 code appears with different binomials")

```
It appears that many of these are spelling differences.  

## Combine the lists

Keep all distinct sp6 codes.  When the sp6 code appears in the published BCI 50 list, use the binomial that appears there, regardless of what appears in the plots list.  

Keeping `nrow(lianasbci50pub)` codes from the published BCI 50-ha species list (even though in some cases more than one code corresponds to the same binomial), and adding `length(lianasplots$sp6[!lianasplots$sp6 %in% lianasbci50pub$sp6])` codes with their binomials from the other plots list.  

Remove the record for BAUH1 = Bauhinia sp., because this lacks a full species identification. 

 
Manually edit one name, changing genus Chondodendron to Chondrodendron, to avoid matching with the wrong taxon in WCVP.  


```{r combinelists}

lianasbci50pub$isbci = TRUE

lianascombined <- bind_rows(lianasbci50pub,lianasplots[!lianasplots$sp6 %in% lianasbci50pub$sp6,]) %>% 
  filter(sp6!="BAUH1")

origlist <- lianascombined %>%
  rename(orig_name=binomial,
         orig_family=Family,
         orig_genus=Genus,
         orig_species=SpeciesOnly,
         orig_author=Authority) %>%
  mutate(orig_infraspecies_rank=word(Subspecies,1),
         orig_infraspecies=word(Subspecies,2),
         orig_genus=ifelse(!is.na(orig_genus),orig_genus,word(orig_name,1)),
         orig_species=ifelse(!is.na(orig_species),orig_species,word(orig_name,2)),
         # remove extra spaces inside name
          orig_name = str_squish(orig_name), 
        # add columns for the simple species name (not including "sp." or such), and simple binomial
          orig_binomial = str_squish(paste0(orig_genus, " ", 
                               ifelse(substr(orig_species, 1, 3) == "sp.", "", orig_species)))
         ) %>% 
  arrange(orig_name)

origlist$orig_genus[origlist$sp6=="CHONTO"] <- "Chondrodendron"
origlist$orig_name[origlist$sp6=="CHONTO"] <- "Chondrodendron tomentosum"

# find duplicates
dups <- origlist %>% group_by(orig_name) %>% filter(n() > 1) %>% select(sp6, orig_name)

if(nrow(dups) > 0){
  knitr::kable(dups, caption="Duplicated names in original")
}

# leaving them because this have different codes

```


Working with `r nrow(origlist)` species. 

 
### Match WCVP names

```{r mergewcvp}
# Helper function to merge and format accepted names from WCVP backbone
matchaccepted_names <- function(matcheswcvp, wcvp_backbone){
  matcheswcvp %>% 
    left_join(
      wcvp_backbone %>% 
        select(
          plant_name_id,
          # rename immediately
          wcvp_accepted_ipni_id  = ipni_id,
          wcvp_accepted_powo_id  = powo_id,
          wcvp_accepted_family   = family,
          wcvp_accepted_name_rank = taxon_rank,
          wcvp_accepted_name_status = taxon_status,
          wcvp_lifeform_description = lifeform_description,

          # rename after the mutate for readability 
          genus, species, infraspecific_rank, infraspecies,
          parenthetical_author, primary_author
        ),
      by = c("wcvp_accepted_id" = "plant_name_id")
    ) %>% 
    mutate( 
      wcvp_accepted_name = case_when(
        is.na(species) | species == "" ~ genus,
        !is.na(infraspecies) & infraspecies != "" ~ paste(genus, species, infraspecific_rank, infraspecies),
        TRUE ~ paste(genus, species)
      ),
      
      wcvp_accepted_authority =
        paste0(
          ifelse(!is.na(parenthetical_author), paste0("(", parenthetical_author, ") "),""),
          coalesce(primary_author, "")
        )
    ) %>% 
    rename(
      wcvp_accepted_author = primary_author,
      wcvp_accepted_name_genus = genus,
      wcvp_accepted_name_species = species,
      wcvp_accepted_name_infraspecific_rank = infraspecific_rank,
      wcvp_accepted_name_infraspecies = infraspecies
    )
}

# --- 1. Run or Load Match ---
if(file.exists(matches_mid_path) & !redoWCVP){
  matchresult <- readRDS(matches_mid_path)
} else {
  # Run the matching process
  matchresult <- wcvp_match_names(origlist, wcvp_names, name_col = "orig_name")

  # Save as RDS for speed
  saveRDS(matchresult, matches_mid_path)
}

# do the same for the orig_binomial in cases where this is not equal to orig_name
if(file.exists(matches_mid_path2) & !redoWCVP){
  matchresult2 <- readRDS(matches_mid_path2)
} else {
  matchresult2 <- wcvp_match_names(subset(origlist,orig_name!=orig_binomial), wcvp_names, 
                                   name_col = "orig_binomial")

  # Save as RDS for speed
  saveRDS(matchresult2, matches_mid_path2)
}

# --- 2. Sanity Checks & Duplicate Resolution ---
# Ensure all species are accounted for
stopifnot(all(origlist$orig_name %in% matchresult$orig_name))

# Identify names that mapped to multiple 'Accepted' records
manyaccepted <- matchresult %>%
  group_by(orig_name) %>%
  filter(sum(wcvp_status == "Accepted", na.rm = TRUE) > 1) %>%
  ungroup()

# If duplicates exist, force use of the original ID to avoid grouping explosion
if (nrow(manyaccepted) > 0) {
  matchresult <- matchresult %>%
    mutate(wcvp_accepted_id = ifelse(orig_name %in% manyaccepted$orig_name, wcvp_id, wcvp_accepted_id))
}

# --- 3. Clean Match Results (1 row per species) ---
matchresult_clean <- matchresult %>%
  mutate(
    status_priority = case_when(
      wcvp_status == "Accepted"     ~ 1,
      wcvp_status == "Synonym"      ~ 2,
      wcvp_status == "Illegitimate" ~ 3,
      wcvp_status == "Unplaced"     ~ 4,
      TRUE                          ~ 5
    )
  ) %>%
  group_by(orig_name) %>%
  slice_min(status_priority, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  matchaccepted_names(wcvp_names)

matchresult2_clean <- matchresult2 %>%
  mutate(
    status_priority = case_when(
      wcvp_status == "Accepted"     ~ 1,
      wcvp_status == "Synonym"      ~ 2,
      wcvp_status == "Illegitimate" ~ 3,
      wcvp_status == "Unplaced"     ~ 4,
      TRUE                          ~ 5
    )
  ) %>%
  group_by(orig_name) %>%
  slice_min(status_priority, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  matchaccepted_names(wcvp_names)

```

### Fix mismatches

```{r spot_mismatch}

mismatches <- matchresult_clean %>% 
  filter(orig_name != wcvp_name | is.na(wcvp_name) | is.na(wcvp_accepted_name)) %>%
  left_join(matchresult2_clean %>% select(-orig_binomial) %>% rename_with(~paste0(., "_binomial"), -orig_name),
            by = "orig_name") %>%
  mutate(
    # Extract wcvp genus, species, infraspecies rank and infraspecies
    wcvp_genus = word(wcvp_name, 1),
    wcvp_species = word(wcvp_name, 2),
    wcvp_infraspecies_rank = word(wcvp_name, 3),
    wcvp_infraspecies = word(wcvp_name, 4),
    
   # Add booleans for mismatches
    genus_mismatch = !is.na(orig_genus) & 
                      (orig_genus != wcvp_genus | is.na(wcvp_genus)),
    species_mismatch = !is.na(orig_species) & 
                        (orig_species != wcvp_species | is.na(wcvp_species)),
    infraspecies_rank_mismatch = !is.na(orig_infraspecies_rank) & 
                                  (orig_infraspecies_rank != wcvp_infraspecies_rank | 
                                   is.na(wcvp_infraspecies_rank)),
    infraspecies_mismatch = !is.na(orig_infraspecies) & 
                             (orig_infraspecies != wcvp_infraspecies | 
                              is.na(wcvp_infraspecies))
  )


printmismatches <- mismatches %>% 
  select(
    orig_name, wcvp_name, wcvp_accepted_name, match_similarity, 
    orig_binomial, wcvp_name_binomial, wcvp_accepted_name_binomial, match_similarity_binomial,
    genus_mismatch, species_mismatch, infraspecies_rank_mismatch, infraspecies_mismatch
  ) %>% 
  arrange(match_similarity, match_similarity_binomial)

datatable(printmismatches,options = list(scrollX = TRUE, pageLength = 10))


colsprint = c("orig_name", "wcvp_name","wcvp_accepted_name" ,"orig_binomial", "wcvp_name_binomial", "wcvp_accepted_name_binomial")


mismatches$preferred = "full"
```

#### Case 1:
```{r printmisses}
# both binomial and full mismatched
mismatchcase1 <- mismatches %>% filter( 
  !is.na(wcvp_accepted_name) &
  orig_name != wcvp_name   &
                        (orig_binomial != wcvp_name_binomial | is.na(wcvp_name_binomial)) )

datatable( mismatchcase1 %>% select(all_of(colsprint)),
           caption="both binomial and full name mismatched",
options = list(scrollX = TRUE, pageLength = 10))

```

In this cases the original spelling differs slightly from the name matched in WCVP. 


#### Case 2:
```{r case3}
# full name matched except infraspecies rank
mismatchcase3 <- mismatches %>% filter( infraspecies_rank_mismatch & !genus_mismatch & !species_mismatch & !infraspecies_mismatch) 
  
datatable( mismatchcase3 %>% select(all_of(colsprint)),
           caption="full name matched except infraspecies rank",options = list(scrollX = TRUE, pageLength = 10))

```

The mismatch is a difference only in infraspecific rank. Use the WCVP accepted name for the full match. 

#### Case 3:
```{r findingcase2}
# full name mismatched, binomial didin't. And genus and species did not mismatch
# this is because the infraspecies is messing the fuzzy matching
mismatchcase2 <- mismatches %>% filter( orig_name != wcvp_name &
                       orig_binomial == wcvp_name_binomial &
                      (genus_mismatch | species_mismatch) &
                       !is.na(orig_infraspecies) & orig_infraspecies!= ""
                         ) 
datatable( mismatchcase2 %>% select(all_of(colsprint)),
           caption="Mismatch for full name involving a mismatch on genus or species, and at the same time perfect match when just on binomial.",
           options = list(scrollX = TRUE, pageLength = 10)
           )


```
In this cases the match to binomial is better, so use that.  (Note also the infraspecies is the same as the species in all these cases except the one without a species name at all.) 

```{r fixingcase3}

inc <- with(mismatches, orig_name %in% mismatchcase2$orig_name)
if (any(inc, na.rm = TRUE)) {
  tmp <- mismatches[inc, ]
  tmp$preferred = "binomial"
  mismatches[inc, ] <- tmp
}

```

#### Case 4:
```{r case4}
# genus, species and rank matched but infraspecies didint 
mismatchcase4 <- mismatches %>% filter( !genus_mismatch & !species_mismatch & infraspecies_mismatch & !infraspecies_rank_mismatch)

datatable( mismatchcase4 %>% select(all_of(colsprint)), caption="genus, species and infraspecific rank matched but infraspecies name did not ",options = list(scrollX = TRUE, pageLength = 10))


```

In all of this cases the subspecies did not match in wcvp. Thus using the simple binomial match instead (dropping the infraspecies).

```{r fixingcase4}
inc <- with(mismatches, orig_name %in% mismatchcase4$orig_name)
if (any(inc, na.rm = TRUE)) {
  tmp <- mismatches[inc, ]
  tmp$preferred = "binomial"
  mismatches[inc, ] <- tmp
}
```

#### Other cases:
```{r case5}
# species not considered in the previous cases:
missrest <- mismatches %>% 
  filter(!orig_name %in% c(mismatchcase1$orig_name, mismatchcase2$orig_name, mismatchcase3$orig_name, mismatchcase4$orig_name))

datatable( missrest %>% select(all_of(colsprint)), caption="Species not considered in previous cases",options = list(scrollX = TRUE, pageLength = 10))

```
For all of these, there is no WCVP match to the full name (Except for Phyllanthus sp. and Trema micrantha), but in most cases there is a match to the binomial name.  Thus use the binomial name.  And in the cases where there is no match, it is because of a lack of taxonomic resolution in the original ID (just to family).  


```{r fixingcase5}

inc <- with(mismatches, orig_name %in% missrest$orig_name)
if (any(inc, na.rm = TRUE)) {
  tmp <- mismatches[inc, ]
  tmp$preferred = "binomial"
  mismatches[inc, ] <- tmp
}


```

#### Summary of mismatches after fixing:

List of wcvp columns to keep
```{r wcvpcolumns}

# list of wcvp columns we are interested to keep at the end.
wcvp_columns <- c(
        # columns of matched name
        "wcvp_name", "wcvp_authors", "wcvp_rank", "wcvp_status", "wcvp_id", "wcvp_ipni_id", "match_similarity",
        
        # columns of accepted name
        "wcvp_accepted_id","wcvp_accepted_powo_id", "wcvp_accepted_ipni_id",
        "wcvp_accepted_name", "wcvp_accepted_name_genus",  "wcvp_accepted_name_species",
        "wcvp_accepted_name_rank", "wcvp_accepted_family", 
        "wcvp_accepted_name_infraspecific_rank", "wcvp_accepted_name_infraspecies",
        "wcvp_accepted_author", "wcvp_accepted_authority",
        "wcvp_lifeform_description"
        )
```

Now all the mismatches should make more sense
```{r fixingmismatches}

# summary of mismatches
knitr::kable(mismatches %>% group_by(preferred) %>% summarise(species = n()))

# correct the mismatches with the binomial  matches when necessary 
mismatches <- mismatches %>% 
  mutate(
    across(
      all_of(c(wcvp_columns)),
      ~ifelse(
        preferred == "binomial",
        get(paste0(cur_column(), "_binomial")),
        .x
      )
    )
  )

datatable(mismatches %>% select(orig_name, wcvp_name, wcvp_accepted_name),
          options = list(scrollX = TRUE, pageLength = 10),
          caption="Species after setting the wcvp matched name to a closer match")

```



## Joining mismatch fixes to the main list

```{r matcheshandling}

# --- 4. Handle  matches ---

# joins the corrected mismatches back to the original wcvp matched list
inc <- with(matchresult_clean, orig_name %in% mismatches$orig_name)
if (any(inc, na.rm = TRUE)) {
  corrections <- mismatches %>%
    select(all_of(c("orig_name", wcvp_columns)))
  
  matchresult_clean <- rows_update(matchresult_clean, corrections, 
                                   by = "orig_name")
}

# --- 5. Final Join and Rename ---
# This merges the WCVP column with the original columns of the list
processlist <- origlist %>%
  left_join(matchresult_clean %>% 
              select(all_of(c("orig_name", wcvp_columns))),
            by = "orig_name") %>%
  rename(
    # rename matched columns
    wcvp_matched_name = wcvp_name,
    wcvp_matched_authors = wcvp_authors,
    wcvp_matched_rank = wcvp_rank,
    wcvp_matched_status=wcvp_status,
    wcvp_matched_plant_name_id = wcvp_id,
    wcvp_matched_ipni_id = wcvp_ipni_id,
    
    wcvp_accepted_plant_name_id =  wcvp_accepted_id,
  ) %>%
  mutate(
  origmatches = orig_name == wcvp_matched_name
  )

# Review Results
knitr::kable(processlist %>% count(wcvp_matched_status))

knitr::kable(subset(processlist,
             is.na(wcvp_accepted_name) | !wcvp_matched_status %in% c("Accepted", "Synonym") ) %>% 
        select(orig_family,orig_name,wcvp_matched_status,wcvp_accepted_name),
      caption = "Species not listed as Accepted or Synonym in WCVP")
```


## Save output of all species.

Showing columns saved in the final dataset:
```{r saveoutput}

# Save species list with wcvp columns
write_xlsx(processlist, out_wcvp_list)

print(paste(colnames(processlist), collapse=", "))
```

Saved `r nrow(processlist)` species with their names verified with WCVP. 


## Save output of all changes
output the taxa that have any differences between accepted name and orig_name as a separate file of changes.  
```{r saveoutput_changes}

list_wcvp_changes <- processlist %>% filter(orig_name != wcvp_accepted_name)
  
# Save species list with wcvp columns
write_xlsx(list_wcvp_changes, out_wcvp_changes_list)


```

Saved `r nrow(list_wcvp_changes)` species where the original name is different from the wcvp accepted name. 



